---
title: "Tok Pisin"
author: "Justin Bai"
bibliography: tpi.bib
output: html_document
---

Last Updated: 2020-07-02

# Background

**Language Family:** Creole / English based / Pacific

* Although the "Pisin" of the language's name does indeed come from the English word
"pidgin," the language is more accurately characterized as a creole as there are
native speakers and the system has become grammaticized [@Smith2008, p. 192].

# Phonology

* There is much variation in the phonemic inventory as many speakers learn Tok Pisin as a second language or also speak English [@Smith2008, p. 199]. The analysis below reflects the core (somewhat idealized) phonemic inventory.

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial  Alveolar  Postalveolar  Palatal   Velar   Glottal
Stops                      "p b"   "t d"         ""           ""       "k ɡ"   ""
Affricates                 ""      ""            "dʒ"         ""       ""      ""
Fricatives                 "f v"   "s"           ""           ""       ""      "h"
Nasals                     "m"     "n"           ""           ""       "ŋ"     ""
Trills                     ""      "r"           ""           ""       ""      ""
Approximants               "w"     "l"           ""           "j"      ""      ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  footnote("Note: For phonemes that share a cell, those on the left are voiceless, whereas those on the right are voiced.", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

## Vowels

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
          Front   Central   Back
High      "i"     ""        "u"
Mid       "e"     ""        "o"
Low       ""      "a"       ""
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme
"a"              "/a/"
"b"              "/b/"
"d"              "/d/"
"e"              "/e/"
"f"              "/f/"
"g"              "/ɡ/"
"h"              "/h/"
"i"              "/i/"
"j"              "/dʒ/"
"k"              "/k/"
"l"              "/l/"
"m"              "/m/"
"n"              "/n/"
"o"              "/o/"
"p"              "/p/"
"r"              "/r/"
"s"              "/s/"
"t"              "/t/"
"u"              "/u/"
"v"              "/v/"
"w"              "/w/"
"y"              "/j/"
**Digraph**       ""
"ng"             "/ŋ/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

#Lenition Rules

* /dʒ/ spirantizes to [s] non-word-initally [@Boer2017, p.60].

# Misc. Rules

* Epenthetic vowels are often inserted between consonant clusters [@Smith2008, p. 204].

# References
