---
title: "Qeqchi"
author: "Bill Mizgerd"
bibliography: kek.bib
output: html_document
---

Last Updated: 2019-08-07

**Slightly Compromised: orthographic ambiguity between ejective stops and stop-glottal stop sequences**

# Background

**Language Family**: Mayan / Core Mayan / Quichean-Mamean / Greater Quichean / Qeqchi

* The language has several different names, including Kekchi, K'ekchi', and Q'eqchi'.
* Qeqchi is spoken in Guatemala and Brazil.

# Phonology

## Consonants

* @KekStewart2015 describes the dorsal fricative as both velar (p. 16) and uvular (p. 12) at different points in the document. I have followed the lead of @KekPinkerton1976 in preferring to describe it as velar (p. 10).

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial    Alveolar    Postalveolar     Palatal     Velar     Uvular    Glottal
Stops                     "p  ɓ"     "t  t’"         ""             ""       "k  k’"   "q  q’"      "ʔ"
Affricates                   ""      "ts  ts’"    "tʃ  tʃ’"         ""         ""        ""         ""
Fricatives                   ""        "s"           "ʃ"            ""         "x"       ""         "h"
Nasals                      "m"        "n"           ""             ""         ""        ""         ""
Trills                      ""         "r"           ""             ""         ""        ""         ""
Approximants                "w"        "l"           ""             "j"        ""        ""         ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Uvular", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 7)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: Phonemes that share a cell match in voicing, but those on the right are ejectives. The exception is the labial stops, which contrast in voicing (/ɓ/ is an implosive).", general_title = "")
```

## Vowels

* Vowel length is contrastive in Qeqchi [@KekStewart2015, p. 12]. Long vowels are written by doubling the vowel character.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         ""           "u"
Mid       "e"         ""           "ɔ"
Low        ""         "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```   

# Alphabet

* /ʔ/ can follow plain stops, which creates possible ambiguity. For example, ⟨t'⟩ might represent an ejective consonant or a sequence of a plain stop and a glottal stop. In such cases, I have opted to prefer the ejective-consonant analysis, as it is significantly more common in Qeqchi than the stop/glottal sequence is [@KekDeChicchis2011, p. 3].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme              Comment
"a"           "/a/"                  ""
"b\'"         "/ɓ/"                  ""
"e"           "/e/"                  ""
"h"           "/h/"                  ""
"i"           "/i/"                  ""
"j"           "/x/"                  ""
"k"           "/k/"                  ""
"k\'"         "/k’/"                 ""
"l"           "/l/"                  ""
"m"           "/m/"                  ""
"n"           "/n/"                  ""
"o"           "/ɔ/"                  ""
"p"           "/p/"                  ""
"q"           "/q/"                  ""
"q\'"         "/q’/"                 ""
"r"           "/r/"                  ""
"s"           "/s/"                  ""
"t"           "/t/"                  ""
"t\'"         "/t’/"                 ""
"u"           "/u/"                  ""
"w"           "/w/"                  ""
"x"           "/ʃ/"                  ""
"y"           "/j/"                  ""
"\'"          "/ʔ/"       "⟨-⟩ is also used on occasion"
**Digraph**    ""                    ""
"ch"         "/tʃ/"                  ""
"ch\'"       "/tʃ’/"                 ""
"tz"         "/ts/"                  ""
"tz\'"       "/ts’/"                 ""
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* /h/ voices to [ɦ] intervocalically, and sometimes deletes entirely in non-final postconsonantal positions [@KekStewart2015, p. 16].

# Misc. Rules

* Preconsonantal and word-final plain stops (besides /ʔ/) are realized with aspiration [@KekStewart2015, p. 13].
* Non-pulmonic consonants are realized as preglottalized when in postvocalic position (e.g. /k’/ → [ʔk’], /ɓ/ - [ʔb’]) [@KekStewart2015, pp. 14-15; @KekPinkerton1976, p. 12].
* /n/ assimilates to [ŋ] before dorsal consonants [@KekStewart2015, p. 17].
* @KekPinkerton1976 argues that /r/ is typically realized as [ʂ], with voicing and trilling occurring intermittently for some speakers (p. 16).
* Glides "obstruentize" word-initially and in postconsonantal syllable-initial positions; this results in what @KekPinkerton1976 describes as [ẘ] and [j̊], but what some other sources describe as [kʷ] and [tʲ] (pp. 17, 26).

# References
