---
title: "Mari"
author: "Abi Creighton"
bibliography: chm.bib
output: html_document
---

Last Updated: 2020-06-10

**COMPROMISED: conflation between /n/ and /ɲ/ and between /l/ and /ʎ/; skepticism related to the orthographic reliability of the reduced vowels /ə/ and /æ/ (i.e. is it exhaustive)**

# Background

**Language Family:** Uralic / Finno-Ugric / Finno-Permic / Finno-Cheremisic / Cheremisic

* Mari is spoken in Russia.
* Crúbadán appears to represent Hill (western) Mari rather than Meadow Mari, with the inclusion of graphemes ⟨ӓ⟩ (/æ/) and ⟨ӹ⟩ (/ə/), so where dialects diverge, I will favor Hill Mari.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"         Bilabial  Labiodental Alveolar Postalveolar Alveopalatal Palatal Velar     
Stops                             "p b"       ""        "t d"       ""           ""         ""    "k ɡ"    
Affricates                         ""         ""        "ts"        ""          "tɕ"        ""     ""
Fricatives                         "β"       "f"        "s z"      "ʃ ʒ"        "ɕ"         ""     "x"
Nasals                             "m"        ""        "n"         ""           ""         "ɲ"    "ŋ"
Trills                             ""         ""        "r"         ""           ""         ""     ""
Approximants                       ""         ""        "l"         ""           ""        "j ʎ"   ""           
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Bilabial", "Labiodental", "Alveolar", "Postalveolar", "Alveopalatal", "Palatal", "Velar"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 7)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced. Both palatal approximants are voiced, but the one on the right is lateral.", general_title = "")
```

## Vowels

* Although, I have chosen to stick with the analysis of /ə/, given that other sources attest its reduced nature without contradiction [@Kangasmaa-Minn1998, p. 223; @Campbell2012, p. 1082], I have veered away from the analysis of /ə̂/. @Alhoniemi1993 [as cited in @Bradley2017, p. 371] highlights the classification of this reduced mid-central vowel in Mari; however, there is some skepticism related to this. @Estill2012 and @Bradley2017 point out that this interpretation is suspect given that the description of the phoneme warrants a place more back than that of schwa (p. 229; p. 371). I have chosen to use /ɤ/ just as @Bradley2017 does (p. 371).

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
          Front  Central   Back
High      "i y"    ""       "u"
Mid       "e ø"    "ə"      "ɤ o"
Low        "æ"     ""       "ɑ"
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For vowels that share a cell, those on the left are unrounded and those on the right are rounded.", general_title = "")

```  

# Alphabet

* The palatalization of ⟨н⟩ and ⟨л⟩ is not indicated orthographically preceding vowels /e/ and /i/, which somewhat compromises the language [@Riese2012, p. 32].
* @Kangasmaa-Minn1998 states that the reduced vowels /ə/ and /æ/ are not always marked orthographically (p. 223).
* Recent Russian loanwords may be realized according to their Russian pronunciation rather than adapted to Mari phonology [@Bradley2017, p. 377].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme                           Comment
"а"            "/ɑ/"                              ""
"ӓ"            "/æ/"                              ""
"б"            "/b/"                              ""
"в"            "/β/"                              ""
"г"            "/ɡ/"                              ""
"д"            "/d/"                              ""
"е"          "/e/; /je/"                        "/je/: word-initially, following vowels, and following both the hard and soft sign"
"ё"            "/jo/"                           "primarily appears in Russian loanwords"
"ж"            "/ʒ/"                              ""
"з"            "/z/"                              ""
"и"            "/i/"                              ""
"й"            "/j/"                              ""
"к"            "/k/"                              ""
"л"            "/l/"                              ""
"м"            "/m/"                              ""
"н"            "/n/"                              ""
"ҥ"            "/ŋ/"                              ""
"о"            "/o/"                              ""
"ӧ"            "/ø/"                              ""
"п"            "/p/"                              ""
"р"            "/r/"                              ""
"с"            "/s/"                              ""
"т"            "/t/"                              ""
"у"            "/u/"                              ""
"ӱ"            "/y/"                              ""
"ф"            "/f/"                              "primarily appears in Russian loanwords"
"х"            "/x/"                              "primarily appears in Russian loanwords"
"ц"            "/ts/"                             "primarily appears in Russian loanwords"
"ч"            "/tɕ/"                             ""
"ш"            "/ʃ/"                              ""
"щ"            "/ɕ/"                              "primarily appears in Russian loanwords"
"ъ"             ""                              "referred to as a hard sign: indicates palatalization at word boundaries and when preceding specific vowels"
"ы"            "/ɤ/"                              ""
"ӹ"            "/ə/"                              ""
"ь"             ""                              "referred to as a soft sign: indicates palatalization of specific consonants and when preceding specific vowels"
"э"            "/e/"                              ""
"ю"            "/ju/; /u/"                      "/ju/: word-initially, following vowels, and following both the hard and soft sign"
"я"            "/jɑ/; /ɑ/"                      "/jɑ/: word-initially, following vowels, and following both the hard and soft sign"
**Digraph**     ""                                ""
"нь"           "/ɲ/"                            "phoneme occurs if ⟨н⟩ is also followed by ⟨ю⟩ or ⟨я⟩"
"ль"           "/ʎ/"                            "phoneme occurs if ⟨л⟩ is also followed by ⟨ю⟩ or ⟨я⟩"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Misc. Rules

* /d/ devoices when followed by a consonant [@Riese2012, p. 38].
* Voiced stops and /β/ devoice syllable-finally (ibid.).
* Mari has vowel harmony. Suffixes match the stressed vowel in terms of frontness and roundedness [@Riese2012, pp. 39-40].
* /β/ may be realized as [v], especially among younger generations [@Riese2012, pp. 21-22]

# References
