---
title: "Rotokas"
author: "Bill Mizgerd"
bibliography: roo.bib
output: html_document
---

Last Updated: 2019-07-18

# Background

**Language Family:** East Papuan / Bougainville / West / Rotokas

* The Aita dialect of Rotokas has a few significant differences in its phoneme inventory; I will note these differences in this file, but my ruleset will only reflect the phonology of Central Rotokas.
* Rotokas is spoken in the Bougainville province of Papua New Guinea.

# Phonology

## Consonants

* The voiced consonants of Rotokas undergo a considerable degree of free variation, making the choice of underlying form - in the words of @RooRobinson2006 - "somewhat arbitrary" (p. 207). The bilabial and velar consonants are realized as either a stop, fricative, or nasal; the alveolar voiced consonants are realized as either a stop, nasal, flap, or lateral approximant. @RooFirchow1969 list /β ɾ ɡ/ as the underlying voiced phonemes at each place of articulation (pp. 273-274); I have chosen to follow this analysis.
* For a more detailed explanation of the role of nasals in Rotokas, see the Misc. Rules section.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
              Bilabial   Alveolar   Velar
Stops         "p"        "t"        "k ɡ"
Fricatives    "β"        ""         ""
Flaps         ""         "ɾ"        ""
'), TRUE)

kable(consonants, align = ('c')) %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 3)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: Phonemes to the left of each cell are voiceless and phonemes to the right of each cell are voiced.", general_title = "")
```

## Vowels

* Vowel length is contrastive in Rotokas [@RooFirchow1969, p. 271]; however, this contrast has been lost in the Aita dialect [@RooRobinson2006, p. 209].
    - Vowel length is written by doubling the vowel character. [@RooFirchow1992, p. 1]
* @RooFirchow1992 argues for /ɑ/ rather than /a/ (p. 2).

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
       Front   Central   Back
High   "i"     ""        "u"
Mid    "e"     ""        "o"
Low    ""      "a"       ""
'), TRUE)

kable(consonants, align = ('c')) %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme
"a"              "/a/"  
"e"              "/e/"  
"g"              "/ɡ/"  
"i"              "/i/"  
"k"              "/k/"
"o"              "/o/"  
"p"              "/p/"  
"r"              "/ɾ/"  
"s"              "/t/"  
"t"              "/t/"  
"u"              "/u/"  
"v"              "/β/"
'), TRUE)

kable(alphabet, align = "c") %>%
    kable_styling("bordered")
```

# Syllable Structure

* The structure of Rotokas syllables is (C)V(V) [@RooFirchow1969, pp. 274-275].

# Lenition Rules

* /d/ is most commonly realized as [ɾ] [@RooRobinson2006, p. 207].

# Misc. Rules

* /t/ spirantizes to [s], or sometimes goes to [ts], before /i/ [@RooFirchow1969, p. 274].
    - In recent years, however, this process has become less frequent; @RooRobinson2006 theorizes that this trend is because of bilingualism with Tok Pisin, which has contrastive /t/ and /s/ (p. 207).
* In the Aita dialect of Rotokas, voiced consonants are nasalized so often that the nasals /m n ŋ/ are held to be the underlying forms of those sounds; however, other dialects of Rotokas rarely nasalize voiced consonants unless deliberately trying to replicate foreign speech patterns [@RooFirchow1969, p. 274].
* @RooRobinson2006, meanwhile, actually finds the nasals to be contrastive with the voiced stops in the Aita dialect (p. 208); in his view, there was a historical contrast between those two series that has since collapsed in standard Central Rotokas.

# References
