---
title: "Sinaugoro"
author: "Emily Strand"
bibliography: snc.bib
output: html_document
---

Last Updated: 2020-06-23

# Background

**Language Family:** Austronesian / Malayo-Polynesian / Central-Eastern / Eastern Malayo-Polynesian / Oceanic / Western / Papuan Tip / Peripheral / Central Papuan / Sinagoro-Keapara

* Sinaugoro is spoken in the Central province of Papua New Guinea.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial  Alveolar     Velar   
Stops                      "b"      "t d"     "k kʷ ɡ ɡʷ"  
Fricatives                "f v"      "s"        "ɣ ɣʷ"     
Nasals                     "m"       "n"          ""    
Trills                     ""        "r"          ""     
Approximants               ""        "l"          ""    
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Velar"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 3)) %>%
  footnote("Note: For phonemes that share a cell, those on the left are voiceless, whereas those on the right are voiced. Phonemes with the diacritic (ʷ) are labialized.", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* The diphthongs listed below are interpreted as complex phonemes [@tauberschmidt_grammar_1999, p. 4].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
          Front   Central   Back
High      "i"     ""        "u"
Mid       "e"     ""        "o"
Low       ""      "a"       ""
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)


diphthongs <- read.table(textConnection('
Diphthongs
"/ae/, /ai/, /ao/, /au/, /ei/, /ou/, /oi/"
'), TRUE)

kable(diphthongs, align = 'c') %>%
  kable_styling("bordered")
```

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme
"a"              "/a/"  
"b"              "/b/"  
"d"              "/d/"  
"e"              "/e/"  
"f"              "/f/"  
"g"              "/ɡ/"  
"ḡ"              "/ɣ/"  
"i"              "/i/"  
"k"              "/k/"  
"l"              "/l/"  
"m"              "/m/"  
"n"              "/n/"  
"o"              "/o/"  
"r"              "/r/"  
"s"              "/s/"  
"t"              "/t/"  
"u"              "/u/"  
"v"              "/v/"  
**Digraph**       ""
"gw"             "/ɡʷ/"  
"kw"             "/kʷ/"  
"ḡw"             "/ɣʷ/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```   

# Syllable Structure

* Syllables in Sinaugoro require empty codas and prohibit consonant clusters, leaving CV and V as the inherent structures [@tauberschmidt_grammar_1999, p. 4].

# Misc. Rules

* /i/ is realized as [j] in syllable onsets [@tauberschmidt_grammar_1999, p. 4].
* /ɣ/ palatalizes preceding front vowels [@tauberschmidt_grammar_1999, p. 7].

# References
