---
title: "Tamil"
author: "Becky Mathew"
bibliography: ta.bib
output: html_document
---

Last Updated: 2019-11-21

# Background

**Language family**: Dravidian / Southern / Tamil-Kannada / Tamil-Kodagu / Tamil-Malayalam / Tamil

* Tamil is a language primarily spoken in the Indian state Tamil Nadu, as well in the countries Sri Lanka, Malayasia, and Singapore. Tamil exhibits diglossia, wherein the written variety of the language is formal or literal, and the spoken variety of the language (which is more colloquial) is used in most other contexts. We used Standard Spoken Tamil for this project [@Keane2004, p. 111].

# Phonology

## Consonants

* Unlike other related languages, Tamil doesn't have voiced obstruents or aspirated consonants [@Keane2004, pp. 111-112].
* Like other Dravidian languages, Tamil has multiple rhotic sounds and makes use of the retroflex place of articulation (ibid.).
* All consonants carry an inherent /a/ following the consonant [@Shulman2016, p. 22].
* /dʒ/, /h/, /ʂ/, /ʃ/, /s/, /f/, and /z/ appear in Sanskirt loanwords, their corresponding graphemes of which come from the Grantha writing system [@Keane2004, p. 112]. In terms of standard Tamil, @Keane2004 explains them to be rather marginal, so they won't be accounted for.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial  Dental  Alveolar  Postalveolar  Retroflex  Palatal   Velar
Stops                      "p"      "t̪"       ""         ""            "ʈ"      ""       "k"
Affricates                 ""        ""       ""        "tʃ"           ""       ""       ""  
Nasals                     "m"       ""       "n"        ""            "ɳ"      ""       ""  
Flaps                      ""        ""       "ɾ"        ""            ""       ""       ""  
Approximants               "ʋ"       ""       "l"        ""          "ɻ  ɭ"     "j"      ""  
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Dental", "Alveolar", "Postalveolar", "Retroflex", "Palatal", "Velar"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 7)) %>%
  footnote("Note: The retroflex approximant on the right is lateral.", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* Vowel length is contrastive in Tamil [@Keane2004, pp. 114-115].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
          Front   Central   Back
High      "i"     ""        "u"
Mid       "e"      ""       "o"
Low       ""      "a"       ""
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered")


diphthongs <- read.table(textConnection('
Diphthongs
"/aɪ/, /aʊ/"
'), TRUE)

kable(diphthongs, align = 'c') %>%
  kable_styling("bordered")
```

# Alphabet

* The alphabet is somewhat phonetic, but I have transcribed it such that the underlying phonemes are represented.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
"Consonant Grapheme"         Phoneme                                   Comment
"க"                         "/k a/"                                      ""
"ங"                         "/n a/"        "always precedes nasal consonants, so it represents the allophone [ŋ]"
"ச"                         "/tʃ a/"                                     ""
"ஞ"                        "/n a/"                         "represents the allophone [ɲ]"
"ட"                         "/ʈ a/"                                      ""
"ண"                       "/ɳ a/"                                       ""
"த"                         "/t̪ a/"                                      ""
"ந"                         "/n a/"                        "represents the allophone [n̪]"
"ப"                        "/p a/"                                       ""
"ம"                        "/m a/"                                       ""
"ற"                        "/ɾ a/"               "in some dialects, this is pronounced with the trill [r]"
"ன"                       "/n a/"                                       ""
"ர"                        "/ɾ a/"                                       ""
"ல"                        "/l a/"                                      ""
"ள"                       "/ɭ a/"                                       ""
"ழ"                       "/ɻ a/"                                       ""
"ய"                       "/j a/"                                       ""
"வ"                       "/ʋ a/"                                       ""
"ஃ"                       "/k a/"       "in some dialects, this is pronounced as [x] but in Standard Spoken Tamil [x] is an allophone of /k/ (often used to show aspiration from English loanwords)"
"**(Word-initial) Vowel Grapheme**"   ""                                 ""
"அ"                       "/a/"                                         ""
"ஆ"                      "/aː/"                                         ""
"இ"                       "/i/"                                         ""
"ஈ"                       "/iː/"                                        ""
"உ"                       "/u/"                                         ""
"ஊ"                     "/uː/"                                         ""
"எ"                      "/e/"                                          ""
"ஏ"                      "/eː/"                                         ""
"ஐ"                      "/aɪ/"                                         ""
"ஒ"                      "/o/"                                          ""
"ஓ"                      "/oː/"                                         ""
"ஔ"                    "/aʊ/"                                         ""
"**Vowel Diacritic**"      ""                    "vowel diacritics attach to consonants, changing the inherent /a/ to different vowels"
"ா"                     "/aː/"                                          ""
"ி"                      "/i/"                                           ""
"ீ"                      "/iː/"                                          ""
"ு"                     "/u/"                                           ""
"ூ"                    "/uː/"                                          ""
"ெ"                    "/e/"                                           ""
"ே"                    "/eː/"                                          ""
"ை"                   "/aɪ/"                                          ""
"ொ"                  "/o/"                                           ""
"ோ"                  "/oː/"                                          ""
"ௌ"                 "/aʊ/"                                          ""
"்"                      "∅"                          "erases the inherent vowel /a/ after a consonant"
'), TRUE)

kable(alphabet, col.names = c("Consonant Grapheme", "Phoneme", "Comment"), align = 'c') %>%
  kable_styling("bordered")
```  

# Lenition Rules

* Voiceless obstruents become voiced folling nasals [@Keane2004, p. 112].
    - Geminate voiceless obstruents remain voiceless.
* Intervocalically [@Schiffman1999, p. 10]:
    - /t̪/ fricates
    - /tʃ/ reduces to /s/
    - /p/ becomes voiced and may spirantize or become a glide
    - /ʈ/ becomes voiced
    - /t/ becomes a flap
    - /k/ debuccalizes to /h/
* Following a nasal, the above stops and affricates become voiced (ibid.).
* Occasionally, /ʋ/ and /k/ delete intervocalically (ibid.).

# Misc. Rules

* All consonants except ⟨ர⟩ and ⟨ழ⟩ can geminate. Although ⟨ர⟩ is underlying the same phoneme as ⟨ற⟩, ⟨ற⟩ is said to be able to undergo gemination [@Schiffman1999, pp. 11-12].

# References
