---
title: "Batak Karo"
author: "Daniela Wiepert"
edited by: "Emily Strand"
bibliography: btx.bib
output: html_document
---

Last Updated: 2020-06-26

**COMPROMISED: conflation among /e/, /ɘ/, and /ɯ/**

# Background

**Language Family:** Austronesian / Malayo-Polynesian / Western Malayo-Polynesian / Sundic / Sumatra / Batak / Northern

* Batak Karo is spoken in the Aceh and North Sumatra provinces of Indonesia.

# Phonology

## Consonants

* @Woollams1996 argues for /t̪/ rather than /t/ (p. 15).

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"     Labial  Alveolar  Alveopalatal  Palatal  Velar  Glottal
Stops                        "p b"    "t d"        ""          ""     "k ɡ"    ""
Affricates                    ""       ""        "tɕ dʑ"       ""      ""      ""
Fricatives                    ""       "s"         ""          ""      ""      "h"
Nasals                        "m"      "n"         ""          ""      "ŋ"     ""
Trill                         ""       "r"         ""          ""      ""      ""
Approximants                  "w"      "l"         ""          "j"     ""      ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Alveopalatal", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* @Woollams1996 specifies a high central vowel /ɨ/, but represents it as the high back unrounded vowel /ɯ/ (pp. 18-19). In his 2004 book, he also specifies it as back rather than central [@Woollams2004, p. 536]. I have chosen to use /ɯ/, as it is more consistent with the literature.
* @Woollams1996 also specifies /ə/ as a phoneme (pp. 19, 22). Schwa, however, doesn't seem to be appropriate in this case as it doesn't represent a truly reduced vowel (i.e. it sometimes appears in stressed syllables). I have chosen to use /ɘ/ instead.
* Adjacent vowels are interpreted as sequences rather than complex nuclei [@Woollams1996, p. 27].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
           Front      Central       Back
High        "i"         ""          "ɯ u"
Mid         "e"         "ɘ"          "o"
Low         ""          "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: The high back vowel on the left is unrounded and the one on the right is rounded.", general_title = "")
```

# Alphabet

* Although @Woollams2004 specifies one-to-one correspondences between the graphemes ⟨é⟩, ⟨e⟩, and ⟨ë⟩ and the respective phonemes /e/, /ɘ/, and /ɯ/ (p. 536), he also states that all three phonemes are represented by ⟨e⟩ [@Woollams1996, p. 23]. Therefore, the language is compromised.
    - Although the Crúbadán corpus does not include the graphemes ⟨é⟩ and ⟨ë⟩, I have included rules for them in the rule set. I have decided to transcribe ⟨e⟩ to /ɘ/ as the default rule, given that /ɘ/ seems to be the most frequently occurring phoneme of the three [based on the phonemic representations in @Woollams1996].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme                 Comment
"a"              "/a/"                    ""
"b"              "/b/"                    ""
"c"              "/tɕ/"                   ""
"d"              "/d/"                    ""
"e"          "/e/; /ɯ/; /ɘ/"      "/ɘ/: default in the rules"
"g"              "/ɡ/"                    ""
"h"              "/h/"                    ""
"i"              "/i/"                    ""
"j"              "/dʑ/"                   ""
"k"              "/k/"                    ""
"l"              "/l/"                    ""
"m"              "/m/"                    ""
"n"              "/n/"                    ""
"o"              "/o/"                    ""
"p"              "/p/"                    ""
"r"              "/r/"                    ""
"s"              "/s/"                    ""
"t"              "/t/"                    ""
"u"              "/u/"                    ""
"w"              "/w/"                    ""
"y"              "/j/"                    ""
**Digraph**       ""                      ""
"ng"             "/ŋ/"                    ""
'), TRUE)
kable(alphabet, align = "c") %>%
  kable_styling("bordered")
```

# Lenition Rules

* /h/ is normally elided word-initially [@Woollams1996, p. 30].
* /k/ is realized as voiceless glottal stop word-finally and preceding consonants [@Woollams1996, p. 16].

# Misc. Rules

* Vowels are often elided in fast speech [@Woollams1996, p. 30].
* Voiceless stops tend to be unreleased word-finally [@Woollams1996, p. 16].
* /h/ is sometimes realized as /x/ syllable-finally [@Woollams2004, p. 536].
* Syllable-initially, /r/ may be realized as a flap (ibid.).
* Consonant gemination occurs with voiceless stops (including /tɕ/), nasals, /s/ and /r/ [@Woollams1996, p. 26].
* [j] and [w] epenthesis may occur intervocalically [@Woollams1996, p. 29].

# References
