---
title: "Jola-Fogny"
author: "Abi Creighton"
bibliography: dyo.bib
output: html_document
---

Last Updated: 2020-07-17

**COMPROMISED: unclear how reliably +ATR vowels are marked**

# Background

**Language Family:** Niger-Congo / Atlantic-Congo / Atlantic / Northern / Bak / Jola / Jola Proper / Jola Central / Jola-Fogny

* Jola-Fogny, also known as Jola-Fonyi, Diola-Fogny, Joola, or Jola, is spoken primarily in the Casamance region of Senegal with a smaller population of speakers residing in Gambia.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial   Alveolar  Palatal  Velar   Glottal
Stops                      "p b"     "t d"     "c ɟ"   "k ɡ"     ""   
Fricatives                  "f"       "s"       ""      ""       "h"
Nasals                      "m"       "n"       "ɲ"     "ŋ"      ""       
Trills                      ""        "r"       ""      ""       ""  
Laterals                    ""        "l"       ""      ""       ""
Approximants                "w"       ""        "j"     ""       ""   
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 5)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced", general_title = "")
```  

## Vowels

* Jola-Fogny has 10 vowels: five +ATR/tense (advanced tongue root) and five -ATR/lax. +ATR vowels are transcribed orthographically with an acute accent (´) [@gero_1993, p. 80].
* Length is phonemic for all vowels and is marked by duplicate graphemes [@wade_2005].
* Diphthongs exist in Jola-Fogny, but they are rare [@easterday_2019, p. 385]. Only /iu/ appears in the Crúbadán corpus.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
             Front      Central        Back
High          "i"         ""           "u"
Near-high     "ɪ"         ""           "ʊ"
High-Mid      "e"         "ɘ"          "o"
Low-Mid       "ɛ"         ""           "ɔ"
Low           ""          "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)


diphthongs <- read.table(textConnection('
Diphthongs
"/eʊ/, /iu/, /ɪe/, /ɔa/, /eɪ/"
'), TRUE)

kable(diphthongs, align = 'c') %>%
  kable_styling("bordered")
```     

# Alphabet

* Accented vowels are not considered to be part of the alphabet by Jola-Fogny speakers, but because they correspond to different phonemes compared to their non-accented counterparts, I have chosen to include them below.
* Apostrophes are used intervocalically to distinguish between two independent short vowels and long vowels [@wade_2005, p. 3].
* Looking at other Niger-Congo languages with similar vowel inventories (e.g. Igbo), +ATR vowels, which are more common, don't generally occur in words with -ATR vowels [@Ringen1979, p. 248]. This is ultimately reflective of the vowel harmony present in the language [@Sapir1975, p. 6]. Due to this, all vowels following a marked +ATR vowel (generally the first vowel in a word) are unmarked (i.e. they don't have an acute accent) [@wade_2005, p. 3].
    - E.g. ⟨ébe⟩ = /ebe/
* The transcription of the Crúbadán corpus is somewhat suspect as the -ATR vowels appear more frequently than the +ATR vowels, which is contrary to what is expected. Therefore, the marking of +ATR vowels is suspect. I have opted to classify this language as compromised as it is unclear how reliably +ATR vowels are marked.


```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme           
"a"              "/ɘ/"   
"á"              "/a/"
"b"              "/b/"
"c"              "/c/"
"d"              "/d/"
"e"              "/ɛ/"
"é"              "/e/"
"f"              "/f/"
"g"              "/ɡ/"              
"h"              "/h/"              
"i"              "/ɪ/"
"í"              "/i/"
"j"              "/ɟ/"
"k"              "/k/"
"l"              "/l/"
"m"              "/m/"              
"n"              "/n/"
"ñ"              "/ɲ/"
"ŋ"              "/ŋ/"
"o"              "/ɔ/"
"ó"              "/o/"
"p"              "/p/"              
"r"              "/r/"              
"s"              "/s/"              
"t"              "/t/"              
"u"              "/ʊ/"
"ú"              "/u/"
"w"              "/w/"   
"y"              "/j/"   
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Syllables in Jola-Fogny follow the order of (C)V(C)(C) [@sapir_1965, pp. 6-9].

# Misc. Rules

* Geminate consonants are present throughout Jola languages, which are marked by repeated graphemes [@wade_2005, p. 2]. Jola-Fogny is specifically said to only have nasal geminates. However, in the rules, I have accounted for all possible geminate consonants.
* Consonants tend to prenasalize preceding nasals (ibid.).
* Velar stops are realized as post-velar preceding /u/ and palatal preceding front vowels [@sapir_1965, p. 5].
* /ɘ/ reduces to schwa when unstressed [@sapir_1965, p. 6].
* Vowels nasalize preceding nasal consonants [@wade_2005, p. 3].

# References
