---
title: "Saint Lucian Creole French"
author: "Emily Strand"
bibliography: acf.bib
output: html_document
---

Last Updated: 2019-09-11

# Background

**Language family**: French Creole

* This creole language is spoken in Saint Lucia, an island within the Caribbean.

# Phonology

## Consonants

```{r echo = FALSE, message = FALSE, warning = FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('

"Manner of Articulation"    Labial     Alveolar    Postalveolar     Palatal       Velar         Glottal
Stops                        "p  b"     "t  d"          ""             ""         "k  ɡ"           ""
Affricates                     ""         ""         "tʃ  dʒ"          ""           ""             ""
Fricatives                   "f  v"     "s  z"        "ʃ  ʒ"           ""           ""             "h"
Nasals                        "m"         "n"           ""             ""           "ŋ"            ""
Approximants                   ""       "l  ɹ"          ""             "j"          "w"            ""
'), header = TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced. The alveolar approximants are both voiced, but the one on the left is lateral.", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* There is some debate as to how many vowels are included in the vowel inventory. @bhatt_creole_2006 as well as @mondesir_dictionary_1992 argue for an eight-vowel inventory (p. 9; p. 3); however, @crosbie_kweyol_2001 as well as @frank_sources_2007 argue for a seven-vowel inventory (p. xi; p. 3). The debate centers around the question of whether /a/ and /ɑ/ are contrastive. Given the fact that @lefebvre_creoles_2011 indicates that /ɑ/ has only been marginally attested and seldom appears in modern day St. Lucian Creole, I have opted for the seven-vowel distribution (pp. 109-110).
* Three nasal vowels are also classified as phonemes, indicated by a following ⟨n⟩ [@crosbie_kweyol_2001, p. xvii].

```{r echo = FALSE}

vowels <- read.table(textConnection('
               Front       Central         Back
High            "i"         ""             "u"
"High-Mid"    "e  ẽ"        ""            "o  õ"
"Low-Mid"       "ɛ"         ""             "ɔ"
Low             ""         "a  ã"           ""
'), header = TRUE)

kable(vowels, align = 'c') %>%
 kable_styling("bordered") %>%
 footnote(general = "Note: For phonemes that share a cell, those on the left are oral and those on the right are nasal.", general_title = "") %>%
 column_spec(1, bold = TRUE)
```


# Alphabet

* @mondesir_dictionary_1992 has /ɛ/ represented by ⟨e⟩ (p. 4), whereas most other sources use ⟨è⟩ as the convention [@crosbie_kweyol_2001, p. xi; @frank_sources_2007, pp. 4-11]. Crubadan seems to use ⟨è⟩; however I will account for both forms due to /e/ being distinguishable.

```{r echo = FALSE}

alphabet <- read.table(textConnection('
Grapheme      Phoneme                 Comment                            
"a"            "/a/"                    ""                               
"b"            "/b/"                    ""                               
"d"            "/d/"                    ""                               
"é"            "/e/"                    ""                               
"è;  e"        "/ɛ/"                    ""                              
"f"            "/f/"                    ""                               
"g"            "/ɡ/"                    ""                               
"h"            "/h/"                    ""                               
"i"            "/i/"                    ""                               
"j"            "/ʒ/"                    ""                               
"k"            "/k/"                    ""                                
"l"            "/l/"                    ""                                
"m"            "/m/"                    ""                               
"n"            "/n/"                    ""                                
"o"            "/o/"                    ""                                
"ò"            "/ɔ/"                    ""                                
"p"            "/p/"                    ""                               
"r"            "/ɹ/"     "only appears in English loanwords"             
"s"            "/s/"                    ""                                
"t"            "/t/"                    ""                              
"v"            "/v/"                    ""                               
"w"            "/w/"                    ""                               
"y"            "/j/"                    ""                                
"z"            "/z/"                    ""                                
**Digraph**     ""                      ""
"ch"           "/ʃ/"                    ""
"dj"           "/dʒ/"                   ""
"ng"           "/ŋ/"                    ""
"ou"           "/u/"                    ""
"tj"           "/tʃ/"                   ""
"an"           "/ã/"                    ""
"en"           "/ẽ/"                    ""
"on"           "/õ/"                    ""

'), header = TRUE)

knitr::kable(alphabet, align = 'c') %>%
kable_styling("bordered")
```  

# References
