---
title: "Chuvash"
author: "Emily Strand"
bibliography: cv.bib
output: html_document
---

Last Updated: 2019-08-06

# Background

**Language Family**: Altaic / Turkic / Bolgar

* Chuvash, also called Bulgar, is spoken in the Chuvashia republic in western Russia.

# Phonology

## Consonants

* /b/, /d/, /g/, /f/, /z/, /ʒ/, /ts/, and /dʒ/ are preserved only in loanwords (Russian) [@johanson_chuvash_2006, p. 434]. Therefore, voicing is not phonemic within the Chuvash language.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial    Alveolar    Alveolo-Palatal  Postalveolar  Palatal   Velar
Stops                       "p"         "t"           ""               ""         ""       "k"
Affricates                   ""         ""            ""              "tʃ"        ""       ""  
Fricatives                  "v"         "s"           "ɕ"             "ʃ"         ""       "χ"
Nasals                       "m"        "n"           ""               ""         ""       ""
Trills                       ""         "r"           ""               ""         ""       ""
Approximants                 ""         "l"           ""               ""         "j"      ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Alveolo-Palatal", "Postalveolar", "Palatal", "Velar"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE)
```


## Vowels

* There is some debate concerning /ӗ/ and /ӑ/ in terms of what phonemes they actually represent. They are said to be reduced, with /ӑ/ being a back-to-central vowel and /ӗ/ being a front to central vowel [@johanson_chuvash_2006, p. 434]. I have opted for these transcriptions given that @johanson_chuvash_2006 and @kruger_chuvash_1997 did as well.
* /o/ does not seem to be included in the phonemic inventory; however, it is often mentioned as appearing heavily through Russian loanwords (and in the Crubadan corpus) [@johanson_chuvash_2006, p. 434], thus, I will include it in the ruleset.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front       Back
High     "i  y"     "ɯ  u"
Mid      "e  ӗ"       ""
Low        ""       "a  ӑ"
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: The high front and back vowel are ordered unrounded and rounded. Both the mid front and low back vowels are unrounded.", general_title = "")
```    

# Alphabet

* Much of the grapheme to phoneme mapping was inferred from @kruger_chuvash_1997.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme                      Comment
"a"            "/a/"                          ""
"ӑ"            "/ӑ/"                          ""
"в"            "/v/"                          ""
"e"         "/je/; /e/"      "/je/: word-initially; /e/: elsewhere"           
"ӗ"	           "/ӗ/"                          ""                           
"и"            "/i/"                          ""
"й"            "/j/"                          ""
"к"            "/k/"                          ""
"л"            "/l/"                          ""
"м"            "/m/"                          ""
"н"            "/n/"                          ""
"п"            "/p/"                          ""
"о"             "o"               "appears heavily in loanwords"
"р"            "/r/"                          ""
"с"            "/s/"                          ""
"ҫ"            "/ɕ/"                          ""
"т"            "/t/"                          ""
"у"            "/u/"                          ""
"ӳ"            "/y/"                          ""
"х"            "/χ/"                          ""
"ч"            "/tʃ/"                         ""
"ш"            "/ʃ/"                          ""
"ы"            "/ɯ/"                          ""
"э"            "/e/"                          ""
"ю"            "/ju/"                         ""
"я"	           "/ja/"                         ""
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* Voiceless consonants are voiced intervocalically, after sonorants, and before vowels [@johanson_chuvash_2006, p. 435]     
* Consonants palatalize preceding and following front vowels [@johanson_chuvash_2006, p. 435].

# Misc. Rules

* The Chuvash language requires vowel harmony [@johanson_chuvash_2006, p. 436].    
* Gemination can occur with any consonant (specification only on native phonemes) [@dobrovolsky_chuvash_1996, p. 31].    
* Geminate obstruents are voiceless [@dobrovolsky_chuvash_1996, p. 31].

# References
