---
title: "Upper Sorbian"
author: "Allison Shao"
bibliography: hsb.bib
output: html_document
---

Last updated: 2020-03-21

# Background

**Language family**: Indo-European / Slavic / West / Sorbian

* Also referred to as Wendish or Upper Lusatian, it is spoken in eastern Germany and in areas bordering Poland and the Czech Republic.

# Phonology

## Consonants

* /ɡ/, /f/, and /v/ are typically found in loanwords [@Howson2017, p. 4].
* While @Howson2017 represents Upper Sorbian as having an /n~nʲ/ contrast (p. 3), @Iskarous2018 state that Upper Sorbian more accurately has an /n~ɲ/ contrast (p. 52). I have opted for Howson's approach because it is more consistent with how the other palatalized sounds are represented.
* According to @Brijnen1994, there is some debate as to whether the Sorbian languages truly preserve the palatal-nonpalatal distinction (p. 66). The dialects of Sorbian vary in how they preserve the distinction.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"      Labial    Alveolar  Postalveolar   Palatal    Velar   Uvular    Glottal
"Stops (plain)"               "p  b"    "t  d"        ""           ""      "k  ɡ"     ""        ""
"Stops (palatalized)"        "pʲ  bʲ"     ""          ""           ""        ""       ""        ""
Affricates                      ""       "ts"      "tʃ  dʒ"        ""        ""       ""        ""
Fricatives                    "f  v"    "s  z"      "ʃ  ʒ"         ""        "x"      ""        "h"
Nasals                        "m  mʲ"   "n  nʲ"       ""           ""        ""       ""        ""
Trills                          ""        ""          ""           ""        ""     "ʀ  ʀʲ"     ""
Approximants                  "w  wʲ"    "l"          ""           "j"       ""       ""        ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Uvular", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 7)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes (without a diacritic) that share a cell, those on the left are voiceless and those on the right are voiced. For phonemes (with the diacritic (ʲ)) that share a cell, those on the left are (voiced) plain and those on the right are (voiced) palatalized.", general_title = "")
```

## Vowels

* @Howson2017 explains some uncertainty over the vowel system of Upper Sorbian, which has been described as having seven, eight, or ten vowels (p. 5).
    - The main point of controversy is the phonemic status of /ɨ/, and whether it's an allophone of /i/. @Jentsch1980 [as cited in @Brijnen1994, p. 67] states that they are not in complementary distribution, and should be treated as separate phonemes.
    - I have chosen to use the the eight vowel system just as @Howson2017 does (pp. 5-6).

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         "ɨ"          "u"   
Near-High "ɪ"         ""           "ʊ"      
Mid       "e"         ""           "o"
Low       ""          "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```   

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme
"a"	           "/a/"
"b"	           "/b/"
"c"            "/ts/"
"č"            "/tʃ/"
"ć"            "/tʃ/"
"d"	           "/d/"
"e"	           "/e/"
"ě"            "/ɪ/"
"f"            "/f/"
"g"	           "/ɡ/"
"h"	           "/h/"
"i"	           "/i/"
"j"            "/j/"
"k"	           "/k/"
"ł"            "/w/"
"l"	           "/l/"
"m"	           "/m/"
"n"	           "/n/"
"ń"            "/nʲ/"
"o"	           "/o/"
"ó"	           "/ʊ/"
"p"	           "/p/"
"r"            "/ʀ/"
"ř"            "/ʃ/"
"s"	           "/s/"
"š"            "/ʃ/"
"t"	           "/t/"
"u"	           "/u/"
"v"	           "/v/"
"w"	           "/w/"
"y"	           "/ɨ/"
"z"            "/z/"
"ž"            "/ʒ/"
**Digraph**     ""
"dź"	         "/dʒ/"
"ch"	         "/x/"
"bj"           "/bʲ/"
"pj"           "/pʲ/"
"mj"           "/mʲ/"
"nj"           "/nʲ/"
"wj"           "/wʲ/"
"rj"           "/ʀʲ/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* /ʀ/ and /ʀʲ/ are often realized as fricatives [ʁ] and [ʁʲ], respectively. Also, they are sometimes elided [@Howson2017, p. 4].  
* /x/ is often realized as trill [ʀ̥] (ibid.).
* When palatalized, /t/ and /d/ are realized as affricates [tʃʲ] and [dʒʲ], respectively [@Iskarous2018, p. 52].

# Misc. Rules

* Both voiced and voiceless stops can be aspirated [@Howson2017, p. 4].
* Stress is realized as a higher pitch on the stressed syllable; syllables with secondary stress exhibit vowel lengthening [@Howson2017, pp. 6-7].
* In word-final positions, palatal /nʲ/ is decomposed into [jn] [@Howson2017, p. 7].
* All consonants undergo palatalization preceding vowels /e/, /i/, and /ɪ/. However, @Jocz2013 [as cited in @Howson2017, p. 7] states that this is an optional process, given that the contrast is neutralized. Based on this, I have chosen to transcribe the phonemic palatalized consonants only using the digraphs listed above [@Howson2017, p. 4].

# References  
