---
title: "Bislama"
author: "Becky Mathew"
bibliography: bi.bib
output: html_document
---

# Background

**Language family**: Creole, English-based, Pacific

* Bislama is a relatively new language (from the mid 19th century) that is the national language of Vanuatu, a Pacific island country. Bislama is a creole language that has some basis in English [@crowley2004, pp. 1-2].

## Consonants

* Not all speakers make a contrast between voiced and voiceless stops. Not all speakers make a contrast between /v/ and /f/ [@crowley2004, p. 14].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"    Labial     Alveolar   Postalveolar  Palatal    Velar      Glottal
Stops                       "p  b"      "t  d"       ""          ""       "k  ɡ"        ""
Affricates                    ""          ""        "tʃ"         ""         ""          ""
Fricatives                  "f  v"        "s"        ""          ""         ""          "h"
Nasals                        "m"         "n"        ""          ""         "ŋ"         ""
Flaps                         ""          "ɾ"        ""          ""          ""         ""        
Approximants                  "w"         "l"        ""          "j"         ""         ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```

## Vowels

* Although diphthongs are apparent in Bislama, there is much controversy around them being seen as complex phonemes or vowel sequences. I have chosen not to include them in the inventory based on the argument that they are vowel sequences [@Lynch1975, p. 195].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
       Front    Central     Back
High    "i"       ""        "u"
Mid     "e"       ""        "o"
Low     ""        "a"       ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme
"a"            "/a/"
"b"            "/b/"
"d"            "/d/"
"e"            "/e/"
"f"            "/f/"
"g"            "/ɡ/"
"h"            "/h/"
"i"            "/i/"
"j"           "/tʃ/"
"k"           "/k/"
"l"           "/l/"
"m"           "/m/"
"n"           "/n/"
"o"           "/o/"
"p"           "/p/"
"r"           "/ɾ/"
"s"           "/s/"
"t"           "/t/"
"u"           "/u/"
"v"           "/v/"
"w"           "/w/"
"y"           "/j/"
**Digraphs**   ""
"ng"           "ŋ"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* /h/ is frequently deleted intervocalically [@crowley2004, p. 14].

# Misc. Rules

* /h/ may often be epenthesized preceding vowels in word-initial positions [@crowley2004, p. 14].

# References
