---
title: "Kayabí"
author: "Emily Strand"
bibliography: kyz.bib
output: html_document
---

Last Updated: 2020-06-30


**SLIGHTLY COMPROMISED: possible conflation between /i/ and /j/ due to syllable structure**


# Background

**Language Family:** Tupi / Tupi-Guarani / Kayabi

* Kayabí is spoken in Brazil.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial  Alveolar  Palatal   Velar    Glottal
Stops                      "p"     "t"         ""     "k kʷ ɡ"    "ʔ"
Fricatives                 "f"     "s"         ""       ""        ""
Nasals                     "m"     "n"         ""       "ŋ"       ""
Flaps                      ""      "ɾ"         ""       ""        ""
Approximants               "w"     ""          "j"      ""        ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 5)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: The velar stop with the diacritic (ʷ) is labialized.", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* Kayabí makes a distinction between oral and nasal vowels [@Souza2004, p. 24].
* It appears that like-vowel sequences appear [based on vocab in @Dobson1997]; however, vowel length is never mentioned as a distinctive feature. They will be treated as sequences.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
          Front   Central   Back
High      "i"     "ɨ"       "u"
Mid       "e"     ""        "o"
Low       ""      "a"       ""
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

* **/ŋ/ is not represented in the Crúbadán corpus.**

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme            Comment
"a"              "/a/"               ""
"ã"              "/ã/"               ""
"e"              "/e/"               ""
"ẽ"              "/ẽ/"               ""
"f"              "/f/"               ""
"g"              "/ɡ/"               ""
"i"            "/i/; /j/"    "/j/: syllable-finally"
"ĩ"              "/ĩ/"               ""
"j"              "/j/"               ""
"k"              "/k/"               ""
"m"              "/m/"               ""
"n"              "/n/"               ""
"o"              "/o/"               ""
"õ"              "/õ/"               ""
"p"              "/p/"               ""
"r"              "/ɾ/"               ""
"s"              "/s/"               ""
"t"              "/t/"               ""
"u"              "/u/"               ""
"ũ"              "/ũ/"               ""
"w"              "/w/"               ""
"y"              "/ɨ/"               ""
"ỹ"              "/ɨ̃/"               ""
"\'"             "/ʔ/"               ""
**Digraph**       ""                 ""
"kw"             "/kʷ/"              ""
"ng"             "/ŋ/"               ""
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Syllables in Kayabí follow the order of (C)V(C) [@Souza2004, p. 26].
    * Given the varying syllable structures, /j/ can't always be transcribed accurately. The rules take into account the environment in which ⟨i⟩ goes to /j/ when followed by CV; however, this doesn't encompass all the environments in which ⟨i⟩ becomes /j/. Thus, the language is slightly compromised.

# Misc. Rules

* /p/, /t/, and /k/ become [w], [ɾ], and [ɡ], respectively, when preceding a vowel-initial morpheme [@Souza2004, p. 27].
* Consonants may also weaken when morpheme-initial [@Souza2004, pp. 27-28]:
    * /p/, /f/, and /m/ may become [w].
    * /n/ and /t/ may become [ɾ].
    * /k/ may become [ŋ].
* Morpheme-final consonants may become elided when preceding a morpheme-initial consonant [@Souza2004, p. 28].

# References
