---
title: "Amele"
author: "Bill Mizgerd"
bibliography: aey.bib
output: html_document
---

Last Updated: 2019-09-23

# Background

**Language Family**: Trans-New Guinea / Madang / Croisilles / Mabuso / Gum / Amele

* Amele is spoken on the Madang Province of Papua New Guinea.

# Phonology

## Consonants

* The Huar dialect includes /r/, which does not exist in any other dialect of Amele [@AeyRoberts2016, p. 81].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial  Alveolar  Palatal  Velar  Labiovelar  Glottal
Stops                       "b"     "t d"     ""       "k ɡ"  "ɡb"        "ʔ"
Fricatives                  "f"     "s"       ""       ""     ""          "h"
Nasals                      "m"     "n"       ""       ""     ""          ""
Approximants                ""      "l"       "j"      ""     "w"         ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Palatal", "Velar", "Labiovelar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* Amele contains the following diphthongs: /ai/, /au/, /ei/, /eu/, /oi/, /ou/; @AeyRoberts2016 interprets these as sequences of consecutive vowels rather than as complex phonemes. Other sequences of vowels can also occur (including consecutive identical vowels), but only across syllable boundaries [@AeyRoberts2016, pp. 66-68]. (@AeyRoberts1998 also argues for the existence of /ae/ and /ao/ as diphthongs (p. 3), but he seems to have moved away from that analysis by 2016.)
* Mid vowels sometimes lengthen, but @AeyRoberts2016 finds the length to be a realization of an underlying sequence of identical vowels (p. 75).


```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
       Front   Central   Back
High   "i"     ""        "u"
Mid    "e"     ""        "o"
Low    ""      "a"       ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

* The alphabet used is of the modern Amele orthography (developed ca. 1990), as the older orthography (developed in the 1920s) has several ambiguities; the modern orthography is also the one used by Crubadan, making it all the more preferable.
* Older orthography also uses ⟨p⟩, but because [p] is non-phonemic in Amele, I have excluded ⟨p⟩ from my ruleset.


```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme                Comment
  "a"          "/a/"                   ""
  "b"          "/b/"                   ""
  "c"          "/ʔ/"                   ""
  "d"          "/d/"                   ""
  "e"          "/e/"                   ""
  "f"          "/f/"                   ""
  "g"          "/ɡ/"                   ""
  "h"          "/h/"                   ""
  "i"          "/i/"                   ""
  "j"          "/j/"                   ""
  "k"          "/k/"                   ""
  "l"          "/l/"                   ""
  "m"          "/m/"                   ""
  "n"          "/n/"                   ""
  "o"          "/o/"                   ""
  "q"          "/ɡb/"                  ""
  "r"          "/r/"       "only used in Huar dialect"  
  "s"          "/s/"                   ""
  "t"          "/t/"                   ""
  "u"          "/u/"                   ""
  "w"          "/w/"                   ""
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* In some words, /s/ alternates with [h] when adjacent to high vowels [@AeyRoberts2016, p. 56].

# Misc. Rules

* Word-final non-alveolar stops devoice, if the word is multisyllabic; contrast between /ɡb/ and /b/ also collapses in such position, with both phonemes being realized as [p] [@AeyRoberts2016, pp. 52-53, 64].
* /j/ goes to [ʒ] in certain circumstances: word-initially, when surrounded on both sides by /a/, word-finally after /i/, and intervocalically before /i/ [@AeyRoberts2016, p. 57].
* Word-final /w/ is realized as [ʋ] [@AeyRoberts2016, p. 64].
* Consonant clusters can only occur word-medially [@AeyRoberts2016, p. 65].
* Phonetically lax /a/ lengthens before voiced stops in monosyllabic words [@AeyRoberts2016, p. 76].

# References
