---
title: "Belarusan"
author: "Emily Strand"
bibliography: be.bib
output: html_document
---

Last Updated: 2019-08-28

# Background

**Language family**: Indo-European / Slavic / East

* Also referred to as Belarusian or White Russian, it is spoken in Belarus, parts of Russia, Poland, and Ukraine.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"        Labial          Alveolar       Postalveolar     Palatal        Velar
"Stops (plain)"                "p  b"           "t  d"             ""             ""            "k"            
"Stops (palatalized)"          "pʲ  bʲ"           ""               ""             ""            "kʲ"
"Affricates (plain)"             ""             "ts  dz"        "tʃ  dʒ"          ""            ""     
"Affricates (palatalized)"       ""            "tsʲ  dzʲ"          ""             ""            ""
"Fricatives (plain)"           "f  v"            "s  z"          "ʃ  ʒ"           ""           "x  ɣ"       
"Fricatives (palatalized)"     "fʲ  vʲ"          "sʲ  zʲ"          ""             ""          "xʲ  ɣʲ"
"Nasals (plain)"                "m"               "n"              ""             ""            ""
"Nasals (palatalized)"          "mʲ"              "nʲ"             ""             ""            ""
Trills                          ""                "r"              ""             ""            ""
"Approximants (plain)"          ""                "l"              ""             "j"           ""
"Approximants (palatalized)"    ""               "lʲ"              ""             ""            ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 5)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```

## Vowels

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         ""           "u"
Mid       "ɛ"         ""           "o"
Low       ""          "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```


# Alphabet

* ⟨Ґ⟩, representing /ɡ/, was officially removed from the Belarusan language during the orthography reform in 1933 [@Bulk_2008, p. 2].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme                  Comment
"а"            "/a/"                     ""
"б"            "/b/"                     ""
"в"            "/v/"                     ""
"г"            "/ɣ/"                     ""
"д"            "/d/"                     ""
"е"            "/jɛ/"                    ""
"ё"            "/jo/"                    ""
"ж"            "/ʒ/"                     ""
"з"            "/z/"                     ""
"i"            "/ji/"                    ""
"й"            "/j/"                     ""
"к"            "/k/"                     ""
"л"            "/l/"                     ""
"м"            "/m/"                     ""
"н"            "/n/"                     ""
"о"            "/o/"                     ""
"п"            "/p/"                     ""
"р"            "/r/"                     ""
"с"            "/s/"                     ""
"т"            "/t/"                     ""
"у"            "/u/"                     ""
"ў"            "/u/" "realized as [w], but understood as an allophone of /u/ [@stolz_human_2012, p. 73]"  
"ф"            "/f/"                     ""
"х"            "/x/"                     ""
"ц"           "/ts/"                     ""
"ч"           "/tʃ/"                     ""
"ш"            "/ʃ/"                     ""
"ы"            "/i/"  "realized as [ɨ], but understood as an allophone of /i/ [@stolz_human_2012, p. 73]"
"ь"             ""       "indicates palatalization of the preceding consonant"    
"э"            "/ɛ/"                     ""
"ю"            "/ju/"                    ""
"я"            "/ja/"                    ""
"\'"              ""         "indicates separation of sounds (non-palatalization of preceding consonant)"
**Digraph**     ""                       ""
"дж"           "/dʒ/"                    ""        
"дз"           "/dz/"                    ""
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```   

# Lenition Rules

* Within the Slavic languages, it's common for /ɡ/ to lenite to /ɣ/ [@klein_86_2018, p. 1607]; although it seems as if Belarusan has permanently implemented this change.

# Misc Rules

* Palatalized consonants precede ⟨я⟩, ⟨е⟩, ⟨ё⟩, ⟨ю⟩, ⟨i⟩ [@stolz_human_2012, p. 73].  
* Gemination of some consonants occurs [@bocek_relationship_2012, p. 166].

# References
