---
title: "Abkhaz"
author: "Bill Mizgerd"
bibliography: ab.bib
output: html_document
---

Last Updated: 2019-07-05

# Background

**Language Family**: Northwest Caucasian / Northwest / Abkhaz-Abazin

* Abkhaz is spoken within Abkhazia, located in Georgia along the Black Sea.
* Where dialects diverge, my ruleset will prefer Standard Abkhaz.

# Phonology

## Consonants

* @AbChirikba2003 depicts the alveolar series of consonants as dental (p. 18).
* @AbChirikba2003 argues for the plain voiceless obstruents as aspirated (p. 19); @AbHewitt1979 observes aspiration as well, but regards it as a non-distinctive part of the phoneme (p. 256).
* @AbKoryakov2002 and @AbHewitt1979 argue for velar rather than uvular fricatives (p. 7; p. 258).
* The sources have diverged somewhat on what sibilants are present in Abkhaz. @AbCatford1977 lists both the /ʃ ʒ/ and /ɕ ʑ/ series as present in Abkhaz (p. 291); however, the consensus among other sources seems to be that /ɕ ʑ/ has only been preserved in a subset of Abkhaz dialects, and has been lost in Standard Abkhaz. @AbChirikba2003 and @AbHewitt1979 describe the postalveolar sibilants as being produced as retroflex (p. 19; pp. 256, 257); however, this view is not widely attested elsewhere, so I have preferred the more common /ʃ ʒ/ analysis. @AbHewitt1979 also follows Catford in regarding /tɕ dʑ/ as phonemic (p. 256), but nevertheless I have continued to follow the more recent analyses that regard those phonemes as lost in Standard Abkhaz.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial        Alveolar        Postalveolar     Palatal       Velar        Uvular     Pharyngeal
"Stops (plain)"          "p  b  pʼ"     "t  d  tʼ"           ""             ""       "k  ɡ  kʼ"        "qʼ"        "ʡ"
"Stops (labialized)"         ""       "tʷ  dʷ  tʷʼ"          ""             ""      "kʷ  ɡʷ  kʷʼ"     "qʷʼ"        ""
"Stops (palatalized)"        ""            ""                ""             ""      "kʲ  ɡʲ  kʲʼ"     "qʲʼ"        ""
"Affricates (plain)"         ""       "ts  dz  tsʼ"     "tʃ  dʒ  tʃʼ"       ""           ""             ""         ""
"Affricates (palatalized)"   ""            ""          "tʃʲ  dʒʲ  tʃʲʼ"     ""           ""             ""         ""
"Fricatives (plain)"       "f  v"        "s  z"            "ʃ  ʒ"           ""           ""           "χ  ʁ"       "ħ"
"Fricatives (labialized)"    ""            ""             "ʃʷ  ʒʷ"          ""           ""          "χʷ  ʁʷ"      "ħʷ"
"Fricatives (palatalized)"   ""            ""             "ʃʲ  ʒʲ"          ""           ""          "χʲ  ʁʲ"      ""
"Nasals"                    "m"           "n"                ""             ""           ""             ""         ""
"Trills"                     ""           "r"                ""             ""           ""             ""         ""
"Approximants"              "w"           "l"                ""             "j"          ""             ""         ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Uvular", "Pharyngeal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 7)) %>%
  footnote(general = "Note: From left to right, phonemes that share a cell are voiceless, voiced, and ejective. If a cell only contains two phonemes, they are voiceless and voiced.", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* Some authors argue for the existence of /aː/, but @AbChirikba2003 and @AbHewitt1979 assert that [aː] is just a realization of the sequence /ħ a/ (p. 20; p. 263).
* [o e i u] all surface allophonically in certain environments [@AbChirikba2003, p. 21; @AbHewitt1979, pp. 259-260].
* The Tapanta dialect "marginally preserves" a tonal system from older forms of Abkhaz, but all other dialects - including Standard Abkhaz - have lost it [@AbChirikba1996, p. iii].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

consonants <- read.table(textConnection('
        Central
  Mid    "ə"
  Low    "a"
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered")%>%
  column_spec(1, bold = TRUE)
```

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('

Grapheme     Phoneme     Comment   
"а"           "/a/"        ""
"б"           "/b/"        ""
"в"           "/v/"        ""
"г"           "/ɡ/"        ""
"ӷ"           "/ʁ/"        "⟨ҕ⟩ may be used instead"
"д"           "/d/"        ""
"е"           "/a/"        ""
"ж"           "/ʒ/"        ""
"з"           "/z/"        ""
"ӡ"           "/dz/"       ""
"и"           "/j/"        ""
"к"           "/kʼ/"       ""
"қ"           "/k/"        ""
"ҟ"           "/qʼ/"       ""
"л"           "/l/"        ""
"м"           "/m/"        ""
"н"           "/n/"        ""
"о"           "/a/"        ""
"п"           "/pʼ/"       ""
"ԥ"           "/p/"        "⟨ҧ⟩ may be used instead"
"р"           "/r/"        ""
"с"           "/s/"        ""
"т"           "/tʼ/"       ""
"ҭ"           "/t/"        ""
"у"           "/w/"        ""
"ф"           "/f/"        ""
"х"           "/χ/"        ""
"ҳ"           "/ħ/"        ""
"ц"           "/ts/"       ""
"ҵ"           "/tsʼ/"      ""
"ч"           "/tʃʲ/"      ""
"ҷ"           "/tʃʲʼ/"     ""
"ҽ"           "/tʃ/"       ""
"ҿ"           "/tʃʼ/"      ""
"ш"           "/ʃ/"        ""
"ы"           "/ə/"        ""
"ҩ"           "/ʡ/"        ""
"џ"           "/dʒ/"       ""
"ь"             ""   "marks palatalization"  
"ә"             ""   "marks labialization"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* Aspiration and ejectivization of consonants are only weakly realized [@AbChirikba2003, p. 19].
* /qʼ/ debuccalizes to [ʔ] for many speakers (ibid.).
* Some speakers spirantize the consonant in the word /apʼa/, yielding [afʼa] [@AbHewitt1979, p. 257].
* For some speakers, /ħʷ/ approximantizes to the labiopalatal approximant, [ɥ] [@AbHewitt1979, p. 258].

# Misc. Rules

* Labialized alveolar stops are often realized with full lip closure, causing them to be produced as [db], [tp], [tpʼ] [@AbChirikba2003, p. 19].
* The semivowels /w/ and /j/ have vocalic allophones ([u] and [i] respectively) (ibid.).

# References
