---
title: "Kannada"
author: "Bill Mizgerd"
bibliography: kn.bib
output: html_document
---

Last Updated: 2019-08-01

# Background

**Language Family**: Dravidian / South Dravidian / Tamil-Kannada / Badaga-Kannada / Kannada

* Kannada is spoken in southwester India, specifically in the states of Goa, Karnataka, Andhra Pradesh, Anantapur, Maharashtra, Tamil Nadu, and Kerala.

# Phonology

## Consonants

* /f/ and /z/ appear in loanwords [@KnKrishnamurti2003, p. 67].
* /r/ seems to exist in free variation with [ɾ] [@KnSridhar1990, p. 294; @KnSchiffman1983, p. 15].
* Consonants have contrastive gemination [@KnSridhar1990, pp. 299-301; @KnSchiffman1983, p. 8].
* @KnAndronov1969 argues for /ɲ/ and /ŋ/ as phonemic (p. 21), but this view seems to have fallen out of favor.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial     Dental   Alveolar    Retroflex   Postalveolar    Palatal      Velar    Glottal
"Stops (plain)"            "p  b"     "t̪  d̪"      ""        "ʈ  ɖ"          ""            ""       "k  ɡ"      ""
"Stops (aspirated)"        "pʰ  bʱ"   "t̪ʰ  d̪ʱ"    ""        "ʈʰ  ɖʱ"        ""            ""       "kʰ  ɡʱ"    ""
"Affricates (plain)"          ""         ""        ""          ""        "tʃ  dʒ"          ""         ""        ""
"Affricates (aspirated)"      ""         ""        ""          ""        "tʃʰ  dʒʱ"        ""         ""        ""
Fricatives                    ""         ""        "s"         "ʂ"          "ʃ"            ""         ""        "h"  
Nasals                       "m"         ""        "n"         "ɳ"          ""             ""         ""        ""        
Trills                        ""         ""        "r"          ""          ""             ""         ""        ""
Approximants                  "ʋ"        ""        "l"         "ɭ"          ""             "j"        ""        ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Dental", "Alveolar", "Retroflex", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 8)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```

## Vowels

* /æ/ appears in some loanwords [@KnKrishnamurti2003, p. 67], as does /ɔ/ [@KnSchiffman1983, p. 3].
* All native vowels can be contrastively lengthened [@KnKrishnamurti2003, p. 66].
* /aj/ and /aʋ/ occur in Kannada, but they are sequences of discrete phonemes rather than diphthongs [@KnSridhar1990, p. 296].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         ""           "u"
Mid       "e"         ""           "o"
Low        ""         "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```   

# Alphabet

* There is no diacritic for /a/ because it is the inherent vowel on consonant characters.
* In loanwords with /æ/ or /ɔ/, those vowels are written with the characters for /eː/ and /oː/ respectively [@KnSchiffman1983, p. 4].
* Some Sanskrit loanwords feature syllabic /r/; however, in Kannada the /r/ is not considered syllabic. Rather, it's followed by either /u/ or /i/ [@KnKrishnamurti2003, p. 84]. I have opted to prefer to add /u/, since in my experience it is likelier; however, it is not represented in the rule set.


```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
"Consonant Grapheme"      Phoneme                    Comment
"ಕ"                       "/ka/"                       ""
"ಖ"                       "/kʰa/"                      ""
"ಗ"                       "/ɡa/"                       ""
"ಘ"                       "/ɡʱa/"                      ""
"ಙ"                       "/na/"    "allophonically velar [@KnSchiffman1983, p. 9]"  
"ಚ"                      "/tʃa/"                       ""
"ಛ"                      "/tʃʰa/"                      ""
"ಜ"                      "/dʒa/"                       ""
"ಝ"                    "/dʒʱa/"                       ""
"ಞ"                      "/na/"    "allophonically palatal [@KnSchiffman1983, p. 10]"  
"ಟ"                      "/ʈa/"                        ""
"ಠ"                      "/ʈʰa/"                       ""
"ಡ"                      "/ɖa/"                        ""
"ಢ"                     "/ɖʱa/"                        ""
"ಣ"                      "/ɳa/"                        ""
"ತ"                      "/t̪a/"                        ""
"ಥ"                     "/t̪ʰa/"                        ""
"ದ"                     "/d̪a/"                         ""
"ಧ"                     "/d̪ʱa/"                        ""
"ನ"                     "/na/"                         ""
"ಪ"                     "/pa/"                         ""
"ಫ"                     "/pʰa/"                        ""
"ಬ"                     "/ba/"                         ""
"ಭ"                     "/bʱa/"                        ""
"ಮ"                     "/ma/"                        ""
"ಯ"                    "/ja/"                        ""
"ರ"                     "/ra/"                         ""
"ಲ"                     "/la/"                         ""
"ವ"                     "/ʋa/"                        ""
"ಶ"                     "/ʃa/"                        ""
"ಷ"                     "/ʂa/"                        ""
"ಸ"                     "/sa/"                        ""
"ಹ"                    "/ha/"                         ""
"ಳ"                     "/ɭa/"                        ""
"**Vowel Grapheme**"      ""                          ""
"ಅ"                     "/a/"                         ""
"ಆ"                     "/aː/"                        ""
"ಇ"                     "/i/"                         ""
"ಈ"                     "/iː/"                        ""
"ಉ"                     "/u/"                        ""
"ಊ"                    "/uː/"                       ""
"ಋ"                    "/ru/"              "occurs only in loanwords"  
"ೠ"                   "/ruː/"           "obsolete; occurs only in loanwords"  
"ಎ"                     "/e/"                         ""
"ಏ"                    "/eː/"                         ""
"ಐ"                    "/aj/"                         ""
"ಒ"                     "/o/"                         ""
"ಓ"                    "/oː/"                         ""
"ಔ"                    "/aʋ/"                         ""
"**Vowel Diacritic**"     ""                           ""
"್"                      "∅"              "marks consonant-final syllables"  
"ಾ"                    "/aː/"                         ""
"ಿ"                     "/i/"                          ""
"ೀ"                    "/iː/"                         ""
"ು"                    "/u/"                          ""
"ೂ"                   "/uː/"                         ""
"ೃ"                    "/ru/"               "occurs only in loanwords"  
"ೄ"                   "/ruː/"              "occurs only in loanwords"  
"ೆ"                    "/e/"                           ""
"ೇ"                   "/eː/"                          ""
"ೈ"                    "/aj/"                         ""
"ೊ"                  "/o/"                           ""
"ೋ"                  "/oː/"                         ""
"ೌ"                  "/aʋ/"                          ""
"**Diacritic Variations**"    ""                    ""
"ಂ"                   ""          "anusvara - nasal with the same place as the following consonant [@KnKrishnamurti2003, p. 84]"  
"ಃ"                 "/h/; /hV/"               "visarga - /h/: after long vowels; /hV/: after short vowels"
'), TRUE)

kable(alphabet, col.names = c("Consonant Grapheme", "Phoneme", "Comment"), align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Kannada syllable structure is (C)(C)(C)V(C)(C)(C) [@KnSridhar1990, p. 298].

# Lenition Rules

* Word-final short vowels often delete [@KnSridhar1990, p. 307].
* Word-medial short vowels sometimes delete, especially in casual speech; this happens most often in a penultimate syllable, and never in the first or last syllable of a word (ibid.).
* Consonants degeminate when followed by an unlike consonant [@KnSridhar1990, p. 308].
* Preconsonantal stops delete after a nasal with the same place of articulation (ibid.).
* Intervocalically or post-nasally, /d̪/ spirantizes to [ð] [@KnSchiffman1983, p. 12].

# Misc. Rules

* Epenthesis of [u] typically occurs after word-final consonants (although if the consonant is /j/, then [i] will be inserted instead) [@KnSridhar1990, p. 296].
* The final consonants of monosyllabic words undergo gemination [@KnSridhar1990, p. 310].

# References
