---
title: "Nobonob"
author: "Emily Strand"
bibliography: gaw.bib
output: html_document
---

Last updated: 2019-11-06

# Background

**Language Family**: Trans-New Guinea / Madang-Adelbert Range / Madang / Mabuso / Hanseman

* Nobonob is spoken in the Madang province of Papua New Guinea.

# Phonology

## Consonants

* /r/ and /s/ appear only in loanwords [@aeschliman_nobonob_1992, p. 4].
* Nobonob makes an interesting four way distinction between stops. That is, not only does it have the standard voiceless and voiced series, it also has a voiceless geminate series and a voiced implosive series [@aeschliman_nobonob_1992, p. 1].
* @reesink_structures_1987 states that prenasalized stops are present in the language, which is consistent with others within the language family (p. 12); however it seems that such consonants are very marginal. Also, given that other documentation on the matter could not be found, I have opted for the inventory described in @aeschliman_nobonob_1992.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial     Alveolar    Palatal     Velar    Glottal
"Stops (voiceless)"        "p  pː"     "t  tː"      ""       "k  kː"     ""     
"Stops (voiced)"           "b  ɓ"      "d  ɗ"       ""       "ɡ  ɠ"      ""
Fricatives                  ""           ""         ""         ""        "h"
Nasals                      "m"         "n"         ""         ""        ""
Approximants                "w"         "l"         "j"        ""        ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 5)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: The voiceless stops are ordered plain and geminate and the voiced stops are ordered plain and implosive.", general_title = "")
```

## Vowels

* Vowel length is contrastive in Nobonob [@aeschliman_nobonob_1992, p. 4]. Long vowels are indicated by repeated graphemes.
* Just like other Trans-New Guinea languages, Nobonob has an abundance of vowel sequences [@aeschliman_nobonob_1992, p. 3]. It is, however, unclear whether they are realized as complex phonemes or independent vowels. Therefore, they will not be transcribed in the rules.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         ""           "u"
Mid       "e"         ""           "o"
Low        ""         "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```   

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme
"a"             "a"
"b"             "b"
"q"             "ɓ"
"d"             "d"
"ḏ"             "ɗ"
"e"             "e"
"g"             "ɡ"
"g̱"             "ɠ"
"h"             "h"
"i"             "i"
"k"             "k"
"ḵ"             "kː"
"l"             "l"
"m"             "m"
"n"             "n"
"o"             "o"
"p"             "p"
"p̱"             "pː"
"t"             "t"
"ṯ"             "tː"
"u"             "u"
"w"             "w"
"y"             "j"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* Following a vowel or a nasal consonant, /ɓ/ debuccalizes [@aeschliman_nobonob_1992, p. 3].

# Misc. Rules

* /l/ is rounded preceding /u/ [@aeschliman_nobonob_1992, p. 4].
* /ɗ/ is realized as [t̪ʰ] in some dialects (ibid.).

# References
