---
title: "Usarufa"
author: "Jordan Schuster"
bibliography: usa.bib
output: html_document
---

Last Updated: 2019-11-15

# Background

**Language Family:** Trans-New Guinea, East New Guinea Highlands, Eastern (Kainantu), Gadsup-Auyana-Awa

* Usarufa is spoken in the Eastern Highlands province of Papua New Guinea.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"    Labial       Alveolar    Palatal     Velar      Glottal    
Stops                        "p"           "t"        ""          "k"         "ʔ"
Nasals                       "m"           "n"        ""          ""          ""
Trills                       ""            "r"        ""          ""          ""
Approximants                 "w"           ""         "j"         ""          ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 5)) %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* There are three phonemic tones superimposed on vowels [@bee1965, p. 59]:
    - High: marked by an acute accent (´)
    - Low: unmarked
    - Falling: marked by a circumflex (ˆ)

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
             Front      Central       Back
High          "i"         ""           "u"
Mid           "e"         ""           "o"
Low           ""          "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```     

# Alphabet

* /ʔ/ is left unwritten when occurring before /p/, /t/ and /k/ because these sounds occur word medially only after the glottal stop [@chenoveth2004, p. 3]. Disagreement exists around these series as clusters (i.e., /ˀp/, /ˀt/, /ˀk/) or as independent phonemes. I have chosen to see them as independent, allowing the syllable structure to be left unaffected [@bee1965, pp. 42-44].
* Similarly, there is disagreement about nasal clusters being long (/mː/, /nː/) or separate. Again, I have chosen to see them as separate [@bee1965, pp. 42-44].
* /p/ is written ⟨p⟩ word-initially and after a glottal stop; /p/ is written ⟨b⟩ when it occurs intervocalically [@chenoveth2004, p. 3].
* /k/ is written ⟨k⟩ word initially and after a glottal stop; /k/ is written ⟨g⟩ when it occurs intervocalically [@chenoveth2004, p. 3].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme
"a"              "/a/"
"b"              "/p/"
"p"              "/p/"
"e"              "/e/"
"g"              "/k/"
"k"              "/k/"
"i"              "/i/"
"m"              "/m/"
"n"              "/n/"
"o"              "/o/"
"q"              "/ʔ/"
"r"              "/r/"
"t"              "/t/"
"u"              "/u/"
"w"              "/w/"
"y"              "/j/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Syllables in Usarufa include the following structures [@chenoveth2004, p. 3]:
    - V
    - VC
    - CV
    - CVC
* Vowel sequences occur as separate syllables [@chenoveth2004, p. 2].
* Consonants do not occur word-finally [@chenoveth2004, p. 3].

# Lenition Rules

* /p/ and /k/ spirantizes to [β] and [ɣ], respectively, intervocalically [@chenoveth2004, p. 3].
* In some words /t/ varies freely with [s] (ibid.).

# Misc. Rules

* /r/ is realized as [l] following /i/ and /a/ [@chenoveth2004, p. 3].

# References
