---
title: "Amanab"
author: "Bill Mizgerd"
bibliography: amn.bib
output: html_document
---

Last Updated: 2019-07-24

# Background

**Language Family**: (Trans-New Guinea - debated) / Border / Waris / Amanab

* Amanab is spoken on the Sandaun province of Papua New Guinea.

# Phonology

## Consonants

* @AmnMinch1991 argues against the presence of /ⁿd/ (p. 107).
* @AmnSIL2001 argues for /ɸ/ rather than /f/, as well as for a series of /ᵑɡ/ and /ɣ/ rather than /ŋ/ and /ᵑɡ/ (p. 1).
* @AmnPalmer2017 presents a general consonant inventory of the Border family, rather than for Amanab specifically, so his inventory features a handful of differences from the other sources.
* @AmnSIL2001 includes /l/ as a phoneme (p. 1), but it seems to actually be an allophone of /ɾ/ - even later in the same paper, the author lists [l] and [ɾ] as occurring in free variation (p. 4).

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"    Labial   Alveolar   Palatal   Velar   Glottal
Stops                       "p  ᵐb"   "t  ⁿd"     ""     "k  ᵑɡ"    ""
Fricatives                   "f"       "s"        ""       ""       "h"
Nasals                       "m"       "n"        ""       "ŋ"      ""
Flaps                        ""        "ɾ"        ""       ""       ""
Approximants                 "w"       ""         "j"      ""       ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Palatal", "Velar", "Glottal", align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 5)) %>%
  footnote(general = "Phonemes to the left of each cell are voiceless and phonemes to the right of each cell are voiced", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* @AmnSIL2001 argues for /ɑ/ and /ɛ/ rather than /a/ and /e/ (p. 1).

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
vowels <- read.table(textConnection('
       Front   Central   Back
High   "i"     ""        "u"
Mid    "e"     ""        "o"
Low    ""      "a"       ""
'), TRUE)


kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
alphabet <- read.table(textConnection('

Grapheme     Phoneme       Comment
  "a"        "/a/"           ""
  "b"        "/ᵐb/"          ""
  "d"        "/ⁿd/"         "rare"  
  "e"        "/e/"           ""
  "f"        "/f/"           ""
  "g"        "/ᵑɡ/"          ""
  "h"        "/h/"           ""
  "i"        "/i/"           ""
  "k"        "/k/"           ""
  "l"        "/ɾ/"           ""
  "m"        "/m/"           ""
  "n"        "/n/"           ""
  "o"        "/o/"           ""
  "p"        "/p/"           ""
  "r"        "/ɾ/"           ""
  "s"        "/s/"           ""
  "t"        "/t/"           ""
  "u"        "/u/"           ""
  "w"        "/w/"           ""
  "y"        "/j/"           ""
  Digraph     ""             ""
  "ng"       "/ŋ/"           ""
  '), TRUE)


  kable(alphabet, align = 'c') %>%
    kable_styling("bordered") %>%
    pack_rows("Digraph", 21)
  ```

# Lenition Rules

* Word-medial /ᵑɡ/ sometimes spirantizes to [ɣ] [@AmnSIL2001, p. 4].
* Word-medial /f/ sometimes voices to [v] (ibid.).
* /x/, which is attested in other languages of the Border family [@AmnPalmer2017, p. 327], has historically debuccalized to [h] in Amanab [@AmnSIL2001, p. 4].

# Misc. Rules

* Word-final /k/ has a lateral release [@AmnSIL2001, p. 4].

# References
