---
title: "Fula"
author: "Bill Mizgerd"
bibliography: ff.bib
output: html_document
---

Last Updated: 2019-07-17

# Background

**Language Family**: Niger-Congo / Atlantic-Congo / Atlantic / Northern Atlantic / Senegambian / Fula-Wolof / Fula-Serer / Fula

* The Fula language comprises a broad dialect continuum; I have opted to put together information that, as much as possible, reflects the most common cross-dialectal patterns.
* Other names for the language include Fulani and Fulfulde.
* Fula is spoken in parts of Western and Central Africa.

# Phonology

## Consonants

* @FfPelletier1981 argues for /tʃ/ and /dʒ/ (p. 3), while @FfMukoshy1984 argues for /c/ and /ɟ/ (p. 46). I have chosen to prefer Mukoshy's analysis, since his work analyzes a greater range of Fula dialects.
* @FfMukoshy1984 interprets the prenasalized stops as simply sequences of two consonants, rather than as single distinct phonemes (pp. 48-49).
* /ᵑɡ/ and /ŋ/ are mutually exclusive between dialects; a dialect of Fula may have one or the other of those phonemes, but none have both [@FfMukoshy1984, p. 49].
* Some western dialects debuccalize /w/ to [ʔ]; the Adamawa dialect realizes /w/ as [v] [@FfMukoshy1984, p. 54].
* The Adamawa dialect collapses the distinction between /ʔʲ/ and /ʔ/ [@FfMukoshy1984, p. 57].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"    Labial     Alveolar  Palatal   Velar     Glottal
Stops                       "p b ᵐb"   "t d ⁿd"  "c ɟ ᶮɟ"  "k ɡ ᵑɡ"  "ʔ ʔʲ"
Implosives                  "ɓ"        "ɗ"       ""        ""        ""
Fricatives                  "f"        "s"       ""        ""        "h"
Nasals                      "m"        "n"       "ɲ"       "ŋ"       ""
Trills                      ""         "r"       ""        ""        ""
Approximants                "w"        "l"       "j"       ""        ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 5)) %>%
  footnote(general = "Note: From left to right, phonemes in a cell are voiceless, voiced, and prenasalized; the exception is in the glottal column, where phonemes are plain (left) and palatalized (right).", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* Vowel length is contrastive in Fula [@FfMukoshy1984, p. 49].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
       Front   Central   Back
High   "i"     ""        "u"
Mid    "e"     ""        "o"
Low    ""      "a"       ""
'), TRUE)

kable(consonants, align = ('c')) %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

* Long vowels and geminate consonants are written by doubling the appropriate character [@FfPelletier1981, pp. 14, 23].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme               Comment
"a"            "/a/"                  ""
"b"            "/b/"                  ""
"ɓ"	           "/ɓ/"                  ""
"c"            "/c/"                  ""
"d"            "/d/"                  ""
"ɗ"            "/ɗ/"                  ""
"e"            "/e/"                  ""
"f"            "/f/"                  ""
"g"            "/ɡ/"                  ""
"ɠ"            "/ɠ/"  "only present in the Pular dialect"  
"h"            "/h/"                  ""
"i"            "/i/"                  ""
"j"            "/ɟ/"                  ""
"k"            "/k/"                  ""
"l"            "/l/"                  ""
"m"            "/m/"                  ""
"n"            "/n/"                  ""
"ŋ"            "/ŋ/"                  ""
"o"            "/o/"                  ""
"p"            "/p/"                  ""
"r"            "/r/"                  ""
"s"            "/s/"                  ""
"t"            "/t/"                  ""
"u"            "/u/"                  ""
"w"            "/w/"                  ""
"y"            "/j/"                  ""
"ƴ"            "/ʔʲ/"                 ""
"\'"            "/ʔ/"                  ""
**Digraph**     ""                    ""
"mb"           "/ᵐb/"                 ""
"nd"           "/ⁿd/"                 ""
"ng"           "/ᵑɡ/"                 ""
"nj"           "/ᶮɟ/"                 ""
"ny"           "/ɲ/"                  ""
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* In some dialects, /s/ approximantizes to [j] [@FfMukoshy1984, p. 53].
* Hausa-influenced dialects tend to spirantize /c/ to [ʃ] [@FfMukoshy1984, p. 56].

# References
