---
title: "Cha'palaa"
author: "Bill Mizgerd"
bibliography: cbi.bib
output: html_document
---

Last Updated: 2019-07-25

# Background

**Language Family**: Barbacoan / Cayapa-Colorado

* Also known as Chachi or Cayapa, Cha'palaa is spoken in norther Ecuador.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"     Labial   Alveolar  Postalveolar  Palatal   Velar  Glottal
Stops                         "p b"    "t d"     "tʲ dʲ"       ""        "k ɡ"  "ʔ"
Affricates                    ""       "ts"      "tʃ"          ""        ""     ""
Fricatives                    "f v"    "s"       "ʃ"           ""        ""     "h"
Nasals                        "m"      "n"       ""            "ɲ"       "ŋ"    ""
Flaps                         ""       "ɾ"       ""            ""        ""     ""
Approximants                  "w"      "l"       ""            "j ʎ"     ""     ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  footnote(general = "Note: Phonemes to the left of each cell are voiceless and phonemes to the right of each cell are voiced; the exception is the approximant row, where phonemes on the right are lateral.", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* Vowel length is contrastive [@CbiFloyd2014, p. 80]. Vowel length is written by doubling the vowel character.
* @CbiLindskoog1962 argue that vowel clusters do not form diphthongs (pp. 38-39), so I have opted not to treat them as such.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
       Front   Central   Back
High   "i"     ""        "u"
Mid    "e"     ""        ""
Low    ""      "a"       ""
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme     Phoneme
"a"           "/a/"  
"b"           "/b/"    
"d"           "/d/"    
"e"           "/e/"  
"f"           "/f/"  
"g"           "/ɡ/"  
"i"           "/i/"  
"j"           "/h/"  
"k"           "/k/"  
"l"           "/l/"    
"m"           "/m/"  
"n"           "/n/"   
"ñ"           "/ɲ/"  
"p"           "/p/"  
"r"           "/ɾ/"  
"s"           "/s/"    
"t"           "/t/"    
"u"           "/u/"  
"v"           "/v/"  
"w"           "/w/"  
"y"           "/j/"  
"\'"          "/ʔ/"  
**Digraph**    ""
"ch"          "/tʃ/"
"dy"          "/dʲ/"
"ll"          "/ʎ/"
"ng"          "/ŋ/"
"sh"          "/ʃ/"
"ty"          "/tʲ/"  
"ts"          "/ts/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* The syllable structure of Cha'palaa is CV(V)(C) [@CbiFloyd2014, p. 80; @CbiLindskoog1962, p. 37].

# Lenition Rules

* Stops voice post-nasally [@CbiFloyd2014, p. 79].
* /h/ sometimes deletes word-medially [@CbiTapuyo2009, p. xvii].

# Misc. Rules

* /w/ becomes [v] before front vowels [@CbiFloyd2014, p. 79].
* /n/ adapts to the place of following bilabials [@CbiFloyd2014, p. 80].
* The contrast between alveolar consonants and their palatalized equivalents collapses before /i/ [@CbiLindskoog1962, p. 34].
* /f/ exists in free variation with [ɸ] [@CbiLindskoog1962, p. 36].

# References
