---
title: "Nigerian Fulfulde"
author: "Emily Strand"
bibliography: fuv.bib
output: html_document
---

Last updated: 2020-02-07

**SLIGHTLY COMPROMISED: glottal stop not always represented; ambiguity concerning the transcription of the palatalized glottal stop**

# Background

**Language Family:** Niger-Congo / Atlantic-Congo / Atlantic / Northern / Senegambian / Fula-Wolof / Fulani / East Central

* Nigerian Fulfulde is spoken in Nigeria.

# Phonology

## Consonants

* Length is a contrastive feature for consonants intervocalically [@Ahmad2014, p. 5]. Geminate consonants are indicated by duplicated graphemes.
    - Fricatives, /ⁿdʒ/, and /ŋ/ do not geminate.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial  Alveolar  Postalveolar  Palatal  Velar   Glottal   
"Stops (plain)"            "p b"    "t d"        ""          ""     "k ɡ"   "ʔ ʔʲ"
"Stops (implosives)"        "ɓ"      "ɗ"         ""          ""      ""       ""
"Stops (prenasalized)"      "ᵐb"     "ⁿd"        ""          ""      "ᵑɡ"     ""
Affricates                  ""       ""      "tʃ dʒ ⁿdʒ"     ""      ""       ""
Fricatives                  "f"      "s"         ""          ""      ""       "h"
Nasals                      "m"      "n"         ""          "ɲ"     "ŋ"      ""
Trills                      ""       "r"         ""          ""      ""       ""
Approximants                ""       "l"         ""          "j"     "w"      ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced. The postalveolar affricate on the right is prenasalized. The glottal stop on the right is palatalized.", general_title = "")
```

## Vowels

* Vowel length is contrastive in Nigerian Fulfulde [@Ahmad2014, p. 8]. Long vowels are indicated by duplicate graphemes.
* @McIntosh1982 explains diphthongs as phonetic realizations of vowel glide sequences (p. 22); thus, I have chosen not to include them below.
* Fulfulde is not a tonal language; rather, it exhibits pitch contours based off predictable intonational patterns [@Ahmad2014, p. 13].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
           Front      Central       Back
High        "i"         ""           "u"
Mid         "e"         ""           "o"
Low         ""          "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

* Glottal stops are not always represented word-initially, which compromises the language slightly [@McIntosh1982, p. 25].
* There is also some ambiguity related to how the palatalized glottal stop (including the geminate form) is represented. I've listed the different forms below. I have gone with the transcription represented in @Ahmad2014 (pp. 5, 7), given that such forms (non-geminate) appear in the Crúbadán corpus; the forms based on the transcription of @McIntosh1982 do not appear (p. 23).

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme       Phoneme          Comment
"a"             "/a/"             ""
"b"             "/b/"             ""
"ɓ"             "/ɓ/"             ""
"c"             "/tʃ/"            ""
"d"             "/d/"             ""
"ɗ"             "/ɗ/"             ""
"e"             "/e/"             ""
"f"             "/f/"             ""
"g"             "/ɡ/"             ""
"h"             "/h/"             ""
"i"             "/i/"             ""
"j"             "/dʒ/"            ""
"k"             "/k/"             ""
"l"             "/l/"             ""
"m"             "/m/"             ""
"n"             "/n/"             ""
"ŋ"             "/ŋ/"             ""
"o"             "/o/"             ""
"p"             "/p/"             ""
"r"             "/r/"             ""
"s"             "/s/"             ""
"t"             "/t/"             ""
"u"             "/u/"             ""
"w"             "/w/"             ""
"y"             "/j/"             ""
"\'"            "/ʔ/"             ""
"ƴ; y\'"        "/ʔʲ/"  "geminate form: ⟨\'y\'y⟩ or ⟨ƴƴ⟩ [@Ahmad2014, pp. 5, 7]"
"\'y"           "/ʔʲ/"  "geminate form: ⟨\'\'y⟩ [@McIntosh1982, p. 23] - **Not represented in the rules**"
**Digraph**      ""               ""
"mb"            "/ᵐb/"            ""
"nd"            "/ⁿd/"            ""
"nj"            "/ⁿdʒ/"           ""
"ny"            "/ɲ/"             ""
"ng"            "/ᵑɡ/"            ""
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Syllables in Nigerian Fulfulde follow the order of CV(V)(C)(C) [@Ahmad2014, p. 11].

# Misc. Rules

* Long vowels may undergo degemination when occurring in word-final unstressed syllables (i.e. the length distinction in this position is neutralized) [@McIntosh1982, p. 25].

# References
