---
title: "Kirghiz"
author: "Emily Strand"
bibliography: ky.bib
output: html_document
---

Last Updated: 2019-12-19

# Background

**Language Family:** Turkic / Kipchak / Kyrgyz-Kipchak

* Also spelled Kyrgyz, speakers reside in a multitude of places, including but not limited to Kyrgyzstan, Afghanistan, China, and Tajikistan.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial     Alveolar    Postalveolar    Palatal     Velar
Stops                      "p  b"      "t  d"         ""            ""        "k  ɡ"
Affricates                   ""          ""        "tʃ  dʒ"         ""         ""
Fricatives                   "v"       "s  z"       "ʃ  ʒ"          ""         ""
Nasals                       "m"         "n"          ""            ""         "ŋ"
Trills                        ""         "r"          ""            ""         ""
Approximants                  ""         "l"          ""            "j"        ""       
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 5)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```

## Vowels

* Vowel length is phonemic [@johanson_kirghiz_2006, p. 344].    
* Long vowels are represented by repeated graphemes in Cyrillic, or macrons (¯) in the Romanized alphabet [@johanson_kirghiz_2006, p. 345].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front        Back
High     "i  y"      "ɯ  u"
Mid      "e  ø"       "o"
Low        ""         "ɑ"
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "For phonemes that share a cell, those on the left are unrounded and those on the right are rounded.", general_title = "")
```    

# Alphabet

* Although a Roman-based alphabet was introduced following the Soviet era, the Cyrillic-based script is primarily used [@brown_kirghiz_2009, p. 611].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme                                        Comment
"a"            "/ɑ/"                                           ""
"б"            "/b/"                                           ""
"в"            "/v/"                                           ""
"г"            "/ɡ/"                                           ""
"д"            "/d/"                                           ""
"е"            "/e/"                                           ""
"ё"            "/jo/"                                          ""
"ж"            "/dʒ/"                                          ""
"з"            "/z/"                                           ""
"и"            "/i/"                                           ""
"й"            "/j/"                                           ""
"к"            "/k/"                                           ""
"л"            "/l/"                                           ""
"м"            "/m/"                                           ""
"н"            "/n/"                                           ""
"ң"            "/ŋ/"                                           ""
"о"            "/o/"                                           ""
"ө"            "/ø/"                                           ""
"п"            "/p/"                                           ""
"р"            "/r/"                                           ""
"с"            "/s/"                                           ""
"т"            "/t/"                                           ""   
"у"            "/u/"                                           ""
"ү"            "/y/"                                           ""
"ф"            "/f/"                                 "only used in loanwords"
"х"            "/χ/"                                           ""
"ц"            "/ts/"                                "only used in loanwords"  
"ч"            "/tʃ/"                                          ""
"ш"            "/ʃ/"                                           ""
"щ"            "/ʃtʃ/"                               "only used in loanwords"      
"ъ"             ""        "indicates non-palatalization of the preceding consonant - only used in loanwords"      
"ы"            "/ɯ/"                                           ""
"ь"             ""        "indicates palatalization of the preceding consonant - only used in loanwords"     
"э"            "/e/"                                           ""
"ю"            "/ju/"                                          ""
"я"            "/jɑ/"                                          ""
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* Spirantization occurs with most stops intervocalically [@johanson_kirghiz_2006, pg. 345].

# Misc. Rules

* /k/, /ɡ/, and /l/ are realized as [q], [ɢ], and [ɫ] respectively preceding back vowels [@johanson_kirghiz_2006, pg. 345].    
* The Kirghiz language implements vowel harmony (specifically two different types) [@Hebert1997, pg. 7].

# References
