---
title: "Mekeo"
author: "Emily Strand"
bibliography: mek.bib
output: html_document
---

Last updated: 2020-06-30

**SLIGHTLY COMPROMISED: glottal stops inconsistently marked**

# Background

**Language Family:** Austronesian / Malayo-Polynesian / Central-Eastern / Eastern Malayo-Polynesian / Oceanic / Wester Oceanic / Papuan Tip / Peripheral / Central Papuan / West Central Papuan / Nuclear

* Mekeo is spoken in the Central province of Papua New Guinea.
* The East Mekeo (standard) dialect is represented below [@Chung2009, p. 9].

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial    Alveolar   Velar   Glottal
Stops                        "p"       "t"       "k"      "ʔ"
Fricatives                   "f"       "s"       ""       ""
Nasals                       "m"       "n"       "ŋ"      ""
Approximants                 ""        "l"       ""       ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 4)) %>%
  column_spec(1, bold = TRUE)
```  

## Vowels

* Vowel length is phonemic in Mekeo [@Chung2009, p. 16]. Long vowels are indicated by duplicate graphemes.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
           Front      Central       Back
High        "i"         ""          "u"
Mid         "e"         ""          "o"
Low         ""          "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

* The glottal stop is said to be written only when it poses contrast [@Chung2009, p. 4]. Due to this, the language is somewhat compromised.
* ⟨v⟩ may be used orthographically, primarily in loanwords. I have chosen to transcribe ⟨v⟩ as /f/ given that it varies freely with [v] (ibid.).

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme       Phoneme
"a"             "/a/"
"e"             "/e/"
"g"             "/ŋ/"
"f"             "/f/"
"i"             "/i/"
"k"             "/k/"
"l"             "/l/"
"m"             "/m/"
"n"             "/n/"
"o"             "/o/"
"p"             "/p/"
"s"             "/s/"
"t"             "/t/"
"u"             "/u/"
"\'"            "/ʔ/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Syllables in Mekeo consist of the following structures [@SIL, p. 2]:
    - V
    - VV
    - CV
    - CVV

# Lenition Rules

* The glottal stop is frequently omitted, especially by younger generations [@Chung2009, p. 4].

# Misc. Rules

* /t/ is realized as [ts] preceding /i/ [@Chung2009, p. 3].

# References
