---
title: "Shipibo-Konibo"
author: "Bill Mizgerd"
bibliography: shp.bib
output: html_document
---

Last Updated: 2020-07-01

# Background

**Language Family:** Panoan / Mainline Panoan / Nawa / Chama / Shipibo-Konibo

* The language is often simply referred to as Shipibo and is spoken in Brazil and Peru.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial  Alveolar  Postalveolar  Retroflex  Palatal  Velar   Glottal
Stops                      "p"     "t"       ""            ""         ""       "k"     ""
Affricates                 ""      "ts"      "tʃ"          ""         ""       ""      ""
Fricatives                 "β"     "s"       "ʃ"           "ʂ"        ""       ""      "h"
Nasals                     "m"     "n"       ""            ""         ""       ""      ""
Approximants               "w"     ""        ""            "ɻ"        "j"      ""      ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Retroflex", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 7)) %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* Each vowel has both an oral and a nasal phoneme, according to @ShpValenzuelaEtAl2001 (p. 282); @ShpValenzuela2003, however, analyzes nasal vowels as /Vn/ sequences (p. 95). Because @ShpEliasUlloa2006 also finds nasal vowels to only exist through a nasalization process from consonants (p. 14), I have opted to prefer the analysis that excludes nasal vowels.
* Long vowels are realized in some words. @ShpEliasUlloa2006 describes vowel length as phonemic, but @ShpValenzuela2003 finds it never to be contrastive (p. 95).
* @ShpValenzuelaEtAl2001 argue for /ɯ/ rather than /ɨ/.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
          Front   Central   Back
High      "i"     "ɨ"       ""
Mid       ""      ""        "o"
Low       ""      "a"       ""
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

* Apostrophes are used to represent the glottal stop, but because it is not phonemic, I will be omitting it from the ruleset [@ShpValenzuela2003, p. 138].
* Normally postvocalic ⟨n⟩ is used to indicate vowel nasalization, so ⟨nh⟩ is used to indicate when a nasal vowel is actually followed by /n/ [@ShpWise1993, p. 18]. However, because my analysis does not treat vowel nasalization as contrastive, ⟨nh⟩ behaves like any other ⟨n⟩ in my ruleset.
* **Note:** /ʂ/ is not represented in the Crúbadán corpus.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme           Comment
"a"              "/a/"              ""
"b"              "/β/"              ""
"c"              "/k/"              ""
"e"              "/ɨ/"              ""
"i"              "/i/"              ""
"j"              "/h/"              ""
"k"              "/k/"              ""
"m"              "/m/"              ""
"n"              "/n/"              ""
"o"              "/o/"              ""
"p"              "/p/"              ""
"r"              "/ɻ/"              ""
"s"              "/s/"              ""
"t"              "/t/"              ""
"w"              "/w/"              ""
"x"              "/ʂ/"    "according to @ShpValenzuela2003"  
"y"              "/j/"              ""
**Digraph**       ""                ""
"ch"             "/tʃ/"             ""
"hu"             "/w/"              ""
"nh"             "/n/"              ""
"qu"             "/k/"              ""
"sh"             "/ʃ/"              ""
"s̈h"             "/ʂ/"    "according to @ShpWise1993"
"ts"             "/ts/"             ""
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Shipibo has a (C)V(C) syllable structure [@ShpValenzuela2003, p. 104].
* No consonants are syllabic. Only sibilants and /n/ occur in codas [@ShpValenzuela2003, p. 105].

# Lenition Rules

* /k/ lenites, either to [ɣ] or to [ɰ], in certain high-frequency morphemes [@ShpValenzuelaEtAl2001, p. 282; @ShpValenzuela2003, p. 103].
* The middle marker /-t/ often deletes before consonant-initial suffixes [@ShpValenzuela2003, pp. 116-117].

# Misc. Rules

* Epenthesis of [ʔ] occurs at the beginning of vowel-initial utterances and the end of vowel-final utterances, as well as at morpheme boundaries with vowels on both sides [@ShpValenzuelaEtAl2001, p. 282; @ShpValenzuela2003, pp. 95-96].
* Vowels nasalize after nasal consonants; vowels nasalized in this way behave differently than contrastively nasalized vowels (ibid.).
* Word-initial /β/ despirantizes to [b] in stressed syllables [ibid.; @ShpValenzuela2003, p. 97].
* /ɻ/ tends to be variable: intervocalically it is sometimes pronounced [ɹ], and in other cases it is realized as [ʐ] or [ɾ]. In the onsets of stressed syllables, it tends to be pre-stopped, yielding [dz] or [dɹ] [@ShpValenzuelaEtAl2001, pp. 282-283].
* Approximants nasalize before nasal vowels [@ShpValenzuelaEtAl2001, p. 283].
* /w/ becomes the labiopalatal approximant [ɥ] before high front vowels, and unrounds to [ɰ] before high back vowels (ibid.).
* /ɯ/ fronts to [ɨ] after coronals (ibid.).
* Nasalized vowels spread their nasalization to preceding vowels, if the vowels are adjacent or only separated by approximants (ibid.).
* /o/ raises to [u] before /i/ [@ShpValenzuela2003, p. 94].
* Vowels centralize somewhat in closed syllables [@ShpValenzuelaEtAl2001, p. 283].
* Word-initial /h/ sometimes deletes before /a/ [@ShpValenzuela2003, p. 103].
* Between voiceless obstruents, vowels may devoice or delete (ibid.).

# References
