---
title: "Samoan"
author: "Emily Strand"
bibliography: sm.bib
output: html_document
---

Last Updated: 2020-07-02

**SLIGHTLY COMPROMISED: suspect marking of vowel length and glottal stops**

# Background

**Language Family:** Austronesian / Malayo-Polynesian / Central-Eastern / Eastern Malayo-Polynesian / Oceanic / Central-Eastern Oceanic / Remote Oceanic / Central Pacific / East Fijian-Polynesian / Polynesian / Nuclear / Samoic-Outlier / Samoan

* Samoan is spoken throughout the Samoan Islands.
* The formal style is represented below.

# Phonology

## Consonants

* /k/, /h/, and /ɹ/ appear in loanwords [@Zuraw2014, p. 277].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial    Alveolar    Velar   Glottal
Stops                        "p"       "t"        ""       "ʔ"
Affricates                   ""        ""         ""       ""
Fricatives                  "f v"      "s"        ""       ""
Nasals                       "m"       "n"        "ŋ"      ""
Approximants                 ""        "l"        ""       ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 4)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```  

## Vowels

* Vowel length is phonemic [@Kernan1974, p. 107]. Long vowels are marked with a macron (¯).

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         ""           "u"
Mid       "e"         ""           "o"
Low        ""         "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)

```

# Alphabet

* @Alderete2012 state that macrons and apostrophes (marking long vowels and glottal stops respectively) aren't generally used in orthography (p. 9). However, the Crúbadán corpus appears to use the conventions quite frequently. Because of this unexpected outcome (where the observation is not in line with the documentation), the marking of vowel length and glottal stops is somewhat suspect. Therefore, I have decided to classify the language as compromised.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme  
"a"             "/a/"
"e"             "/e/"
"f"             "/f/"
"g"             "/ŋ/"
"i"             "/i/"
"l"             "/l/"
"m"             "/m/"
"n"             "/n/"
"o"             "/o/"
"p"             "/p/"
"s"             "/s/"
"t"             "/t/"
"u"             "/u/"
"v"             "/v/"
"‘"             "/ʔ/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* In fast speech, voiceless stops (including /k/), are voiced intervocalically [@Alderete2012, p. 3].
* Glottal stops are often deleted intervocalically in fast speech (ibid.).
* /f/ may debuccalize to /h/ (ibid.).

# Misc. Rules

* In the colloquial style [@Kernan1974, p. 108]:
    - /t/ is realized as [k]
    - /n/ is realized as [ŋ]
* /t/ tends to become [ts] preceding /i/ [@Alderete2012, p. 3].
* /k/ may palatalize in the environment of /i/ or /e/ (ibid.).
* In fast speech, /s/ may become [ts] or [ʃ] (ibid.).
* Short vowels may be deleted word-finally [@Alderete2012, p. 6].

# References
