---
title: "Uzbek"
author: "Bill Mizgerd"
bibliography: uz.bib
output: html_document
---

Last Updated: 2019-08-02

# Background

**Language Family**: Turkic / Common Turkic / Karluk / Uzbek

* Uzbek is spoken throughout Uzbekistan and Central Asia.

# Phonology

## Consonants

* /ʒ/ and /ts/ appear in Russian loanwords [@UzIsmatulla1992, pp. vii-viii]; @UzSjoberg1963, pp. 7-8].
* /ʔ/ appears in some Arabic loanwords [@UzSjoberg1963, pp. 20-21].
* @UzSjoberg1963 argues for /ɸ/ rather than /f/; /f/, in her view, is limited to loans (pp. 7-8).
* @UzSjoberg1963 writes /x/ rather than /χ/, but describes it as "postvelar, almost uvular" (p. 11), so I have chosen to follow @UzBoeschoten1998 in interpreting the sound as /χ/ (p. 359).
* /w/ varies with voiced bilabial and labiodental fricatives [@UzSjoberg1963, p. 13].
* @UzWaterson1980 argues for /ʁ/ rather than /ɣ/ (p. xviii).
* Most consonants have contrastive gemination [@UzSjoberg1963, p. 41].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial  Alveolar  Postalveolar  Palatal  Velar  Uvular  Glottal
Stops                      "p b"   "t d"     ""            ""       "k ɡ"  "q"     ""
Affricates                 ""      ""        "tʃ dʒ"       ""       ""     ""      ""
Fricatives                 "f"     "s z"     "ʃ"           ""       "ɣ"    "χ"     "h"
Nasals                     "m"     "n"       ""            ""       "ŋ"    ""      ""
Flaps                      ""      "ɾ"       ""            ""       ""     ""      ""
Approximants               "w"     "l"       ""            "j"      ""     ""      ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Uvular", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 7)) %>%
  footnote("Note: For phonemes that share a cell, those on the left are voiceless, whereas those on the right are voiced.", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* Some sources list the grapheme ⟨ъ⟩ as marking length on preceding vowels, but none of my sources have made any indication that length is contrastive in Uzbek.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
          Front   Central   Back
High      "i"     ""        "u"
High-Mid  ""      ""        "o"
Low-Mid   "ɛ"     ""        "ɔ"
Low       ""      "a"       ""
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered")
```

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme               Comment
"а"              "/a/"                  ""
"б"              "/b/"                  ""
"в"              "/w/"          "/v/ in some loanwords"  
"г"              "/ɡ/"                  ""
"ғ"              "/ɣ/"                  ""
"д"              "/d/"                  ""
"е"            "/ɛ/; /jɛ/"    "/ɛ/: word-medially; /jɛ/: word-initially"  
"ё"              "/jɔ/"                 ""
"ж"              "/dʒ/"          "/ʒ/ in some loanwords"  
"з"              "/z/"                  ""
"и"              "/i/"                  ""
"й"              "/j/"                  ""
"к"              "/k/"                  ""
"қ"              "/q/"                  ""
"л"              "/l/"                  ""
"м"              "/m/"                  ""
"н"              "/n/"                  ""
"о"              "/ɔ/"                  ""
"п"              "/p/"                  ""
"р"              "/ɾ/"                  ""
"с"              "/s/"                  ""
"т"              "/t/"                  ""
"у"              "/u/"                  ""
"ў"              "/o/"                  ""
"ф"              "/f/"                  ""
"х"              "/χ/"                  ""
"ҳ"              "/h/"                  ""
"ч"              "/tʃ/"                 ""
"ш"              "/ʃ/"                  ""
"ъ"              "/j/"  "before ⟨е⟩; marks (non-phonemic) vowel length (postvocalically) - see note above"
"э"              "/ɛ/"                  ""
"ю"              "/ju/"                 ""
"я"              "/ja/"                 ""
"ь"                ""   "marks palatalization of preceding consonant; only occurs in Russian loans"  
"ц"              "/ts/"         "only occurs in Russian loans"
**Digraph**        ""                   ""
"нг"             "/ŋ/"                  ""
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```  

# Lenition Rules

* Many speakers spirantize /k/ and /q/ [@UzBoeschoten1998, p. 359].
* /j/ is produced with lax articulation next to high front vowels [@UzSjoberg1963, p. 13].
* /ɡ/ sometimes approximantized to [j] in casual speech [@UzSjoberg1963, p. 21].

# Misc. Rules

* Velar stops palatalize before front vowels, especially rounded ones [@UzSjoberg1963, p. 10].
* Some speakers lower /ɔ/ to [a] in various words [@UzSjoberg1963, p. 39].

# References
