---
title: "Au"
author: "Emily Strand"
bibliography: avt.bib
output: html_document
---

Last updated: 2019-10-29

# Background

**Language family**: Torricelli / Wapei-Palei / Wapei

* Au is spoken in the Sandaun province of Papua New Guinea.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"  Bilabial    Alveolar  Palatal     Velar    
Stops                       "p"        "t"        ""         "k"      
Fricatives                  ""         "s"        ""         "ɣ"        
Nasals                      "m"        "n"        ""         ""       
Trills                      ""         "r"        ""         ""        
Approximants                "w"        ""        "j"         ""        
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Bilabial", "Alveolar", "Palatal", "Velar"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 4)) %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* There are four possible vowel combinations (/ai/, /əi/, /au/, /iu/); however, they are seen as separate units rather than complex phonemes [@scorza_phonemes_1971, p. 6].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"        "ɨ"           "u"
Mid       ""         "ə"           "o"
Low       ""       "a  aː"         ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: The low central vowels are ordered short and long.", general_title = "")
```    

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme
"a"             "a"
"e"             "ə"
"h"             "ɣ"
"i"             "i"
"ɨ"             "ɨ"
"k"             "k"
"m"             "m"
"n"             "n"
"o"             "o"
"p"             "p"
"r"             "r"
"s"             "s"
"t"             "t"
"u"             "u"
"w"             "w"
"y"             "j"
**Digraph**     ""
"aa"            "aː"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* Voiceless stops may become voiced word-medially; in such contexts the voiced and voiceless counterparts are said to alternate freely [@scorza_phonemes_1971, pp. 7-8].

# Misc. Rules

* /n/ varies freely with /l/ word-medially [@scorza_au_1992, p. 3].

# References
