---
title: "Nomatsiguenga"
author: "Emily Strand"
bibliography: not.bib
output: html_document
---

Last Updated: 2020-07-01

**SLIGHTLY COMPROMISED: conflation among placeless /N/, /m/, and /n/**

# Background

**Language Family:** Arawakan / Maipuran / Southern Maipuran / Campa

* Nomatsiguenga is spoken in the Junin region in central Peru.

# Phonology

## Consonants

* There is mention of a placeless /N/ phoneme that occurs in word-medial codas, preceding stop consonants [@Chen2019, p. 3; @Lawrence2013, pp. 8, 9]. Both such sources also indicate /m/ and /n/ as phonemes of Nomatsiguenga, stating that the placeless /N/ behaves differently where it deletes preceding vowels [@Chen2019, p. 3]. Orthographically, /N/ is represented "as an ⟨m⟩ before bilabials and ⟨n⟩ elsewhere" [@Lawrence2013, p. 8]. Due to this conflation of /N/ with /n/ and /m/, the language is somewhat compromised. I have opted to use /m/ and /n/ in the ruleset, not accounting for the placeless nasal.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Bilabial   Dental   Alveolar  Postalveolar  Palatal  Velar   Glottal
Stops                       "p b"      "t̪"        ""         ""          ""     "k ɡ"     ""
Affricates                   ""        "t̪s"       ""         "tʃ"        ""      ""       ""
Fricatives                   ""        "s̪"        ""         "ʃ"         ""      ""       "h"
Nasals                       "m"       "n̪"        ""         ""          ""      "ŋ"      ""
Flaps                        ""        ""         "ɾ"        ""          ""      ""       ""
Approximants                 ""        ""         ""         ""          "j"     ""       ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Bilabial", "Dental", "Alveolar", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 7)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced. Nomatsiguenga also has a placeless /N/ (described above the chart).", general_title = "")
```  

## Vowels

* Tone is contrastive in Nomatsiguenga (high tone vs. no tone). High tone is marked over the vowel with an acute accent (´), whereas no tone is unmarked [@Chen2019, p. 68; @Lawrence2013, p. 15].
* Length is phonemic in Nomatsiguenga. Long vowels are indicated by the duplication of vowels (either one is marked with a high tone) (ibid.).
* /ɨ/ does not have a long counterpart [@Lawrence2013, p. 8].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
           Front      Central       Back
High        "i"         "ɨ"          ""
High-Mid    "e"         ""           "o"
Low         ""          "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)


diphthongs <- read.table(textConnection('
Diphthongs
"/ai/, /ei/, /oi/"
'), TRUE)

kable(diphthongs, align = 'c') %>%
  kable_styling("bordered")
```  

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme         Comment
"a"              "/a/"            ""
"b"              "/b/"            ""
"c"              "/k/"            ""
"e"              "/e/"            ""
"ë"              "/ɨ/"            ""
"g"              "/ɡ/"            ""
"i"              "/i/"            ""
"j"              "/h/"            ""
"m"              "/m/"            ""
"n"            "/n̪/; /ŋ/"  "/ŋ/: preceding ⟨c⟩ or ⟨qu⟩"  
"o"              "/o/"            ""
"p"              "/p/"            ""
"q"              "/k/"            ""
"r"              "/ɾ/"            ""
"s"              "/s̪/"            ""
"t"              "/t̪/"            ""
"y"              "/j/"            ""
**Multigraph**    ""              ""
"ch"             "/tʃ/"           ""
"ng"             "/ŋ/"     "intervocalically"
"sh"             "/ʃ/"            ""
"ts"             "/t̪s/"           ""
"gu"             "/ɡ/"  "preceding ⟨e⟩, ⟨ë⟩, and ⟨i⟩"
"ngu"            "/ŋ/"  "preceding ⟨e⟩, ⟨ë⟩, and ⟨i⟩"     
"qu"             "/k/"   "preceding ⟨e⟩ or ⟨i⟩"
'), TRUE)

kable(alphabet, align = "c") %>%
  kable_styling("bordered")
```


# Lenition Rules

* /h/ may delete preceding a vowel marked with high tone [@Lawrence2013, p. 20].

# Misc. Rules

* /q/ is palatalized preceding front vowels (spelled by ⟨qu⟩) [@Chen2019, p. 13].
* Stops at morpheme boundaries are realized as [h] preceding /k/ as a suffix-onset [@Lawrence2013, pp. 20, 21].

# References
