---
title: "Tatar"
author: "Bill Mizgerd"
bibliography: tt.bib
output: html_document
---

Last Updated: 2020-07-02

# Background

**Language Family:** Turkic / Common Turkic / Kipchak / North Kipchak / Tatar

* Although this language is usually simply called Tatar, it is sometimes also known as Kazan Tatar or Volga Tatar, in order to differentiate it from the Crimean Tatar and Siberian Tatar languages.
* Tatar is spoken in Russia, specifically in Saint Petersburg, Moscow, and the Bashkortostan and Tatarstan republics.

# Phonology

## Consonants

* Bilabial and labiodental consonants are non-contrastive in Turkic languages [@TtBaitchura1975, p. 244], so /f/ could theoretically also be /ɸ/. @TtComrie1997 explicitly describes this consonant as bilabial (p. 901), but he diverges from a general consensus of using /f/. I have opted for the labiodental /f/ in the phonemic inventory below.
* /ɕ/ and /ʑ/ should not be replaced with affricates [@TtComrie1997, p. 902].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial  Alveolar  Postalveolar  Alveopalatal      Palatal   Velar   Glottal
Stops                      "p b"   "t d"     ""            ""                 ""       "k ɡ"   "ʔ"
Fricatives                 "f"     "s z"     "ʃ ʒ"         "ɕ ʑ"              ""       "x"     "h"
Nasals                     "m"     "n"       ""            ""                 ""       "ŋ"     ""
Trills                     ""      "r"       ""            ""                 ""       ""      ""
Approximants               "w"     "l"       ""            ""                 "j"      ""      ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Alveolo-Palatal", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 7)) %>%
  footnote("Note: For phonemes that share a cell, those on the left are voiceless, whereas those on the right are voiced.", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

## Vowels

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
          Front   Central   Back
High      "i y"   ""        "ɯ u"
Mid       "e ø"   ""        "ɤ o"
Low       ""      "a"       "ɑ"
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered") %>%
  footnote("Note: For phonemes that share a cell, those on the left are unrounded, whereas those on the right are rounded.", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

* Tatar can be represented by both the Cyrillc and Latin scripts. However, the Latin alphabet was made illegal by the Tatarstani government [@Wertheim2012, p. 75]. I have chosen to include both alphabets below as well as both rulesets.

## Cyrillic

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

cyrillic_alphabet <- read.table(textConnection('
Grapheme        Phoneme              Comment
"а"              "/ɑ/"                 ""
"ә"              "/a/"       "some authors prefer to use /æ/"  
"б"              "/b/"                 ""  
"в"              "/w/"         "/v/ in Russian loanwords"  
"г"              "/ɡ/"                 ""  
"д"              "/d/"                 ""
"е"           "/e/; /je/"  "/je/: word-initially [@TtPoppe1963, p. 20]"  
"ё"           "/o/; /jo/"  "/jo/: word-initially (only appears in Russian loanwords)"  
"ж"              "/ʒ/"        "only appears in Russian loanwords"  
"җ"              "/ʑ/"                 ""
"з"              "/z/"                 ""
"и"              "/i/"                 ""
"й"              "/j/"                 ""
"к"              "/k/"                 ""  
"л"              "/l/"                 ""
"м"              "/m/"                 ""
"н"              "/n/"                 ""
"ң"              "/ŋ/"                 ""
"о"              "/o/"                 ""
"ө"              "/ø/"                 ""
"п"              "/p/"                 ""
"р"              "/r/"                 ""
"с"              "/s/"                 ""
"т"              "/t/"                 ""
"у"              "/u/"                 ""
"ү"              "/y/"                 ""  
"ф"              "/f/"                 ""  
"х"              "/x/"                 ""
"һ"              "/h/"                 ""
"ц"              "/ts/"  "only appears in Russian loanwords"  
"ч"              "/ɕ/"                 ""
"ш"              "/ʃ/"                 ""
"щ"              "/ʃɕ/"  "only appears in Russian loanwords"  
"ъ"              "/ʔ/"   "indicates phonetic backing of associated vowels"  
"ы"              "/ɤ/"                 ""
"ь"              "/ʔ/"   "indicates phonetic fronting of associated vowels"  
"э"              "/e/"                 ""
"ю"              "/ɯ/" "phonemic status is controversial; some writers prefer /ju/ or /jy/"
"я"              "/jɑ/"                ""
'), TRUE)

kable(cyrillic_alphabet, align = 'c') %>%
  kable_styling("bordered")
```  

## Latin

* The capital of ⟨i⟩ is ⟨İ⟩ and the capital of ⟨ı⟩ is ⟨I⟩.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

latin_alphabet <- read.table(textConnection('
Grapheme        Phoneme               Comment
"a"              "/ɑ/"                  ""
"ä"              "/a/"                  ""
"b"              "/b/"                  ""
"c"              "/ʑ/"                  ""
"ç"              "/ɕ/"                  ""
"d"              "/d/"                  ""
"e"              "/e/"                  ""
"f"              "/f/"                  ""
"g"              "/ɡ/"                  ""
"ğ"              "/ɡ/"                  ""
"h"              "/h/"                  ""
"i"              "/i/"                  ""
"ı"              "/ɤ/"                  ""
"j"              "/ʒ/"                  ""
"k"              "/k/"                  ""
"l"              "/l/"                  ""
"m"              "/m/"                  ""
"n"              "/n/"                  ""
"ñ"              "/ŋ/"                  ""
"o"              "/o/"                  ""
"ö"              "/ø/"                  ""
"p"              "/p/"                  ""
"q"              "/k/"                  ""
"r"              "/r/"                  ""
"s"              "/s/"                  ""
"ş"              "/ʃ/"                  ""
"t"              "/t/"                  ""
"u"              "/u/"                  ""
"ü"              "/y/"                  ""
"v"              "/w/"                  ""
"w"              "/w/"                  ""
"x"              "/x/"                  ""
"y"              "/j/"                  ""
"z"              "/z/"                  ""
"\'"             "/ʔ/"                  ""
**Digraph**       ""                    ""
"ts"             "/ts/"    "only appears in loanwords"
"şç"             "/ʃɕ/"    "only appears in loanwords"
"yu"             "/ɯ/"                  ""
"ya"             "/ja/"                 ""
'), TRUE)

kable(latin_alphabet, align = 'c') %>%
  kable_styling("bordered")
```  

# Syllable Structure

* Syllables in Tatar follow the order of (C)V(C)(C) [@TtComrie1997, pp. 905-906]. However, such structures could also be described as CV(C)(C), since syllables without an initial consonant only occur word-initially and are phonetically introduced by a [ʔ].

# Lenition Rules 

* Stem-final voiceless labial stops become voiced when followed by a vowel [@TtPoppe1963, p. 18].
* Stem-final voiceless dorsal stops become voiced intervocalically (ibid.).

# Misc. Rules

* Velar stops adjacent to back vowels become backed [@TtPoppe1963, pp. 10-11].
    - /k/ goes to [q].
    - /ɡ/ is said to go to [ʁ]; however it is often indicated as [ɣ]. Because it is a "post-velar" sound that acts as a farther-back contrast to /ɡ/, I felt that [ʁ] more accurately described the character of the phoneme.
* /l/ becomes velarized in the presence of back vowels [@TtPoppe1963, p. 11].
* /w/ is realized as a fricative word-initially [@TtComrie1997, p. 902].
* Velar/uvular contrasts (/k/ vs. /q/ and /g/ vs. /ʁ/) occur only in words of Arabic or Persian origin [@TtComrie1997, p. 901].
* Voiced consonants become voiceless word-finally [@TtPoppe1963, p. 18].
    - Stem-final dental stops devoice when followed by a vowel.
* Palatal harmony in consonants: velar stops occur in the context of front vowels, while their uvular counterparts occur in the context of back vowels [@TtComrie1997, pp. 904-905].
* Palatal harmony in vowels: if a word's first vowel is front, all the subsequent vowels must also be front; likewise, if a word's first vowel is back then all subsequent vowels must be back [@TtComrie1997, p. 903].
* Voicing assimilation occurs in clusters of obstruents, defaulting to voiceless [@TtComrie1997, p. 980].

# References
