---
title: "Maltese"
author: "Bill Mizgerd"
bibliography: mt.bib
output: html_document
---

Last Updated: 2018-10-21

# Background

**Language Family**: Afro-Asiatic / Semitic / Central Semitic / Arabic / Western Arabic / Maltese

* Maltese is spoken in Malta and throughout the European Union.

# Phonology

## Consonants

* /ʒ/ and /dz/ are commonly included in the phonemic inventory of Maltese through loanword integration; however, they will not be included in the ruleset as their native voiceless counterparts are represented by the same graphemes [@MtBorg1997, p. 248].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial     Dental   Alveolar    Postalveolar   Palatal     Velar     Pharyngeal     Glottal
Stops                      "p  b"     "t̪  d̪"      ""           ""           ""       "k  ɡ"         ""           "ʔ"
Affricates                   ""        ""      "ts  **dz**"     "tʃ  dʒ"        ""         ""           ""           ""
Fricatives                 "f  v"      ""       "s  z"       "ʃ  **ʒ**"         ""         ""           "ħ"          ""
Nasals                       "m"       ""         "n"          ""           ""         ""           ""           ""
Trills                       ""        ""         "r"          ""           ""         ""           ""           ""
Approximants                 ""        ""         "l"          ""           "j"        "w"          ""           ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Dental", "Alveolar", "Postalveolar", "Palatal", "Velar", "Pharyngeal", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 8)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```  

## Vowels

* Vowel length is contrastive [@MtBorg1997, p. 251].
* Authors seem to vary in terms of the vowel inventory:
    - @MtHume1996:             /a/ /e/ /i/ /o/ /u/  
    - @MtOrtho1996:            /a/ /ɛ/ /i/ /o/ /u/  
    - @MtBorg1997:             /a/ /e/ /i/ /o/ /u/  
    - @MtHumeEtAl2009:         /ɐ/ /ɛ/ /i/ /ɔ/ /u/  
    - @MtPuech2016:            /a/ /ɛ/ /ɪ/ /ɔ/ /ʊ/
  * (NB: @MtHumeEtAl2009 only uses phonetic transcriptions, rather than phonemic, so it is ambiguous whether her choices of vowels reflect phonemes rather than realizations.)

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High    "i  iː"       ""         "u  uː"
""        "ɪː"        ""           ""
Mid       "e"         ""           "o"
Low        ""         "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)


diphthongs <- read.table(textConnection('
Diphthongs
"/aj/, /ej/, /oj/, /aw/, /ew/, /iw/, /ow/"
'), TRUE)

kable(diphthongs, align = 'c') %>%
  kable_styling("bordered")
```

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme                   Comment
"a"            "/a/"             "some authors use /ɐ/"
"b"            "/b/"                      ""
"ċ"            "/tʃ/"                     ""
"d"            "/d̪/"                      ""
"e"            "/e/"             "some authors use /ɛ/"
"f"            "/f/"                      ""
"ġ"            "/dʒ/"                     ""
"g"            "/ɡ/"                      ""  
"h"            "/ː/"         "lengthens associated vowels"  
"ħ"            "/ħ/"   "realized by some speakers as [x] or [h]"  
"ie"           "/ɪː/"           "some authors use /ie/"  
"j"            "/j/"                      ""
"k"            "/k/"                      ""
"l"            "/l/"                      ""
"m"            "/m/"                      ""
"n"            "/n/"                      ""
"o"            "/o/"             "some authors use /ɔ/"  
"p"            "/p/"                      ""
"q"            "/ʔ/"                      ""
"r"            "/r/"                      ""
"s"            "/s/"                      ""
"t"            "/t̪/"                      ""
"u"         "/u/, /uː/"          "some authors use /ʊ/"
"v"            "/v/"                      ""
"w"            "/w/"                      ""
"x"            "/ʃ/"         "/ʒ/ appears in certain loanwords"  
"ż"            "/z/"                      ""
"z"            "/ts/"       "/dz/ appears in certain loanwords"
**Digraph**     ""                        ""
"għ"           "/ː/"         "lengthens associated vowels"
"i"         "/i/; /iː/"                   ""
"aj"          "/aj/"                      ""
"aw"          "/aw/"                      ""
"ew"          "/ew/"         "occasionally realized as [ow]"  
"għi"         "/ej/"                      ""
"għu"         "/ow/"                      ""
"iw"          "/iw/"                      ""
"oj"          "/oj/"                      ""
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Processes

* Long vowels are shortened word-finally [@MtBorg1997, p. 267].
* ⟨għ⟩ represents the Old Arabic phones [ʕ] or [ɣ], which were deleted at some point during the evolution of modern Maltese [@MtBorg1997, p. 261].
* ⟨h⟩ represents the Old Arabic phone [h], which was deleted at some point during the evolution of modern Maltese [@MtBorg1997, p. 259].
* Elision of short vowels occurs in unstressed open syllables [@MtBorg1997, p. 273].

# More Rules

* Clusters of obstruents assimilate to be either fully voiced or fully voiceless [@MtBorg1997, p. 250].
* Consonants geminate [@MtBorg1997, pp. 251-252].
    - Word-final long consonants are sometimes shortened (ibid.).
* In the definite article, ⟨il-⟩, the /l/ typically assimilates to subsequent coronal segments (except /dʒ/) [@MtBorg1997, p. 255].
* Metathesis of the definite article occurs before Italian-derived words with initial sibilants [@MtBorg1997, p. 273].
* The ⟨h⟩ in the suffixed pronouns (⟨-ha⟩, ⟨-hom⟩) is replaced by a glide after stressed stem-final vowels [@MtBorg1997, p. 261].
* CCC clusters are pronounced as CVCC when the middle C is a sonorant [@MtBorg1997, p. 274].

# References
