---
title: "Macedonian"
author: "Jordan Schuster"
bibliography: mk.bib
output: html_document
---

Last Updated: 2020-06-15

# Background

**Language Family:** Indo-European / Balto-Slavic / Slavic / South / Eastern

* Macedonian is spoken in a multitude of places, including, but not limited to, Macedonia, Albania, Montenegro, Bulgaria, and Greece.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial   Dental   Alveolar    Postalveolar    Palatal     Velar
Stops                      "p  b"   "t̪  d̪"     ""             ""          "c  ɟ"     "k  ɡ"
Affricates                   ""    "ts̪  dz̪"    ""          "tʃ  dʒ"         ""          ""
Fricatives                 "f  v"   "s̪  z̪"     ""           "ʃ  ʒ"          ""         "x"
Nasals                       "m"      "n̪"      ""             ""           "ɲ"         ""
Trills                       ""       ""       "r"            ""            ""         ""
Approximants                 ""       "ɫ̪"      "l"            ""            "j"        ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Dental", "Alveolar", "Postalveolar", "Palatal", "Velar"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```


## Vowels

* /ə/ is phonemic in many dialects, but very marginal in the standard language [@Friedman2001, p. 10], thus it will not be included in the phonemic inventory.
* Where adjacent like-vowels occur, they're either interpreted as independent vowels or one long vowel (ibid.). Where three adjacent vowels occur, the second and third are often separated with /j/. Thus, vowel length is not phonemic.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         ""           "u"
Mid       "ɛ"         ""           "ɔ"
Low        ""         "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```  

# Alphabet

* There is much variation regarding the transcription of ⟨л⟩ and ⟨љ⟩ [@Friedman2001, pp. 11-12; @Kramer2011, pp. 3-4]. ⟨л⟩ is said to represent /l/ preceding front vowels and /j/, and represent /ɫ/ elsewhere. This would suggest that the two represented phonemes are actually allophones. However, younger generations have started to be more liberal with using /ɫ/. ⟨љ⟩ represents /l/ (before back vowels, consonants (except /j/) and word-finally). Younger speakers extend velarization to this environment as well, making it so /l/ and /ɫ/ appear in like-environments. It is unclear if this is reflective of a diachronic change with /l/ losing phonemic status in the language, but as of now, I have opted to keep/reflect the distinction.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
"Cyrllic Grapheme"    "Romanized Grapheme/Multigraph"    Phoneme               Comment
      "а"	                     "a"                       "/a/"                  ""
      "б"	                     "b"                       "/b/"                  ""
      "в"	                     "v"                       "/v/"                  ""
      "г"	                     "g"                       "/ɡ/"                  ""
      "д"	                     "d"                       "/d̪/"                  ""
      "ѓ"                      "gj"                      "/ɟ/"                  ""
      "е"	                     "e"                       "/ɛ/"                  ""
      "ж"	                     "zh"                      "/ʒ/"                  ""
      "з"	                     "z"                       "/z̪/"                  ""
      "s"                      "dz"                      "/dz̪/"                 ""
      "и"	                     "i"                       "/i/"                  ""
      "j"                      "j"                       "/j/"                  ""
      "к"	                     "k"                       "/k/"                  ""
      "л"	                     "l"                    "/ɫ̪/; /l/"         "/l/: before front vowels and /j/"
      "љ"                      "lj"                      "/l/"           "/l/: before back vowels, consonants (except /j/), and word-finally"
      "м"	                     "m"                       "/m/"                  ""
      "н"	                     "n"                       "/n̪/"                  ""
      "њ"	                     "nj"                      "/ɲ/"                  ""
      "о"	                     "o"                       "/ɔ/"                  ""
      "п"	                     "p"                       "/p/"                  ""
      "р"	                     "r"                       "/r/"                  ""
      "с"	                     "s"                       "/s̪/"                  ""
      "т"	                     "t"                       "/t̪/"                  ""
      "ḱ"                      "kj"                      "/c/"                  ""
      "у"	                     "u"                       "/u/"                  ""
      "ф"	                     "f"                       "/f/"                  ""
      "х"	                     "h"                       "/x/"                  ""
      "ц"	                    "ts; c"                    "/ts̪/"                 ""
      "ч"	                     "ch"                      "/tʃ/"                 ""
      "џ"                    "dj; dzh"                   "/dʒ/"                 ""
      "ш"	                     "sh"                      "/ʃ/"                  ""
      '), TRUE)

      kable(alphabet, col.names = c("Cyrillic Grapheme", "Romanized Grapheme/Multigraph", "Phoneme", "Comment"), align = 'c') %>%
        kable_styling("bordered")
    ```

# Lenition Rules

* Voiceless consonants becomes voiced preceding voiced consonants [@Friedman2001, p. 12].

# Misc. Rules

* Voiced consonants devoice word-finally and before voiceless consonants [@Friedman2001, p. 12].
* /t̪/, /d̪/, /s̪/, /z̪/, /l/, /m/, and /j/ can geminate at morpheme boundaries (ibid.).

# References
