---
title: "Telugu"
author: "Becky Mathew"
edited by: "Emily Strand"
bibliography: te.bib
output: html_document
---

Last Updated: 2019-1-10

# Background

**Language family**: Dravidian / South-Central / Telugu

* Telugu is one of the most popular languages in India, particularly in the southern states of Andhra Pradesh and Telangana. Telugu is one of the official classical languages of India, and has significant influence from Sanskrit and Prakrit, as well as Perso-Arabic and English [@JIPA2016, p. 231].

# Phonology

## Consonants

* @JIPA2016 claims the existence of affricates /ts/ and /dz/ (p. 232), but other sources do not include them [@Jagganath1981, p. 4; @Krishnamurti1985, p. xviii]. I have opted not to included them as well.
* Consonants contrast for gemination (excluding /f/, /ʂ/, /ʃ/, /h/, and /ɳ/) [@JIPA2016, p. 234].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial     Dental   Alveolar   Retroflex   Postalveolar  Palatal   Velar    Glottal
"Stops (plain)"             "p b"     "t̪ d̪"       ""        "ʈ ɖ"         ""          ""      "k ɡ"      ""
"Stops (aspirated)"        "pʰ bʱ"   "t̪ʰ d̪ʱ"      ""       "ʈʰ ɖʱ"        ""          ""     "kʰ ɡʱ"     ""
"Affricates (plain)"         ""        ""         ""         ""         "tʃ dʒ"       ""       ""        ""
"Affricates (aspirated)"     ""        ""         ""         ""        "tʃʰ dʒʱ"      ""       ""        ""
Fricatives                   "f"       ""         "s"        "ʂ"          "ʃ"         ""       ""        "h"
Nasals                       "m"       ""         "n"        "ɳ"          ""          ""       ""        ""
Trills                       ""        ""         "r"        ""           ""          ""       ""        ""
Approximants                 "w"       ""         "l"        "ɭ"          ""          "j"      ""        ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Dental", "Alveolar", "Retroflex", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 8)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```  

## Vowels

* All Telugu vowel phonemes contrast in length [@Jagganath1981, p. 4].
* @JIPA2016 claims that /ɛ/ and /æ ː/ are phonemes because they contrast in English loanwords (pp. 235-236). Older works on Telugu phonology do not include /ɛ/ and /æ ː/ as vowel phonemes [@Jagganath1981, p. 4; @Krishnamurti1985, p. xviii]. I've also chosen not to include /ɛ/ and /æ ː/.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         ""           "u"
Mid       "e"         ""           "o"
Low       ""          "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

* Like other Dravidian languages that have an abugida alphabet, Telugu consonants have a following inherent vowel [@Daniels1996, p. 416].
* The candrabindu (ఁ; ఀ) generally indicates vowel nasalization [@NBGP2019, 6]; however, because vowel length is not phonemic and given that (present day) speakers don't use it, I have opted not to include it.  

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
"Consonant Grapheme"      Phoneme          Comment
"క"                        "/ka/"            ""
"చ"                       "/tʃa/"           ""
"ట"                       "/ʈa/"            ""
"త"                       "/t̪a/"            ""
"ప"                       "/pa/"            ""
"య"                      "/ja/"            ""
"శ"                        "/ʃa/"            ""
"ఖ"                       "/kʰa/"           ""
"ఛ"                      "/tʃʰa/"           ""
"ఠ"                       "/ʈʰa/"            ""
"థ"                       "/t̪ʰa/"           ""
"ఫ"                       "/pʰa/"           ""
"ర"                       "/ra/"            ""
"ష"                       "/ʂa/"            ""
"గ"                        "/ɡa/"           ""
"జ"                       "/dʒa/"           ""
"డ"                       "/ɖa/"            ""
"ద"                       "/d̪a/"            ""
"బ"                       "/ba/"            ""
"ల"                       "/la/"            ""
"స"                       "/sa/"            ""
"ఘ"                      "/ɡʱa/"           ""
"ఝ"                     "/dʒʱa/"           ""
"ఢ"                       "/ɖʱa/"           ""
"ధ"                       "/d̪ʱa/"           ""
"భ"                       "/bʱa/"           ""
"వ"                       "/wa/"            ""
"హ"                       "/ha/"            ""
"**Independent Vowel**"    ""               ""
"అ"                       "/a/"             ""
"ఇ"                       "/i/"             ""
"ఉ"                       "/u/"             ""
"ఋ"                     "/ru"              ""
"ఌ"                      "/lu/"             ""
"ఎ"                       "/e/"              ""
"ఐ"                       "/aj/"            ""
"ఓ"                       "/oː/"            ""
"ఆ"                       "/aː/"            ""
"ఈ"                       "/iː/"            ""
"ఊ"                       "/uː"             ""
"ౠ"                     "/ruː/"           ""
"ౡ"                      "/luː/"           ""
"ఏ"                       "/eː/"            ""
"ఒ"                       "/o/"             ""
"ఔ"                       "/aw/"            ""
**Diacritic**               ""              ""
"ా"                        "/aː/"           ""
"ి"                        "/i/"            ""
"ీ"                        "/iː/"           ""
"ు"                       "/u/"            ""
"ూ"                      "/uː/"           ""
"ె"                        "/e/"            ""
"ే"                        "/eː/"           ""
"ై"                        "/aj/"           ""
"ొ"                        "/o/"            ""
"ో"                        "/oː/"            ""
"ౌ"                        "/aw/"            ""
"ం"                       "/m/"       "called an anusvara, this diacritic represents a homoorganic nasal with the following consonant, but /m/ is used underlyingly [@Daniels1996, p. 415]"
"ః"                        "/h/"       "called the visarga, this diacritic occurs predominantly in Sanskrit loanwords, but also in Telugu (which is why I kept it in transcription) (ibid.)"
"్"                         ""          "called the virama, this diacritic cancels the inherent vowel [@Daniels1996, p. 416]"
'), TRUE)

kable(alphabet, col.names = c("Consonant Grapheme", "Phoneme", "Comment"), align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* /r/ is a flap [ɾ] intervocalically [@JIPA2016, p. 235].

# References
