---
title: "Mamasa"
author: "Emily Strand"
bibliography: mqj.bib
output: html_document
---

Last updated: 2020-03-03

# Background

**Language family**: Austronesian / Malayo-Polynesian / Western Malayo-Polynesian / Sulawesi / South Sulawesi / Northern / Toraja-Sa'dan

* Mamasa is spoken in the South Sulawesi and West Sulawesi provinces of Indonesia.

# Phonology

## Consonants

* Voiceless stops, nasals, /s/, /l/, and /r/ geminate [@Matti1991, p. 57].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial    Alveolar  Palatal     Velar    Glottal
Stops                      "p b"     "t d"     ""          "k ɡ"    "ʔ"
Fricatives                 ""        "s"       ""          ""       ""
Nasals                     "m"       "n"       ""          "ŋ"      ""
Trills                     ""        "r"       ""          ""       ""
Approximants               "w"       "l"       "j"         ""       ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 5)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```

## Vowels

* Vowel length is not considered to be a phonemic feature [@Matti1991, p. 58]. Rather, adjacent like-vowels are interpreted separately with an intervening epenthetic glottal stop.
* Additionally, adjacent vowels do not constitute complex nuclei (ibid.).

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High     "i"          ""           "u"         
Mid      "e"          ""           "o"
Low      ""           "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```  

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme
"a"            "/a/"
"b"            "/b/"
"d"            "/d/"
"e"            "/e/"
"g"            "/ɡ/"
"i"            "/i/"
"k"            "/k/"
"l"            "/l/"
"m"            "/m/"
"n"            "/n/"
"o"            "/o/"
"p"            "/p/"
"r"            "/r/"
"s"            "/s/"
"t"            "/t/"
"u"            "/u/"
"w"            "/w/"
"y"            "/j/"
"\'"           "/ʔ/"
**Digraph**    ""
"ng"           "/ŋ/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Syllables in Mamasa follow the order of (C)V(C) [@Matti1991, p. 69].

# Lenition Rules

* /t/ varies freely with [tʃ] [@Matti1991, p. 58].

# References
