---
title: "Komba"
author: "Emily Strand"
bibliography: kpf.bib
output: html_document
---

Last Updated: 2019-08-08

# Background

**Language Family**: Trans-New Guinea / Main Section / Central and Western / Huon-Finisterre / Huon / Western

* Komba is spoken in Papua New Guinea.

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial    Alveolar   Palatal   Velar
"Stops"                     "p b"     "t d"       ""   "k kʷ ɡ ɡʷ"
"Affricates"                 ""        ""         ""       ""
Fricatives                   ""       "s z"       ""       ""
Nasals                       "m"       "n"        ""       "ŋ"
Approximants                 "w"       "l"        "j"      ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Palatal", "Velar"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 4)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced. Phonemes with the diacritic (ʷ) are labialized.", general_title = "")
```  

## Vowels

* Vowel length is phonemic in Komba [@Southwell1972, p. 5]. Long vowels are represented by duplicate graphemes.
* Adjacent like-vowels can represent one of two series word-initially, which are contrastive [@Southwell1972, p. 5; @1981, p. 2]:
    - Voiceless-voiced (represented by an ⟨h⟩ preceding the vowel)
        * These will be transcribed as long voiceless vowels (ex. /ḁ ː/)
    - Voiced-voiced
* Adjacent vowels are interpreted as a complex nucleus only when a high vowel (/u/ or /i/) follows a mid or low vowel [@Southwell1972, p. 7]. This allows for the combinations listed in the table below. /ou/ is not represented in Crúbadán.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
           Front      Central       Back
High        "i"         ""           "u"
High-Mid    "e"         ""           "o"
Low-Mid     ""          "ɐ"          ""
Low         ""          "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)

diphthongs <- read.table(textConnection('
Diphthongs
"/ɐi/, /ɐu/, /ai/, /au/, /ei/, /eu/, /oi/, /ou/"
'), TRUE)

kable(diphthongs, align = 'c') %>%
  kable_styling("bordered")
```

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme          Comment
"a"              "/a/"             ""
"â"              "/ɐ/"             ""
"b"              "/b/"             ""
"d"              "/d/"             ""
"e"              "/e/"             ""
"g"              "/ɡ/"             ""
"i"              "/i/"             ""
"k"              "/k/"             ""
"l"              "/l/"             ""
"m"              "/m/"             ""
"n"              "/n/"             ""
"ŋ"              "/ŋ/"             ""
"o"              "/o/"             ""
"p"              "/p/"             ""
"r"              "/d/"    "written intervocalically"
"s"              "/s/"             ""
"t"              "/t/"             ""
"u"              "/u/"             ""
"w"              "/w/"             ""
"y"              "/j/"             ""
"z"              "/z/"             ""
**Digraph**       ""               ""
"gw"             "/ɡʷ/"            ""
"kw"             "/kʷ/"            ""
'), TRUE)

kable(alphabet, align = "c") %>%
  kable_styling("bordered")
```

# Lenition Rules

* Voiced stops /b/, /d/, and /ɡ/ are realized as /β/, /ɾ/, and /ɣ/ intervocalically [@Southwell1972, p. 10; @1981, p. 4].

# Misc. Rules

* Voiced stops may me realized as prenasalized word-initially (only in some words) [@Southwell1972, p. 10; @1981, p. 4].
* /s/ and /z/ may be realized as [ts] and [dz] respectively [@1981, p. 4].

# References
