---
title: "Pangasinan"
author: "Abi Creighton"
edited by: "Emily Strand"
bibliography: pag.bib
output: html_document
---

Last Updated: 2020-06-18

**Slightly Compromised: Possible conflation between /ŋ/ and /nɡ/**

# Background

**Language Family:** Austronesian / Malayo-Polynesian / Western Malayo-Polynesian / Northern Philippine / Northern Luzon / South-Central Cordilleran / Southern Cordilleran / Pangasinic

* Pangasinan is spoken in the Philippines.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"         Bilabial Dental Alveolar Postalveolar Palatal Velar Glottal      
Stops                             "p b"   "t̪ d̪"     ""        ""         ""    "k ɡ"   ""       
Affricates                         ""      ""       ""       "tʃ"        ""     ""     ""
Nasals                             "m"     "n̪"      ""        ""         ""     "ŋ"    ""   
Fricatives                         ""      ""       "s"       ""         ""     ""     "h"    
Trills                             ""      "r̪"      ""        ""         ""     ""     ""   
Approximants                       ""      "l̪"      ""        ""         "j"    "w"    ""   
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Bilabial", "Dental", "Alveolar", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 7)) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

## Vowels

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
          Front    Central  Back
High      "i"        ""     "u"
Mid       "ɛ"        ""     "ɔ"
Low       ""         "a"     ""
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

* Stress is sometimes marked with an acute accent (´) over a vowel [@BentonP1971, p. 7].
* Some graphemes appearing in Spanish loanwords are often pronounced as they would be in Spanish. For example, ⟨f⟩ may be realized as [f] instead of [p] [@BentonP1971, p. 10].
* There is some ambiguity as to whether ⟨ng⟩ represents phoneme /ŋ/ or sequence /nɡ/. I will default to /ŋ/.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme                           Comment
"a"           "/a/"                               ""
"b"           "/b/"                               ""
"c"           "/k/; /s/"   "/s/: preceding front vowels; /k/: preceding back vowels"
"d"           "/d̪/"                               ""
"e"           "/ɛ/"                               ""
"f"           "/p/"                   "appears in Spanish loanwords"
"g"           "/ɡ/"                               ""
"k"           "/k/"                               ""
"h"           "/h/"                               ""
"i"           "/i/"                               ""
"j"           "/h/"                   "appears in Spanish loanwords"
"l"           "/l̪/"                               ""
"m"           "/m/"                               ""
"n"           "/n̪/"                               ""
"o"           "/ɔ/"                               ""
"p"           "/p/"                               ""
"q"           "/k/"                               ""
"r"           "/r̪/"                               ""
"s"           "/s/"                               ""
"t"           "/t̪/"                               ""
"u"           "/u/"                               ""
"v"           "/b/"                  "appears in Spanish loanwords"
"w"           "/w/"                               ""
"y"           "/j/"                               ""
"z"           "/s/"                  "appears in Spanish loanwords"
**Digraph**     ""                                ""
"ch"          "/tʃ/"                              ""
"ng"          "/ŋ/"             "also written ⟨ñg⟩ to distinguish /ŋ/ from /nɡ/ - however, this is not present in the Crúbadán corpus"
"qu"          "/k/"                               ""
'), TRUE)

kable(alphabet) %>%
  kable_styling("bordered")
```

# Lenition Rules

* /d/ may be realized intervocalically as a flap [@BentonP1971, p. 16]; /d/ and [ɾ] vary freely word-finally.
* /t/ and /k/ may be neutralized to [ʔ] syllable-finally, preceding a consonant [@BentonP1971, pp. 17-18].
* /k/ may be realized as a fricative intervocalically [@BentonP1971, pp. 18-19].

# Misc. Rules

* Nasals assimilate in place to the following consonant [@BentonP1971, p. 21].
* /l/ may be realized as [d] word-finally [@BentonP1971, p. 17].
* A non-phonemic glottal stop is epenthesized word-initially and can be epenthesized intervocalically [@BentonP1971, p. 13].

# References
