---
title: "Kwoma"
author: "Bill Mizgerd"
bibliography: kmo.bib
output: html_document
---

Last Updated: 2019-07-24

**Slightly Compromised: glottal stop not orthographically marked**

# Background

**Language Family**: Sepik / Middle Sepik / Nukuma / Kwoma

* Also referred to as Washkuk, Kwoma is spoken in the East Sepik province of Papua New Guinea.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial     Alveolar    Postalveolar     Palatal     Velar     Glottal
Stops                       "ᵐb"      "t  ⁿd"         ""             ""       "k  ᵑɡ"      "ʔ"
Affricates                  ""           ""        "tʃ  ⁿdʒ"         ""         ""         ""
Fricatives                 "ɸ  β"       "s"           "ʃ"            ""         ""         "h"
Nasals                      "m"         "n"           ""             "ɲ"        ""         ""
Flaps                       ""          "ɾ"           ""             ""         ""         ""
Approximants                "w"         ""            ""             "j"        ""         ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced. Phonemes that contain a diacritic /ᵐ/, /ⁿ/, or /ᵑ/ are prenasalized.", general_title = "")
```

## Vowels

* Based on the description of @KmoBowden1997, there appears to be three diphthongs (p. xxx). Based on the syllable structure (see Syllable Structure below), it would seem that they'd be realized as complex nuclei.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         "ɨ"         "u"
High-Mid  "e"         ""          "o"
Low-Mid   "ɛ"         ""          ""
Low       ""          "a"         ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)


diphthongs <- read.table(textConnection('
Diphthongs
"/ai/, /ei/, /ou/"
'), TRUE)

kable(diphthongs, align = 'c') %>%
  kable_styling("bordered")
```    

# Alphabet

* Although @KmoBowden1997 indicates that the glottal stop is represented by an apostrophe (p. xxx), @1981 states that it is not marked (p. 1). The Crúbadán corpus does not mark it, so the language is somewhat compromised.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme  
"a"            "/a/"
"b"            "/ᵐb/"    
"d"            "/ⁿd/"  
"e"            "/e/"  
"g"            "/ᵑɡ/"  
"h"            "/h/"  
"i"            "/i/"  
"j"            "/ⁿdʒ/"  
"k"            "/k/"  
"m"            "/m/"  
"n"            "/n/"   
"o"            "/o/"  
"p"            "/ɸ/"  
"r"            "/ɾ/"  
"s"            "/s/"    
"t"            "/t/"  
"u"            "/u/"  
"v"            "/β/"  
"w"            "/w/"  
"y"            "/j/"  
"\'"            "/ʔ/"  
**Digraph**     ""
"ch"           "/tʃ/"
"ny"           "/ɲ/"
"sh"           "/ʃ/"
"ay"           "/ai/"  
"ee"           "/ɛ/"  
"ey"           "/ei/"  
"ii"           "/ɨ/"  
"iy"           "/i/"  
"ow"           "/ou/"  
"uw"           "/u/"  
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Syllables in Kwoma follow the order of CV(C)(C) [@1981, p. 3].

# Lenition Rules

* The suffix /-ta/ spirantizes to /-tʃa/ after certain morphemes [@KmoKooyers1974, p. 9].
* The suffix /-tawa/ collapses to one of four shortened versions, depending on preceding morphemes [@KmoKooyers1974, p. 10].

# Misc. Rules

* The glottal stop only occurs word-initially or intervocalically [@1981, p. 3].

# References
