---
title: "Mountain Koiali"
author: "Emily Strand"
bibliography: kpx.bib
output: html_document
---

Last updated: 2020-02-07

# Background

**Language family**: Trans-New Guinea / Main Section / Eastern / Central and Southeastern / Koiarian / Koiaric

* Umanakaina is spoken in the Central province of Papua New Guinea.

# Phonology

## Consonants

* Rather than including /ɸ/ and /x/ [@1981, p. 1], @Garland2008 and @Garland1975 use /v/ and /h/ respectively (p. 3; p. 416). I have chosen to stick with /ɸ/ and /x/ because @Garland2008 state that the use of /v/ and /h/ merely makes for simpler phonemic symbolization. The descriptions of the phonemes later on in the paper are consistent with the use of /ɸ/ and /x/ rather than the alternative (pp. 6-7).

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial   Alveolar  Retroflex   Velar         
Stops                       "b"      "t d"       ""       "k ɡ"   
Fricatives                  "ɸ"       ""         "ʂ"       "x"    
Nasals                      "m"       "n"        ""        ""  
Approximants                ""        "l"        ""        ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Retroflex", "Velar"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 4)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```

## Vowels

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
           Front      Central       Back
High        "i"         ""           "u"
High-Mid    ""          ""           "o"
Low-Mid     "ɛ"         ""           ""
Low         ""          "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme       Phoneme
"a"             "/a/"
"b"             "/b/"
"d"             "/d/"
"e"             "/ɛ/"
"g"             "/ɡ/"
"h"             "/x/"
"i"             "/i/"
"k"             "/k/"
"l"             "/l/"
"m"             "/m/"
"n"             "/n/"
"o"             "/o/"
"s"             "/ʂ/"
"t"             "/t/"
"u"             "/u/"
"v"             "/ɸ/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Syllables in Mountain Koiali follow the order of (C)V [@Garland2008, p. 17].

# Lenition Rules

* Fricatives /ʂ/, /ɸ/, and /x/ become voiced word-medially [@Garland2008, pp. 6-8].
* /ʂ/ varies freely with [ɾ] word-medially [@Garland2008, p. 8].

# References
