---
title: "Nahuatl"
author: "Bill Mizgerd"
bibliography: nhe.bib
output: html_document
---

Last Updated: 2020-07-01

# Background

**Language Family:** Uto-Aztecan / Southern Uto-Aztecan / Aztecan / General Aztec / Aztec

* Nahuatl has 31 varieties, many of which are mutually non-intelligible; I have chosen to focus on Huasteca Nahuatl, since it's the most widely spoken variety. Huasteca Nahuatl is sometimes divided into three languages (Eastern, Western, and Central), but according to Ethnologue the Huasteca Nahuatl languages have about 85% mutual intelligibility; where differences have been drawn, I have preferred Eastern Huasteca Nahuatl (ISO code [nhe]), since that is the most spoken variant.

# Phonology

## Consonants

* @NheKimball1990 uses /λ/, which indicates /dɮ/, in place of /tɬ/ (p. 196); however, this seems to be a minority view. For instance, @NheBeller1979 outright states that /t͡ɬ/ is voiceless (p. 204).
* @NheBeller1979 include /ɡ/ as a phoneme that occurs, although rarely, in native words (p. 204); however, the Crúbadán corpus does not feature ⟨g⟩ except in loaned proper nouns, so based on current evidence I have opted to omit it.
* /ʔ/ is phonemic, but only in rare contexts [@NheBeller1979, p. 204; @NheKimball1990, p. 197].

```{r echo = FALSE, message = FALSE, warning = FALSE, results = 'asis'}
library(dplyr)
library(knitr)
library(kableExtra)
consonants <- read.table(textConnection('
"Manner of Articulation"   Bilabial   Alveolar   Postalveolar   Palatal    Velar  Labiovelar   Glottal
Stops                      "p"        "t"        ""             ""         "k"    "kʷ"         "ʔ"
Affricates                 ""         "ts tɬ"    "tʃ"           ""         ""     ""           ""
Fricatives                 ""         "s"        "ʃ"            ""         ""     ""           "h"
Nasals                     "m"        "n"        ""             ""         ""     ""           ""
Trills                     ""         "r"        ""             ""         ""     ""           ""
Approximants               ""         "l"        ""             "j"        ""     "w"          ""
'), header = TRUE)
kable(consonants, col.names = c("Manner of Articulation", "Bilabial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Labiovelar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 7)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote("Note: The alveolar affricate on the right is lateral.", general_title = "")
```

## Vowels

* Vowel length (short vs. half-long) has become rather marginal, the distinction not often retained, except in minimal pairs, which speakers try to avoid using [@NheBeller1979, p. 205; @NheKimball1990, p. 197]. Where vowel length is contrasted, it is represented by duplicate vowel characters (due to the lack of use, the frequency in Crúbadán is quite low).

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
       Front    Central   Back
High   "i"      ""        ""
Mid    "e"      ""        "o"
Low    ""      "a"        ""
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

* The orthography used will be that of SIL International; the extant Nahuatl corpora are largely based on Bible translations, which were generally done by SIL.
* The "saltillo" is the term used to collectively refer to the glottal consonants (/ʔ/ and /h/) in Nahuatl. Although they are typically considered interchangeable, both consonants are considered phonemic in this dialect. There are three characters sometimes used to indicate the saltillo: ⟨h⟩, ⟨j⟩, or ⟨'⟩. In assigning each character a specific phoneme, I have worked from my understanding of which is more common, represented below in the table.
    - A grave accent (\`) on a vowel indicates that the vowel is followed by a glottal stop.
* The SIL orthography suggests a few alternate characters/digraphs that I have removed due to their extreme scarcity in the Crúbadán corpus:
    - ⟨b⟩ was only attested in proper names, and in loanwords with fewer than 50 appearances.
    - ⟨v⟩ and ⟨'⟩ were unattested.
    - ⟨ts⟩ was only attested in a single loaned proper noun, which had fewer than 10 appearances in the corpus.
    - Some characters (⟨k⟩ and ⟨w⟩, and the digraph ⟨uh⟩) were similarly scarce in the Crúbadán corpus, but were attested in other sources, so I have preserved them in the rule set.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme                   Comment
"a"            "/a/"                       ""
"c"          "/s/; /k/"          "/s/: before a front vowel; /k/: before a back vowel"  
"ç"            "/s/"                 "rather uncommon"  
"e"            "/e/"                       ""
"h"            "/ʔ/"                       ""
"i"            "/i/"                       ""
"j"            "/h/"                       ""
"k"            "/k/"                       ""
"l"            "/l/"                       ""
"m"            "/m/"                       ""
"n"            "/n/"                       ""
"o"            "/o/"                       ""
"p"            "/p/"                       ""
"r"            "/r/"                       ""
"s"            "/s/"                       ""
"t"            "/t/"                       ""
"w"            "/w/"                       ""  
"x"            "/ʃ/"                       ""
"y"            "/j/"                 "occasionally [i]"
"z"            "/s/"                       ""
"\'"           "/ʔ/"                       ""
**Multigraph**  ""                        ""
"ch"           "/tʃ/"                      ""
"cu"           "/kʷ/"                      ""
"hu"           "/w/"                "syllable-initially"  
"ku"           "/kʷ/"                      ""
"qu"        "/k/; /kʷ/"      "/k/: before a front vowel; /kʷ/: before a back vowel"  
"tl"           "/tɬ/"                      ""
"ts"           "/ts/"                      ""
"tz"           "/ts/"                      ""
"uh"           "/w/"                 "syllable-finally"  
"cuh"          "/kʷ/"                "syllable-finally"  
"uhc"          "/kʷ/"                "syllable-finally"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* According to @NheBeller1979 (p. 205):
    - /w/ becomes [h] when followed by a stop.
    - /w/ and /j/ become [ʔ] when word-final.
    - Degemination of /ll/.

# Misc. Rules

* Glottal stops are appended after word-final vowels [@NheKimball1990, p. 197].
* Glottal stops occur intervocalically that straddle morpheme boundaries (ibid.).
* Syllable-final /l/ devoices [@NheBeller1979, p. 205].
* /w/ devoices when preceded by /l/ (ibid.).


# References
