---
title: "Kuku-Yalanji"
author: "Bill Mizgerd"
bibliography: gvn.bib
output: html_document
---

Last Updated: 2019-10-17


**Note: Roughly 30% of Crúbadán not transcribed - possibly due to loanwords or phonetic adaptation of the orthography**

# Background

**Language Family**: Pama-Nyungan / Yalandyic / Kuku-Yalanji

* This language has a number of different names, including Gugu-Yalanji and Guugu Yalandji.
* Kuku-Yalanji is spoken in the Queensland state of Northern Australia.

# Phonology

## Consonants

* Kuku-Yalanji does not have a phonemic voicing distinction for stops [@GvnPatz1982, p. 34]. The voiced and voieless allophones vary freely; however, stops tend to be voiced in intervocalic positions and in some consonant clusters, and voiceless elsewhere.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial    Alveolar   Retroflex    Palatal     Velar
Stops                        "p"       "t"         ""          "c"        "k"
Nasals                       "m"       "n"         ""          "ɲ"        "ŋ"
Trills                        ""       "r"         ""          ""         ""
Approximants                 "w"       "l"         "ɻ"         "j"        ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Retroflex", "Palatal", "Velar"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 5)) %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* Kuku-Yalanji vowels do not have contrastive length, and do not form diphthongs [@GvnPatz1982, p. 36].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"        ""           "u"
Low       ""         "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```      

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme     
"a"           "/a/"  
"b"           "/p/"  
"d"           "/t/"  
"i"           "/i/"  
"j"           "/c/"  
"k"           "/k/"  
"l"           "/l/"  
"m"           "/m/"  
"n"           "/n/"   
"r"           "/ɻ/"   
"u"           "/u/"  
"w"           "/w/"  
"y"           "/j/"  
**Diphthong**  ""
"ng"          "/ŋ/"  
"ny"          "/ɲ/"
"rr"          "/r/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Kuku-Yalanji has a (C)CV(C) syllable structure [@GvnOates1964, p. 23].

# Lenition Rules

* Intervocalic /t/ often taps to [ɾ] [@GvnPatz1982, p. 32].
* /c/ is typically realized as an affricate [@GvnPatz1982, p. 33].
* Intervocalic /r/ typically taps to [ɾ] [@GvnPatz1982, p. 34].
* Word-initial /j/ can be dropped before /i/ or /a/; word-initial /w/ can be dropped before /u/ [@GvnPatz1982, p. 36].

# Misc. Rules

* /l/ tends to velarize after /a/ or /u/ (ibid.).

# References
