---
title: "Inga"
author: "Emily Strand"
bibliography: inb.bib
output: html_document
---

Last updated: 2020-03-06

# Background

**Language Family**: Quechuan / Quechua II / B

* Inga is spoken in Colombia.

# Phonology

## Consonants

* /b/, /d/, /ɡ/, and /f/ appear in Spanish loanwords [@Caudmont1953, p. 363].
    - @Levinsohn1976 (p. 23) and @McDowell2014 (p. 194) include the voiced stops in the phonemic inventory, most likely because the language has been majorly influenced by them (voiced stops make up 44.5% of the Crúbadán corpus).

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial    Dental  Postalveolar  Palatal     Velar    
Stops                      "p"       "t̪"     ""            ""          "k"     
Affricates                 ""        ""      "tʃ"          ""          ""         
Fricatives                 ""        "s̪"     "ʃ"           ""          "x"       
Nasals                     "m"       "n̪"     ""            "ɲ"         ""          
Flaps                      ""        "ɾ"     ""            ""          ""        
Approximants               "w"       "l̪"     ""            "j ʎ"       ""        
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Dental", "Postalveolar", "Palatal", "Velar"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 5)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: The palatal approximant on the right is lateral.", general_title = "")
```

## Vowels

* /e/ and /o/ appear in Spanish loanwords [@Caudmont1953, p. 363].
* Based on the syllable structure of the language (see Syllable Structure below), adjacent vowels are interpreted as independent nuclei (one of which generally bears stress) [@Levinsohn1976, p. 21].  

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High     "i"          ""           "u"         
Low      ""           "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```  

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme        Comment  
"a"            "/a/"           ""
"c"            "/k/"     "preceding back vowels"
"i"            "/i/"           ""
"j"            "/x/"           ""
"k"            "/k/"           ""
"l"            "/l̪/"           ""
"m"            "/m/"           ""
"n"            "/n̪/"           ""
"ñ"            "/ɲ/"           ""
"p"            "/p/"           ""
"r"            "/ɾ/"           ""
"s"            "/s̪/"           ""
"t"            "/t̪/"           ""
"u"            "/u/"           ""
"w"            "/w/"           ""
"y"            "/j/"           ""
**Digraph**    ""              ""
"ch"           "/tʃ/"          ""
"ll"           "/ʎ/"           ""
"qu"           "/k/"     "preceding front vowels"
"sh"           "/ʃ/"           ""
"hu"           "/w/"           ""
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Syllables in Inga follow the order of (C)(C)V(C) [@Levinsohn1976, p. 21].

# Lenition Rules

* Voiceless stops become voiced following nasal consonants [@Bilbao1976, p. 21].
* /p/ may spirantize preceding /x/ (ibid.).
* /s/ may become voiced preceding /n/ [@Bilbao1976, p. 22].
* In loanwords [@Levinsohn1976, p. 24]:
    - /b/ becomes [ɸ] preceding voiceless consonants.
    - /b/ varies freely with [v] word-initially.
    - /d/ varies freely with [ð] intervocalically.
    - /ɡ/ becomes [ɣ] preceding /w/.
    - /ɡ/ becomes [x] word-initially (preceding high vowels), word-finally, intervocalically, or preceding voiceless consonants.

# Misc. Rules

* /n/ assimilates in place preceding velar consonants [@Bilbao1976, p. 22].
* /i/ and /u/ are realized as [j] and [w] postvocalically [@Bilbao1976, p. 23].
* Vowels may become nasalized preceding nasal consonants [@Levinsohn1976, p. 25].

# References
