---
title: "Wolof"
author: "Bill Mizgerd"
bibliography: wo.bib
output: html_document
---

Last Updated: 2019-07-02

# Background

**Language Family**: Niger-Congo / Atlantic-Congo / Atlantic / Senegambian / Fula-Wolof / Wolof

* Wolof is spoken throughout Senegal.

# Phonology

## Consonants

* The authors I have read disagree about which of the prenasalized consonants exist in Wolof; I have opted to include all of the prenasalized voiced consonants in my analysis, but to exclude the prenasalized voiceless consonants, as those were less widely attested.
    - Voiceless prenasalized consonants do not occur word-initially [@Camara2006, p. 2].
* Consonants geminate, which is indicated by repetition of the grapheme [@Camara2006, pp. 1-2].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial       Alveolar     Postalveolar     Palatal     Velar      Uvular    
Stops                    "p  b  ᵐb"    "t  d  ⁿd"        ""             ""     "k  ɡ  ᵑɡ"     "q"
Affricates                   ""            ""       "tʃ  dʒ  ᶮdʒ"       ""         ""         ""
Fricatives                  "f"           "s"             ""            ""         "x"        ""
Nasals                      "m"           "n"             ""            "ɲ"        "ŋ"        ""
Flaps                       ""            "ɾ"             ""            ""         ""         ""
Approximants                "w"           "l"             ""            "j"        ""         ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Uvular"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: Where three phonemes share a cell, they are ordered voiceless, voiced, and pre-nasal.", general_title = "")
```

## Vowels

* Vowel length is contrastive in Wolof. Long vowels are written by doubling the vowel character [@Diop2012, p. 5].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
             Front      Central       Back
"High"        "i"         ""           "u"
"High-Mid"    "e"         ""           "o"
"Mid"         ""          "ə"          ""
"Low-Mid"     "ɛ"         ""           "ɔ"
"Low"         ""          "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```    

# Alphabet

* Some older orthographies use ⟨ny⟩ instead of ⟨ñ⟩ [@Gamble1985, p. v].
* When accented vowels are doubled, the diacritic is sometimes omitted on the second [@Diop2012, p. 5].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme   
"a"               "/a/"  
"à"               "/aː/"     
"b"               "/b/"  
"c"               "/tʃ/"  
"d"               "/d/"  
"e"               "/ɛ/"  
"é"               "/e/"  
"ë"               "/ə/"  
"f"               "/f/"  
"g"               "/ɡ/"  
"i"               "/i/"  
"j"               "/dʒ/"  
"k"               "/k/"  
"l"               "/l/"  
"m"               "/m/"  
"n"               "/n/"  
"ñ"               "/ɲ/"  
"ŋ"               "/ŋ/"  
"o"               "/ɔ/"  
"ó"               "/o/"  
"p"               "/p/"  
"q"               "/q/"  
"r"               "/ɾ/"  
"s"               "/s/"  
"t"               "/t/"  
"u"               "/u/"  
"w"               "/w/"  
"x"               "/x/"  
"y"               "/j/"
**Digraph**        ""
"mb"              "/ᵐb/"        
"mp"              "/ᵐp/"  
"nc"              "/ᶮtʃ/"  
"nd"              "/ⁿd/"  
"ng"              "/ᵑɡ/"  
"nj"              "/ᶮdʒ/"  
"nk"              "/ᵑk/"  
"nt"              "/ⁿt/"  
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* Short /u/ is generally realized as lax [ʊ] [@Camara2006, p. 3].
* /a/ reduces to [ə] in the final syllable of a word, unless the preceding syllable has a low or mid-low vowel [@Camara2006, p. 4].

# Misc. Rules

* Epenthesis of [ə] occurs after word-final geminated consonants; if the consonant is a voiceless stop, that [ə] is also realized as voiceless [@Camara2006, p. 1].
* Vowels nasalize before nasal or prenasalized consonants [@Camara2006, p. 4].

# References
