---
title: "Totontepec Mixe"
author: "Bill Mizgerd"
bibliography: mto.bib
output: html_document
---

Last Updated: 2020-07-01

# Background

**Language Family:** Mixe-Zoque / Mixe / Western Mixe

* Totontepec Mixe is spoken within 10 villages in the state of Oaxaca, located in Southern Mexico.

# Phonology

## Consonants

* /ʐ/ is said to be a phoneme by @MtoCrawford1963, occurring intervocalically and following nasals (p. 44). However, I feel that occurrences of this is a due to a voicing process. @MtoSchoenhals1965 state that the phoneme only occurs in loanwords (p. 300). For these reasons, I have chosen not to include it below.

```{r echo = FALSE, message = FALSE, warning = FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)

consonants <- read.table(textConnection('
"Manner of Articulation"   Labial   Alveolar   Retroflex   Palatal    Velar   Glottal
Stops                      "p"      "t  d"     ""          ""         "k  ɡ"  "ʔ"
Affricates                 ""       "ts"       ""          ""         ""      ""
Fricatives                 "v"      "s"        "ʂ"         ""         ""      "h"
Nasals                     "m"      "n"        ""          ""         ""      ""
Approximants               ""       ""         ""          "j"        ""      ""
'), header = TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Retroflex", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote("Note: For phonemes that share a cell, those on the left are voiceless, whereas those on the right are voiced.", general_title = "")
```

## Vowels

* Vowel length is contrastive in Totontepec Mixe [@MtoCrawford1963, p. 32]. Long vowels are shown by doubling the vowel characters [@MtoSchoenhals1965, p. 301].
* Stress is denoted by accented vowels; however, we do not account for stress, so accented vowels will be transcribed to their plain representations.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
       Front    Central   Back
High   "i"      "ɨ"       "u ʊ"
Mid    "e"      ""        "o"
Low    "æ"      "a"       "ʌ"
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE) %>%
  footnote("Note: The high back vowels are ordered tense and lax.", general_title = "")
```

# Alphabet

* ⟨b⟩, ⟨f⟩, ⟨k⟩, ⟨l⟩, ⟨ll⟩, ⟨r⟩, and ⟨z⟩ appear in loanwords [@MtoSchoenhals1965, pp. 299-300].  

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme                      Comment
"a"            "/a/"                         ""
"a̱"            "/ʌ/"                         ""
"c"            "/k/"                         ""
"d"            "/d/"                         ""
"e"            "/e/"                         ""
"e̱"            "/æ/"                         ""
"g"            "/ɡ/"                         ""
"i"            "/i/"                         ""
"j"            "/h/"                         ""
"k"            "/k/"                         ""
"m"            "/m/"                         ""
"n"            "/n/"                         ""
"ñ"            "/n/"             "since [ɲ] is not contrastive"
"o"            "/ʊ/"                         ""
"o̱"            "/o/" "described by @MtoSchoenhals1965 as more open than ⟨o⟩ (p. 300)"  
"p"            "/p/"                         ""
"s"            "/s/"                         ""
"t"            "/t/"                         ""
"\'"           "/ʔ/"                         ""
"u"            "/u/"                         ""
"u̱"            "/ɨ/"                         ""
"v"            "/v/"                         ""
"x"            "/ʂ/"                         ""
"y"            "/j/"                         ""
**Digraph**     ""                           ""
"ch"           "/ʂ/"                         ""
"qu"           "/k/"                         ""
"ts"           "/ts/"                        ""
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* /v/ approximates to [w] in some contexts [@MtoCrawford1963, p. 37].
* /p/ and /ts/ voice syllable-initially when preceded by nasals, or when in syllable-initial clusters [@MtoCrawford1963, pp. 38, 42].
* When intervocalic, following a long vowel, or as part of a syllable-initial cluster, /ʔ/ reduces to a laryngealization of the preceding vowel [@MtoCrawford1963, p. 45].

# Misc. Rules

* Words in Totontepec Mixe must be consonant-initial; words written with initial vowels are preceding with an unwritten yet predictable glottal stop [@MtoSchoenhals1965, p. 302].
* /v/ devoices in some coda clusters, and syllable-finally after a non-low vowel [@MtoCrawford1963, p. 38].
* Postconsonantal /j/ is usually realized as palatalization of the prior consonant [@MtoCrawford1963, p. 39].
    - The most dramatic case of this is that the sequence /ts j/ is realized as [tʃ] [@MtoCrawford1963, p. 41].
* Voiceless stops aspirate when word-final or in the coda of a stressed syllable [@MtoCrawford1963, p. 42].
* Voiceless stops at the beginning of a stop-final coda cluster aspirate [@MtoCrawford1963, p. 43].
* /n/ assimilates to the place of following velar or palatal consonants [@MtoCrawford1963, p. 44].
* Nasals devoice when in the same cluster as /h/, or when in clusters of three or more consonants [@MtoCrawford1963, pp. 44-45].
* Central vowels tend to be fronted near /j/ or /ts/ [@MtoCrawford1963, p. 47].
* In unstressed syllables, the contrast collapses between each of the following pairs of vowels: /u ʊ/ /ɨ ʌ/ /e æ/ [@MtoCrawford1963, p. 40].
* In the coda of unstressed non-final syllables, voiceless stops are unreleased [@MtoCrawford1963, p. 42].
* At the beginning of a non-stop-final coda cluster, voiceless stops are unreleased [@MtoCrawford1963, p. 43].

# References
