---
title: "Awara"
author: "Becky Mathew"
bibliography: awx.bib
output: html_document
---

Last Updated: 2020-04-13

**COMPROMISED: conflation between /nd/, /mb/, /nɡ/ and /d/, /b/, /ɡ/, respectively**

# Background

**Language Family:** Trans-New Guinea / Main Section / Central and Western / Huon-Finisterre / Finisterre / Wantoat

* Awara is spoken in the Morobe province of Papua New Guinea.

# Phonology

## Consonants

* @Quigley2002 and @Quigley2003 disagree slightly on the consonant inventory of Awara (p. 4; p. 14); I have chosen to predominantly follow the more recent source.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
library(dplyr)
library(knitr)
library(kableExtra)

consonants <- read.table(textConnection('
"Manner of Articulation"     Labial     Alveolar  Palatal     Velar      Glottal
Stops                        "p b"      "t d"     ""       "k kʷ ɡ ɡʷ"     ""
Fricatives                   "β"        "s"       ""           "ɣ"        "h"
Nasals                       "m"        "n"       ""         "ŋ ŋʷ"        ""
Approximants                 ""         "l"       "j"          ""          ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 5)) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced. Velar consonants that have the diacritic (ʷ) are labialized.", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* Both @Quigley2002 and @Quigley2003 indicate a mid central vowel (p. 4; p. 35, 37); however, they represent it as /ʌ/. I have chosen to use /ɘ/, as it's more reflective of the description.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
       Front   Central     Back
High   "i"       ""        "u"
Mid    "e"       "ɘ"       "o"
Low    ""        "a"       ""
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

* Surface level prenasalization of consonants /b/, /d/, and /ɡ/ is orthographically represented intervocalically [@Quigley2003, p. 155]. This compromises the language because there is uncertainty whether the sequence ⟨ambi⟩, for example, always represents /abi/ (with an [ambi] realization) or /ambi/ in some instances. I have chosen to transcribe the language how it appears; that is, every ⟨mb⟩ sequence, for example, transcribes to /mb/.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme               Comment
"a"             "/a/"                 ""
"ä"             "/ɘ/"                 ""
"b"             "/b/"                 ""
"d"             "/d/"                 ""
"e"             "/e/"                 ""
"g"             "/ɡ/"                 ""
"h"             "/h/"                 ""
"i"             "/i/"                 ""
"k"             "/k/"                 ""
"l; r"          "/l/"          "intervocalically"
"m"             "/m/"                 ""
"n"             "/n/"                 ""
"o"             "/o/"                 ""
"p"             "/p/"                 ""
"s"             "/s/"                 ""
"t"             "/t/"                 ""
"u"             "/u/"                 ""
"w"             "/β/"                 ""
"x"             "/ɣ/"                 ""
"y"             "/j/"                 ""
**Multigraph**   ""                   ""
"gw"            "/ɡʷ/"                ""
"kw"            "/kʷ/"                ""
"ng"            "/ŋ/"                 ""
"ngw"           "/ŋʷ/"                ""
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Syllables in Awara follow the order of (C)V(C) [@Quigley2003, p. 175].

# Lenition Rules

* Voiceless stops (including /kʷ/) lenite intervocalically at morpheme boundaries [@Quigley2003, p. 26].
* /t/ and /k/ may be realized as [l] and [ɣ] word-initially [@Quigley2003, pp. 20-21].
* /k/, /p/, and /t/ are realized as [ɣ], [w], and [l] intervocalically [@Quigley2002, p. 6].
* /pu/ may be realized as [βu] or [wu] word-initially [@Quigley2003, p. 22].

# Misc. Rules

* Voiced stops have prenasalized variants [@Quigley2003, pp. 16-17].
    - Word-initially, they are realized as prenasalized consonants.
    - Syllable-initially, following open syllables, the homorganic nasal is realized as the coda of the preceding syllable and the voiced stop is realized as the onset of the following syllable.
* Voiceless stops tend to be aspirated word-initially and syllable-initially intervocalically [@Quigley2003, p. 17].
* Glottal stop epenthesis may occur word-initially, preceding vowels [@Quigley2003, p. 18].

# References
