---
title: "Somba-Siawari"
author: "Emily Strand"
bibliography: bmu.bib
output: html_document
---

Last Updated: 2019-11-19

**COMPROMISED: ambiguity due to phonetic transcription of /p/, /t/, and /k/ intervocalically (ex. ⟨kapa⟩ and ⟨kawa⟩ both transcribe to /kapa/)**

# Background

**Language Family:** Nuclear Trans New Guinea / Finisterre-Huon / Huon / Western Huon / Kosorong-Burum-Mindik

* Also referred to as Burum, Somba-Siawari is spoken in the Morobe province of Papua New Guinea.

# Phonology

## Consonants

* /f/ is not a native phoneme of Somba-Siawari [@Olkkonen1985, p. 33]. However, after much influence from other languages (Kâte and Pidgin), it has become prevalent in the language. I have chosen to include it in the phonemic inventory.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial       Alveolar    Palatal        Velar     
Stops                      "p  b"        "t  d"       ""       "k  kʷ  ɡ  ɡʷ"
Affricates                   ""         "ts  dz"      ""            ""
Fricatives                  "f"           "s"         ""           "ɣ"  
Nasals                      "m"           "n"         ""           "ŋ"  
Trills                       ""           "r"         ""            ""
Approximants                "w"           "l"         "j"           ""   
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Palatal", "Velar"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 4)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced. The velar stops that have the diacritic (ʷ) are labialized.", general_title = "")
```

## Vowels

* Both @Olkkonen1985 and @Gasaway1997 represent the mid central vowel as /ə/ (p. 35; p. 59); however, I have chosen to use /ɵ/ as @Olkkonen1985 describes it as a close mid-high central rounded vowel.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
             Front      Central       Back
High          "i"         ""           "u"
Mid           "e"         "ɵ"          "o"
Low           ""          "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```  

# Alphabet

* Intervocalically, /p/, /t/, and /k/ are written ⟨w⟩, ⟨r⟩, and ⟨h⟩, respectively (see Lenition Rules below) [@Olkkonen1994, p. 4].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme         Phoneme
"a"               "/a/"
"b"               "/b/"
"d"               "/d/"
"e"               "/e/"
"f"               "/f/"
"g"               "/ɡ/"
"h"               "/ɣ/"
"i"               "/i/"
"j"               "/dz/"
"k"               "/k/"
"l"               "/l/"
"m"               "/m/"
"n"               "/n/"
"ŋ"               "/ŋ/"
"o"               "/o/"
"ö"               "/ɵ/"
"p"               "/p/"
"q"               "/kʷ/"
"r"               "/r/"
"s"               "/s/"
"t"               "/t/"
"u"               "/u/"
"w"               "/w/"
"y"               "/j/"
"z"               "/ts/"
**Digraph**       ""
"gw"              "/ɡʷ/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Syllables in Somba-Siawari have the following structures [@Olkkonen1985, pp. 35-36]:
    - V
    - VC
    - CV
    - CVC
    - VV
    - VVC
    - CVV
    - CVVC

# Lenition Rules

* Intervocalically /p/, /t/, and /k/ are realized as [w], [r], and [h] [@Gasaway1997 argues for [ɣ] (p. 63)], respectively [@Olkkonen1994, p. 4].
* Voiceless stops become voiced following voiced consonants [@Gasaway1997, p. 67].
* Degemination of consonants occurs through affixation [@Gasaway1997, p. 62].

# Misc. Rules

* /ɣ/ is realized as [h] word-initially [@Olkkonen1994, p. 4].
* /w/ becomes [β] preceding front vowels [@Gasaway1997, p. 60].
* Voiced stops become voiceless following voiceless consonants [@Gasaway1997, p. 67].
* [j] epenthesis occurs intervocalically at morpheme boundaries if the second vowel is not high [@Gasaway1997, p. 69].
* /i/ and /u/ are realized as glides when occurring in syllable onsets and /w/ and /j/ are realized as vowels when occurring in syllable nuclei [@Olkkonen1985, p. 30].
* Voiceless stops are aspirated word-initially and unreleased word-finally [@Olkkonen1985, p. 31].
* /kʷ/ and /ɡʷ/ can be realized as [kp] and [ɡb] in the Yaknge and Siawari dialects [@Olkkonen1994, p. 4].
* The velar and labiovelar consonants are backed preceding central and back vowels and velar consonants (unless found word-finally) [@Olkkonen1985, p. 31].

# References
