---
title: "Daga"
author: "Justin Bai"
bibliography: dgz.bib
output: html_document
---

Last updated: 2019-12-16

# Background

**Language family**: Trans–New Guinea / Dagan

* Daga is spoken in Papua New Guinea, specifically in the Central, Milne Bay, and Oro provinces.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Biabial    Alveolar    Palatal     Velar
Stops                       "p  b"     "t  d"       ""        "k  ɡ"
Nasals                       "m"        "n"         ""         ""
Trills                       ""         "r"         ""         ""
Approximants                 "w"        ""          "j"        ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Bilabial", "Alveolar", "Palatal", "Velar"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 4)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```

## Vowels

* Like many Trans-New Guinea languages, Daga includes a large amount of diphthongs (and some triphthongs) [@sil2011daga, pp. 2-3]. However, these can also be interpreted as vowel sequences [@John1972, p. 24], so they won't be accounted for in the rules.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         ""           "u"
Mid       "e"         ""           "o"
Low       ""          "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

* There is some debate as to how ⟨a⟩ is realized: @Murane1974 transcribes it as /a/ whereas @sil2011daga uses /ɑ/ (p. 5; p. 1). I have opted to use /a/ in the ruleset.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme
"a"             "/a/"
"b"             "/b/"
"d"             "/d/"
"e"             "/e/"
"g"             "/ɡ/"
"i"             "/i/"
"k"             "/k/"
"m"             "/m/"
"n"             "/n/"
"o"             "/o/"
"p"             "/p/"
"r"             "/r/"
"s"             "/t/"
"t"             "/t/"
"u"             "/u/"
"v"             "/w/"
"w"             "/w/"
"y"             "/j/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Syllables in Daga include the following structures [@sil2011daga, p. 3]:
    - V
    - VV
    - VC
    - CV
    - CVV
    - CVC
    - CVVC

# Lenition

* /t/ spirantizes to [s] preceding front vowels [@sil2011daga, p. 3].


# Misc. Rules

* There is free variation between /r/, [ɾ], and [l] [@John1972, p. 20].
* /n/ assimilates to velar stops (ibid.).
* /w/ becomes [β] preceding front vowels (ibid.).
* A phonetic glottal stop is epenthesized preceding word-initial vowels, between like-vowels, and following all word-final vowels (ibid.).

# References
