---
title: "Karachay-Balkar"
author: "Allison Shao"
bibliography: krc.bib
output: html_document
---

Last Updated: 2020-07-28

# Background

**Language Family:** Turkic / Western / Ponto-Caspian

# Phonology

## Consonants

* While @Seegmiller1996 and @Seegmiller1988 include the voiceless velar fricative /x/ (p. 6; p. 502), @CampbellKing2013 include its voiced counterpart /ɣ/ (p. 832). I have chosen to follow the  analyses with the voiceless fricative.
* Additionally, @Seegmiller1988 includes /h/ (p. 502). I have chosen to follow the more recent sources and not include this phoneme.
* Many instances of /ŋ/ phonetically result from adjacent /n/ and /ɡ/, but /ŋ/ is still considered phonemic because it occurs in other environments as well [@Seegmiller1996, p. 7].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial  Alveolar  Postalveolar  Palatal   Velar   
Stops                      "p b"    "t d"         ""         ""     "k ɡ"    
Affricates                  ""       ""        "tʃ dʒ"       ""      ""      
Fricatives                  ""      "s z"        "ʃ"         ""      "x"     
Nasals                      "m"      "n"         ""          ""      "ŋ"     
Trills                      ""       "r"         ""          ""      ""      
Approximants                "w"      "l"         ""          "j"     ""      
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 5)) %>%
  footnote("Note: For phonemes that share a cell, those on the left are voiceless, whereas those on the right are voiced.", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* Karachay-Balkar has vowel harmony based on backness and rounding [@Seegmiller1996, p. 8].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front       Back
High     "i y"      "ɯ u"         
Mid      "e ø"       "o"
Low       ""         "ɑ"
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE) %>%
  footnote("Note: For phonemes that share a cell, those on the left are unrounded, whereas those on the right are rounded.", general_title = "")
```  

# Alphabet

* Though ⟨ў⟩ is included in the alphabet to represent /w/ [@CampbellKing2013, p. 833], the Crúbadán corpus does not contain this grapheme. @Orthography2019 lists ⟨у⟩ as representing /w/ in addition to /u/, provided specific environments (e.g. when preceded by ⟨а⟩, ⟨e⟩, ⟨ё⟩, ⟨и⟩, ⟨о⟩, ⟨ы⟩, ⟨ю⟩, ⟨я⟩, ⟨у⟩). The ruleset accounts for this.
* Some graphemes may be transcribed differently in Russian loanwords. For example, grapheme ⟨ю⟩ represents /y/ in Karachay-Balkar but /ju/ in Russian. The ruleset defaults to the native Karachay-Balkar phonemes.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme          Comment
"а"              "/ɑ/"             ""
"б"              "/b/"             ""
"в"              "/v/"             "appears in loanwords"
"г"              "/ɡ/"             ""
"д"              "/d/"             ""
"e"              "/e; je/"         "/je/: word-initially"
"ё"              "/ø; jo/"         "/jo/: appears in loanwords"
"ж"              "/ʒ/"             "appears in loanwords"
"з"              "/z/"             ""
"и"              "/i/"             ""
"й"              "/j/"             ""
"к"              "/k/"             ""
"л"              "/l/"             ""
"м"              "/m/"             ""
"н"              "/n/"             ""
"о"              "/o/"             ""
"п"              "/p/"             ""
"р"              "/r/"             ""
"с"              "/s/"             ""
"т"              "/t/"             ""
"у"              "/u/; /w/"        "see above explanation"
"ў"              "/w/"             ""
"ф"              "/f/"             "appears in loanwords"
"х"              "/x/"             ""
"ц"              "/ts/"            "appears in loanwords"
"ч"              "/tʃ/"            ""
"ш"              "/ʃ/"             ""
"щ"              "/ʃtʃ/"           "appears in loanwords"
"ъ"              ""                "appears in loanwords"
"ы"              "/ɯ/"             ""
"ь"              ""                "appears in loanwords"
"э"              "/e/"             "word-initially"
"ю"              "/y; ju/"         "/ju/: appears in loanwords"
"я"              "/jɑ/"            ""
**Digraph**       ""               ""
"гъ"             "/ɡ/"             "represents [ɣ], an allophone of /ɡ/"
"дж"             "/dʒ/"            ""
"къ"             "/k/"             "represents [q], an allophone of /k/"
"нг"             "/ŋ/"             ""
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* For some speakers, /x/ is realized as [h] before front vowels. Other speakers articulate [h] in all positions [@Seegmiller1996, p. 11].
* Sequences of vowels are not allowed in Karachay-Balkar. This is resolved via vowel elision (ibid.).

# Misc. Rules

* Like other Turkic languages, /k/ and /ɡ/ have a range of different realizations, especially with respect to front and back vowels [@Seegmiller1996, p. 7].
    - Next to front vowels, /k/ and /ɡ/ are realized generally as just that, [k] and [ɡ]. However, next to back vowels, their realizations are more uvular or spirantized; /k/ may be realized as [q], and /ɡ/ may be realized as [ɣ]. Depending on the environment, /ɡ/ may also be realized as [x] or [ɢ].
* Voiced stops and /r/ devoice word-finally and when preceding other consonants [@Seegmiller1996, p. 9]. Voiced stops also frequently devoice when following voiceless consonants.
* Metathesis is common in words that have /x/ followed by a sibilant [@Seegmiller1996, p. 11].
* Consonants generally geminate through affixation (ibid.).
    - The sequence /nɡ/ is often realized as the geminate /ŋŋ/ [@Seegmiller1996, p. 10].
    - With adjacent nasal stops, depending on the status of the affix (e.g. denotes case marking), simplification/degemination occurs [@Seegmiller1996, p. 11].
* /n/ sometimes assimilates to [m] when preceding /b/, /p/, or /m/ [@Seegmiller1996, p. 10].
* /b/ surfaces as [m] preceding nasal stops (still occurs if a vowel is present between the /b/ and the nasal) [@Seegmiller1996, p. 10].
* A stem-final nasal will be realized as [l] when the adjacent suffix begins with /l/ [@Seegmiller1996, p. 10].

# References
