---
title: "Tarahumara"
author: "Bill Mizgerd"
bibliography: tac.bib
output: html_document
---

Last Updated: 2019-06-24

# Background

**Language Family**: Uto-Aztecan / Southern Uto-Aztecan / Tarahumaran / Tarahumara

* Where dialects diverge, I will be preferring Western Tarahumara, which is only the second most widely spoken dialect but is the dialect with the most robust Crubadan corpus.
* Western Tarahumara is spoken in the Chihuahua state in northern Mexico.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial  Alveolar  Postalveolar  Retroflex  Palatal  Velar   Glottal
Stops                      "p"     "t"       ""            ""         ""       "k"     "ʔ"
Affricates                 ""      ""        "tʃ"          ""         ""       ""      ""
Fricatives                 "β"     "s"       ""            ""         ""       "ɣ"     "h"
Nasals                     "m"     "n"       ""            ""         ""       ""      ""
Flaps                      ""      ""        ""            "ɽ"        ""       ""      ""
Approximants               "w"     ""        ""            "ɭ"        "j"      ""      ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Retroflex", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 7)) %>%
  column_spec(1, bold = TRUE)
```

## Vowels

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
          Front   Central   Back
High      "i"     ""        "u"
Mid       "e"     ""        "o"
Low       ""      "a"       ""
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered")
```

# Alphabet

* Loanwords sometimes contain ⟨f q z⟩ [@TacThordGray1955, pp. 26-27].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme
"a"              "/a/"  
"b"              "/β/"    
"e"              "/e/"  
"g"              "/ɣ/"  
"h"              "/h/"  
"i"              "/i/"  
"k"              "/k/"  
"l"              "/ɭ/"  
"m"              "/m/"  
"n"              "/n/"  
"o"              "/o/"  
"p"              "/p/"  
"r"              "/ɽ/"  
"s"              "/s/"  
"t"              "/t/"  
"u"              "/u/"  
"w"              "/w/"  
"y"              "/j/"  
"\'"             "/ʔ/"
**Digraph**       ""
"ch"             "/tʃ/"
"**(Spanish) Orthographic Variations**"   "[@TacThordGray1955, p. 1157]"
"c"              "/k/"  
"hu"             "/w/"  
"j"              "/h/"  
"qu"             "/k/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```   

# Lenition Rules

* When between stressed high vowels and glottal stops, word-initial /a/ reduces to [ə] [@TacBurgess1984, p. 8].
* Unstressed vowels often delete in fast speech (ibid.).
* Unstressed vowels in non-initial syllables delete when preceded by a nasal and followed by a stop (ibid.).
* Word-initial stops are sometimes voiced [@TacThordGray1955, pp. 26-27].
* /t/ sometimes approximantizes to [ɽ] (ibid.).
* Before high or front vowels, /tʃ/ spirantizes to [ʃ] [@TacValdezJara2013, p. 20].

# Misc. Rules

* /β/ and /ɣ/ despirantize word-initially to [b] and [ɡ] respectively [@TacBurgess1984, p. 7].
* Phrase-initial /ɽ/ is often produced as [r] or, less commonly, [ɾ] (ibid.).
* In phrase-initial syllables, unstressed vowels lengthen unless they're followed by a vowel or glottal consonant [@TacBurgess1984, p. 8].
* /i/ lowers in unstressed non-initial syllables (ibid.).
* Epenthesis of [j] typically occurs before word-initial vowels [@TacThordGray1955, pp. 26-27].
* /h/ is often realized as [x] [@TacValdezJara2013, p. 21].

# References
