---
title: "Anjam"
author: "Daniela Wiepert"
bibliography: boj.bib
output: html_document
---

Last Updated: 2020-06-26

# Background

**Language Family:** Trans-New Guinea / Madang-Adelbert Range / Madang / Rai Coast / Mindjim

* Anjam is spoken in the Madang province of Papua New Guinea.

# Phonology

## Consonants

* /j/ and /w/ are considered consonants in non-syllabic positions (i.e. in the onset or coda) [@rucker1990, p. 2].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"     Bilabial   Alveolar    Palatal   Velar      Uvular
Stops                        "p b"      "t d"       ""        "k ɡ"      "q"
Affricates                   ""         "dʑ"        ""        ""         ""
Fricatives                   ""         "s"         ""        ""         ""
Nasals                       "m"        "n"         "ɲ"       "ŋ"        ""
Trills                       ""         "r"         ""        ""         ""
Approximants                 ""         "l"         "j"       "w"        ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Bilabial", "Alveolar", "Palatal", "Velar", "Uvular"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 5)) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

## Vowels

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
           Front      Central       Back
High        "i"         ""           "u"
Mid         "e"         ""           "o"
Low         ""          "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

* A large percentage of the corpus (roughly 20%) seems to consist of ⟨ý⟩; however, it is unclear what this grapheme represents. At first glance, we thought it indicated stress or phonetic vowel length; however, stress is not orthographically marked and the acute accent only appears atop this one grapheme [@rucker2000, p. 4]. Looking deeper, we found that it's seemingly present only in scripture [@Anjam2013], so we have opted not to include it.
* **Note:** /ŋ/ is not represented in the Crúbadán corpus.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme                 
"a"              "/a/"                    
"b"              "/b/"                    
"d"              "/d/"                    
"e"              "/e/"                    
"g"              "/ɡ/"                    
"i"              "/i/"                    
"j"              "/dʑ/"                   
"k"              "/k/"                    
"l"              "/l/"                    
"m"              "/m/"                    
"n"              "/n/"                    
"ñ"              "/ɲ/"                    
"ŋ"              "/ŋ/"                    
"o"              "/o/"                    
"p"              "/p/"                    
"q"              "/q/"                      
"r"              "/r/"                    
"s"              "/s/"                    
"t"              "/t/"                    
"u"              "/u/"                    
"w"              "/w/"                    
"y"              "/j/"                    
'), TRUE)

kable(alphabet, align = "c") %>%
  kable_styling("bordered")
```

# Syllable Structure

* Syllables in Anjam follow the order of (C)(C)V(V)(C)(C) [@rucker2000, p. 4].

# Misc. Rules

* /p/ and /b/ neutralize to /p/ word finally [@rucker2000, p. 4].
* /dʑ/ devoices word-finally [@rucker2000, p.4].
* /i/ deletes when followed by /r/ across a morpheme boundary [@rucker2000, p. 28].

# References
