---
title: "Anjam"
author: "Daniela Wiepert"
bibliography: boj.bib
output: html_document
---

Last Updated: 2020-02-19

# Background

**Language Family**: Trans New Guinea / Madang-Adelbert Range / Mindjim

* Anjam is spoken in the Madanp province of Papua New Guinea.

# Phonology

## Consonants

* /j/ and /w/ are considered consonants in non-syllabic positions (i.e. the onset or coda) [@rucker1990, p. 2].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"     Bilabial   Alveolar    Palatal   Velar      Uvular
Stops                        "p b"      "t d"       ""        "k ɡ"      "q"
Affricates                   ""         "dʑ"        ""        ""         ""
Fricatives                   ""         "s"         ""        ""         ""
Nasals                       "m"        "n"         "ɲ"       "ŋ"        ""
Trills                       ""         "r"         ""        ""         ""
Approximants                 ""         "l"         "j"       "w"        ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Bilabial", "Alveolar", "Palatal", "Velar", "Uvular"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 5)) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

## Vowels

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
           Front      Central       Back
High        "i"         ""           "u"
Mid         "e"         ""           "o"
Low         ""          "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

* /ŋ/ does not appear in the Crúbadán corpus.
* A large percentage of the corpus (roughly 20%) seems to consist of ⟨ý⟩; however, it is not clear what this grapheme represents. I first thought it indicated stress or phonetic vowel length, however stress is not orthographically marked and the acute accent only appears atop this grapheme [@rucker2000, p. 4]. It seems as if it's present only in scripture [@Anjam2013], so I have opted not to include it.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme                 
"a"              "/a/"                    
"b"              "/b/"                    
"d"              "/d/"                    
"e"              "/e/"                    
"g"              "/ɡ/"                    
"i"              "/i/"                    
"j"              "/dʑ/"                   
"k"              "/k/"                    
"l"              "/l/"                    
"m"              "/m/"                    
"n"              "/n/"                    
"ñ"              "/ɲ/"                    
"ŋ"              "/ŋ/"                    
"o"              "/o/"                    
"p"              "/p/"                    
"q"              "/q/"                      
"r"              "/r/"                    
"s"              "/s/"                    
"t"              "/t/"                    
"u"              "/u/"                    
"w"              "/w/"                    
"y"              "/j/"                    
'), TRUE)

kable(alphabet, align = "c") %>%
  kable_styling("bordered")
```

# Syllable Structure

* Syllables in Anjam follow the order of (C)(C)V(V)(C)(C) [@rucker2000, p. 4].

# Misc. Rules

* /p/ and /b/ neutralize to /p/ word finally [@rucker2000, p. 4].
* /dʑ/ devoices word-finally [@rucker2000, p.4].

# References
