---
title: "Nehan"
author: "Emily Strand"
bibliography: nsn.bib
output: html_document
---

Last updated: 2020-01-24

# Background

**Language family**: Austronesian / Malayo-Polynesian / Central-Eastern / Eastern Malayo-Polynesian / Oceanic / Western Oceanic / Meso Melanesian / New Ireland / South New Ireland-Northwest Solomonic / Nehan-North Bougainville / Nehan

* Nehan is spoken in the Bougainville province of Papua New Guinea.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial    Alveolar   Velar   Glottal
Stops                       "p b"     "t d"     "k ɡ"     ""
Fricatives                   ""        "s"       ""      "h"
Nasals                       "m"       "n"       "ŋ"      ""
Trills                       ""        "r"       ""       ""
Approximants                 ""        "l"       ""       ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 4)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```  

## Vowels

* Vowels often appear adjacent to one another in Nehan; however, based on the syllable structure discussed below, it appears that they are interpreted as separate [@SIL, p. 3].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
           Front      Central       Back
High        "i"         ""          "u"
Mid         "ɛ"         ""          "ɔ"
Low         ""          "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme
"a"              "/a/"
"b"              "/b/"
"d"              "/d/"
"e"              "/ɛ/"
"g"              "/ɡ/"
"h"              "/h/"
"i"              "/i/"
"k"              "/k/"
"l"              "/l/"
"m"              "/m/"
"n"              "/n/"
"o"              "/ɔ/"
"p"              "/p/"
"r"              "/r/"
"s"              "/s/"
"t"              "/t/"
"u"              "/u/"
**Digraph**       ""
"ng"             "/ŋ/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Syllables in Nehan consist of the following structures [@SIL, p. 3]:
    - V
    - VC
    - CV
    - CVC

# Misc. Rules

* All stops, /m/, /s/ and /h/ labialize preceding /uV/ [@SIL, p. 3].

# References
