---
title: "Bashkir"
author: "Emily Strand"
bibliography: ba.bib
output: html_document
---

Last Updated: 2019-08-15

# Background

**Language Family**: Turkic / Western / Uralian

* Bashkir is spoken around central and western Russia.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"    Labial     Dental     Alveolar    Postalveolar   Palatal     Velar   Uvular    Glottal
Stops                       "p  b"       ""       "t  d"          ""           ""        "k  ɡ"    "q"       "ʔ"
Affricates                    ""         ""         "ts"         "tʃ"          ""          ""       ""       ""
Fricatives                  "f v"      "θ  ð"      "s  z"       "ʃ  ʒ"         ""        "x  ɣ"     ""       "h"
Nasals                       "m"         ""         "n"           ""           "ŋ"         ""       ""       ""
Trills                        ""         ""         "r"           ""           ""          ""       ""       ""
Approximants                 "w"         ""         "l"           ""           "j"         ""       ""       ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Dental", "Alveolar", "Postalveolar", "Palatal", "Velar", "Uvular", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 8)) %>%
  column_spec(1, bold = TRUE)%>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```

## Vowels

* Vowels do not occur consecutively [@poppe_bashkir_1997, p. 12].  
* /e/ and /ɔ/ only appear in loanwords [@poppe_bashkir_1997, p. 6].
* Vowel harmony also occurs [@poppe_bashkir_1997, pg. 19].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High    "i  y"        ""           "u"
Mid        ""        "ɘ ɵ"        "ɤ  o"
Low        ""         "a"          "ɑ"
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are unrounded and those on the right are rounded.", general_title = "")
```

# Alphabet

* Bashkir appears to be heavily influenced by Russian [@poppe_bashkir_1997], such that the phonemic inventory includes phonemes that are only present through Russian loanwords (p. 6). Although loanwords (graphemes) themselves are not native to Bashkir, the sounds are, so the transcription of them will be included.
* Following standard Russian orthography, ⟨ъ⟩ and ⟨ь⟩ do indicate non-palatalization and palatalization, respectively, of the preceding consonant; however, this is not native to Bashkir, so it will not be represented in the rules.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme                                    Comment
"а"            "/ɑ/"                                       ""
"б"            "/b/"                                       ""
"в"         "/w/; /v/"                        "/w/: word-initially; /v/: elsewhere and in loanwords"
"г"            "/ɡ/"                                       ""
"ғ"            "/ɣ/"                                       ""
"д"            "/d/"                                       ""
"ҙ"            "/ð/"                                       ""
"е"            "/jɘ/"                          "/jɘ/: word initially; /ɘ/ elsewhere"  
"ё"            "/jo/"                            "only occurs in loanwords"  
"ж"            "/ʒ/"                             "only occurs in loanwords"
"з"            "/z/"                             "only occurs in loanwords"
"и"            "/i/"   "/i/: word initially and within the first syllable; /ɘj/ word finally and appearing in non-initial syllables"  
"й"            "/j/"                                       ""
"к"            "/k/"                                       ""
"ҡ"            "/q/"                                       ""
"л"            "/l/"                                       ""
"м"            "/m/"                                       ""
"н"            "/n/"                                       ""
"ң"            "/ŋ/"                                       ""
"о"            "/o/"                                       ""
"ө"            "/ɵ/"                                       ""
"п"            "/p/"                                       ""
"р"            "/r/"                                       ""
"с"            "/s/"                                       ""
"ҫ"            "/θ/"                                       ""
"т"            "/t/"                                       ""
"у"            "/u/"                                       ""
"ү"            "/y/"                                       ""
"ф"            "/f/"                                       ""
"х"            "/x/"                                       ""
"һ"            "/h/"                                       ""
"ц"            "/ts/"                           "only occurs in loanwords"
"ч"            "/tʃ/"                           "only occurs in loanwords"
"ш"            "/ʃ/"                                       ""
"щ"            "/ʃtʃ/"                          "only occurs in loanwords"
"ъ"            "/ʔ/"                            "following back vowels"      
"ы"            "/ɤ/"                                       ""
"ь"            "/ʔ/"                            "following front vowels"      
"э"            "/ɘ/"                                       ""
"ә"            "/a/"                                       ""
"ю"            "/ju/"                                      ""
"я"            "/jɑ/"                                      ""
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Bashkir seems to favor complex codas over onsets, given the possible syllable structures below [@poppe_bashkir_1997, p. 12]:  
    - V  
    - VC  
    - VCC  
    - CV  
    - CVC  
    - CVCCC  

# Lenition Rules

* /b/ can undergo spirantization in fast speech [@poppe_bashkir_1997, p. 8].
* /b/ becomes a glide /w/ word-initially (ibid.).  
* /d/ undergoes spirantization in suffix onsets [@poppe_bashkir_1997, p. 23].

# Misc. Rules

* Depending on the syllable structure, mid-vowels are consistently reduced or unpronounced [@poppe_bashkir_1997, p.22].  
* /n/ assimilates to /m/ or /ŋ/ when preceding bilabial or velar consonants respectively [@poppe_bashkir_1997, p. 24].  
* There is apparent consonant harmony between /g/ and /k/ in words containing front vowels and between /ɣ/ and /q/ in words containing back vowels [@poppe_bashkir_1997, p. 22].  
* /l/ and /n/ are often realized as /t/, /d/, or /ð/ in the presence of suffixation [@poppe_bashkir_1997, p. 23].
* In the presence of vowels (e.g. intervocalically, or following a vowel), /u/ and /y/ are realized as /w/ [@poppe_bashkir_1997, pp. 28-29].

# References
