---
title: "Carib"
author: "Bill Mizgerd"
bibliography: car.bib
output: html_document
---

Last Updated: 2020-06-26

# Background

**Language Family**: Carib / Northern / Galibi

* Also known as Galibi, Carib is spoken in the north end of South America in French Guiana, Suriname, Guyana, and Venezuela.

# Phonology

## Consonants

* /ʔ/ only occurs in syllable codas, while no other stops appear in that position [@CarCourtz2008, p. 32]. It seems likely that /ʔ/ historically entered the language through a debuccalization process; however, because it is unclear where any specific instance of /ʔ/ may have occurred, and because /ʔ/ is distinctively marked in the orthography, I have opted to treat it as a full phoneme.
* Syllable-final nasals assimilate to the place of following stops, but before /s/ they are realized as [ŋ] rather than the alveolar [n] [@CarCourtz2008, p. 32]. For this reason, my ruleset will be treating /ŋ/ as a distinct phoneme that occurs in syllable codas.
* Per @CarPeasgood1972, /m/ and /ŋ/ are the only nasals that occur word-finally (p. 40); otherwise, they agree with Courtz's analysis.
* @CarHoff1968 and @CarPeasgood1972 see /b d ɡ/ as contrasting with the voiceless series of stops, but he sees the contrast as marginal, so I have found @CarCourtz2008's argument for allophonic voicing to be more compelling.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"    Labial   Alveolar  Retroflex  Palatal   Velar  Glottal
Stops                        "p"      "t"       ""         ""        "k"    "ʔ"
Fricatives                   ""       "s"       ""         ""        ""     ""
Nasals                       "m"      "n"       ""         ""        "ŋ"    ""
Flaps                        ""       ""        "ɽ"        ""        ""     ""
Approximants                 "w"      ""        ""         "j"       ""     ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Retroflex", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* In diphthongs, /i/ can follow any non-identical vowel; the other permissible diphthong is /au/ [@CarCourtz2008, p. 23].
* Older texts record vowel length as significant, but @CarCourtz2008 argues that it is unnecessary because vowel length can be predicted based on stress, which is itself predictable (p. 38).

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
       Front   Central   Back
High   "i"     "ɨ"       "u"
Mid    "e"     ""        "o"
Low    ""      "a"       ""
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)


diphthongs <- read.table(textConnection('
Diphthongs
"/ai/, /ei/, /ɨi/, /ui/, /oi/, /au/"
'), TRUE)

kable(diphthongs, align = 'c') %>%
  kable_styling("bordered")
```

# Alphabet

* Some orthographies use ⟨b d g⟩ for postnasal stops [@CarCourtz2008, p. 37], which I have accounted for in the rules.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme     Phoneme        Comment
"a"           "/a/"           ""
"e"           "/e/"           ""
"i"           "/i/"           ""
"j"           "/j/"           ""
"k"           "/k/"           ""
"m"           "/ŋ/; /m/"   "/ŋ/: in word-medial syllable codas; /m/: elsewhere"
"n"           "/ŋ/; /n/"   "/ŋ/: in syllable codas; /n/: elsewhere"  
"o"           "/o/"           ""
"p"           "/p/"           ""
"r"           "/ɽ/"           ""
"s"           "/s/"           ""
"t"           "/t/"           ""
"u"           "/u/"           ""
"w"           "/w/"           ""
"y"           "/ɨ/"        "⟨ï⟩: used in older orthographies"
"\'"          "/ʔ/"           ""
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Carib has a (C)V(V)(C) syllable structure [@CarCourtz2008, p. 23].
* Any consonant can occur syllable-initially, but the sequence /ji/ never occurs (ibid.).

# Lenition Rules

* Stops voice after unstressed nasals or /CV/ sequences  [@CarCourtz2008, p. 31].
* Syllable-final nasals sometimes delete, leaving behind just a nasalization of the preceding vowel [@CarCourtz2008, p. 32].
* /ʔ/ spirantizes to [h] before voiceless consonants (ibid.).
* @CarHoff1968, conversely, sees /ʔ/ as going to [x] instead before voiceless consonants (p. 48).
* Nasals delete before /j/; when this happens, the /j/ undergoes compensatory lengthening, and the preceding vowel may be nasalized [@CarCourtz2008, p. 33].
* /ʔ/ is produced less distinctly before /s/ [@CarCourtz2008, pp. 44-45].

# Misc. Rules

* Epenthesis of glottal consonants occurs between stressed vowels and following syllable-initial stops: if the stressed vowel is short, [ʔ] is inserted, while if long [h] is inserted [@CarCourtz2008, p. 31].
* Unstressed word-initial /ɨ/ deletes (ibid).
* High non-back vowels sometimes delete word-finally when preceded by /m/ (ibid.).
* /s/ palatalizes to [ʃ] near /i/ [@CarCourtz2008, p. 32].
* /ɽ/ becomes [ɾ] after front vowels [@CarCourtz2008, p. 32; @CarPeasgood1972, p. 37].
* /Nɽ/ sequences collapse to [n] [@CarCourtz2008, pp. 33-34].
* /j/ is sometimes realized as [w], most often between a back vowel and /a/ [@CarCourtz2008, p. 34].
* Root-initial /a/ is realized as [o] when stressed; similarly, some verb stems with stressed /o/ in the first syllable reduce it to [a] when unstressed [@CarCourtz2008, pp. 35-36].
* Epenthesis of [ʔ] occurs between adjacent vowels in different syllables [@CarCourtz2008, p. 43; @CarPeasgood1972, p. 38].
* Consonants palatalize when in a syllable with /i/ or a diphthong that includes /i/ [@CarHoff1968, pp. 31-32; @CarPeasgood1972, pp. 36-37].
* Non-labial obstruents are backed after back vowels [@CarPeasgood1972, p. 37].
* /o/ centralizes to [ɵ] after palatalized consonants [@CarPeasgood1972, p. 39].


# References
