---
title: "Chʼortiʼ"
author: "Emily Strand"
bibliography: caa.bib
output: html_document
---

Last updated: 09-12-2019

# Background

**Language family**: Mayan / Cholan-Tzeltalan / Cholan / Chorti

* Chʼortiʼ is spoken in eastern Guatemala, specifically within the Chiquimul and Zacapa departments and along the border with Honduras.

# Phonology

## Consonants

* There is some ambiguity around the phonemes /h/ and /x/. Sources indicate either as being represented by ⟨j⟩, but I have opted to follow @hull2016 and @dugan2013 in using /h/ (p. 7; p. 18).
* /b/, /d/, and /ɡ/ are said to appear in Spanish loanwords [@dugan2013, p. 18].


```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"      Bilabial   Alveolar  Postalveolar   Palatal   Velar      Glottal
Stops                          "p ɓ"     "t tʼ"        ""          ""       "k kʼ"       "ʔ"
Affricates                      ""      "ts  tsʼ"   "tʃ  tʃʼ"      ""        ""          ""  
Fricatives                      ""        "s"         "ʃ"          ""        ""          "h"
Nasals                          "m"       "n"         ""           ""        ""          ""
Trills                          ""        "r"         ""           ""        ""          ""
Approximants                    ""        "l"         ""           "j"       "w"         ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Bilabial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are plain and those on the right are ejective. The bilabial stop on the right is implosive.", general_title = "") %>%
  column_spec(1, bold = TRUE)
```  

## Vowels

* Vowel length is not necessarily contrastive in Chʼortiʼ. Although long vowels can arise in the environment where a glottal stop follows, it's not obligatory [@marhenke2019, p. 8]. In fact, the glottal stop in such positions is pronounced lenis, often followed by rearticulation of the vowel (ibid.). Therefore, I will be treating all vowels (⟨a'⟩, ⟨e'⟩, ⟨i'⟩, ⟨o'⟩, ⟨u'⟩) as the vowel followed by a glottal stop.
* Vowel length is sometimes indicated by an accent mark over the vowel or by a postvocalic ⟨h⟩ [@Fought2012, p. 45]. However, given that vowel length is not contrastive, such accented vowels will be transcribed as short vowels. In addition, the postvocalic ⟨h⟩ is purely used to separate segments, thus it has no phonemic realization (ibid.).  
* Diphthongs are said to be present in Chʼortiʼ [@marhenke2019, p. 10], which are always written with a vowel and a glide [@dugan2013, p. 28]. @dugan2013 expresses the fact that they are largely uncommon in the language (p. 18); so, given this and the fact that there isn't a wide consensus on the inventory, I have chosen not to include them in the analysis.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
       Front   Central   Back
High    "i"       ""      "u"
Mid     "e"       ""      "o"
Low     ""        "a"     ""
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme     Phoneme                        Comment
"a"           "/a/"                           ""
"b"           "/b/"                           ""
"b\'"         "/ɓ/"                           ""
"d"           "/d/"                           ""
"e"           "/e/"                           ""
"g"           "/ɡ/"                           ""
"i"           "/i/"                           ""
"j"           "/h/"                           ""
"k"           "/k/"                           ""
"k\'"         "/kʼ/"                          ""
"l"           "/l/"                           ""
"m"           "/m/"                           ""
"n"           "/n/"                           ""
"o"           "/o/"                           ""
"p"           "/p/"                           ""
"r"           "/r/"                           ""
"s"           "/s/"                           ""
"t"           "/t/"                           ""
"t\'"         "/tʼ/"                          ""
"u"           "/u/"                           ""
"w"           "/w/"                           ""
"x"           "/ʃ/"                           ""
"y"           "/j/"                           ""
"\'"          "/ʔ/"                           ""
**Digraph**   ""                              ""
"ch"          "/tʃ/"                          ""
"ch\'"        "/tʃʼ/"                         ""
"tz"          "/ts/"                          ""
"tz\'"        "/tsʼ/"                         ""
"**Orthographic Variation**"  ""              ""
"c"           "/k/"              "realized as /s/ in Spanish loanwords"
"c\'"         "/kʼ/"                          ""
"qu"          "/k/"                           ""
"q\'u"        "/kʼ/"                          ""
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* /h/ is often deleted word-initially [@hull2016, p. 8].
* Glottal stops are often unpronounced within the first syllable and word-finally [@hull2016, pp. 8-9].

# Misc. Rules

* /l/ and /r/ often appear in free variation [@hull2016, p. 7].

# References
