---
title: "Gapapaiwa"
author: "Allison Shao"
bibliography: pwg.bib
output: html_document
---

Last-updated: 2020-02-06

# Background

**Language family**: Austronesian / Malayo-Polynesian / Central-Eastern / Eastern Malayo-Polynesian / Oceanic / Western Oceanic / Papuan Tip / Nuclean / North Papuan Mainland-D'Entrecasteaux / Are-Taupota / Are

* Gapapaiwa is spoken in the Milne Bay province of Papua New Guinea.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"    Labial   Alveolar   Palatal   Velar
"Stops (plain)"             "p  b"   "t  d"       ""      "k  ɡ"
"Stops (labialized)"          ""       ""         ""      "kʷ  ɡʷ"
Fricatives                    "v"      "s"        ""        "ɣ"
Nasals                        "m"      "n"        ""        ""
Flaps                         ""       "ɾ"        ""        ""
Approximants                  "w"      ""         "j"       ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Palatal", "Velar"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 4)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* @McGuckin2008 describe /ɨ/ more accurately as a glide from /ɨ/ to /i/ (p. 5).
    - /ɨ/ is not as fully distributed compared to the other vowels: it is not found word-initially, following approximants, or in closed syllables.
    - @McGuckin2001 explains that the contrast between /i/ and /ɨ/ may be or may be becoming lost (p. 299). This is probably why @McGuckin2011 doesn't include it as a phoneme and rather as a phonetic realization (p. 2). I, however, have chosen to include it below.
* /ε/ can also be realized as [e]; the realization of /ε/ seems to vary by speaker (ibid.).

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
       Front   Central   Back
High   "i"     "ɨ"       "u"
Mid    "ɛ"     ""        "o"
Low    ""      "a"       ""
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme     Phoneme    
"a"           "/a/"  
"b"           "/b/"  
"d"           "/d/"  
"e"           "/ε/"  
"g"           "/ɡ/"  
"i"           "/i/"  
"k"           "/k/"  
"m"           "/m/"  
"n"           "/n/"  
"o"           "/o/"  
"p"           "/p/"  
"r"           "/ɾ/"  
"s"           "/s/"  
"t"           "/t/"  
"u"           "/u/"  
"v"           "/v/"  
"w"           "/w/"  
"y"           "/j/"   
**Digraph**    ""     
"gh"          "/ɣ/"   
"ii"          "/ɨ/"   
"gw"          "/ɡʷ/"  
"kw"          "/kʷ/"  
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```  

# Syllable Structure

* Syllables in Gapapaiwa follow the order of (C)V(V)(C) [@McGuckin2008, p. 7].
    - Only /m/, /n/, and /s/ can appear syllable-finally.

# Lenition Rules

* /ɾ/ may be realized as [l] [@McGuckin2001, p. 297].
* /p/ may be realized as a bilabial fricative (ibid.).

# References
