---
title: "Rabinal Achi'"
author: "Emily Strand"
bibliography: acr-x-rabinal.bib
output: html_document
---

Last Updated: 2020-03-30

**Slightly Compromised: suspect marking of vowel length**

# Background

**Language family**: Mayan / Quichean-Mamean / Greater Quichean / Quichean / Quiche-Achi

* Rabinal Achi' is spoken in Guatemala. It is one of two dialects of Achi. The other dialect is known as Cubulco.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"    Labial   Alveolar  Postalveolar  Palatal  Velar  Uvular  Glottal
Stops                       "p ɓ"     "t t’"       ""          ""    "k k’"  "q q’"    "ʔ"
Affricates                   ""      "ts ts’"   "tʃ tʃ’"       ""      ""     ""       ""  
Fricatives                   ""        "s"        "ʃ"          ""      ""     ""       "h"  
Nasals                       "m"       "n"         ""          ""      ""     ""       ""
Flaps                        ""        "ɾ"         ""          ""      ""     ""       ""  
Approximants                 "w"       "l"         ""          "j"     ""     ""       ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Uvular", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 7)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are plain and those on the right are ejectives. The labial stop on the right is an implosive.", general_title = "")
```

## Vowels

* Vowel length is contrastive in Rabinal Achi' [@Pellicer2005, p. 15]. Long vowels are indicated by duplicate vowel graphemes.
    - All the literature pertaining to the language suggests the marking of vowel length; however, the output of the Crúbadán corpus (only 33 accounts reflecting vowel length) makes this suspect.
    - Unlike the languages in the Western branch of Proto-Mayan that have undergone neutralization of vowel length, languages in the Eastern branch (including Achi) have not [@England2017].
    - I have chosen to categorize this language as compromised, given the suspect nature of the vowel length marking.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         ""           "u"
Mid       "e"         ""           "o"
Low        ""         "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme       Phoneme
"a"             "/a/"
"b; b\'"        "/ɓ/"
"e"             "/e/"
"i"             "/i/"
"j"             "/h/"
"k"             "/k/"
"k\'"           "/k’/"
"l"             "/l/"
"m"             "/m/"
"n"             "/n/"
"o"             "/o/"
"p"             "/p/"
"q"             "/q/"  
"q\'"           "/q’/"
"r"             "/ɾ/"
"s"             "/s/"
"t"             "/t/"
"t\'"           "/t’/"
"u"             "/u/"
"w"             "/w/"
"x"             "/ʃ/"
"y"             "/j/"
"\'"            "/ʔ/"
**Digraph**     ""
"ch"            "/tʃ/"
"ch\'"          "/tʃ’/"
"tz"            "/ts/"
"tz\'"          "/ts’/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Misc. Rules

* Voiceless stops are aspirated word-finally and following consonants [@Solares2008, pp. 18-20].
* Phonetic glottal stops precede vowels in word-initial positions [@Solares2008, p. 21].

# References
