---
title: "Ukrainian"
author: "Bill Mizgerd"
bibliography: uk.bib
output: html_document
---

Last Updated: 2019-08-05

# Background

**Language Family**: Indo-European / Balto-Slavic / Slavic / East Slavic / Ukrainian

* Ukrainian is spoken throughout Ukraine and in several other Baltic states.

# Phonology

## Consonants

* @UkMarschall2016 argue for /ʋ/ rather than /v/ (p. 350).
* Any Ukrainian consonant can be allophonically palatalized, but this palatalization is only phonemic for alveolar consonants [@UkMarschall2016, p. 350].
* Consonant gemination exists, but it is controversial whether consonant length is contrastive; @UkMarschall2016 dispute it, arguing instead for simple consonant sequences (p. 353).

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial  "Alveolar (plain)"  "Alveolar (palatalized)"  Postalveolar  Palatal  Velar  Glottal
Stops                      "p b"   "t d"               "tʲ dʲ"                   ""            ""       "k ɡ"  ""
Affricates                 ""      "ts dz"             "tsʲ dzʲ"                 "tʃ dʒ"       ""       ""     ""
Fricatives                 "f v"   "s z"               "sʲ zʲ"                   "ʃ ʒ"         ""       "x"    "ɦ"
Nasals                     "m"     "n"                 "nʲ"                      ""            ""       ""     ""
Trills                     ""      "r"                 "rʲ"                      ""            ""       ""     ""
Approximants               ""      "l"                 "lʲ"                      ""            "j"      ""     ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar (plain)", "Alveolar (palatalized)", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 7)) %>%
  footnote("Note: For phonemes that share a cell, those on the left are voiceless, whereas those on the right are voiced.", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* Vowel length is not contrastive in Ukrainian. [@UkMarschall2016, p. 353]

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
          Front   Central   Back
High      "i ɪ"   ""        "u"
Mid       "ɛ"     ""        "ɔ"
Low       ""      "a"       ""
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered") %>%
  footnote("Note: The high front vowels are tense and lax, respectively..", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

* **NOTE:** During the Soviet period, the character ⟨ґ⟩ was not used, and ⟨г⟩ represented both /ɦ/ and /ɡ/. Although /ɡ/ is relatively rare in Ukrainian, this conflict still compromises the language for corpora that are based on Soviet-era texts [@UkPugh1999, p. 20; @UkHumesky1980, p. 17].
* Acute accents are sometimes used to mark stress, but these seem to be rare outside of academic/didactic sources [@UkPugh1999, p. 24].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme                Comment
"а"              "/a/"                   ""
"б"              "/b/"                   ""
"в"              "/v/"                   ""
"г"              "/ɦ/"                   ""
"ґ"              "/ɡ/"                   ""
"д"              "/d/"                   ""
"е"              "/ɛ/"                   ""
"є"              "/jɛ/"       "palatalizes preceding consonant"  
"ж"              "/ʒ/"                   ""
"з"              "/z/"                   ""
"и"              "/ɪ/"                   ""
"і"              "/i/"        "palatalizes preceding consonant"  
"ї"              "/ji/"   "always indicates two separate sounds [@UkMedushevsky1963, p. 11]"  
"й"              "/j/"                   ""
"к"              "/k/"                   ""
"л"              "/l/"                   ""
"м"              "/m/"                   ""
"н"              "/n/"                   ""
"о"              "/ɔ/"                   ""
"п"              "/p/"                   ""
"р"              "/r/"                   ""
"с"              "/s/"                   ""
"т"              "/t/"                   ""
"у"              "/u/"                   ""
"ф"              "/f/"                   ""
"х"              "/x/"                   ""
"ц"              "/ts/"                  ""
"ч"              "/tʃ/"                  ""
"ш"              "/ʃ/"                   ""
"щ"              "/ʃtʃ/"                 ""
"ь"                ""        "marks palatalization of preceding consonant"  
"ю"              "/ju/"      "palatalizes preceding consonant"  
"я"              "/ja/"      "palatalizes preceding consonant"  
"\'"               ""        "marks non-palatalization of preceding consonant in an otherwise palatalizing environment"
**Digraph**        ""                    ""
"дж"             "/dʒ/"                  ""
"дз"             "/dz/"                  ""
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```  

# Lenition Rules

* The palatalized trill, /rʲ/, tends to flap to [ɾʲ] [@UkMarschall2016, p. 352].
* /v/ approximantizes to [w] before rounded back vowels (ibid.).
* /v/ approximantizes to [w] when preconsonantal or word-final [@UkPugh1999, p. 27; @UkHumesky1980, p. 18].

# Misc. Rules

* Non-palatalized /l/ tends to velarize to [ɫ] [@UkMarschall2016, p. 352].
* /ɦ/ fronts to [x] before /k/ (ibid.).
* Syllable-final glides become the non-syllabic vowels [i̯] and [u̯] (ibid.).

# References
