---
title: "Ossetic"
author: "Bill Mizgerd"
bibliography: os.bib
output: html_document
---

Last Updated: 2019-12-02

**COMPROMISED: conflation among /u/, /w/, /ʷ/; some orthographic ambiguity related to consonant gemination**

# Background

**Language Family:** Indo-European / Indo-Iranian / Iranian / Eastern / Northeastern / Ossetic

* Also known as Ossetian, it is spoken in Russia and Georgia.
* Where dialects diverge, I will be following the more widely spoken Iron dialect.

# Phonology

## Consonants

* A small number of southern dialects include /ʃ/ and /ʒ/ [@OsTesten1997, p. 710].
* @OsAbaev1964 and @OsTesten1997 regard /j/ and /w/ as distinct phonemes (p. 6; p. 711), while @OsThordarson1989 argues that they are just allophones of [i] and [u] respectively (p. 462). I have chosen to include them in the inventory below.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial    Alveolar     Postalveolar  Palatal  Velar     Labiovelar    Uvular  Labiouvular
Stops                      "p b pʼ"  "t d tʼ"     ""            ""       "k ɡ kʼ"  "kʷ ɡʷ kʷʼ"   "q"     "qʷ"
Affricates                 ""        "ts dz tsʼ"  "tʃ dʒ tʃʼ"   ""       ""        ""            ""      ""
Fricatives                 "f v"     "s z"        ""            ""       ""        ""            "χ ʁ"   "χʷ ʁʷ"
Nasals                     "m"       "n"          ""            ""       ""        ""            ""      ""
Trills                     ""        "r"          ""            ""       ""        ""            ""      ""
Approximants               ""        "l"          ""            "j"      ""        "w"           ""      ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Labiovelar", "Uvular", "Labiouvular"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 8)) %>%
  footnote("Note: From left to right, phonemes that share a cell are voiceless, voiced, and ejective.", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* @OsThordarson1989 argues that /æ/ is actually realized closer to [ɐ] (p. 459).

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
          Front   Central   Back
High      "i"     "ɨ"        "u"
Mid       "e"     ""         "o"
Low       "æ"     "a"        ""
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

* There is some amount of orthographic confusion regarding ⟨y⟩, which can be /u/, /w/, or simply a marker of labialization. Therefore, the language is compromised. In the ruleset, I use ⟨y⟩ to mark labialization wherever possible, and default to /u/ otherwise.
    - **Note:** Due to this, /w/ is not represented.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme                  Comment
"а"              "/a/"                      ""
"ӕ"              "/æ/"                      ""
"б"              "/b/"                      ""
"в"              "/v/"                      ""
"г"              "/ɡ/"                      ""
"д"              "/d/"                      ""
"е"              "/e/"                      ""
"з"              "/z/"                      ""
"и"              "/i/"                      ""
"й"              "/j/"                      ""
"к"              "/k/"                      ""
"л"              "/l/"                      ""
"м"              "/m/"                      ""
"н"              "/n/"                      ""
"о"              "/o/"                      ""
"п"              "/p/"                      ""
"р"              "/r/"                      ""
"с"              "/s/"                      ""
"т"              "/t/"                      ""
"у"              "/u/; /w/"            "/u/: default"
"ф"              "/f/"                      ""
"х"              "/χ/"                      ""
"ц"              "/ts/"                     ""
"ч"              "/tʃ/"                     ""
"ы"              "/ɨ/"                      ""
**Multigraph**   ""                         ""
"гу"             "/ɡʷ/"                     ""
"гъ"             "/ʁ/"                      ""
"гъу"            "/ʁʷ/"                     ""
"дж"             "/dʒ/"                     ""
"дз"             "/dz/"                     ""
"ку"             "/kʷ/"                     ""
"къ"             "/kʼ/"                     ""
"къу"            "/kʷʼ/"                    ""
"пъ"             "/pʼ/"                     ""
"тъ"             "/tʼ/"                     ""
"ху"             "/χʷ/"                     ""
"хъ"             "/q/"                      ""
"хъу"            "/qʷ/"                     ""
"цъ"             "/tsʼ/"                    ""
"чъ"             "/tʃʼ/"                    ""
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```  

# Lenition Rules

* Word-medial consonants voice after voiced consonants or vowels [@OsAbaev1964, p. 8].
* In a few specific words, /z/ approximantizes to [r] and /n/ approximantizes to [l] (ibid.).

# Misc. Rules

* Epenthesis of [h] occurs before word-initial vowels [@OsThordarson1989, p. 465].
* Gemination (which systematically occurs morphophonologically) causes stops to devoice [@OsTesten1997, pp. 711, 713-15]. It however is not consistently marked orthographically [@OsAbaev1964, p. 9]; thus, it is not reflected in the rule set.

# References
