---
title: "Aomie"
author: "Abi Creighton"
bibliography: aom.bib
output: html_document
---

Last Updated: 2020-08-04

# Background

**Language Family:** Trans-New Guinea / Main Section / Eastern / Central and Southeastern / Koiarian / Baraic

* Aomie, also called Ömie or Omie, is a Trans-New Guinean language spoken along the Mamama River in the Upper Kumusi Valley.

# Phonology

## Consonants

* The contrast between /p/ and /h/ has neutralized for some speakers [@austing_1975, p. 517].
* Depending on the age and location/dialect of Aomie speakers, contrasts may be made between what have been considered allophones of /p/ and /h/ [@dutton_1969, p. 142]. For example, younger speakers may additionally contrast /pʰ/, /ɸ/, and /f/. I, however, opt to treat these as allophones just as @dutton_1969 as well as @austing_1975 do.
* /k/ is extremely marginal, often replaced with /ʔ/ [@austing_1975, p. 519].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"    Labial   Alveolar   Postalveolar  Velar   Glottal
Stops                       "p b"     "t d"         ""        "k ɡ"     "ʔ"
Affricates                   ""        ""           "dʒ"       ""       ""
Fricatives                   "v"       "s"          ""         ""       "h"  
Nasals                       "m"       "n"          ""         ""       ""
Flaps                        ""        "ɾ"          ""         ""       ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 5)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```  

## Vowels

* @austing_1975 include a phonemic mid-central vowel, interpreted as /ə/ (p. 518). However, @dutton_1969 argues that it is actually an allophone of /o/ (although he recognizes its phonemic status in other dialects - e.g. Asapa). I have opted to treat [ə] as allophonic, and not include it below.
* Based on the syllable structure of Aomie (see Syllable Structure below), vowels that appear adjacent to one another constitute complex phonemes [@austing_1975, p. 518]. Complex nuclei may consist of up to three vowels.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
             Front      Central       Back
High          "i"         ""          "u"
High-Mid      "e"         ""          "o"
Low-Mid       "æ"         ""          "ɔ"
Low           ""         "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)


diphthongs <- read.table(textConnection('
Diphthongs
"/ie/, /iæ/, /ia/, /io/, /iɔ/, /ae/, /ui/, /ue/, /ua/, /uo/, /uæ/"
**Triphthongs**
"/iae/, /iaæ/, /iɔe/, /iɔæ/, /uae/, /uaæ/"
'), TRUE)

kable(diphthongs, align = 'c') %>%
  kable_styling("bordered")
```     

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme                 
"a"            "/a/"                 
"b"            "/b/"
"d"            "/d/"
"e"            "/e/"     
"ë"            "/æ/"
"g"            "/ɡ/"  
"h"            "/h/"
"i"            "/i/"   
"j"            "/dʒ/"
"k"            "/k/"                       
"m"            "/m/"                       
"n"            "/n/"                       
"o"            "/o/"  
"ö"            "/ɔ/"
"p"            "/p/"                       
"r"            "/ɾ/"                 
"s"            "/s/"                       
"t"            "/t/"                       
"u"            "/u/"                       
"v"            "/v/"                       
"\'"           "/ʔ/"                       
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Word-initial syllables consist of either a (C)V or (C)(V)VV structure [@austing_1975, p. 518]. Everywhere else, syllable follow the order C(V)(V)V.

# Misc. Rules

* /v/ is generally realized as [β] or [w] [@austing_1975, pp. 517-518].
* Labialization and palatalization of consonants occurs in the language [@dutton_1969, p. 142].
* Glottal stops, although not orthographically marked, predictably precede word-initial vowels [@SIL2004, p. 4].

# References
