---
title: "Oriya"
author: "Becky Mathew"
bibliography: or.bib
output: html_document
---

# Background
Oriya (now romanized as Odia) is the official of Orissa, an eastern Indian state. There are four major dialects of Oriya, and this project focuses on Standard Oriya.
[@Jain2003]

**Language family**: Indo-European, Indo-Iranian, Indo-Aryan, Eastern zone, Oriya

# Phonology
## Vowels
Although the Oriya script includes characters that indicate vowel lengthening or vowel nasalization, those are not contrastive features. 
```{r echo = FALSE, results = 'asis'}
library(knitr)
front <- c("i","e","")
central <- c("u","","a")
back <- c("","o","ɔ")
vowels <- data.frame(front, central, back)
row.names(vowels) <- c("high", "mid", "low")
kable(vowels)
```
[@Jain2003]

## Consonants 
```{r echo = FALSE, results = 'asis'}
library(knitr)
bilabial <- c("p pʰ","b bʱ","m","","","","w")
dental <- c("t̪ t̪ʰ","d̪ d̪ʱ","n̪","","","","")
alveolar <- c("","","","ɾ","l","s","")
retroflex <- c("ʈ ʈʰ","ɖ ɖʱ","ɳ","","ɭ","","")
palatal <- c("c cʰ","ɟ ɟʱ","","","","","j")
velar <- c("k kʰ","g gʱ","ŋ","","","","")
glottal <- c("","","","","","h","")
consonants <- data.frame(bilabial, dental, alveolar, retroflex, palatal, velar, glottal)
row.names(consonants) <- c("Voiceless plosive", "Voiced plosive", "Nasal", "Tap", "Lateral", "Fricative", "Semivowel")
kable(consonants, caption = "Note: Two sounds may appear in the same cell if they have the same place and manner. Diacritics will distinguish them.")
``` 
[@Jain2003]

# Alphabet 
Oriya is descended from the Brahmi script but does not hang from a horizontal line. It is a left-to-right abugida script without capital letters. As an abugida script, every consonant carries an inherent /ɔ/ following it. 
[@Jain2003]

## Consonants
* <କ> = /k ɔ/
* <ଖ> = /kʰ ɔ/
* <ଗ> =  /g ɔ/
* <ଘ> = /gʱ ɔ/
* <ଙ> = /ŋ ɔ/
* <ଚ> = /c ɔ/
* <ଛ> = /cʰ ɔ/
* <ଜ>, <ଯ> = /ɟ ɔ/
* <ଝ> = /ɟʱ ɔ/
* <ଞ> = /n ɔ/
* <ଟ> = /ʈ ɔ/
* <ଠ> = /ʈʰ ɔ/
* <ଡ> = /ɖ ɔ/
* <ଢ> = /ɖʱ ɔ/
* <ଣ> = /ɳ ɔ/
* <ତ> = /t̪ ɔ/
* <ଥ> = /t̪ʰ ɔ/
* <ଦ> = /d̪ ɔ/
* <ଧ> = /d̪ʱ ɔ/
* <ନ> = /n̪ ɔ/
* <ପ> = /p ɔ/
* <ଫ> = /pʰ ɔ/
* <ବ> = /b ɔ/
* <ଭ> = /bʱ ɔ/
* <ମ> = /m ɔ/
* <ର> = /ɾ ɔ/
* <ଳ> = /ɭ ɔ/
* <ୱ>, <ଵ> =  /w ɔ/
* <ଶ>, <ଷ> , <ସ> = /s ɔ/ 
* <ହ> = /h ɔ/
* <ୟ> = /j ɔ/
* <ଲ> = /l ɔ/

## Independent vowels
* <ଅ> = /ɔ/
* <ଆ> = /a/
* <ଇ>, <ଈ> = /i/
* <ଉ>, <ଊ> = /u/
* <ଋ>, <ୠ> = /ɾ/
* <ଌ>, <ୡ> = /l/
* <ଏ> = /e/
* <ଐ> = /a i/
* <ଓ> = /o/
* <ଔ> = /a u/

## Diacritics
* <ା> = /a/
* <ି>, <ୀ> = /i/
* <ୁ>, <ୂ> = /o/
* <ୃ> = /ɾ/
* <େ> = /e/ 
* <ୈ> = /a i/
* <ୋ> = /o/ 
* <ୌ> = /a u/
* <ଁ> = ∅, represents nasalization which isn't contrastive
* <ଂ> = /ŋ/
* <ଃ> = /h/
* <୍> = ∅, deletes the inherent vowel from a consonant
* <଼> = ∅, represents that a sound originally came from another language

# Lenition rules 
* Aspirated consonants are deaspirated when preceded by /s/ and followed by a vowel. 
* For many speakers, geminate consonants are replaced by a single consonant (so consonant length isn't contrastive). 
[@Jain2003]