---
title: "Azerbaijani"
author: "Bill Mizgerd"
bibliography: az.bib
output: html_document
---

Last Updated: 2019-06-12

# Background

**Language Family**: Turkic / Common Turkic / Oghuz / Western Oghuz / Azerbaijani

* Azerbaijani is spoken in Iran and Azerbaijan, a country located between Eastern Europe and Western Asia.

# Phonology

## Consonants

* Although some sources argue for /c/ and /k/ as distinct phonemes, they do not have overlapping distributions, making it plausible that [c] only manifests as a fronted allophone of /k/. I have included /c/ in my consonant chart below, but do not include it by default in the ruleset.
    - Southern dialects only retain /c/ and /k/ in loanwords [@AzMokari2017, p. 209].
    - In words where northern dialects retain /c/, southern dialect speakers tend to realize the consonant as /ts/ or /tʃ/ (ibid.).
* @AzSalimi1976 argues for /ts/ and /dz/ only existing as fronted allophones of their postalveolar counterparts (p. 72).
* @AzFraenkel1962 posits /ʔ/ as existing as a non-phonemic "juncture feature" that sometimes marks syllable boundaries or appears in Arabic loans (p. 451).
* /ts/ and /dz/ are likely to be peripheral phonemes, as they are significantly less common than any other phonemes in the Crubadan corpus.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"    Labial     Alveolar  Postalveolar  Palatal   Velar     Glottal
Stops                       "p b"      "t d"     ""            "c ɟ"     "k ɡ"     ""
Affricates                  ""         "ts dz"   "tʃ dʒ"       ""        ""        ""
Fricatives                  "f v"      "s z"     "ʃ ʒ"         ""        "x ɣ"     "h"
Nasals                      "m"        "n"       ""            ""        ""        ""
Taps                        ""         "ɾ"       ""            ""        ""        ""
Approximants                ""         "l"       ""            "j"       ""        ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  footnote(general = "Note: Phonemes to the left of each cell are voiceless and phonemes to the right of each cell are voiced.", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* Vowel length is non-contrastive [@AzMokari2017, p. 210].
    - @AzSalimi1976 argues for the existence of underlying long vowels (pp. 89-91), but he describes them as being rather marginal, so I have opted to prefer Mokari's analysis.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
       Front   Back
High   "i y"   "ɯ u"
Mid    "e œ"   "o"
Low    "æ"     "ɑ"
'), TRUE)

kable(consonants, align = ('c')) %>%
  kable_styling("bordered") %>%
  footnote(general = "Note: Phonemes to the left of each cell are unrounded and phonemes to the right of each cell are rounded.", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme     Comment
"a"           "/ɑ/"         ""
"b"           "/b/"         ""
"c"           "/dʒ/"        ""
"ç"           "/tʃ/"        ""
"d"           "/d/"         ""
"e"           "/e/"         ""
"ə"           "/æ/"         ""
"f"           "/f/"         ""
"g"           "/ɟ/"         ""
"ğ"           "/ɣ/"         ""
"h"           "/h/"         ""
"x"           "/x/"         ""
"ı"           "/ɯ/"         ""
"i"           "/i/"         ""
"j"           "/ʒ/"         ""
"k"        "/c/; /k/" "non-overlapping distribution"  
"q"           "/ɡ/"         ""
"l"           "/l/"         ""
"m"           "/m/"         ""
"n"           "/n/"         ""
"o"           "/o/"         ""
"ö"           "/œ/"         ""
"p"           "/p/"         ""
"r"           "/ɾ/"         ""
"s"           "/s/"         ""
"ş"           "/ʃ/"         ""
"t"           "/t/"         ""
"u"           "/u/"         ""
"ü"           "/y/"         ""
"v"           "/v/"         ""
"y"           "/j/"         ""
"z"           "/z/"         ""
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

## Syllable Structure

* Azerbaijani has a (C)V(C) syllable structure [@AzMokari2017, p. 211].

# Lenition Rules

* Word-final affricates tend to weaken: the alveolar affricates go to [ʃ], and /tʃ/ debuccalizes to [h] [@AzMokari2017, p. 209].
* Stem-final /tʃ/ approximantizes to [j] before vowel-initial suffixes (ibid.).
  * Stem-final /x/ voices to [ɣ] in the same circumstances (ibid.).
* Velar stops palatalize when adjacent to a front vowel (ibid.).
* Vowels undergo vowel harmony based on frontness and rounding [@AzMokari2017, pp. 210-11].
* /l/ palatalizes slightly near front vowels, and velarizes slightly near back vowels [@AzSalimi1976, p. 70].
* Postvocalic /v/ sometimes approximantizes to [w] (ibid.).
  * Voiced dorsal stops similarly approximantize to [w] after round vowels (ibid.).
* /ɾ/ sometimes spirantizes initially or before non-low vowels (ibid.).
* Word-final /ɡ/ spirantizes, and sometimes also devoices, depending on dialect [@AzSalimi1976, p. 83].
* /c/ spirantizes to [ç] in preconsonantal and word-final contexts [@AzSalimi1976, p. 85].
* Stops occasionally spirantize intervocalically [@AzFraenkel1962, pp. 447-448].
* /n/ sometimes reduces to a nasalization of its preceding vowel, most commonly when word-final or preceding a continuant [@AzFraenkel1962, p. 449].

# Misc. Rules

* Voiceless stops are realized with aspiration [@AzMokari2017, p. 209].
* Stops devoice word-finally [@AzMokari2017, p. 209]; @AzFraenkel1962 argues that stops devoice word-initially as well (pp. 447-448).
* Word-finally or preconsonantally, /ɾ/ tends to be realized as [ʐ] [@AzSalimi1976, pp. 70-71].
* Many speakers realize /ʒ/ as [dʒ] [@AzFraenkel1962, p. 450].

# References
