---
title: "Kunimaipa"
author: "Emily Strand"
bibliography: kup.bib
output: html_document
---

Last Updated: 2019-10-23

# Background

**Language Family**: Trans-New Guinea / Goilalan

* Kunimaipa is spoken in the Central, Morobe, and Oro provinces of Papua New Guinea.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)

consonants <- read.table(textConnection('
"Manner of Articulation"   Labial   Alveolar    Velar    
Stops                      "p  b"    "t  d"    "k  ɡ"     
Fricatives                  "β"      "s  z"      "ɣ"    
Nasals                      "m"       "n"        "ŋ"     
Trills                      ""        "r"        ""     
Approximants                ""        "l"        ""    
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Velar"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 3)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```

## Vowels

* Although diphthongs are prevalent in the Kunimaipa language, such vowels can also appear in sequence [@Pence1966], thus, diphthongs will not be accounted for nor transcribed (p. 55).

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front   Central     Back
High     "i"       ""        "u"
Mid      "e"       ""        "o"
Low      ""        "a"        ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```    

# Alphabet

* The orthography does include the symbols ⟨ë⟩, ⟨ï⟩, ⟨ö⟩, and ⟨ü⟩, representing silent vowels in word-final positions [@Geary1992, p. 2]. However, this does not appear to be phonemic, and rather a phonological devoicing process, so I will transcribe such graphemes as voiced.
* Repetition of vowel graphemes is also represented in the orthography [@Pence1966, p. 64]. Given that vowel length is non-contrastive, it is unclear what such sequences indicate, whether it be phonetic length or independently pronounced vowels. @Pence1966 does describe such occurrences as vowel sequences, leading me to believe that each vowel is pronounced independently, so I will transcribe them as such.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme
"a"             "a"
"b"             "b"
"d"             "d"
"e"             "e"
"g"             "ɡ"
"h"             "ɣ"
"i"             "i"
"k"             "k"
"l"             "l"
"m"             "m"
"n"             "n"
"ñ"             "ŋ"
"o"             "o"
"p"             "p"
"r"             "r"
"s"             "s"
"z"             "z"
"t"             "t"
"u"             "u"
"v"             "β"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Syllables in Kunimaipa follow the order of (C)V(C) [@Pence1966, p. 56], with stress normally occurring on the penultimate syllable [@Geary1992, p.3].

# Lenition Rules

* /b/, /d/, and /ɡ/ can spirantize word-medially and word-finally in fast speech [@Geary1992, p. 3; @Pence1966, p. 60].
* /p/ and /k/ can be pronounced with frication or fully spirantize word-medially [@Geary1992, p. 4].
* /β/ and /z/ may be realized as glides [w] and [j] [@Pence1966, p. 60].
* /ɣ/ is said to be backed [@Pence1966, p. 61; @Geary1992, p. 4], making way for the deduction that it may debuccalize.

# Misc. Rules

* /b/, /d/, and /ɡ/ may be voiceless word-initially [@Geary1992, p. 3; @Pence1966, p. 60].

# References
