---
title: "Warlpiri"
author: "Bill Mizgerd"
bibliography: wbp.bib
output: html_document
---

Last Updated: 2019-08-09

# Background

**Language Family:** Australian / Pama-Nyungan / South-West / Ngarga

* Warlpiri is spoken in the Northern Territory and Western state of Australia.

# Phonology

## Consonants

* @WbpJagst1975 includes a series of prenasalized stops in his analysis (p. 21).

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial  Alveolar  Retroflex     Palatal  Velar
Stops                      "p"     "t"       "ʈ"           "ɟ"      "k"
Nasals                     "m"     "n"       "ɳ"           "ɲ"      "ŋ"
Flaps                      ""      "ɾ"       "ɽ"           ""       ""
Laterals                   ""      "l"       "ɭ"           "ʎ"      ""
Approximants               "w"     ""        "ɻ"           "j"      ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Retroflex", "Palatal", "Velar"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 5)) %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* Vowel length is contrastive [@WbpNash1980, p. 65]. Long vowels are represented by doubling the vowel character.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
          Front   Central   Back
High      "i"     ""        "u"
Low       ""      "a"       ""
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

* Word-initial /ʈ/, /ɳ/, and /ɭ/ are simply written as ⟨t⟩, ⟨n⟩, and ⟨l⟩ [@WbpNash1983, p. 2]. Because alveolar consonants do not appear word-initially [@WbpHale1977, pp. 2-3], this does not create any ambiguity. [See also @WbpJagst1975, p. 26, for this issue.]

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme  
"a"            "/a/"  
"i"            "/i/"  
"j"            "/ɟ/"  
"k"            "/k/"  
"l"            "/l/"   
"m"            "/m/"  
"n"            "/n/"   
"p"            "/p/"  
"r"            "/ɻ/"   
"t"            "/t/"    
"u"            "/u/"  
"w"            "/w/"  
"y"            "/j/"
**Digraph**     ""
"ly"           "/ʎ/"  
"ng"           "/ŋ/"
"ny"           "/ɲ/"
"rd"           "/ɽ/"
"rl"           "/ɭ/"  
"rn"           "/ɳ/"
"rr"           "/ɾ/"
"rt"           "/ʈ/"  
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Warlpiri has a CV(C) syllable structure, except word-initially, where it becomes CV(V)(C) [@WbpNash1980, p. 78].

# Lenition Rules

* Stops often gain voicing in unstressed syllables [@WbpHale1977, p. 3]; the specific conditions under which these voicing processes occur differ slightly by consonant [@WbpJagst1975, p. 22].
* Retroflex sonorants flap (/ɭ/ to [ɺ], /ɳ/ to [ɾ̃]?) after non-front vowels [@WbpJagst1975, pp. 25-26].

# References
