---
title: "Angor"
author: "Emily Strand"
bibliography: agg.bib
output: html_document
---

Last Updated: 2020-06-26

# Background

**Language Family**: Trans-New Guinea / Senagi

* Angor is spoken in within the Sandaun province in Northern Papua New Guinea.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial     Alveolar    Palatal     Velar   
Stops                    "p  b  ᵐb"  "t  d  ⁿd"     ""      "k  ɡ  ᵑɡ"
Fricatives                  "ɸ"          "s"        ""         "x"
Nasals                      "m"          "n"        ""         "ŋ"
Flaps                       ""           "ɾ"        ""         ""
Approximants                "w"          ""         "j"        ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Palatal", "Velar"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 4)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: Cells that contain three phonemes are ordered voiceless, voiced, and prenasalized.", general_title = "")
```

## Vowels

* Vowel harmony is present in Angor [@litteral_features_2008, p. 25].
* Diphthongs can also be realized as vowel sequences [@litteral_angor_1997, p. 4], so I have chosen not to represent them in the transcription.
* Based on the description of stress by @litteral_features_2008, such that the phonemic mid central vowel present in the language occurs only in unstressed syllables (pp. 23-24), I have chosen to stick with their analysis of using /ə/.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         "ɨ"          "u"
Mid       "e"         "ə"          "o"
Low        ""         "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme       Phoneme
"a"             "/a/"
"b"             "/b/"
"d"             "/d/"
"e"             "/e/"
"f"             "/ɸ/"
"g"             "/ɡ/"
"h"             "/x/"
"i"             "/i/"
"ɨ"             "/ə/"
"k"             "/k/"
"m"             "/m/"
"n"             "/n/"
"o"             "/o/"
"p"             "/p/"
"r"             "/ɾ/"
"s"             "/s/"
"t"             "/t/"
"u"             "/u/"
"ü"             "/ɨ/"
"w"             "/w/"
"y"             "/j/"
**Multigraph**   ""
"mb"            "/ᵐb/"
"nd"            "/ⁿd/"
"ng; ŋ"         "/ŋ/"
"ngg; ŋg"       "/ᵑɡ/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Syllables in Angor have the following structures [@litteral_angor_1997, p. 4]:
    - V  
    - VV  
    - VC  
    - CV  
    - CVV  
    - CVC

# Lenition Rules

* /ɸ/ and /x/ can be voiced word-medially [@litteral_angor_1997, p. 4].

# Misc. Rules

* /ɾ/ may be realized as retroflex following /a/ [@litteral_angor_1997, p. 4].
* Unstressed word-final vowels are often deleted following either voiceless stops, nasals (including pre-nasals), or /x/ (ibid.).

# References
