---
title: "Albanian"
author: "Bill Mizgerd"
bibliography: sq.bib
output: html_document
---

Last Updated: 2020-07-02

# Background

**Language Family:** Indo-European / Albanian / Tosk

* Albanian has two major dialects, Gheg and Tosk. I will be aiming to reflect Standard Albanian, which is predominantly based on the Tosk dialect.
* Albanian is spoken in a multitude of places, but predominantly in southeastern Europe.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial  Dental  Alveolar  Postalveolar  Palatal  Velar   Glottal
Stops                      "p b"   ""      "t d"     ""            "c ɟ"    "k ɡ"   ""
Affricates                 ""      ""      "ts dz"   "tʃ dʒ"       ""       ""      ""
Fricatives                 "f v"   "θ ð"   "s z"     "ʃ ʒ"         ""       ""      "h"
Nasals                     "m"     ""      "n"       ""            "ɲ"      ""      ""
Flaps                      ""      ""      "ɾ"       ""            ""       ""      ""
Trills                     ""      ""      "r"       ""            ""       ""      ""
Approximants               ""      ""      "l"       ""            "j"      "ɫ"     ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Dental", "Alveolar", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 7)) %>%
  footnote(general = "Note: Phonemes to the left of each cell are voiceless and phonemes to the right of each cell are voiced.", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* @SqNewmark1982 argue for /ɛ/ rather than /e/ (p. 10).
* Nasal and long vowels exist in Gheg, but Tosk has lost them. Because my analysis focuses on the Tosk-based Standard Albanian, I will be omitting information on the nasal and long vowels, but it can be found on pp. 5-6 of @SqCamaj1984.
    - In Gheg, all vowel phonemes are nasal equivalents except for /o/ and /ə/. In ambiguous situations, nasal vowels may be indicated with the circumflex (ˆ). As for long vowels, /ə/ once again lacks a long counterpart [@SqCamaj1984 , pp. 5-6].
* Albanian also has four diphthongs, that @SqNewmark1982 state are always interpreted as complex phonemes (p. 12). However, he does state that there is variation in terms of realization. Additionally, @SqCamaj1984 states that new documentation identifies these sequences as separate phonemes. I have, however, chosen to include them in the ruleset.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
          Front   Central   Back
High      "i y"   ""        "u"
Mid       "e"     "ə"       "o"
Low       ""      "a"       ""
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered") %>%
  footnote(general = "Note: The high front phonemes are unrounded and rounded, respectively.", general_title = "") %>%
  column_spec(1, bold = TRUE)


diphthongs <- read.table(textConnection('
Diphthongs
"/ie/, /ua/, /ue/, /ye/"
'), TRUE)

kable(diphthongs, align = 'c') %>%
  kable_styling("bordered")
```

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme
"a"              "/a/"  
"b"              "/b/"  
"c"              "/ts/"  
"ç"              "/tʃ/"  
"d"              "/d/"   
"e"              "/e/"  
"ë"              "/ə/"  
"f"              "/f/"  
"g"              "/ɡ/"   
"h"              "/h/"  
"i"              "/i/"  
"j"              "/j/"  
"k"              "/k/"  
"l"              "/l/"   
"m"              "/m/"  
"n"              "/n/"   
"o"              "/o/"  
"p"              "/p/"  
"q"              "/c/"  
"r"              "/ɾ/"   
"s"              "/s/"   
"t"              "/t/"  
"u"              "/u/"  
"v"              "/v/"  
"x"              "/dz/"   
"y"              "/y/"
"z"              "/z/"  
**Digraph**    ""
"dh"             "/ð/"
"gj"             "/ɟ/"
"ll"             "/ɫ/"  
"nj"             "/ɲ/"
"rr"             "/r/"
"sh"             "/ʃ/"
"th"             "/θ/"  
"xh"             "/dʒ/"
"zh"             "/ʒ/"
"ie"             "/ie/"  
"ua"             "/ua/"  
"ue"             "/ue/"  
"ye"             "/ye/"  
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* The palatal stops spirantize to affricates in many Gheg-speaking regions [@SqCamaj1984, p. 2].
* Nasals tend to assimilate to the place of following voiced stops [@SqCamaj1984, p. 4].
    - In northern Gheg, the stop tends to delete in such environments, leaving just the assimilated nasal behind (ibid.).
* When preceded by /n/, /d/ tends to be released nasally or even deleted entirely (ibid.).

# Misc. Rules

* Stress and tone are non-contrastive in Albanian [@SqNewmark1982, pp. 14-17].
* In the Arbëresh dialects, /l/ tends to palatalize [@SqCamaj1984, p. 3].
* /ɡ/ devoices when followed by /t/ [@SqCamaj1984, p. 5].
* Word-final /ə/ often deletes in casual speech [@SqNewmark1982, p. 11].
* Final /ə/ tends to lengthen the syllable preceding it [@SqNewmark1982, p. 12].
* /n/ assimilates to [ŋ] when followed by /ɡ/ or /k/ [@SqNewmark1982, p. 21].

# References
