---
title: "Malayalam"
author: "Becky Mathew"
bibliography: ml.bib
output: html_document
---

Last Updated: 2020-06-16

**SLIGHTLY COMPROMISED: conflation between dental and alveolar /n/**

# Background

**Language Family:** Dravidian / Southern / Tamil-Kannada / Tamil-Kodagu / Tamil-Malayalam / Malayalam

* Malayalam, the official language of Kerala and Lakshadweep, is a Dravidian language from South India. It is also heavily influenced by Sanskrit, and has become influenced by English with the introduction of /f/.
* It is spoken in the states of Karnataka, Kerala, Lakshadweep, Puducherry, and Tamil Nadu located in southern India.

# Phonology

## Consonants

* Consonant length is contrastive for voiceless (unaspirated) stops (including /tʃ/), nasals, and laterals /l/ and /ɭ/ [@Panel2018, p. 6; @Jiang2010, p. 8].
* /t̪ʰ/, /d̪ʱ/, /ʈʰ/, and /ɖʱ/ are very marginal [@Jiang2010, p. 17].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"         Labial        Dental     Alveolar    Retroflex   Postalveolar   Palatal       Velar       Glottal
"Stops (plain)"                  "p  b"        "t̪  d̪"      ""          "ʈ  ɖ"        ""           "ɟ"         "k  ɡ"         ""
"Stops (aspirated)"              "pʰ  bʱ"      "t̪ʰ  d̪ʱ"    ""         "ʈʰ  ɖʱ"       ""           "ɟʱ"        "kʰ  ɡʱ"       ""
Affricates                       ""            ""          ""            ""        "tʃ tʃʰ"       ""            ""           ""
Fricatives                       ""           ""          "s"          "ʂ"          "ʃ"          ""            ""           "h"
Nasals                           "m"           "n̪"         "n"          "ɳ"          ""           "ɲ"           "ŋ"          ""
Flaps                            ""            ""          "ɾ"          ""           ""           ""            ""           ""
Trills                           ""            ""          "r"          ""           ""           ""            ""           ""
Approximants                     "ʋ"            ""          "l"        "ɻ  ɭ"         ""           "j"           ""           ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Dental", "Alveolar", "Retroflex", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 8)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced. Both retroflex approximants are voiced, but the one on the right is lateral.", general_title = "")
```

## Vowels

* Vowel length is contrastive. The diphthongs listed below do not have a length contrast [@Jiang2010, p. 18].
* Most sources [e.g. @Jiang2010, p. 18] include schwa as part of the vowel inventory; however, its phonemic status is somewhat vague. The only minimal pairs @Jiang2010 lists involve contrast word-finally (pp. 18-19), which some sources argue merely as the reduction of /u/ [@Swenson2016, p. 133; @Malayalam, p. 2]. I have chosen not to include /ə/ below, treating /u/ as the underlying form.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         ""           "u"
Mid       "e"         ""           "o"
Low        ""         "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)


diphthongs <- read.table(textConnection('
Diphthongs
"/ai/, /au/"
'), TRUE)

kable(diphthongs, align = 'c') %>%
  kable_styling("bordered")
```   

## Alphabet

* **Note**: /n/ and /ɟʱ/ are not represented in the Crúbadán corpus.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
"Consonant Grapheme"       Phoneme           Comment
"ക"                       "/ka/"              ""
"ഖ"                       "/kʰa/"             ""
"ഗ"                       "/ɡa/"              ""
"ഘ"                      "/ɡʱa/"             ""
"ങ"                       "/ŋa/"              ""
"ച"                       "/tʃa/"             ""
"ഛ"                     "/tʃʰa/"             ""
"ജ"                       "/ɟa/"              ""
"ഝ"                     "/ɟʱ/"              ""
"ഞ"                     "/ɲa/"             ""
"ട"                       "/ʈa/"              ""
"ഠ"                      "/ʈʰa/"              ""
"ഡ"                      "/ɖa/"              ""
"ഢ"                     "/ɖʱa/"              ""
"ണ"                     "/ɳa/"               ""
"ത"                      "/t̪a/"               ""
"ഥ"                     "/t̪ʰa/"               ""
"ദ"                      "/d̪a/"               ""
"ധ"                     "/d̪ʱa/"               ""
"ന"                   "/n̪a/; /na/"   "/n̪a/: used as default in the rules"
"ഩ"                      "/na/"             "rare"
"പ"                      "/pa/"               ""
"ഫ"                     "/pʰa/"              ""
"ബ"                     "/ba/"               ""
"ഭ"                      "/bʱa/"              ""
"മ"                      "/ma/"               ""
"യ"                      "/ja/"              ""
"ര"                      "/ɾa/"               ""
"റ"                      "/ra/"               ""
"ല"                      "/la/"               ""
"ള"                      "/ɭa/"               ""
"ഴ"                      "/ɻa/"               ""
"വ"                     "/ʋa/"                ""
"ശ"                     "/ʃa/"                ""
"ഷ"                     "/ʂa/"               ""
"സ"                     "/sa/"               ""
"ഹ"                     "/ha/"               ""
"**Consonant Grapheme**"   ""          "called chillaksharams, these characters don\'t carry an inherent vowel [@Panel2018, p. 9]"
"ൺ"                     "/ɳ/"               ""  
"ൻ"                      "/n̪/"               ""
"ർ"                       "/r/"               ""
"ൽ"                      "/l/"               ""
"ൾ"                      "/ɭ/"               ""
"ൿ"                      "/k/"               ""
"**Vowel Grapheme (word-initial)**"  ""        ""
"അ"                     "/a/"                ""
"ആ"                    "/aː/"                ""
"ഇ"                      "/i/"                ""
"ഈ"                    "/iː/"                ""
"എ"                     "/e/"                ""
"ഏ"                     "/eː/"               ""
"ഉ"                      "/u/"                ""
"ഊ"                    "/uː/"               ""
"ഐ"                   "/ai/"               ""
"ഒ"                     "/o/"                ""
"ഓ"                    "/oː/"               ""
"ഔ"                    "/au/"               ""
**Diacritic**             ""                  ""
"ാ"                     "/aː/"               ""
"ി"                     "/i/"                ""
"ീ"                     "/iː/"               ""
"െ"                    "/e/"                ""
"േ"                    "/eː/"                ""
"ു"                     "/u/"                ""
"ൂ"                    "/uː/"                ""
"ൃ"                    "/ai/"                ""
"ൊ"                   "/o/"                ""
"ോ"                   "/oː/"               ""
"ൌ; ൗ"             "/au/"               ""
"്"                    "∅; /u/"     "called a virama, this diacritic either removes the inherent vowel or reduces word-final /u/ to [ə] [@Malayalam, p. 2]"  
"ം"                    "/m/"       "called an anusvara(m), this character follows a vowel, often resulting in assimilation to the following consonant [@Panel2018, p. 9]"  
"ഃ"                    "/h/"       "called a visarga(m), this character follows a vowel (ibid.)"  
'), TRUE)

kable(alphabet, col.names = c("Consonant Grapheme", "Phoneme", "Comment"), align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Syllables in Malayalam follow the order of (C)(C)(C)V(V)(C) [@Jiang2010, p. 24].

# Lenition Rules

* Voiceless stops become voiced intervocalically [@Jiang2010, p. 22].

# Misc. Rules

* Nasals and lateral approximants are unreleased word-finally [@Jiang2010, p. 22].

# References
