---
title: "Pele-Ata"
author: "Emily Strand"
bibliography: ata.bib
output: html_document
---

Last updated: 2019-11-08

# Background

**Language family**: East Papuan / Yele-Solomons-New Britain / New Britain / Wasi

* **Note:** The classification of Pele-Ata carries some controversy. @campbell2017 classifies it as a language isolate, stating that similarities among such languages within the (tentative) family are far too small to warrant any genealogical relation [also noted in @Campbell2010, pp. 6-7].

* Pele-Ata is spoken in the East and West New Britain provinces of Papua New Guinea.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial   Alveolar    Velar   Glottal
Stops                       "p"       "t"        "k"      "ʔ"
Fricatives                  "β"       "s"        "x"      ""
Nasals                      "m"       "n"        "ŋ"      ""
Approximants                ""        "l"        ""       ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 4)) %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* Vowel length is phonemic in Pele-Ata [@hashimoto2003, p. 2]. Long vowels are indicated by duplicate graphemes.
* Pele-Ata has a large number of possible diphthongs, which, depending on the syllable structure, can be interpreted as either complex nuclei or sequences of vowels [@hashimoto2003, p. 3]. Thus, they will not be transcribed in the rules.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front        Back
High      "i"         "u"
Mid       "ɛ"         "o"
Low        ""         "ɑ"
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```   

# Alphabet

* Orthographically, ⟨w⟩ and ⟨j⟩ are sometimes used to represent /u/ and /i/ respectively, which will be transcribed as such in the rules [@hashimoto2003, p. 3].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme  
"a"             "/ɑ/"
"e"             "/ɛ/"
"i"             "/i/"
"k"             "/k/"
"l"             "/l/"
"m"             "/m/"
"n"             "/n/"
"o"             "/o/"
"p"             "/p/"
"s"             "/s/"
"t"             "/t/"
"u"             "/u/"
"v"             "/β/"
"x"             "/x/"
"\'"            "/ʔ/"
**Digraph**     ""
"ng"            "/ŋ/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Syllables in Pele-Ata consist of the following structures [@hashimoto2003, p. 3]:
    - V
    - CV
    - CVV

# Lenition Rules  

* /x/ becomes voiced intervocalically [@hashimoto2003, p. 3].

# Misc. Rules

* /s/ is palatalized preceding /i/ [@hashimoto2003, p. 3].
* /i/ and /u/ become [j] and [w] word-initially preceding /o/ and /ɑ/ [@hashimoto2003, p. 3].
* Epenthetic [j] and [w] are inserted intervocalically (agreeing with the first vowel), in the sequences /io/, /iɑ/, /uo/, and /uɑ/ (ibid.).

# References
