---
title: "Kazakh"
author: "Abi Creighton"
bibliography: kk.bib
output: html_document
---

Last Updated: 2019-09-24

**Compromised: conflation between vowels and glides; widely contrasting phonological documentation of the language**

# Background

**Language family**: Turkic / Common Turkic / Kipchak / Kipchak-Nogai / Kazakh

* Kazakh is spoken within Central Asia, specifically in China, Kazakhstan, Mongolia, Turkmenistan, and Uzbekistan.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial   Dental  Alveolar  Postalveolar  Palatal  Velar   Uvular  Glottal
Stops                      "p b"    "t̪ d̪"      ""         ""          ""    "k  ɡ"    "q"      ""
Fricatives                   ""     "s̪ z̪"      ""        "ʃ ʒ"         ""      ""      "χ ʁ"   "h"
Nasals                       "m"     "n̪"       ""         ""          ""      "ŋ"      ""      ""
Trills                        ""     ""        "r"         ""          ""      ""       ""     ""
Approximants                 "w"     "l̪"       ""         ""          "j"     ""       ""      ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Dental", "Alveolar", "Postalveolar", "Palatal", "Velar", "Uvular", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 8)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```

## Vowels

* No two accounts of the Kazakh vowel inventory (ranging between 9 and 12 vowel phonemes) seem to agree [@McCollum2020, p. 7]. I have chosen to follow the analysis done by @Bowman2013 (p. 1).

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
            Front     Back
High         "i y"   "ɯ u"
Near-High    "ɪ"      "ʊ"
Mid          "e ø"    "o"
Near-Low     "æ"      ""
Low          ""       "ɑ"
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are unrounded and those on the right are rounded.", general_title = "")
```

# Alphabet

* Although Kazakh can be written with the Arabic, Latin, and Cyrillic scripts, I have opted to represent only the Cyrillic script below, as the others pose more conflation between phonemes [@Vajda1994, pp. 606-607].

* ⟨в⟩, ⟨ё⟩, ⟨ф⟩, ⟨х⟩, ⟨ц⟩, ⟨ч⟩, ⟨щ⟩, ⟨ъ⟩, ⟨ь⟩, and ⟨э⟩ appear in loanwords [@Vajda1994, pp. 608, 609].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme         Comment                        
"а"            "/ɑ/"            ""                
"ә"            "/æ/"            ""               
"б"            "/b/"            ""               
"г"            "/ɡ/"            ""               
"ғ"            "/ʁ/"            ""               
"д"            "/d̪/"            ""               
"е"            "/e/"            ""               
"ж"            "/ʒ/"            ""              
"з"            "/z̪/"            ""               
"и"          "/i/; /j/"   "/i/: default in rule set"                   
"й"            "/j/"            ""                   
"к"            "/k/"            ""                
"қ"            "/q/"            ""              
"л"            "/l̪/"            ""               
"м"            "/m/"            ""             
"н"            "/n̪/"            ""               
"ң"            "/ŋ/"            ""               
"о"            "/o/"            ""               
"ө"            "/ø/"            ""               
"п"            "/p/"            ""               
"р"            "/r/"            ""               
"с"            "/s̪/"            ""                
"т"            "/t̪/"            ""              
"у"          "/u/; /w/"    "/u/: default in rule set"                  
"ұ"            "/ʊ/"            ""               
"ү"            "/y/"            ""               
"х"            "/χ/"            ""               
"һ"            "/h/"            ""               
"ш"            "/ʃ/"            ""               
"ы"            "/ɯ/"            ""               
"і"            "/ɪ/"            ""               
"ю"            "/ju/"           ""               
"я"            "/jɑ/"           ""               
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Syllables in Kazakh follow the order of (C)V(C)(C) [@McCollum2020, p. 23].

# Lenition Rules

* /b/, /ɡ/, and /q/ may spirantize intervocalically [@kirchner_kazakh_1998].
    - /d̪/ does spirantize as well to [ð] for some speakers, but it is far more rare [@McCollum2020, p. 5].
* In fast speech /q/ tends to be realized as [χ] [@McCollum2020, p. 4].
    - This process tends to neutralize the contrast between /q/ and /χ/.
* /r/ may be realized as a flap in fast speech [@McCollum2020, p. 6].

# Misc. Rules

* A type of vowel harmony is present with velar and uvular stops; front vowels tend to appear in the vicinity of velar stops, while back vowels tend to appear in the vicinity of uvular stops [@McCollum2020, p. 4].
* /χ/ is sometimes realized as [x] due to Russian influence (ibid.).
* /n̪/ assimilates in place to following obstruents [@McCollum2020, p. 5].
* /l/ becomes velarized in the vicinity of back vowels [@McCollum2020, p. 6].
* An epenthetic vowel tends to appear preceding word-initial /r/ (ibid.).
* Consonant gemination appears to occur [@Dotton2018, p. 16].

# References
