---
title: "Palikúr"
author: "Emily Strand"
bibliography: plu.bib
output: html_document
---

Last updated: 2019-11-04

**SLIGHTLY COMPROMISED: conflation between /ɡ/ and /ɣ/**

# Background

**Language Family:** Arawakan / Maipuran / Eastern Maipuran

* Palikúr is spoken in Brazil and French Guiana.

# Phonology

## Consonants

* /f/, /v/, /l/, /ʒ/ and /z/ appear in loanwords [@launey2003, p. 37].
* @launey2003 and @launey2001 specify Palikúr as having a "light consonant," although the phonemic representation seems somewhat unclear (p. 35; p. 10). They describe it in terms of its realization, which ranges from lenited [ɡ] to [ʁ], but they also explicitly classify it as a velar fricative. For this reason, I have chosen to use /ɣ/.
    - Although /ɡ/ undergoes lenition to [ɣ] word-medially, /ɣ/ is recognized as a separate phoneme.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial   Alveolar  Postalveolar   Palatal    Velar   Glottal   
Stops                      "p b"     "t d"        ""           ""       "k ɡ"     ""
Fricatives                  ""        "s"         "ʃ"          ""        "ɣ"      "h"
Nasals                      "m"       "n"         ""           ""        ""       ""
Approximants                "w"        ""         ""           "j"       ""       ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```

## Vowels

* Nasal vowels are phonemic in Palikúr [@launey2001, p. 6]. They appear in syllable-final positions and are indicated by a following ⟨h⟩ (i.e. ⟨h⟩ indicates vowel nasalization word-finally or preceding consonants) [@launey2003, p. 35].
* There are also a several oral and nasal diphthongs present in the language [@launey2001, pp. 6-7]. The oral diphthongs are composed of a vowel followed by a glide and the nasal diphthongs are composed of a vowel followed by a glide and ⟨h⟩. Like the nasal vowels, these diphthongs are said to be present only in syllable-final positions [@launey2003, p. 30].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High     "i  ĩ"       ""         "u  ũ"
Mid      "e  ẽ"       ""         "o  õ"
Low        ""       "a  ã"         ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: Vowels on the left of each cell are oral and vowels on the right of each cell are nasal.", general_title = "")


diphthongs <- read.table(textConnection('
Oral                                                             Nasal
"/aj/, /ej/, /ij/, /uj/, /aw/, /ew/, /iw/, /ow/, /uw/"     "/ãj/, /ãw/, /ĩw/"
'), TRUE)

kable(diphthongs, align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("Diphthongs" = 2))
```  

# Alphabet

* Depending on the documentation, older sources tend to represent /ɣ/ as ⟨g⟩, whereas newer sources tend to represent it as ⟨r⟩ [@launey2001, p. 10]. Therefore, there is some conflation between /ɣ/ and /ɡ/, compromising the language. In the rule set, I have chosen to transcribe ⟨g⟩ to /ɡ/ and ⟨r⟩ to /ɣ/.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme       Phoneme       Comment
"a"             "/a/"          ""
"b"             "/b/"          ""
"d"             "/d/"          ""
"e"             "/e/"          ""
"g"          "/ɡ/; /ɣ/"     "/ɡ/: default in the rule set"
"h"           "/h/; ∅"   "∅: following vowels in syllable-final positions (indicates nasalization)"
"i"             "/i/"          ""
"k"             "/k/"          ""
"m"             "/m/"          ""
"n"             "/n/"          ""
"o"             "/o/"          ""
"p"             "/p/"          ""
"r"             "/ɣ/"          ""
"s"             "/s/"          ""
"t"             "/t/"          ""
"u"             "/u/"          ""
"w"             "/w/"          ""
"x"             "/ʃ/"          ""
"y"             "/j/"          ""
**Multigraph**  ""             ""
"ah"            "/ã/"          ""
"eh"            "/ẽ/"          ""
"ih"            "/ĩ/"          ""
"oh"            "/õ/"          ""
"uh"            "/ũ/"          ""
"ay"            "/aj/"         ""
"ey"            "/ej/"         ""
"iy"            "/ij/"         ""
"uy"            "/uj/"         ""
"aw"            "/aw/"         ""
"ew"            "/ew/"         ""
"iw"            "/iw/"         ""
"ow"            "/ow/"         ""
"uw"            "/uw/"         ""
"ayh"           "/ãj/"         ""
"awh"           "/ãw/"         ""
"iwh"           "/ĩw/"         ""
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* /p/ often spirantizes and becomes voiced word-medially [@launey2003, p. 39].
* /p/ may become voiced word-medially following a nasal vowel [@launey2003, p. 41].
* /ɡ/ often spirantizes to [ɣ] word-medially (ibid.).
    - /ɡ/ may also be realized as [ŋ] in this position, but it is far more rare.

# Misc. Rules

* /t/ and /d/ often palatalize preceding /i/ and /ĩ/ [@launey2003, pp. 37-38].
* Duplicate consonants (occurring word-initially) are realized separately [@launey2001, p. 9]. This is noted only for nasal-consonant sequences /nn/ and /mm/.
    - It is implied that other nasal-consonant sequences (/nd/, /nt/, /mp/, /mb/, /mt/, and /ms/) are also interpreted in this way (i.e. with syllabic nasals) rather than as complex phonemes.

# References
