---
title: "Naasioi"
author: "Abi Creighton"
bibliography: nas.bib
output: html_document
---

Last Updated: 2020-07-27

# Background

**Language Family:** East Papuan / South Bougainville / East / Nasioi

* Naasioi is also known as Nasioi, Kieta, Kieta Talk, or Aunge. It is spoken in Papua New Guinea in parts of the Kieta District in the Bougainville Province.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial  Alveolar    Velar   Glottal
Stops                      "p b"    "t d"       "k"       "ʔ"
Nasals                      "m"      "n"        ""         ""         
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 4)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```  

## Vowels

* Vowel sequences are prevalent in Naasioi, with words consisting of up to six vowels [@hurd_nasioi_1970, p. 38].
* (Up to four) like-vowels can occur in sequences as well. Length is said to correspond with the number of vowels present in such sequences (i.e. three like-vowels would be longer than two like-vowels). However, instances of these lengthy sequences are not present in the Crúbadán corpus (only sequences of two like-vowels were found).
    - Based on this and the fact that @hurd_nasioi_2004 specifies long vowels as sequences of two vowels (p. 2), it puts into question the phonemic status of vowel length. However, @Crothers1979 and Phoible's listing of the language attest the phonemic status of long vowels [@hurdhurd1966]. I have chosen to treat vowel length as phonemic, with long vowels represented by duplicate graphemes.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
             Front      Central       Back
High          "i"         ""          "u"
Mid           "e"         ""          "o"
Low           ""          "a"         ""    
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```     

# Alphabet

* ⟨v⟩, ⟨s⟩, and ⟨r⟩ represent allophones of /b/, /t/ and /d/ respectively [@hurd_nasioi_2004, p. 3].
* ⟨j⟩ and ⟨l⟩ appear in loanwords [@hurd_nasioi_2004, p. 3].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme           
"a"             "/a/"              
"b"             "/b/"              
"d"             "/d/"               
"e"             "/e/"              
"i"             "/i/"              
"k"             "/k/"              
"m"             "/m/"              
"n"             "/n/"              
"o"             "/o/"              
"p"             "/p/"              
"r"             "/d/"              
"s"             "/t/"              
"t"             "/t/"              
"u"             "/u/"               
"v"             "/b/"               
"\'"            "/ʔ/"               
**Digraph**      ""                   
"ts"            "/t/"               
"ng"            "/n/"               
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Syllables in Naasioi take the form of (C)V(C) or N, a syllabic nasal [@hurd_nasioi_1970, p. 38].

# Lenition Rules

* When /b/ and /d/ are neither word-initial nor following a glottal stop, /b/ spirantizes and /d/ becomes the alveolar flap [@hurd_nasioi_1970, p. 37].

# Misc. Rules

* When followed by /i/, /t/ is pronounced as [s] [@hurd_nasioi_1970, p. 37].
* In slow speech, /d/ can become [r] word-medially [@hurd_nasioi_2004, p. 3]
* /n/ assimilates to [ŋ] preceding /k/ (ibid.).

# References
