---
title: "Ixtatán Chuj"
author: "Bill Mizgerd"
bibliography: cnm.bib
output: html_document
---

Last Updated: 2020-03-05

# Background

**Language Family**: Mayan / Kanjobalan-Chujean / Chujean

* **Note:** The Crúbadán code for the entire Chuj language is cac; however, I will be principally focusing on the Ixtatán dialect of Chuj (cnm) because it has a larger corpus in Crúbadán.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"     Labial     Alveolar  Postalveolar  Palatal   Velar      Uvular  Glottal
Stops                        "p ɓ"      "t tʼ"    ""            ""        "k kʼ"     ""      "ʔ"
Affricates                   ""         "ts tsʼ"  "tʃ tʃʼ"      ""        ""         ""      ""
Fricatives                   ""         "s"       "ʃ"           ""        ""         "χ"      ""
Nasals                       "m"        "n"       ""            ""        "ŋ"        ""      ""
Flaps                        ""         "ɾ"       ""            ""        ""         ""      ""
Approximants                 "w"        "l"       ""            "j"       ""         ""      ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Uvular", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 7)) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are plain and those on the right are implosive (labial stop) or ejective.", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* @CnmWilliams1965 lists /aː/ as phonemic (p. 225), but this does not appear in any of the other sources.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
       Front   Central   Back
High   "i"     ""        "u"
Mid    "e"     ""        "o"
Low    ""      "a"       ""
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

* Chuj has multiple orthographies, which represent some phonemes in different ways; however, these orthographies do not have any overlap that would create ambiguity, meaning that the language is not compromised. I have documented the the variations within the chart below.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme  
"a"             "/a/"
"b’"            "/ɓ/"
"c"             "/k/"
"c’"            "/kʼ/"
"d’"            "/tʼ/"
"e"             "/e/"
"i"             "/i/"
"j"             "/χ/"
"k"             "/k/"
"k’"            "/kʼ/"
"l"             "/l/"
"m"             "/m/"
"n"             "/n/"
"n̈"             "/ŋ/"
"o"             "/o/"
"p"             "/p/"
"r"             "/ɾ/"
"s"             "/s/"
"t"             "/t/"
"t’"            "/tʼ/"
"u"             "/u/"
"v"             "/w/"
"w"             "/w/"
"x"             "/ʃ/"
"y"             "/j/"
"\'"            "/ʔ/"
**Digraph**      ""
"ch"            "/tʃ/"
"ch’"           "/tʃʼ/"
"nh"            "/ŋ/"
"qu"            "/k/"
"q’u"           "/kʼ/"
"tz"            "/ts/"
"tz’"           "/tsʼ/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* Some instances of historic /q/ have spirantized to become /χ/ in modern Chuj [@CnmRaymundo2014, pp. 21-22].

# References
