---
title: "Bugis"
author: "Allison Shao"
bibliography: bug.bib
output: html_document
---

Last Updated: 2020-07-16

# Background

**Language family**: Austronesian / Malayo-Polynesian / Western Malayo-Polynesian / Sulawesi / South Sulawesi / Bugis

* The Bugis language is also known as Buginese and is spoken throughout Sulawesi, Indonesia.

# Phonology

## Consonants

* Not all sources agree on the phonemic inventory of Bugis. I have chosen to include the phonemes that the majority of the sources agreed on. The differences, however, are listed below:
    - @jaya2018 additionally includes /tʃ/ and /dʒ/ but omits /c/, /ɟ/, and /ɲ/.
    - @talaohu2019 additionally includes /f/ and /ɸ/.
    - @wolff2010 omits /ʔ/ and /j/.
    - @valls2014 additionally includes /ʃ/ and /ʒ/.
    - @pedroscaballero2014 lists /ɾ/ instead of /r/ and omits /ʔ/.
* Additionally, @pedroscaballero2014 states that the phonemes presented as /c/ and /ɟ/ may actually be /tʃ/ and /dʒ/ (p. 10). This may explain why @jaya2018 includes /tʃ/ and /dʒ/ instead of /c/ and /ɟ/.
* According to @valls2014, /h/ is only found in Indonesian loanwords (p. 8). However, due to the significant influence, it is recognized as a phoneme of Bugis.
* Geminate consonants /pp/, /bb/, /tt/, /dd/, /kk/, /ss/, /mm/, /nn/, /ll/, /cc/, and /ɟɟ/ are constrastive [@valls2014, p. 9; @pedroscaballero2014, p. 8].
* @talaohu2019 argues that instead of two phonemes /k/ and /ɡ/, [ɡ] is actually an allophone of /k/ (pp. 31-32). However, @pedroscaballero2014 presents minimal pairs for these phonemes, so I will treat them as distinct (p. 11).


```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"    Labial   Alveolar  Palatal  Velar   Glottal
Stops                        "p b"    "t d"     "c ɟ"   "k ɡ"     "ʔ"
Fricatives                    ""       "s"       ""      ""       "h"
Nasals                        "m"      "n"       "ɲ"     "ŋ"      ""
Trills                        ""       "r"       ""      ""       ""
Approximants                  "w"      "l"       "j"     ""       ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 5)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: Where two phonemes share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```

## Vowels

* Vowels /i/, /e/, /a/, /o/, and /u/ also have long counterparts, which are indicated by duplicate graphemes [@valls2014, p. 8]. However, vowel length is not actually contrastive in Bugis, so I have chosen to treat this feature as allophonic and transcribe such instances of long vowels as short in the rules.
* Though sequences of consecutive vowels occur in Bugis, they are not always realized as complex phonemes [@pedroscaballero2014, p. 6]. Therefore, I have not accounted for them as diphthongs in the rules.
* The status of schwa as a phoneme in Bugis is somewhat unclear. All sources recognize /ə/ as a distinct phoneme. However, according to @pedroscaballero2014, [ə] may be a possible allophone of both /a/ and /e/ in unstressed syllables (p. 7). It is important to note that it is not uncommon for a phoneme of a language to also be an allophone, so this lenition of /a/ and /e/ is most likely coincidence rather than evidence against schwa being a phoneme. There do however appear to be words in which /ə/ appears in stressed syllables [@pedroscaballero2014, p. 7], which would serve as evidence against the language having a truly reduced phonemic vowel. Furthermore, @valls2014 describes this mid central vowel as "rounded," which suggests it could actually be /ɵ/ or /ɞ/ (p. 12). Based on this, I have chosen to represent this mid central vowel as /ɵ/ rather than /ə/.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         ""           "u"
Mid       "e"         "ɵ"          "o"
Low        ""         "a"           ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```  

# Alphabet

* /f/ and /v/ can be found in loanwords, and are most often realized as [p] [@jaya2018, p. 62].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme       Phoneme         
"a"             "/a/"          
"b"             "/b/"          
"c"             "/c/"          
"d"             "/d/"          
"e"             "/e/"          
"è"             "/ɵ/"          
"g"             "/ɡ/"          
"h"             "/h/"          
"i"             "/i/"          
"j"             "/ɟ/"          
"k"             "/k/"          
"l"             "/l/"          
"m"             "/m/"          
"n"             "/n/"          
"o"             "/o/"          
"p"             "/p/"          
"r"             "/r/"          
"s"             "/s/"          
"t"             "/t/"          
"u"             "/u/"          
"w"             "/w/"          
"y"             "/j/"          
"\'"            "/ʔ/"          
**Digraph**      ""            
"ng"            "/ŋ/"          
"ny"            "/ɲ/"          
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```  

# Syllable Structure

* Syllables in Bugis are of the type (C)(C)V(C) [@pedroscaballero2014, p. 13].
    - In syllables with a CC onset, the second consonant is often glide /j/.
    - The coda is predominantly either a nasal or /ʔ/.

# Lenition Rules

* Stops tend to vary freely with their fricative (or flap) counterparts [@valls2014, pp. 11-12]:
    - /b/ varies freely with [β] intervocalically and in onset position
    - /p/ varies freely with [f] and [ɸ]
    - /d/ varies freely with [ɾ] and [ð] intervocalically
    - /ɡ/ varies freely with [ɣ] intervocalically
* According to @jaya2018, /p/, /b/, /t/, /d/, /k/, and /h/ are generally realized as [ʔ] word-finally (pp. 57-58).

# Misc. Rules

* /ŋ/ assimilates to the place of the following consonant [@valls2014, p. 12].
* All vowels have nasal allophones that surface when in proximity to nasals (ibid.).
* Nasals /m/, /n/, and most likely /ɲ/ are often realized as [ŋ] word-finally (@jaya2018, p. 59).
* /s/ may be realized as [θ] [@pedroscaballero2014, p. 10].
* /w/ may be realized as [β] [@pedroscaballero2014, p. 11].

# References
