---
title: "Indonesian"
author: "Justin Bai"
edited by: "Emily Strand"
bibliography: id.bib
output: html_document
---

Last Updated: 2020-03-26

**Slightly Compromised: conflation between /e/ and /ə/**

# Background

**Language family**: Austronesian / Malayo-Polynesian / Western Malayo-Polynesian / Sundic / Malayic / Malayan / Local Malay

# Phonology

## Consonants

* /f/, /z/, /x/, and /ʃ/ only appear in loanwords, but due to major influence, they have phonemic status in the language [@tadmor2009malay, p. 795].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"    Labial   Alveolar  Postalveolar   Palatal  Velar   Glottal
Stops                        "p b"    "t d"        ""          ""      "k ɡ"     ""
Affricates                    ""       ""        "tʃ dʒ"       ""       ""       ""
Fricatives                    "f"     "s z"        "ʃ"         ""       "x"      "h"
Nasals                        "m"      "n"         ""          "ɲ"      "ŋ"      ""
Trills                        ""       "r"         ""          ""       ""       ""
Approximants                  "w"      "l"         ""          "j"      ""       ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: Where two phonemes share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```

## Vowels

* Adjacent vowels don't typically form complex nuclei [@tadmor2009malay, p. 796].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         ""           "u"
Mid       "e"         "ə"          "o"
Low        ""         "a"           ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```  

# Alphabet

* The languages is somewhat compromised as ⟨e⟩ can represent two different vowels /e/ and /ə/. The rules use /e/ as the default [@soderberg2008indonesian, p. 211].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme       Phoneme         Comment
"a"             "/a/"            ""
"b"             "/b/"            ""
"c"             "/tʃ/"           ""
"d"             "/d/"            ""
"e"          "/e/; /ə/"     "/e/: default"
"é"             "/e/"            ""
"f"             "/f"       "only in loanwords - varies freely with [p]"
"g"             "/ɡ/"            ""
"h"             "/h/"            ""
"j"             "/dʒ/"           ""
"k"             "/k/"            ""
"l"             "/l/"            ""
"m"             "/m/"            ""
"n"             "/n/"            ""
"p"             "/p/"            ""
"q"             "/k/"      "only in loanwords"
"r"             "/r/"            ""
"s"             "/s/"            ""
"t"             "/t/"            ""
"v"             "/f/"      "only in loanwords - varies freely with [v]"
"w"             "/w/"            ""
"x"             "/ks/"     "only in loanwords"
"y"             "/j/"            ""
"z"             "/z/"      "only in loanwords"
**Digraph**      ""              ""
"ng"            "/ŋ/"            ""
"ny"            "/ɲ/"            ""
"kh"            "/x/"      "only in loanwords - replaced with /k/ in younger generations [@soderberg2008indonesian, p. 211]"
"sy"            "/ʃ/"      "only in loanwords"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```  

# Syllable Structure

* Syllables in Indonesian follow the order of (C)V(C) [@tadmor2009malay, p. 796].

# Lenition Rules

* /k/ is realized as a glottal stop syllable finally [@soderberg2008indonesian, p. 211].

# Misc. Rules

* Epenthesis of glottal stop occurs depending on loanword origin (e.g. Arabic) and affixation [@soderberg2008indonesian, p. 211].
* Word-final voiced stops become voiceless in loanwords (voiced stops word-finally aren't present in native words) (ibid.).

# References
