---
title: "Indonesian"
author: "Justin Bai"
edited by: "Emily Strand"
bibliography: id.bib
output: html_document
---

Last Updated: 2020-06-05

**SLIGHTLY COMPROMISED: conflation between /e/ and /ə/**

# Background

**Language Family:** Austronesian / Malayo-Polynesian / Western Malayo-Polynesian / Sundic / Malayic / Malayan / Local Malay

# Phonology

## Consonants

* /f/, /z/, /x/, and /ʃ/ only appear in loanwords, but due to major influence, they have phonemic status in the language [@tadmor2009malay, p. 795].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"    Labial   Alveolar  Postalveolar   Palatal  Velar   Glottal
Stops                        "p b"    "t d"        ""          ""      "k ɡ"     ""
Affricates                    ""       ""        "tʃ dʒ"       ""       ""       ""
Fricatives                    "f"     "s z"        "ʃ"         ""       "x"      "h"
Nasals                        "m"      "n"         ""          "ɲ"      "ŋ"      ""
Trills                        ""       "r"         ""          ""       ""       ""
Approximants                  "w"      "l"         ""          "j"      ""       ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: Where two phonemes share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```

## Vowels

* Some sources present diphthongs. For example, @soderberg2008indonesian list /ai/, /oi/, and /au/. However, adjacent vowels don't typically form complex nuclei [@tadmor2009malay, p. 796], so I have opted not to account for them.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         ""           "u"
Mid       "e"         "ə"          "o"
Low        ""         "a"           ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```  

# Alphabet

* The language is somewhat compromised as ⟨e⟩ can represent both /e/ and /ə/. Sometimes ⟨é⟩ is used to represent /e/ to disambiguate (which is represented in the rules), but this is not always the case. The rules transcribe ⟨e⟩ to /e/ by default [@soderberg2008indonesian, p. 211].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme       Phoneme         Comment
"a"             "/a/"            ""
"b"             "/b/"            ""
"c"             "/tʃ/"           ""
"d"             "/d/"            ""
"e"          "/e/; /ə/"     "/e/: default in the rules"
"é"             "/e/"            ""
"f"             "/f"       "only appears in loanwords - varies freely with [p]"
"g"             "/ɡ/"            ""
"h"             "/h/"            ""
"i"             "/i/"            ""
"j"             "/dʒ/"           ""
"k"             "/k/"            ""
"l"             "/l/"            ""
"m"             "/m/"            ""
"n"             "/n/"            ""
"o"             "/o/"            ""
"p"             "/p/"            ""
"q"             "/k/"      "only appears in loanwords"
"r"             "/r/"            ""
"s"             "/s/"            ""
"t"             "/t/"            ""
"u"             "/u/"            ""
"v"             "/f/"      "only appears in loanwords - varies freely with [v]"
"w"             "/w/"            ""
"x"             "/ks/"     "only appears in loanwords"
"y"             "/j/"            ""
"z"             "/z/"      "only appears in loanwords"
**Digraph**      ""              ""
"ng"            "/ŋ/"            ""
"ny"            "/ɲ/"            ""
"kh"            "/x/"      "only appears in loanwords - replaced with /k/ in younger generations [@soderberg2008indonesian, p. 211]"
"sy"            "/ʃ/"      "only appears in loanwords"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```  

# Syllable Structure

* Syllables in Indonesian follow the order of (C)V(C) [@tadmor2009malay, p. 796].

# Lenition Rules

* /k/ is realized as a glottal stop syllable finally [@soderberg2008indonesian, p. 211].

# Misc. Rules

* Epenthesis of glottal stop occurs depending on loanword origin (e.g. Arabic) and affixation [@soderberg2008indonesian, p. 211].
* Word-final voiced stops become voiceless in loanwords (voiced stops word-finally aren't present in native words) (ibid.).
* Affix-final /ŋ/ assimilates to the place of articulation of the following root-initial consonant.

# References
