---
title: "Romanian"
author: "Emily Strand"
bibliography: ro.bib
output: html_document
---

Last Updated: 2019-08-21

# Background

**Language family**: Indo-European / Italic / Romance / Easatern

* Romanian is predominantly spoken in Romania and Moldova, but in other countries as well.

# Phonology

## Consonants

* There is some debate regarding the existence of palatal consonants as phonemes. @Petrovici1950 [as cited in @Dezsi2008, p. 439] includes them, but he also includes labialized and labio-palatalized consonants which proves too elaborate an analysis [@Dezsi2008, p. 440]. @Dezsi2008 states that palatalization is predictable (morphological conditioning), with /i/ surfacing as such a marker [ʲ] [p. 447; @Renwick2012, p. 81].
* @Dezsi2008 explains that the phonemic attestation of /j/ and /w/ is rather unclear, given a lack of analysis (p. 438). @Chitoran2002 [as cited in @Dezsi2008, p. 441] states that the /j/ is a surface realization of /i/. @Dezsi2008 seems to support this analysis (p. 447). I have chosen not to include /j/ and /w/ below.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"     Labial   Dental   Postalveolar   Velar  Glottal
Stops                        "p b"    "t d"         ""        "k ɡ"    ""
Affricates                    ""       "ts"      "tʃ dʒ"       ""      ""
Fricatives                   "f v"     "s z"      "ʃ ʒ"        ""      "h"
Nasals                        "m"      "n"          ""         ""      ""
Trills                        ""       "r"          ""         ""      ""
Approximants                  ""       "l"          ""         ""      ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Dental", "Postalveolar", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 5)) %>%
  footnote(general = "Note: From left to right, phonemes in a cell are plain, aspirated, and voiced.", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* Although Romanian has frequent vowel sequences (constituting possible diphthongs and triphthongs), only the ones listed below are always considered as complex nuclei [@Renwick2012, p. 438].
* Several sources use /ə/ to represent the mid-central vowel [@Dezsi2008, p. 438; @Chitoran2002, p. 208]; however, @Renwick2012 explains that the vowel for which this phoneme represents is not reduced, and shouldn't be treated as such. He opts for /ʌ/, but I have chosen to use /ɜ/ as it's more central.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
           Front      Central       Back
High        "i"         "ɨ"          "u"
Mid         "e"         "ɜ"          "o"
Low         ""          "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)


diphthongs <- read.table(textConnection('
Diphthongs
"/ea/, /oa/"
'), TRUE)

kable(diphthongs, align = "c") %>%
  kable_styling("bordered")
```

# Alphabet

* In the sequences ⟨c(i/e)⟩, ⟨g(i/e)⟩, ⟨ch(i/e)⟩, and ⟨gh(i/e)⟩, the front vowels may or may not be transcribed. @Ordean2009 breaks down the environments in which they're transcribed (p. 403). I have included these rules in the rule set.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme                 Comment
"a"              "/a/"                    ""
"ă"              "/ɜ/"                    ""
"â"              "/ɨ/"                    ""
"b"              "/b/"                    ""
"c"            "/k/; /tʃ/"      "/tʃ/: preceding front vowels"
"d"              "/d/"                    ""
"e"              "/e/"                    ""
"f"              "/f/"                    ""
"g"            "/ɡ/; /dʒ/"      "/dʒ/: preceding front vowels"
"h"              "/h/"                    ""
"i"              "/i/"                    ""
"î"              "/ɨ/"                    ""
"j"              "/ʒ/"                    ""
"k"              "/k/"             "appears in loanwords"
"l"              "/l/"                    ""
"m"              "/m/"                    ""
"n"              "/n/"                    ""
"o"              "/o/"                    ""
"p"              "/p/"                    ""
"q"              "/k/"             "appears in loanwords"
"r"              "/r/"                    ""
"s"              "/s/"                    ""
"ș"              "/ʃ/"                    ""
"t"              "/t/"                    ""
"ț"              "/ts/"                   ""
"u"              "/u/"                    ""
"v"              "/v/"                    ""
"w"              "/v/"             "appears in loanwords"
"x"              "/ks/"                   ""
"y"              "/i/"             "appears in loanwords"
"z"              "/z/"                    ""
**Digraph**       ""                      ""
"ea"             "/ea/"                   ""
"oa"             "/oa/"                   ""
"ch"             "/k/"             "preceding front vowels [@Dindelegan2013, p. 14]"
"gh"             "/ɡ/"             "preceding front vowels (ibid.)"
'), TRUE)

kable(alphabet, align = "c") %>%
  kable_styling("bordered")
```

# Lenition Rules

* In spontaneous speech, consonant deletion occurs [@Chitoran, p. 1].

# References
