---
title: "Pangasinan"
author: "Abi Creighton"
edited by: "Emily Strand"
bibliography: pag.bib
output: html_document
---

Last Updated: 2020-01-17

# Background

**Language family**: Austronesian / Malayo-Polynesian / Northern Philippine / Northern Luzon /South-Central Cordilleran / Southern Cordilleran / Pangasinic / Pangasinan

* Pangasinan is spoken in the Philippines.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"         Bilabial Dental Alveolar Postalveolar Palatal Velar Glottal      
Stops                             "p b"   "t̪ d̪"     ""        ""         ""    "k ɡ"   ""       
Affricates                         ""      ""       ""       "tʃ"        ""     ""     ""
Nasals                             "m"     "n̪"      ""        ""         ""     "ŋ"    ""   
Fricatives                         ""      ""       "s"       ""         ""     ""     "h"    
Trills                             ""      "r̪"      ""        ""         ""     ""     ""   
Approximants                       ""      ""       ""        ""         ""     "w"    ""   
Laterals                           ""      "l̪"      ""        ""         "j"     ""    ""                 
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Bilabial", "Dental", "Alveolar", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 7)) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced..", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

## Vowels

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
          Front    Central  Back
High      "i"        ""     "u"
Mid       "ɛ"        ""     "ɔ"
Low       ""         "a"     ""
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme                           Comment
"a"           "/a/"                               ""
"b"           "/b/"                               ""
"c"           "/k/; /s/"   "/s/: preceding front vowels; /k/: preceding back vowels"
"d"           "/d̪/"                               ""
"e"           "/ɛ/"                               ""
"f"           "/p/"                   "appears in Spanish loanwords"
"g"           "/ɡ/"                               ""
"k"           "/k/"                               ""
"h"           "/h/"                               ""
"i"           "/i/"                               ""
"j"           "/h/"                   "appears in Spanish loanwords"
"l"           "/l̪/"                               ""
"m"           "/m/"                               ""
"n"           "/n̪/"                               ""
"o"           "/ɔ/"                               ""
"p"           "/p/"                               ""
"q"           "/k/"                               ""
"r"           "/r̪/"                               ""
"s"           "/s/"                               ""
"t"           "/t̪/"                               ""
"u"           "/u/"                               ""
"v"           "/b/"                  "appears in Spanish loanwords"
"w"           "/w/"                               ""
"y"           "/j/"                               ""
"z"           "/s/"                  "appears in Spanish loanwords"
**Digraph**     ""                                ""
"ch"          "/tʃ/"                              ""
"ng"          "/ŋ/"                               ""
"qu"          "/k/"                               ""
'), TRUE)

kable(alphabet) %>%
  kable_styling("bordered")
```

# Lenition Rules

* /d/ may be realized intervocalically as a flap [@BentonP1971, p. 16]; /d/ and [ɾ] vary freely word-finally.
* /t/ and /k/ may be realized as [ʔ] syllable-finally, preceding a consonant [@BentonP1971, pp. 17-18].
* /k/ may be realized as a fricative intervocalically [@BentonP1971, pp. 18-19].

# Misc. Rules

* Nasals assimilate in place to the following consonant [@BentonP1971, p. 21].
* /l/ may be realized as [d] word-finally [@BentonP1971, p. 17].
* A non-phonemic glottal stop is epenthesized word-initially and can be epenthesized intervocalically [@BentonP1971, p. 13].

# References
