---
title: "Hiligaynon"
author: "Abi Creighton"
bibliography: hil.bib
output: html_document
---

Last Updated: 2020-07-27

# Background

**Language Family:** Austronesian / Malayo-Polynesian / Western Malayo-Polynesian / Meso Philippine / Central Philippine / Bisayan / Central / Peripheral

* Hiligaynon, also known as Ilonggo, is spoken in the Philippines. It is most widely spoken in the regions of Western Visayas and Soccsksargen.  

# Phonology

## Consonants

* @wolfenden_1971 and @casperson_2010 both include /f/ and /v/ (although introduced through loanwords) in their phonemic inventory, while @motus_1971 claims they are allophones of /p/ and /b/, respectively, and are only used by some speakers as a sign of prestige (p. ix). Therefore, I have opted to omit /f/ and /v/ in the analysis below.
* @casperson_2010 reports an alveolar place for /t/, /d/, /n/, and /s/, but other documentation suggests that they are dental [@wolfenden_1971, pp. 17-20; @motus_1971, p. ix].
* The affricates /c/ and /ʝ/ appear in loanwords [@wolfenden_1971, p. 19].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial  Dental Alveolar Palatal Velar Glottal
Stops                      "p b"   "t̪ d̪"     ""      ""    "k ɡ"   "ʔ"           
Fricatives                  ""      "s̪"      ""      ""     ""     "h"
Nasals                      "m"     "n̪"      ""      ""     "ŋ"    ""       
Taps                        ""      ""       "ɾ"     ""     ""     ""       
Approximants                ""      ""       "l"     "j"    "w"    ""   
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Dental", "Alveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```  

## Vowels

* /e/ and /o/ are borrowed from Spanish and English, but only exist in Hiligaynon as allophones of /i/ and /u/ respectively [@motus_1971, p. ix].
* In some orthographies, an acute accent on a vowel marks stress, which is phonemic in Hiligaynon (ibid.). However, stress is not relevant for the purposes of this project and will not be transcribed.
* Possible Hiligaynon diphthongs include /ai/ and /au/ [@manzano_2016, p. 43]. Because they can also be interpreted as sequences of independent vowels rather than diphthongs, however, they will not be accounted for in the rules [@manzano_2016, p. 43].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
             Front      Central       Back
High          "i"         ""          "u"
Low           ""          "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```     

# Alphabet

* The glottal stop can be written with either ⟨q⟩ or an apostrophe depending on which orthographic convention is being followed. Both variations are included here.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme           
"a"              "/a/"            
"b"              "/b/"            
"k"              "/k/"            
"d"              "/d̪/"           
"g"              "/ɡ/"            
"h"              "/h/"            
"i"              "/i/"            
"l"              "/l/"            
"m"              "/m/"            
"n"              "/n̪/"           
"p"              "/p/"            
"q"              "/ʔ/"            
"\'"             "/ʔ/"            
"r"              "/ɾ/"              
"s"              "/s̪/"             
"t"              "/t̪/"             
"u"              "/u/"              
"w"              "/w/"              
"y"              "/j/"              
**Digraph**      ""                 
"ng"             "/ŋ/"              
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* After nasal place assimilation occurs, the consonant following the nasal may be deleted [@wolfenden_1971, p. 28].
* Stem-final /d̪/ becomes the tap [ɾ] when certain suffixes are added [@wolfenden_1971, p. 29].

# Misc. Rules

* Glottal stops are predictably inserted preceding word-initial vowels [@wolfenden_1971, p. 13].
* When a suffix is added, a vowel may be deleted from a stem-final syllable, thus causing two consonants to become adjacent. The consonants may then metathesize [@wolfenden_1971, p. 27].
* Nasals may assimilate to the place of the following consonant (ibid.).

# References
