---
title: "Cape Verdean Creole"
author: "Abi Creighton"
edited by: "Emily Strand"
bibliography: kea.bib
output: html_document
---

Last Updated: 2020-07-28

**COMPROMISED: possible conflation between /a/ and /ɐ/, /e/ and /ɛ/, and /ɾ/ and /ʀ/**

# Background

**Language Family:** Creole / Portuguese based

* Cape Verdean Creole, known by native speakers as Kriolu, Kriol, or Kabuverdianu, is a Portuguese-based creole spoken in Cape Verde, an island country west of Africa.

# Phonology

## Consonants

* The phonemic status of /j/ and /w/ is somewhat in question:
    - @Swolkien2013 treats [j] and [w] as allophones of /i/ and /u/ (and sometimes /o/), respectively (p. 23). They surface when adjacent to other vowels. @Swolkien2019, on the other hand, do include them (p. 182). Since two different varieties of the language are analyzed here [São Vicente in @Swolkien2013 and Santo Antão in @Swolkien2019], it may very well be the case that /j/ and /w/ are phonemic in one variety but not in the other. Regardless, I have chosen to treat them as allophones and not include them below.
* There is also some uncertainty regarding the flaps/trills of the language:
    - @Swolkien2015 includes the alveolar flap /ɾ/ and uvular trill /ʀ/, both represented orthographically by ⟨r⟩ (p. 33). The same author, discussing the same variety (São Vicente), however, identifies the alveolar trill /r/ as opposed to the uvular trill [@Swolkien2013, p. 23].
    - Also in reference to the São Vicente variety, @monteiro_2015 argues for the uvular fricative /ʁ/ as opposed to the trill.
    - Although based on a different variety (Santo Antão), @Swolkien2019 agree with the attestation of the alveolar flap and uvular trill; however, they are noted to be orthographically distinct (/ɾ/ = ⟨r⟩, /ʀ/ = ⟨rr⟩) (p. 182).
    - I have opted to include the alveolar flap and the uvular trill in the inventory below.


```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial  Alveolar Postalveolar Palatal Velar  Uvular
Stops                      "p b"    "t d"        ""        ""    "k ɡ"   ""
Affricates                  ""       ""        "tʃ dʒ"     ""     ""     ""
Fricatives                 "f v"    "s z"       "ʃ ʒ"      ""     ""     ""
Nasals                      "m"      "n"         ""        "ɲ"    ""     ""
Trills                      ""       ""          ""        ""     ""     "ʀ"     
Flaps                       ""       "ɾ"         ""        ""     ""     ""     
Approximants                ""       "l"         ""        "ʎ"    ""     ""   
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Uvular"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```  

## Vowels

* Cape Verdean Creole has eight oral vowels, each with a nasal counterpart [@Swolkien2015, pp. 29-31]. Nasal vowels are represented by a following ⟨n⟩.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
             Front      Central    Back
High          "i"       ""         "u"
High-mid      "e"       ""         "o"
Low-mid       "ɛ"       "ɐ"        "ɔ"
Low           ""        "a"        ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```     

# Alphabet

* Cape Verdean Creole is written with ALUPEC, the Alfabeto Unificado para a Escrita do Caboverdiano (Unified Alphabet for Cape Verdean Writing) [@coonan_2007, p. 40]. Spanish based conventions may also be used (e.g. ⟨c⟩ or ⟨q⟩ for /k/), but I have opted to focus on ALUPEC below.
* To differentiate from nasal vowels, oral vowel-/n/ sequences are indicated by a following ⟨nn⟩ [@Swolkien2019, p. 181].
    - e.g. ⟨an⟩ = /ã/, ⟨ann⟩ = /an/
* /ɾ/ and /ʀ/ are ultimately conflated, as it is unclear how they are orthographically represented (see Consonants above for explanation). Although there is some probability that ⟨r⟩ represents both /ɾ/ and /ʀ/, I have chosen to transcribe ⟨r⟩ to /ɾ/ and ⟨rr⟩ to /ʀ/.
* There is also possible conflation between /a/ and /ɐ/ and between /e/ and /ɛ/, especially in syllables that don't have canonical stress, as the graphemes for /ɐ/ and /ɛ/ (⟨â⟩ and ⟨é⟩, respectively) in these positions are said not to be fully attested [@Swolkien2019, p. 181]. Both graphemes, however, are present in the Crúbadán corpus, but there is still a possibility for conflation.
* Due to the uncertainty around the transcription of the flap/trill and some vowels, the language is deemed compromised.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme              Comment          
"a"              "/a/; /ɐ/"    "/a/: default in ruleset"
"á"              "/a/"                 ""
"â"              "/ɐ/"                 ""
"b"              "/b/"                 ""
"d"              "/d/"                 ""
"e"              "/e/; /ɛ/"    "/e/: default in ruleset"
"é"              "/ɛ/"                 ""
"ê"              "/e/"                 ""
"f"              "/f/"                 ""
"g"              "/ɡ/"                 ""
"i"              "/i/"                 ""
"í"              "/i/"                 ""
"j"              "/ʒ/"                 ""
"k"              "/k/"                 ""
"l"              "/l/"                 ""
"m"              "/m/"                 ""
"n"              "/n/"                 ""
"o"              "/o/"                 ""
"ó"              "/ɔ/"                 ""
"ô"              "/o/"                 ""
"p"              "/p/"                 ""
"r"              "/ɾ/; /ʀ/"     "/ɾ/: default in ruleset"
"s"              "/s/"                 ""
"t"              "/t/"                 ""
"u"              "/u/"                 ""
"ú"              "/u/"                 ""        
"v"              "/v/"                 ""
"x"              "/ʃ/"                 ""
"z"              "/z/"                 ""
**Digraph**       ""                   ""                
"dj"             "/ɟ/"                 ""
"lh"             "/ʎ/"                 ""
"nh"             "/ɲ/"                 ""
"rr"             "/ʀ/"                 ""
"tx"             "/c/"                 ""
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Misc. Rules

* /v/ is frequently realized as [β], especially when preceding /ɔ/ [@Neves2009, p. 6].
* In the Santo Antão variety, /ʃ/ becomes voiced preceding stops or vowels (ibid.). In some cases, it may also be realized as [s].
* In unstressed word-final positions, vowels are neutralized to either [i],  [ɐ], or [u] [@lang_2013].  

# References
