---
title: "Yawa"
author: "Emily Strand"
bibliography: yva.bib
output: html_document
---

Last Updated: 2020-07-06

# Background

**Language Family:** West Papuan / Yapen

* Also referred to as Yava and Mantembu, Yawa is spoken in the Papua province of Indonesia.
* Yawa has incorporated many Indonesian loanwords, which have mostly been adapted to Yawa's phonological system [@jones_yawa_1986, p. 25].
* The linguistic classification of Yawa is somewhat suspect; some speculate it (or Yapen) is an independent language family.

# Phonology

* The only documentation regarding the phonology of Yawa is provided by @jones_yawa_1986, which is said to reflect only the Sarawandori dialect [@jones_dialects_1968, p. 33]; however, he states that all dialects of Yawa are incredibly similar.

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial       Alveolar     Postalveolar   Alveopalatal  Palatal  Velar   
Stops                      "p  b"        "t  d"          ""              ""         ""     "k"
Affricates                   ""            ""            "ʃ"             "dʑ"       ""     ""
Fricatives                   ""            "s"           ""              ""         ""     ""
Nasals                       "m"           "n"           ""              ""         "ɲ"    ""
Flaps                        ""            "ɾ"           ""              ""         ""     ""
Approximants                 "w"           ""            ""              ""         "j"    ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Alveopalatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```

## Vowels

* Most of the diphthongs that occur are realized as vowel-glide sequences. /ui/ is the only exception, where it's realized either as a glide plus a vowel or a vowel plus a glide [@jones_yawa_1986, p. 17].
    - These diphthongs may be realized merely as vowel sequences [@jones_yawa_1986, p. 17]; therefore, no rules will be included to account for them.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         ""           "u"
Mid       "e"         ""           "o"
Low        ""         "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```    

# Alphabet

* The most comprehensive account of the Yawa orthography is provided by @jones_yawa_1986, which is what the alphabet below is heavily patterned after. However, the same author also provides another account about the transcription [@jones_pronunciation_1989]. In addition to the graphemes from the 1986 source, the 1989 source includes two additional graphemes ⟨ng⟩ and ⟨v⟩ (p. xvii). The former is said to represent /ŋ/, but there isn't a clear description about the transcription of ⟨v⟩, which is "pronounced like in English except with both lips close together."
    - Given that this source provides the pronunciation (i.e. not specific about the underlying phonemes), it could be inferred that /ŋ/ is allophonic, just like the realization of [ɡ] [see Misc. Rules; @jones_yawa_1986, p. 6]. Furthermore, looking at the words contained in Crúbadán, all instances of ⟨ng⟩ are followed by a velar stop. Therefore, it is likely that ⟨ng⟩ represents the phoneme /n/ that has assimilated to the place of the following velar stop.
    - Based on the description of ⟨v⟩, it appears that the pronunciation is somewhere between that of /v/ and /w/, leading me to believe that there could be some free variation with the two (ex. Yawa is understood as both Yawa and Yava). Given that @jones_yawa_1986 doesn't provide a corresponding phoneme for ⟨v⟩, I have chosen to use /w/.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme      Comment
"a"            "/a/"         ""
"b"            "/b/"         ""  
"d"            "/d/"         ""  
"e"            "/e/"         ""
"g"            "/dʑ/"         ""  
"i"            "/i/"         ""  
"j"            "/dʑ/"         ""  
"k"            "/k/"         ""
"m"            "/m/"         ""
"n"            "/n/"         ""  
"o"            "/o/"         ""  
"p"            "/p/"         ""  
"r"            "/ɾ/"         ""  
"s"            "/s/"         ""  
"t"            "/t/"         ""  
"u"            "/u/"         ""
"v"            "/w/"         ""  
"w"            "/w/"         ""  
"y"            "/j/"         ""
**Digraph**      ""          ""
"ng"           "/n/"         "precedes velar stops; realized as [ŋ]"
"ny"           "/ɲ/"         ""  
"sy"           "/ʃ/"         ""
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* The syllable structures permitted in Yawa include [@jones_yawa_1986, p. 2]:
    - CV  
    - CCV  
    - CVV  
    - CCVV  
    - V  
    - VV  
* Consonant clusters must consist of a homorganic nasal followed by a stop or a fricative [@jones_yawa_1986, p. 15].

# Misc. Rules

* Non-palatal consonants (excluding /t/) are labialized when preceded by /u/ and followed by another vowel [@jones_yawa_1986, p. 5].
* /w/ occurs in free variation with [β] [@jones_yawa_1986, p. 6].
* /dʑ/ is realized as [ɡ] following /u/ (ibid.).
* Unstressed /u/, word-initially, is deleted preceding a lone consonant [@jones_yawa_1986, p. 8].

# References
