---
title: "Yawa"
author: "Emily Strand"
bibliography: yva.bib
output: html_document
---

Last updated: 09-23-2019

# Background

**Language family**: West Papuan / Yapen

* Also referred to as Yava and Mantembu, Yawa is spoken in the Papua province of Indonesia.

# Phonology

* The only documentation regarding the phonology of Yawa was provided by @jones_yawa_1986, which is said to reflect only the Sarawandori dialect [@jones_dialects_1968, p. 33]; however, he states that all dialects of Yawa are incredibly similar.

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial       Alveolar     Postalveolar   Alveopalatal  Palatal  Velar   
Stops                      "p  b"        "t  d"          ""              ""         ""     "k"
Affricates                   ""            ""            "ʃ"             "dʑ"       ""     ""
Fricatives                   ""            "s"           ""              ""         ""     ""
Nasals                       "m"           "n"           ""              ""         "ɲ"    ""
Flaps                        ""            "ɾ"           ""              ""         ""     ""
Approximants                 "w"           ""            ""              ""         "j"    ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Alveopalatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```

## Vowels

* Most of the diphthongs that occur are realized as a vowel plus a glide. /ui/ is the exception, where it's realized either as a glide plus a vowel or a vowel plus a glide [@jones_yawa_1986, p. 17].
    - These diphthongs may be realized merely as vowel sequences [@jones_yawa_1986, p. 17]; therefore, no rules will be included to account for them.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         ""           "u"
Mid       "e"         ""           "o"
Low        ""         "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)


diphthongs <- read.table(textConnection('
Diphthongs
"/ui/, /oi/, /ai/, /ae/, /ao/, /au/"
'), TRUE)

kable(diphthongs, align = 'c') %>%
  kable_styling("bordered")
```    

# Alphabet

* The most comprehensive account of the Yawa orthography was provided by @jones_yawa_1986, which is reflected largely by the alphabet listed below. However, three years later, in 1989, the same author provided another account about the pronunciation [@jones_pronunciation_1989]. In addition to the graphemes from the 1986 source, the 1989 source included two additional graphemes ⟨ng⟩ and ⟨v⟩ (p. xvii). The former is said to be pronounce as /ŋ/, but there isn't a clear description about the pronunciation of ⟨v⟩ ("pronounced like in English except with both lips close together").
    - Given that this source provides the pronunciation (not specific about the underlying phonemes), it could be inferred that /ŋ/ is allophonic, just like the realization of [g] [@jones_yawa_1986, p. 6]. Therefore, I won't treat /ŋ/ as a phoneme.
    - Based on the description of ⟨v⟩, it appears that the pronunciation is somewhere between that of /v/ and /w/, leading me to believe that there could be some free variation with the two (ex. Yawa is understood as both Yawa and Yava). Given that @jones_yawa_1986 didn't provide a corresponding phoneme for ⟨v⟩, I use /w/.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme
"a"            "/a/"  
"b"            "/b/"  
"d"            "/d/"  
"e"            "/e/"
"g"            "/dʑ/"  
"i"            "/i/"  
"j"            "/dʑ/"  
"k"            "/k/"  
"m"            "/m/"  
"n"            "/n/"  
"o"            "/o/"  
"p"            "/p/"  
"r"            "/ɾ/"  
"s"            "/s/"  
"t"            "/t/"  
"u"            "/u/"
"v"            "/w/"  
"w"            "/w/"  
"y"            "/j/"
**Digraph**    ""
"ny"           "/ɲ/"  
"sy"           "/ʃ/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* The syllable structures permitted in Yawa include [@jones_yawa_1986, p. 2]:
    - CV  
    - CCV  
    - CVV  
    - CCVV  
    - V  
    - VV  
* Consonant clusters must consist of a homorganic nasal followed by a stop or a fricative [@jones_yawa_1986, p. 15].

# Lenition Rules

* Unstressed /u/, word-initially, is deleted preceding a lone consonant [@jones_yawa_1986, p. 8].

# Misc. Rules

* Non-palatal consonants (excluding /t/) are labialized when preceded by /u/ and followed by another vowel [@jones_yawa_1986, p. 5].
* /w/ occurs in free variation with [β] [@jones_yawa_1986, p. 6].
* /dʑ/ is realized as [g] following /u/ (ibid.).

# References
