---
title: "Wayuu"
author: "Emily Strand"
bibliography: guc.bib
output: html_document
---

Last Updated: 2020-06-29

# Background

**Language Family**: Arawakan / Maipuran / Northern Maipuran / Caribbean

* Also referred to as Goajiro and Wayuunaiki, Wayuu is spoken within the Guajira department, along the Guajira Peninsula located in the northern tip of South America.

# Phonology

## Consonants

* Some documentation includes /β/ and /ɲ/ as phonemes corresponding to ⟨v⟩ and ⟨ñ⟩ respectively [@Hildebrandt1963; @Uterga1895, as cited in @sabogal_variable_2018, pp. 50-52]. However, given that other authors omit them from the phonemic inventory, citing them as allophones of /w/ and /n/ respectively, forming no minimal pairs [@alvarez_manual_2017, p. 18; @sabogal_variable_2018, pp. 50, 52], I have also chosen to exclude them.
* Excluding /h/, /r/, /w/, and /j/, consonants can geminate [@holmer_goajiro_1949, p. 49].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial    Alveolar    Postalveolar    Palatal     Velar     Glottal
Stops                        "p"       "t"           ""            ""         "k"        "ʔ"  
Affricates                    ""       ""            "tʃ"          ""         ""         ""
Fricatives                    ""       "s"           "ʃ"           ""         ""         "h"
Nasals                       "m"       "n"           ""            ""         ""         ""
Flaps                         ""       "ɺ"           ""            ""         ""         ""
Trills                        ""       "r"           ""            ""         ""         ""
Approximants                  "w"       ""           ""            "j"        ""         ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* Vowel length is contrastive. Long vowels are indicated by repeated graphemes [@sabogal_variable_2018, p. 49].
* There is some debate as to whether /ɯ/ or /ɨ/ corresponds to ⟨ü⟩. Whereas @mansen_aprendamos_1984 opts for /ɨ/ (p. 4), @alvarez_manual_2017 and @sabogal_variable_2018 opt for /ɯ/ (p. 18; p. 49). I have chosen to use /ɯ/, which is more consistent with recent documentation.
* Rising vowel sequences can be pronounced as diphthongs or independent monophthongs [i.e. the pronunciation is in free variation; @sabogal_variable_2018, p. 50].
* Vowel harmony is present in Wayuu, structured on the basis of fronting and rounding [@sabogal_variable_2018, p. 53].
* Vowels often become nasalized; however, this is not a phonemic feature [@mansen_aprendamos_1984, p. 9].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         ""         "ɯ  u"
Mid       "e"         ""           "o"
Low        ""         "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: The high back vowel on the left is unrounded and the high back vowel on the right is rounded.", general_title = "")
```     

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme
"a"            "/a/"  
"e"            "/e/"  
"i"            "/i/"  
"j"            "/h/"  
"k"            "/k/"  
"l"            "/ɺ/"  
"m"            "/m/"  
"n"            "/n/"  
"o"            "/o/"  
"p"            "/p/"  
"r"            "/r/"  
"s"            "/s/"  
"t"            "/t/"  
"u"            "/u/"  
"ü"            "/ɯ/"  
"w"            "/w/"  
"y"            "/j/"  
"\'"            "/ʔ/"  
**Digraph**     ""
"ch"           "/tʃ/"  
"sh"           "/ʃ/"  
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* /k/ and /p/ become voiced preceding /ɺ/ [@holmer_goajiro_1949, p. 49].

# Misc. Rules

* /ɺ/ becomes /r/ preceding front vowels [@sabogal_variable_2018, p. 60].
* /ɺ/ is unpronounced when preceding a monomaraic (short) /i/ [@sabogal_variable_2018, p. 60].
* Word-final vowels and vowels preceding /s/, /tʃ/, and /ʃ/ are often pronounced lenis or unpronounced [@holmer_goajiro_1949, p. 50].
* Vowel elision and liquid fortition often occur in tandem [@sabogal_variable_2018, p. 59,60]. First, vowels are elided when preceding a liquid and following an obstruent. Then, with the resulting liquid-obstruent sequence, the liquid becomes /t/ (or assimilates to /n/ if the obstruent was an /n/).

# References
