---
title: "Borong"
author: "Allison Shao"
bibliography: ksr.bib
output: html_document
---

Last updated: 2020-03-28

# Background

**Language Family:** Trans-New Guinean / Main Section / Central and Western / Huon-Finisterre / Huon / Eastern

* Also referred to as Kosorong, it is spoken by approximately 2000 speakers in Papua New Guinea.

# Phonology

## Consonants

* /ts/ is rather marginal [@Olkkonen2000, p. 4].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"      Labial    Alveolar   Palatal    Velar  Labiovelar  Glottal
Stops                         "p b"      "t d"       ""       "k ɡ"   "kp ɡb"      ""
Affricates                     ""       "ts dz"      ""        ""       ""         ""
Fricatives                     "f"        "s"        ""        ""       ""         "h"
Nasals                         "m"        "n"        ""        "ŋ"      ""         ""
Trills                         ""         "r"        ""        ""       ""         ""
Approximants                   "w"        "l"        "j"       ""       ""         ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Palatal", "Velar", "Labiovelar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```

## Vowels

* Vowel length is contrastive in Borong [@Olkkonen2000, p. 5]. Long vowels are represented by duplicate graphemes.
* Adjacent vowels are extremely common in Borong; however, they are interpreted as sequences rather than complex nuclei [@Olkkonen2000, p. 5].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         ""          "u"        
Mid       "e"         ""          "o"
Low        ""         "a"         ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```   

# Alphabet

* @Olkkonen2000 uses ⟨ŋ⟩ to represent /ŋ/ (p. 4), while @Olkkonen2001 uses ⟨ng⟩ (p. 1). I have chosen to use ⟨ŋ⟩ given that it is present in the Crúbadán corpus.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme
"a"	           "/a/"
"b"	           "/b/"
"d"	           "/d/"
"e"	           "/e/"
"f"            "/f/"
"g"	           "/ɡ/"
"h"	           "/h/"
"i"	           "/i/"
"j"            "/dz/"
"k"	           "/k/"
"l"	           "/l/"
"m"	           "/m/"
"n"	           "/n/"
"ŋ"            "/ŋ/"
"o"	           "/o/"
"p"	           "/p/"
"q"            "/kp/"
"r"            "/r/"
"s"	           "/s/"
"t"	           "/t/"
"u"	           "/u/"
"w"	           "/w/"
"y"	           "/j/"
"z"            "/ts/"
**Digraph**     ""
"gb"           "/ɡb/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Syllables in Borong follow the order of (C)V(V)(C) [@Olkkonen2000, p. 6].
    - CV and CVC are the most common structures.

# Misc. Rules

* /p/, /t/, and /k/ are aspirated in syllable-initial positions [@Olkkonen2001, p. 5].
* /w/ may be realized as [v] [@Olkkonen2000, p. 4].
* Nasals assimilate in place to following consonants [@Olkkonen2000, p. 8].

# References  
