---
title: "Mufian"
author: "Emily Strand"
bibliography: aoj.bib
output: html_document
---

Last Updated: 2020-01-09

**Compromised: conflation among /a/, /æ/, and /ɘ/, between /ɘ/ and /o/, and between /e/ and /ɘ/; ambiguity among long vowels; ambiguity related to whether phonemic (labialized) clusters are always realized as such**

# Background

**Language Family**: Torricelli / Kombio-Arapesh / Arapesh

* Mufian is spoken in the East Sepik province of Papua New Guinea.

# Phonology

## Consonants

* @Conrad1977 do not include the labialized consonants as part of the phonemic inventory, but notes that they contrast with their non-labialized counterparts (p. 3). Given this and that @1992 and @Conrad1978 include them, I have chosen to include them below.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial    Alveolar   Velar     Glottal
Stops                       "p b"     "t d"  "kʷ k ɡ ɡʷ"   "ʔ ʔʷ"
Affricates                   ""        ""        ""         ""
Fricatives                   "f"       "s"       ""         "h"
Nasals                       "m"       "n"       ""         ""
Approximants                 "w"       "l"       ""         ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 4)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced. Phonemes that have the diacritic (ʷ) are labialized.", general_title = "")
```  

## Vowels

* @1992 as well as @Conrad1977 include /ə/ as a phoneme in Mufian (p. 2; p. 3); however, I've opted for /ɘ/ given that @Conrad1978 describe it as a high to mid-central vowel (p. 90).
* Adjacent vowels are interpreted as sequences [@Conrad1977, pp. 6-7].
* The documentation for long vowels is ambiguous. @Conrad1977 state that, although contrastive, they are rather infrequent, citing length for only /i/, /a/, and /æ/ (contrastive occurrences for only /a/ and /æ/) [@Conrad1977, pp. 9-10, 23]. @1992 also lists long vowels, but for /a/, /e/, and /æ/ (p. 2). @1992 also notes that long vowels are transcribed just as short vowels are. Because they are rather infrequent and have a low functional load, I have opted not to include them.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         ""           "u"
Mid       "e"         "ɘ"          "o"
Low       "æ"         "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

* The digraphs ⟨mb⟩ and ⟨nd⟩ transcribe to word-medial phonetic prenasalization, so I will transcribe them to /b/ and /d/ respectively [@Conrad1977, p. 6]. The prenasalized /ɡ/ is transcribed as ⟨g⟩, so it poses no correction.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme  
"a"        "/a/; /æ/; /ɘ/"
"b"             "/b/"
"d"             "/d/"
"e"           "/e/; /ɘ/"
"f"             "/f/"
"g"             "/ɡ/"
"h"             "/h/"
"i"             "/i/"
"k"             "/k/"
"l"             "/l/"
"m"             "/m/"
"n"             "/n/"
"o"           "/o/; /ɘ/"
"p"             "/p/"
"s"             "/s/"
"t"             "/t/"
"u"             "/u/"
"w"             "/w/"
"\'"            "/ʔ/"
**Digraph**      ""
"mb"            "/b/"
"nd"            "/d/"
"gw"            "/ɡʷ/"
"kw"            "/kʷ/"
"\'w"           "/ʔʷ/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Misc. Rules

* Voiceless consonants are aspirated word-finally [@Conrad1977, pp. 12-13].

# References
