---
title: "Qʼanjobʼal"
author: "Emily Strand"
bibliography: kjb.bib
output: html_document
---

Last updated: 2019-11-01

# Background

**Language family**: Mayan / Kanjobalan-Chujean / Kanjobalan / Kanjobal-Jacaltec

* Qʼanjobʼal is spoken in the Huehuetenango and Quiché departments of Guatemala.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"  Bilabial   Alveolar   Postalveolar   Retroflex  Palatal   Velar     Uvular   Glottal
Stops                      "p  ɓ"     "t  tʼ"        ""           ""        ""      "k  k’"   "q  q’"    "ʔ"
Affricates                  ""       "ts  tsʼ"   "tʃ  tʃ’"     "tʂ   tʂ’"   ""        ""        ""       ""
Fricatives                  ""         "s"          "ʃ"           "ʂ"       ""        "x"       ""       ""
Nasals                      "m"        "n"          ""            ""        ""        ""        ""       ""
Flaps                       ""         "ɾ"          ""            ""        ""        ""        ""       ""
Approximants                "w"        "l"          ""            ""        "j"       ""        ""       ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Bilabial", "Alveolar", "Postalveolar", "Retroflex", "Palatal", "Velar", "Uvular", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 8)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: Phonemes that have the diacritic (ʼ) are ejectives. The bilabial stop on the right is an implosive.", general_title = "")
```

## Vowels

* In general, vowel length and vowel nasalization are not contrastive [@Lichtman2010, p. 13].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         ""           "u"
Mid       "e"         ""           "o"
Low        ""         "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```   

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme           Comment
"a"            "a"                ""
"b\'"          "ɓ"                ""
"e"            "e"                ""
"h"            "Ø"   "represents the absence of a glottal stop before a vowel [@Lichtman2010, p. 3]"
"i"            "i"                ""
"j"            "x"                ""
"k"            "k"                ""
"k\'"          "k’"               ""
"l"            "l"                ""
"m"            "m"                ""
"n"            "n"                ""
"o"            "o"                ""
"p"            "p"                ""
"q"            "q"                ""
"q\'"          "q’"               ""
"r"            "ɾ"                ""
"s"            "s"                ""
"t"            "t"                ""
"t\'"          "tʼ"               ""
"u"            "u"                ""
"w"            "w"                ""
"x"            "ʂ"                ""
"y"            "j"                ""
"\'"           "ʔ"                ""
**Digraph**    ""                 ""
"ch"           "tʃ"               ""
"ch\'"         "tʃ’"              ""
"tx"           "tʂ"               ""
"tx\'"         "tʂ’"              ""
"tz"           "ts"               ""
"tz\'"         "tsʼ"              ""
"xh"           "ʃ"                ""
"**Orthographic Variations**"  "" ""
"ẍ"            "ʃ"                ""
"c"            "k"                ""
"c\'"          "k’"               ""
"b"            "ɓ"                ""
 '), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Syllables in Qʼanjobʼal follow the order of (C)(C)V(C) [@Toledo2017, p. 535].

# Lenition Rules

* Stops tend to become voiced and spirantize between voiced stops [@Lichtman2010, p. 18].
* Nasals and vowels word-finally tend to be deleted [@Lichtman2010, p. 20]. Word-medial vowels may also be deleted.
* /t/ (word-finally) and /ɓ/ (word and syllable-finally) tend to delete (ibid.).

# Misc. Rules

* /w/ fricates, becoming [β] preceding front vowels [@Toledo2017, p. 534].
* Vowels in word-initial positions are preceding by a glottal stop [@Lichtman2010, p. 3].
* Consonant gemination does not occur [@Lichtman2010, p. 19].

# References
