---
title: "Javanese"
author: "Bill Mizgerd"
bibliography: jv.bib
output: html_document
---

Last Updated: 2019-02-08

# Background

**Language Family**: Austronesian / Malayo-Polynesian / Javanese

* Also called Djawa, Javanese is spoken in the Banten, Central Java, and East Java provinces of Indonesia. It is also spoken in Yogyakarta and Sumatra island, which are also located within Indonesia.

# Phonology

## Consonants

* @JvSuharno1982 uses /q/ instead of /ʔ/, but describes the consonant as glottal; @JvNothofer1975 also uses /q/, but in his table gives it the same place of articulation as /h/. For these reasons, I have followed the pattern of the authors (Wolff, Horne, Robson) who prefer /ʔ/.
* I have chosen to include dental and retroflex stops, but not alveolar stops. The authors I've read have disagreed about which of the aforementioned stops do occur in Javanese; all agree, however, that exactly two of the three places of articulation are represented. Dental and retroflex is the pair that I find to be likeliest.

    - @JvHorne1974: Dental, alveolar  
    - @JvNothofer1975: Alveolar, retroflex(?)  
    - @JvRobson2014: Dental, retroflex  
    - @JvSuharno1982: Dental, retroflex  
    - @JvWolffChart1982: Dental, alveolar  
<br/><br/>

* Regarding the prenasalized stops, @JvWolffChart1982 does not depict /ᵑɡ/ as assimilating, although @JvNothofer1975 and @JvHorne1974 do. Given that (at least some) other prenasalized stops feature nasal assimilation, and given that /ŋ/ is phonemic in Javanese, I have followed Wolff's example. Although I have not seen explicit attestation of nasal assimilation in the rest of the prenasalized stops, I find it to be likely enough that I have included it in my ruleset.
* Although @JvWolffChart1982 accounts for the lenis continuants (pharyngealized) /wh/, /nh/, /lh/, /rh/, /jh/, they are unattested elsewhere, so I have not included them in the ruleset.  
* /f/ and /z/ appear in loanwords [@JvWolffChart1982].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial    Dental   Alveolar   Retroflex    Palatal     Velar    Glottal
Stops                      "p  b"    "t̪  d̪"      ""       "ʈ  ɖ"       "c  ɟ"    "k  ɡ"      "ʔ"
"Stops (prenasalized)"      "ᵐb"      "ⁿd̪"       ""        "ⁿɖ"         "ⁿɟ"      "ᵑɡ"       ""
Fricatives                   ""        ""        "s"        ""           ""        ""        "h"  
Nasals                       "m"       ""        "n"        ""           "ɲ"       "ŋ"       ""
Trills                       ""        ""        "r"        ""           ""        ""        ""
Approximants                 "w"       ""        "l"        ""           "j"       ""        ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Dental", "Alveolar", "Retroflex", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 7)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```

## Vowels

* Vowels may occur successively, but they do not form diphthongs when doing so [@JvSuharno1982, page 2].
* Javanese is characterized as having a "neutral" vowel [@JvRobson2014, p. xii], often represented as /ɘ/. However, this neutral vowel in Javanese is not truly reduced, and based on the chart by @JvWolffChart1982, it appears to represent a sound lower than that of schwa (p. 3). For this reason, I have opted to use /ɜ/.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         ""           "u"
High-Mid  "e"         ""           "o"
Low-Mid   "ɛ"         "ɜ"          "ɔ"
Low        ""         "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```  

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme                                Comment
"a"            "/a/"                                   ""
"b"            "/b/"                                   ""
"c"            "/c/"                   "described by @JvHorne1974 as /t̪s/"  
"d"            "/d̪/"                                   ""
"e"            "/ɜ/"                                   ""
"é"            "/e/"                                   ""
"è"            "/ɛ/"                                   ""
"f"            "/f/"  "only in loanwords; sometimes replaced by /p/ (see @JvHorne1974, p. xi)"  
"g"            "/ɡ/"                                   ""
"h"            "/h/"                                   ""
"i"            "/i/"                                   ""
"j"            "/ɟ/"                    "described by @JvHorne1974 as /d̪z/"  
"k"            "/k/"                                   ""
"l"            "/l/"                                   ""
"m"            "/m/"                                   ""
"n"            "/n/"                                   ""
"o"            "/o/"                                   ""
"ô"            "/ɔ/"                   "not present in the Crúbadán corpus"  
"p"            "/p/"                                   ""
"q"            "/ʔ/"                           "only in loanwords"  
"r"            "/r/"                                   ""
"s"            "/s/"                                   ""
"t"            "/t̪/"                                   ""
"u"            "/u/"                                   ""
"v"            "/f/"        "only in loanwords; per @JvHorne1974, is an alternate spelling of ⟨f⟩"
"w"            "/w/"                                   ""
"y"            "/j/"                                   ""
**Multigraph**  ""                                     ""
"dh"           "/ɖ/"                       "some transcriptions use ⟨ḍ⟩"
"mb"           "/ᵐb/"                                  ""
"nd"           "/ⁿd̪/"                                  ""
"ng"           "/ŋ/"                                   ""
"nj"           "/ᶮɟ/"                                  ""
"ny"           "/ɲ/"                                   ""
"th"           "/ʈ/"                        "some transcriptions use ⟨ṭ⟩"
"ndh"          "/ⁿɖ/"                                  ""  
"ngg"          "/ᵑɡ/"                                  ""
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Javanese Script

* A ruleset for the Javanese abugida is also provided (jv-jscript.rules). Because Crúbadán's only Javanese corpus uses the Latin script, the Javanese-script ruleset is not as thoroughly tested as its Latin-script counterpart (no corresponding verify file).
* The phonology of the Javanese script is taken from @JvOmniglot and @JvRomanization.
* Despite being present in the Latin script, /ɛ/ does not seem to be attested in the Javanese script.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
"Consonant Grapheme"      Phoneme                          Comment
"ꦲ"                      "/h/"    "also used as the zero consonant in word-initial contexts"  
"ꦤ"                      "/n/"                              ""
"ꦕ"                      "/c/"                              ""
"ꦫ"                      "/r/"                              ""
"ꦏ"                     "/k/"                               ""
"ꦢ"                     "/d̪/"                               ""
"ꦠ"                     "/t̪/"                               ""
"ꦱ"                     "/s/"                                ""
"ꦮ"                     "/w/"                               ""
"ꦭ"                    "/l/"                                ""
"ꦥ"                     "/p/"                                ""
"ꦣ"                    "/ɖ/"                                ""
"ꦗ"                     "/ɟ/"                               ""
"ꦪ"                    "/j/"                                ""
"ꦚ"                   "/ɲ/"                                ""
"ꦩ"                    "/m/"                                ""
"ꦒ"                   "/ɡ/"                                ""
"ꦧ"                    "/b/"                               ""
"ꦛ"                    "/ʈ/"                                ""
"ꦔ"                    "/ŋ/"                                ""
"**Consonant Diacritic**"   ""      "these predominantly form the codas of their syllables"
ꦀ                     "/m/"                                 ""
ꦁ                      "/ŋ/"                                 ""
ꦂ                      "/r/"                                 ""
ꦃ                     "/h/"                                  ""
ꦽ                     "/rɜ/"                                 ""
ꦾ                     "/j/"                                 ""
ꦿ                      "/r/"                                ""
꧀                       "∅"                 "nullifies the vowel in its syllable"
"**Vowel Grapheme**"     ""                                 ""                  
"ꦄ"                  "/a/"                                 ""
"ꦆ"                  "/i/"                                 ""
"ꦈ"                  "/u/"                                 ""
"ꦌ"                  "/e/"                                 ""
"ꦎ"                  "/o/"                                 ""
"**Vowel Diacritic**"  ""    "In the absence of a vowel diacritic, /a/ is taken as the \'default\' vowel for a syllable"
ꦶ                    "/i/"                                 ""
ꦸ                    "/u/"                                 ""
ꦺ                   "/e/"                                 ""
ꦺ                  "/o/"                                  ""
ꦼ                    "/ɜ/"                                 ""
"**Newer Orthographic Variation (Aksara)**"    ""         ""
"ꦉ"                 "/rɜ/"                                ""
"ꦊ"                 "/lɜ/"                                ""
'), TRUE)

kable(alphabet, col.names = c("Consonant Grapheme", "Phoneme", "Comment"), align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Syllables take the format (CC)V(C) [@JvHorne1974, p. xiv].

# Lenition rules

* /h/ deleted intervocalically if between non-identical vowels [@JvRobson2014, p. xvi].
* word-final /k/ debuccalizes to [ʔ] (ibid.).
* Tense vowels become lax in closed syllables [@JvHorne1974, p. xi].

# Misc. Rules

* The following consonants do not occur word-finally [@JvSuharno1982, p. 6]:
    - Voiced stops (due to word-final devoicing)
    - Palatal and retroflex consonants
    - Glides
    - /h/
    - Any consonant cluster
* Conversely:
    - /e/ and /ɛ/ do not occur word-finally; /a/ only occurs word-finally in the word <ora> [@JvSuharno1982, p. 6].
    - Except in interjections, /ə/ never occurs word-finally or before /ʔ/ [@JvNothofer1975, p. 9].
    - /ʔ/ always appears word-finally, except in interjections where it can be intervocalic [@JvNothofer1975, p. 9].
    - /h/ only occurs word-initially in interjections or loanwords [@JvRobson2014, p. xvi].
    - /ʔ/ inserted before the suffix <-aké> or after the prefix <sa-> [@JvRobson2014, p. xvi].
    - Nasal prefixes added to many verb forms [@JvRobson2014, p. 12].

# References
