---
title: "Czech"
author: "Emily Strand"
bibliography: cs.bib
output: html_document
---

Last updated: 2019-08-29

# Background

**Language family**: Indo-European / Slavic / West / Czech-Slovak

* Speakers of Czech reside in a multitude of places included, but not limited to, the Czech Republic, Bosnia, Croatia, Slovenia, and Serbia.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial    Alveolar  Postalveolar  Palatal     Velar    Glottal
Stops                     "p  b"     "t  d"      ""           "c  ɟ"     "k  ɡ"     ""
Affricates                ""         "ts"        "tʃ"         ""         ""         ""
Fricatives                "f  v"     "s  z"      "ʃ  ʒ"       ""         "x"        "ɦ"
Nasals                    "m"        "n"         ""           "ɲ"        ""         ""
Trills                    ""         "r  r̝"      ""           ""         ""         ""
Approximants              ""         "l"         ""           "j"        ""         ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced. Both alveolar trills are voiced, but the phoneme on the right is pronounced with frication.", general_title = "")
```

## Vowels

* Vowel length is contrastive in Czech. Although each vowel has a long counterpart phonemically identical in terms of place, the long counterpart of /ɪ/ is /iː/. Long vowels are indicated by an acute accent (´). An overring (˚) over /u/ also indicates lengthening [@harkins_modern_1980, p. 7].  
* /eu̯/ and /au̯/ were introduced into the Czech language through loanwords; however, now days they are included in the vowel inventory [@bican_phonotactics_2011, p. 21; @simackova_czech_2012, p. 23].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High     "ɪ"          ""           "u"         
Mid      "ɛ"          ""           "o"
Low      ""           "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```  

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

diphthongs <- read.table(textConnection('
Diphthongs   
"/eu̯/, /au̯/, /ou̯/"
'), TRUE)

kable(diphthongs, align = 'c') %>%
  kable_styling("bordered") %>%
  row_spec(0, bold = TRUE)
```    

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme          Comment
"a"	           "/a/"             ""
"á"	           "/aː/"            ""
"b"	           "/b/"             ""
"c"	          "/ts/"             ""
"č"	          "/tʃ/"             ""
"d"	           "/d/"             ""
"ď"	           "/ɟ/"   "the uppercase form is ⟨Ď⟩"
"e"	           "/ɛ/"             ""
"é"	           "/ɛː/"            ""
"ě"          	 "/ɛ/"             ""
"f"	           "/f/"             ""
"g"	           "/ɡ/"             ""
"h"	           "/ɦ/"             ""
"i"	           "/ɪ/"             ""
"í"	           "/iː/"            ""
"j"	           "/j/"             ""
"k"	           "/k/"             ""
"l"	           "/l/"             ""
"m"	           "/m/"             ""
"n"	           "/n/"             ""
"ň"	           "/ɲ/"             ""
"o"	           "/o/"             ""
"ó"	           "/oː/"   "only used in loanwords"  
"p"	           "/p/"             ""
"q"	           "/kv/"   "only used in loanwords"  
"r"	           "/r/"             ""
"ř"	           "/r̝/"             ""
"s"	           "/s/"             ""
"š"	           "/ʃ/"             ""
"t"	           "/t/"             ""
"ť"	           "/c/"    "the uppercase form is ⟨Ť⟩"
"u"	           "/u/"             ""
"ú"	           "/uː/"            ""
"ů"	           "/uː/"            ""
"v"	           "/v/"     "only used in loanwords"  
"w"	           "/v/"     "only used in loanwords"  
"x"	           "/ks/"            ""
"y"	           "/ɪ/"             ""
"ý"	           "/iː/"            ""
"z"         	 "/z/"             ""
"ž"            "/ʒ/"             ""
**Digraphs**    ""              ""
"ch"	         "/x/"             ""
"au"          "/au̯/"             ""
"eu"          "/eu̯/"             ""
"ou"          "/ou̯/"             ""
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Czech prefers empty codas, but it's not obligatory in the language [@harkins_modern_1980, p. 9].  

# Lenition Rules

* Voiced word-final consonants generally neutralize [@harkins_modern_1980, p. 10].  
* /j/ may delete word-initially if preceding a consonant [@harkins_modern_1980, p. 14].  
* Voiceless consonants become voiced if followed by voiced consonants [@harkins_modern_1980, p. 10].  
* /ɦ/ is present in the language due to the lenition of /ɡ/, which is common within the Slavic languages [@sussex_slavic_2006, p. 143].  

# Misc. Rules

* Glottal epenthesis occurs preceding a vowel in word-initial positions [@harkins_modern_1980, p. 13].  
* Repetition of consonants is typically avoided in Czech [@harkins_modern_1980, p. 9].  
* The voiceless plosives are often aspirated, but it's not contrastive to their non-aspirated counterparts [@harkins_modern_1980, p. 5].

# References  
