---
title: "Asháninka"
author: "Emily Strand"
bibliography: cni.bib
output: html_document
---

Last updated: 2020-04-22

**COMPROMISED: conflation among placeless /N/, /m/, and /n/**

# Background

**Language Family:** Arawakan / Maipuran / Southern Maipuran / Campa

* Asháninka is spoken in Peru.

# Phonology

## Consonants

* Asháninka is said to have a placeless nasal phoneme /N/ that occurs in word-medial codas, assimilating in place to following stops and affricates [@Mihas2019, p. 3583; @Chen2019, p. 3]. In the rules, I transcribe ⟨m⟩ to /m/ and ⟨n⟩ to /n/, meaning there is conflation among /N/, /m/, and /n/, so the language is compromised.
* Accounts of the phonemic inventory somewhat differ, seemingly rooted in the overlap with the closely related language Ashéninka [@Mihas2019, p. 3583]. I have chosen to follow the analysis done by @Chen2019 (p. 5).

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Bilabial  Alveolar  Postalveolar  Palatal  Velar   Glottal
Stops                        "p"       "t"         ""          ""      "k"      ""
Affricates                   ""        "ts"       "tʃ"         ""      ""       ""
Fricatives                   "β"       "s"        "ʃ"          ""      ""       "h"
Nasals                       "m"       "n"         ""          ""      ""       ""
Flaps                        ""        "ɾ"         ""          ""      ""       ""
Approximants                 ""        ""          ""          "j"     ""       ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Bilabial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: Asháninka also has a placeless /N/ (described above the chart).", general_title = "")
```  

## Vowels

* Vowel length is contrastive in Asháninka [@Chen2019, pp. 4-5]. Long vowels are indicated by duplicate graphemes.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
           Front      Central       Back
High        "i"         ""           ""
Mid         "e"         ""           "o"
Low         ""          "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)


diphthongs <- read.table(textConnection('
Diphthongs
"/ai/, /ao/, /oi/, /ei/, /io/, /ea/"
'), TRUE)

kable(diphthongs, align = 'c') %>%
  kable_styling("bordered")

```  

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme         
"a"              "/a/"            
"c"              "/k/"            
"e"              "/e/"            
"i"              "/i/"            
"k"              "/k/"            
"j"              "/h/"            
"m"              "/m/"            
"n"              "/n/"              
"o"              "/o/"            
"p"              "/p/"            
"r"              "/ɾ/"            
"s"              "/s/"            
"t"              "/t/"            
"v"              "/β/"
"y"              "/j/"            
**Multigraph**    ""              
"ch"             "/tʃ/"           
"sh"             "/ʃ/"            
"qu"             "/k/"            
"ts"             "/ts/"           
'), TRUE)

kable(alphabet, align = "c") %>%
  kable_styling("bordered")
```

# Lenition Rules

* Voiceless stops may become voiced following nasals and intervocalically [@Buttner1991, p. 57; @Mihas2019, p. 3584].

# Misc. Rules

* /n/ assimilates in place preceding /k/ [@Buttner1991, p. 57].
* /t/, /n/, and /s/ may palatalize preceding front vowels [ibid.; @Chen2019, p. 8].

# References
