---
title: "Kalaallisut"
author: "Bill Mizgerd"
bibliography: kl.bib
output: html_document
---

Last Updated: 2020-06-30

# Background

**Language Family**: Eskimo-Aleut / Eskimo / Inuit

* Kalaallisut is the autonym of the Greenlandic language; it is also used specifically to refer to the West Greenlandic dialect of the language. Therefore, where dialects diverge, I intend to follow the phonology of West Greenlandic.
* Kalaallisut is spoken in Greenland, but also in Denmark.

# Phonology

## Consonants

* @KlRischel1974 describes /ɴ/ as being of a "marginal status" (p. 22).
* @KlRischel1974 posits /h/ as appearing in some interjections, but not normally being contrastive (p. 23).
* /l/ is sometimes rendered as [ʟ] [@KlRischel1974, p. 246].
* [ts] is the only non-geminate consonant cluster to be attested in modern Greenlandic [@KlCollis1990, p. 311].
* /ʃ/ is attested in some dialects, but because of its limited usage and lack of orthographic attestation [@KlCollis1990, p. 301], I have opted to omit it.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial     Alveolar    Palatal     Velar     Uvular
Stops                       "p"         "t"         ""         "k"       "q"
Fricatives                  "v"         "s"         ""         "ɣ"       "ʁ"
Nasals                      "m"         "n"         ""         "ŋ"       "ɴ"
Approximants                ""          "l"         "j"        ""        ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Palatal", "Velar", "Uvular"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 5)) %>%
  column_spec(1, bold = TRUE)
```  

## Vowels

* Vowel length is contrastive [@KlCollis1990, p. 301]. Long vowels are indicated by duplicate graphemes.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         ""           "u"
Low        ""         "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```    

# Alphabet

* Geminated consonants are written with double letters [@KlCollis1990, p. 301].
    - Geminated /ŋ/ is written ⟨nng⟩ [@KlCollis1990, p. 317].
* ⟨e⟩ and ⟨o⟩ were used more widely in earlier orthographies, but in the modern orthography they only survive in pre-uvular contexts [@KlCollis1990, p. 301].
* ⟨f⟩ is sometimes used to represent /v/.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme            Comment
"a"            "/a/"               ""
"e"            "/i/"    "only appears before uvulars"  
"g"            "/ɣ/"               ""
"i"            "/i/"               ""
"j"            "/j/"               ""
"k"            "/k/"               ""
"l"            "/l/"               ""
"m"            "/m/"               ""
"n"            "/n/"               ""
"o"            "/u/"     "only appears before uvulars"
"p"            "/p/"               ""
"q"            "/q/"               ""
"r"            "/ʁ/"               ""
"s"            "/s/"               ""
"t"            "/t/"               ""
"u"            "/u/"               ""
"v"            "/v/"               ""
**Digraph**     ""                 ""
"ng"           "/ŋ/"               ""   
"rn"           "/ɴ/"               ""
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* Stem-final uvulars delete when followed by /ɣ/ or /ʁ/ [@KlRischel1974, p. 36].
* Long consonants shorten after long vowels, especially for speakers in southern dialect areas [@KlRischel1974, p. 59].
* Intervocalic /q/ is often closed incompletely, leading to some degree of spirantization [@KlRischel1974, p. 131].

# Misc. Rules

* In the diphthongs /ai/ and /au/, the second vowel assimilates to yield [aa]; however, /ai/ is preserved word-finally [@KlRischel1974, p. 32].
* In consonant clusters, the first fully assimilates to the second, yielding a geminate [@KlRischel1974, p. 35].
* Vowels pharyngealize before uvular consonants [@KlRischel1974, p. 38].
* /ŋ/ nasalizes preceding vowels, sometimes at the expense of its own realization; this process is more advanced in East Greenlandic than in other dialects [@KlRischel1974, p. 132].
* In some dialects, /ɣ/ is replaced entirely by /ŋ/; geminates of /ɣ/ tend to collapse instead to [kk] [@KlRischel1974, p. 167].
* Suffix-initial consonants alternate between stops and fricatives; they are stops when preceded by a consonant, fricatives when preceded by a vowel [@KlRischel1974, p. 242].
* Intervocalic, stem-internal single fricatives tend to despirantize when they become geminates [@KlRischel1974, p. 242].
* For the aforementioned stop/fricative alternations, the fricative seems to be the underlying form [@KlRischel1974, p. 249].
* /t/ becomes an affricate when followed by /i/ [@KlCollis1990, p. 300].

# References
