---
title: "Arabela"
author: "Emily Strand"
bibliography: arl.bib
output: html_document
---

Last Updated: 2020-06-26

# Background

**Language family**: Zaparoan

* Also referred to as Chiripuno, Arabela is spoken in Peru, specifically in two villages of the Napo tributary.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"    Bilabial    Alveolar    Postalveolar    Palatal  Velar    Glottal
Stops                         "p"          "t"           ""           ""      "k"       ""
Fricatives                     ""          "s"           "ʃ"          ""      ""        "h"
Nasals                        "m"          "n"           ""           ""      ""        ""
Trills                        ""           "r"           ""           ""      ""        ""
Approximants                  "w"           ""           ""           "j"     ""        ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Bilabial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* Vowel length is phonemic [@Carvalho2013, p. 95]. Long vowels are represented by duplicate vowel graphemes.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"          ""          "u"
Mid       "e"          ""          "o"
Low       ""          "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme
"a"           "/a/"  
"c"           "/k/"  
"e"           "/e/"   
"i"           "/i/"  
"j"           "/h/"  
"m"           "/m/"  
"n"           "/n/"  
"o"           "/o/"  
"p"           "/p/"  
"r"           "/r/"  
"s"           "/s/"  
"t"           "/t/"  
"u"           "/u/"  
"y"           "/j/"  
**Digraph**   ""
"qu"          "/k/"  
"hu"          "/w/"  
"sh"          "/ʃ/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```  

# Syllable Structure

* The syllables permitted in Arabela follow the order of CV and CCV [@Topintzi2016, p. 84].


# Lenition Rules

* /t/ often becomes a flap [ɾ] [@Rich1999, p. 14].
* /k/ often spirantizes or becomes voiced [@Rich1963, pp. 194, 195].
* Vowels tend to be deleted between homorganic consonants [@Rich1999, p. 14].

# Misc. Rules

* /n/ and /r/ occur in free variation in some instances [@Rich1963, p. 194].
* /t/, /n/, /ʃ/, /r/, and /k/ geminate allophonically between homorganic consonants [@Rich1963, p. 195].
* /w/ and /j/ are nasalized following nasals or nasal-vowel sequences [@Rich1963, pp. 196, 197].
* Vowels are nasalized following nasals and /h/ (most often) [@Rich1963, p. 197].

# References
