---
title: "Oriya"
author: "Becky Mathew"
edited by: "Emily Strand"
bibliography: or.bib
output: html_document
---

Last Updated: 2020-02-25

**COMPROMISED: diacritics not always marked and used interchangeably**

# Background

**Language Family:** Indo-European / Indo-Iranian / Indo-Aryan / Eastern zone / Oriya

* Oriya (now Romanized as Odia) is the official language of Orissa, an eastern Indian state. There are four major dialects of Oriya, and this project focuses on Standard Oriya [@Ray2003, p. 485].

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Bilabial    Dental    Alveolar   Retroflex    Palatal   Velar    Glottal
"Stops (plain)"             "p b"      "t̪ d̪"       ""         "ʈ ɖ"       "c ɟ"    "k ɡ"      ""
"Stops (aspirated)"        "pʰ bʱ"    "t̪ʰ d̪ʱ"      ""        "ʈʰ ɖʱ"     "cʰ ɟʱ"  "kʰ ɡʱ"     ""
Fricatives                   ""         ""         "s"         ""          ""       ""        "h"
Nasals                       "m"        "n̪"        ""          "ɳ"         ""       "ŋ"       ""
Flaps                        ""         ""         "ɾ"         ""          ""       ""        ""
Approximants                 "w"        ""         "l"         "ɭ"         "j"      ""        ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Bilabial", "Dental", "Alveolar", "Retroflex", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 7)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```  

## Vowels

* Nasal vowels are contrastive in Oriya [@Ray2003, p. 448]; however, they don't always contrast with their oral counterparts [@Pattanayak1959 as cited in @Ray2003, p. 448]. /o/ does not have a nasal counterpart.
* Vowel length is said to be a phonetic feature rather than a phonemic one [@Ray2003, p. 447]. Long vowels will be transcribed as short.
* The two diphthongs generally associated with Indo-Aryan languages /ɔi/ and /ɔu/ (variants of /ai/ and /au/), may be interpreted as separate monophthongss so, I have chosen not to transcribe them as complex nuclei [@Ray2003, p. 447].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
           Front      Central       Back
High        "i"         ""           "u"
Mid         "e"         ""           "o"
Low-Mid     ""          ""           "ɔ"
Low         ""          "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)

```

# Alphabet

* Oriya descends from the Brahmi script, but it does not hang from a horizontal line. It is a left-to-right abugida script without capital letters. As an abugida script, every consonant carries an inherent /ɔ/ following it [@Ray2003, p. 487].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
"Consonant Grapheme"        Phoneme         Comment
"କ"                         "/kɔ/"            ""
"ଖ"                         "/kʰɔ/"           ""
"ଗ"                         "/ɡɔ/"            ""
"ଘ"                         "/ɡʱɔ/"           ""
"ଙ"                         "/ŋɔ/"            ""
"ଚ"                         "/cɔ/"            ""
"ଛ"                         "/cʰɔ/"           ""
"ଜ; ଯ"                      "/ɟɔ/"            ""
"ଝ"                         "/ɟʱɔ/"           ""
"ଟ"                         "/ʈɔ/"            ""
"ଠ"                         "/ʈʰɔ/"           ""
"ଡ"                         "/ɖɔ/"            ""
"ଢ"                         "/ɖʱɔ/"           ""
"ଣ"                         "/ɳɔ/"            ""
"ତ"                         "/t̪ɔ/"            ""
"ଥ"                         "/t̪ʰɔ/"           ""
"ଦ"                         "/d̪ɔ/"            ""
"ଧ"                         "/d̪ʱɔ/"           ""
"ନ; ଞ"                      "/n̪ɔ/"            ""
"ପ"                         "/pɔ/"            ""
"ଫ"                         "/pʰɔ/"           ""
"ବ"                          "/bɔ/"           ""
"ଭ"                          "/bʱɔ/"          ""
"ମ"                          "/mɔ/"           ""
"ର"                          "/ɾɔ/"           ""
"ଳ"                          "/ɭɔ/"           ""
"ୱ; ଵ"                       "/wɔ/"           ""
"ଶ; ଷ; ସ"                   "/sɔ/"            ""
"ହ"                          "/hɔ/"           ""
"ୟ"                          "/jɔ/"           ""
"ଲ"                          "/lɔ/"           ""
"**Independent Grapheme**"     ""      "vowels and syllabic consonants"
"ଅ"                           "/ɔ/"           ""
"ଆ"                           "/a/"          ""
"ଇ; ଈ"                        "/i/"          ""
"ଉ; ଊ"                        "/u/"          ""
"ଋ; ୠ"                        "/ɾu/"          ""
"ଌ; ୡ"                         "/lu/"          ""
"ଏ"                            "/e/"          ""
"ଐ"                           "/ɔi/"          ""
"ଓ"                            "/o/"          ""
"ଔ"                           "/ɔu/"          ""
**Diacritic**                   ""            ""
"ା"                            "/a/"          ""
"ି; ୀ"                         "/i/"          ""
"ୁ; ୂ"                         "/u/"          ""
"ୃ; ୄ"                         "/ɾu/"          ""
"ୢ; ୣ"                         "/lu/"         ""
"େ"                           "/e/"          ""
"ୈ"                           "/ɔi/"         ""
"ୋ"                           "/o/"         ""
"ୌ"                           "/ɔu/"        ""
"ଁ"                              ""       "called a candrabindu, this diacritic indicates vowel nasalization [@Mahapatra1996, p. 405]; however, @Giri2011 states that it may be used in place of the anusvara, which represents a separate phonemic feature, compromising the language. It also denotes nasalization of preceding consonants [@NBGP2018, p. 5], but given that nasal consonants aren\'t contrastive, the diacritic in these instances will be removed in the rules."
"ଂ"                            "/m/"     "called an anusvara, this diacritic represents a nasal, either word-finally or in homorganic sequences (/m/ appears to be underlying) [@Mahapatra1996, p. 404; @Mohapatra2016, p. 23]"
"ଃ"                            "/h/"     "called a visarga, this diacritic represents a word-final /h/ [@Mohapatra2016, p. 23]"
"୍"                              ""       "called a halanta, this diacritic removes the inherent vowel from a consonant [@Mahapatra1996, p. 404]; however, it is not always marked, which compromises the language [@Ray2003, p. 447]"
"◌़"                              ""      "called a nukta, this diacritic indicates non-native sounds [@Ishida2002, p. 4]. Given this, the diacritic will not be accounted for in the rules."
'), TRUE)

kable(alphabet, col.names = c("Consonant Grapheme", "Phoneme", "Comment"), align = "c") %>%
  kable_styling("bordered")
```

# Lenition Rules  

* Consonant gemination occurs in Odia [@Ray2003, pp. 449-450].
    - For many speakers, duplicate consonants undergo degemination [@Ray2003].
* /ɖ/ and /ɖʱ/ may become flaps intervocalically and word-finally [@Ray2003, p. 449].

# Misc. Rules

* Aspirated consonants are de-aspirated when preceded by /s/ and followed by a vowel [@Ray2003, p. 449].

# References
