---
title: "Abau"
author: "Allison Shao"
bibliography: aau.bib
output: html_document
---

Last Updated: 2020-08-04

# Background

**Language Family:** Sepik-Ramu / Sepik / Upper Sepik / Abau

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial   Alveolar   Palatal   Velar   Glottal
Stops                       "p"        ""        ""       "k"      ""
Fricatives                  ""         "s"       ""       ""       "h"
Nasals                      "m"        "n"       ""       ""       ""
Flaps                       ""         "ɾ"       ""       ""       ""
Approximants                "w"        ""        "j"      ""       ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 5)) %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* @Bailey1975 proposes a six-vowel system with additional vowel /ɒ/ (p. 8). However, @Lock2007 states that this back vowel is not supported by the evidence (p. 14).
* Some analyses of Abau include a seven-vowel system with additional vowels /ɪ/ and /ʊ/. However, @Lock2011 believes that they are allophones of sequences /ij/ and /uw/, respectively (p. 17).
    - @Foley2018 describes Lock's remarks as "phonetically quite odd" and concludes that the debate over the Abau vowel system is not yet resolved (pp. 288-289). Nevertheless, I have chosen not to include /ɪ/ and /ʊ/ below.
* @Bailey1975 claims that Abau nouns exhibit lexical tone (p. 33). However, Lock refutes this observation, stating that there is no evidence of tone [@Lock2007, p. 21; @Lock2011, p. 20].


```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         ""           "u"         
Mid       "e"         ""           "o"
Low       ""          "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```  

# Alphabet

* /ɾ/ is represented by ⟨l⟩ word-initially, ⟨d⟩ when preceding /n/, and ⟨r⟩ elsewhere [@Lock2011, p. 22].
* Word-initial vowel sequences /iV/ and /uV/ are represented orthographically with a preceding ⟨y⟩ and ⟨w⟩, respectively (ibid.).
    - E.g. /ia-/ = ⟨yia-⟩
    - This only applies when the second vowel is not high.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme          Comment
"a"              "/a/"             ""
"d"              "/ɾ/"             "preceding /n/"
"e"              "/e/"             ""
"h"              "/h/"             ""
"i"              "/i/"             ""
"k"              "/k/"             ""
"l"              "/ɾ/"             "word-initially"
"m"              "/m/"             ""
"n"              "/n/"             ""
"o"              "/o/"             ""
"p"              "/p/"             ""
"r"              "/ɾ/"             ""
"s"              "/s/"             ""
"u"              "/u/"             ""
"w"              "/w/"             ""
"y"              "/j/"             ""
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* /p/ varies freely with [b] [@Lock2007, p 8].

# Misc. Rules

* /ɾ/ has several different realizations [@Lock2007, pp. 8-9]:
    - \[d]: adjacent to /i/ and /j/; preceding non-back vowels; preceding non-velar and non-fricative consonants
    - \[t]: preceding /s/
    - \[ɺ]: preceding non-back vowels
* /h/ also varies in realization [@Lock2007, p. 11]:
    - \[ɸ]: word-finally; adjacent to /i/ and /j/; generally preceding consonants
    - \[m̊] or [n̊]: preceding the respective voiced nasal counterpart
* Across syllable boundaries, /n/ assimilates in place when preceding /k/ [@Lock2007, p. 10].

# References
