---
title: "Tz'utujil"
author: "Emily Strand"
bibliography: tzj.bib
output: html_document
---

Last updated: 2020-02-28

**COMPROMISED: suspect marking of glottal stops; uncertainty surrounding the orthography (the Crúbadán corpus includes symbols not present in the alphabet)**

# Background

**Language Family:** Mayan / Quichean-Mamean / Greater Quichean / Quichean / Tzutujil

* Tz'utujil is spoken in Guatemala.

# Phonology

## Consonants

* /b/, /d/, and /ɡ/ appear in loanwords [@Dayley1985, p. 22].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial  Alveolar Postalveolar  Palatal  Velar  Uvular Glottal
Stops                      "p pʼ"  "t tʼ"   ""            ""       "k kʼ" "q qʼ" "ʔ"
Affricates                 ""      "ts tsʼ" "tʃ tʃʼ"      ""       ""     ""     ""
Fricatives                 ""      "s"      "ʃ"           ""       ""     "χ"    ""
Nasals                     "m"     "n"      ""            ""       ""     ""     ""
Trills                     ""      "r"      ""            ""       ""     ""     ""
Approximants               "w"     "l"      ""            "j"      ""     ""     ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Uvular", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 7)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are ejective.", general_title = "")
```

## Vowels

* Vowel length is contrastive in Tz'utujil [@Dayley1985, p. 25]. Long vowels are represented by duplicate graphemes.
* Diphthongs /ie/ and /uo/ are said to be present in the Santiago Atitlan dialect [@Dayley1985, p. 27].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High     "i"          ""           "u"         
Mid      "e"          ""           "o"
Low      ""           "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```  

# Alphabet

* **Note:** /tʼ/ is not represented in the Crúbadán corpus.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme
"a"            "/a/"
"b\'"          "/pʼ/"
"d\'"          "/tʼ/"
"e"            "/e/"
"i"            "/i/"
"j"            "/χ/"
"k"            "/k/"
"k\'"          "/kʼ/"
"l"            "/l/"
"m"            "/m/"
"n"            "/n/"
"o"            "/o/"
"p"            "/p/"
"q"            "/q/"
"q\'"          "/qʼ/"
"r"            "/r/"
"s"            "/s/"
"t"            "/t/"
"u"            "/u/"
"w"            "/w/"
"x"            "/ʃ/"
"y"            "/j/"
"\'"           "/ʔ/"
**Digraph**    ""
"ch"           "/tʃ/"
"ch\'"         "/tʃʼ/"
"tz"           "/ts/"
"tz\'"         "/tsʼ/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* /pʼ/ and /tʼ/ become voiced and implosive preceding vowels [@Dayley1985, p. 15]. Ejective /q/ may become voiced and implosive in the same environment, but it's optional.
* /χ/ is realized as [h] syllable-internally (i.e. following a vowel and preceding a consonant) [@Dayley1985, p. 19].

# Misc. Rules

* Word-initial vowels are often preceded by a phonetic glottal stop [@Dayley1985, pp. 17-18].
    - Some word-initial vowels are preceded by a phonemic glottal stop [@Dayley1985, p. 19], and it is unclear whether these are marked consistently, which compromises the language to some extent.
* /l/, /r/, /w/, /j/, /m/, and /n/ devoice word-finally [@Dayley1985, p. 20].
    - /m/ and /n/ are also voiceless preceding consonants.
    - Word-final /w/, /j/, /l/, and /r/ from Spanish loanwords contrast with the same consonants in native words, due to the fact that the Spanish phonemes don't undergo the same devoicing rule that the native phonemes do [@Dayley1985, p. 23]. This is only specific to the Santiago Atitlan dialect; in all other dialects, these Spanish phonemes fully assimilate to the native phonemes.
* /w/ is realized as [β] preceding front vowels [@Dayley1985, p. 21].
* /a/ varies freely with /e/ in several words [@Dayley1985, p. 26].

# References
