---
title: "Tajik"
author: "Bill Mizgerd"
bibliography: tg.bib
output: html_document
---

Last Updated: 2020-07-02

# Background

**Language Family:** Indo-European / Indo-Iranian / Iranian / Western / Southwestern / Persian

* Tajik is predominantly spoken in Tajikistan and Uzbekistan.

# Phonology

## Consonants

* @TgKhojayori2009 argue for /ɽ/ rather than /r/ (p. 4).

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial  Alveolar  Postalveolar  Palatal  Velar  Uvular  Glottal
Stops                      "p b"   "t d"     ""            ""       "k ɡ"  "q"     "ʔ"
Affricates                 ""      ""        "tʃ dʒ"       ""       ""     ""      ""
Fricatives                 "f v"   "s z"     "ʃ ʒ"         ""       ""     "χ ʁ"   "h"
Nasals                     "m"     "n"       ""            ""       ""     ""      ""
Trills                     ""      "r"       ""            ""       ""     ""      ""
Approximants               ""      "l"       ""            "j"      ""     ""      ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Uvular", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 7)) %>%
  footnote("Note: For phonemes that share a cell, those on the left are voiceless, whereas those on the right are voiced.", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* Some older sources describe /iː/ as a separate phoneme, but modern scholars tend to argue that length is wholly non-contrastive in Tajik [@TgIdo2005, p. 11].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
          Front   Central   Back
High      "i"     ""        "u"
Mid       "e"     "ɵ"       "ɔ"
Low       ""      "a"       ""
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme                  Comment
"а"              "/a/"                      ""
"б"              "/b/"                      ""
"в"              "/v/"                      ""
"г"              "/ɡ/"                      ""
"ғ"              "/ʁ/"                      ""
"д"              "/d/"                      ""
"е"            "/e/; /je/"   "/je/: word-initially [@TgBaizoyev2004, p. 2]"  
"ё"              "/jɔ/"                     ""
"ж"              "/ʒ/"                      ""
"з"              "/z/"                      ""
"и"              "/i/"                      ""
"ӣ"              "/i/"        "word-finally [@TgBaizoyev2004, p. 2]"  
"й"              "/j/"                      ""
"к"              "/k/"                      ""
"қ"              "/q/"                      ""
"л"              "/l/"                      ""
"м"              "/m/"                      ""
"н"              "/n/"                      ""
"о"              "/ɔ/"                      ""
"п"              "/p/"                      ""
"р"              "/r/"                      ""
"с"              "/s/"                      ""
"т"              "/t/"                      ""
"у"              "/u/"                      ""
"ӯ"              "/ɵ/"                      ""
"ф"              "/f/"                      ""
"х"              "/χ/"                      ""
"ҳ"              "/h/"                      ""
"ч"              "/tʃ/"                     ""
"ҷ"	             "/dʒ/"                     ""
"ш"              "/ʃ/"                      ""
"ъ"              "/ʔ/"                      ""
"э"              "/e/"    "used word-initially where initial glides are absent [@TgBaizoyev2004, p. 2]"  
"ю"              "/ju/"                     ""        
"я"              "/ja/"                     ""
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```  

# Syllable Structure

* Tajik has a (C)V(C)(C) syllable structure [@TgIdo2005, p. 14].

# Lenition Rules

* Loanwords with the affricate /ts/ typically reduce it to /s/ [@TgIdo2005, p. 13].
* /q/ spirantizes in some words (ibid.).
* The middle consonant is often deleted in three-consonant clusters [@TgKhojayori2009, p. 7].
  * However, sometimes [ə] is inserted between the second and third consonant instead (ibid.).
* In word-final consonant clusters, the last consonant is sometimes deleted (ibid.).
* /ʔ/ deletes intervocalically; /v/ does as well, if the vowels are identical [@TgKhojayori2009, p. 8].
* Word-final /h/ deletes (ibid.).

# Misc. Rules

* Obstruents devoice word-finally [@TgKhojayori2009, p. 3].
* Consonant clusters assimilate to the voicing of the final one [@TgKhojayori2009, p. 3].
* /a u i/ reduce to [ə] in unstressed syllables [@TgKhojayori2009, p. 6].
* /n/ assimilates to the place of following bilabials [@TgIdo2005, p. 13; @TgKhojayori2009, p. 7].
* Glides, typically /j/, are often added between vowel clusters [@TgKhojayori2009, p. 8; @TgIdo2005, p. 12].

# References
