---
title: "Haitian Creole"
author: "Justin Bai"
bibliography: ht.bib
output: html_document
---

Last Updated: 2019-12-19

# Background

**Language family**: French Creole

* Speakers of Haitian Creole reside in a multitude of places, including, but not limited to, Haiti, Canada, the United States, France, the Dominican Republic, Cuba, and the Bahamas.

# Phonology

## Consonants

* /ɥ/ is a widely debated phoneme in Haitian Creole, as it only appears in succession with /i/ and can appear in free variation with /w/ [@Kraemer2016, p. 67; @Pindziak2012, pp. 3-4]. I have chosen to treat it as a variant of /w/ and not include it in the phoneme inventory.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial     Alveolar   Postalveolar   Palatal   Velar      Glottal
Stops                      "p  b"     "t  d"         ""           ""      "k  ɡ"       ""
Fricatives                 "f  v"     "s  z"       "ʃ  ʒ"         ""       "ɣ"         "h"  
Nasals                       "m"        "n"          ""           ""       "ŋ"         ""
Approximants                 "w"        "l"          ""           "j"      ""          ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```

## Vowels

* Haitian Creole makes a distinction between oral and nasal vowels. Nasal vowels are orthographically represented with ⟨n⟩ after the vowel [@faraclas2010orthography, p. 97].
* Nasal vowels may also be indicated when preceding two ⟨n⟩s [@faraclas2010orthography, p. 101].
* /an/, /en/, /on/, and /un/ appear as vowel-consonant sequences when followed by a vowel.
* The grave accent (\`) on vowels is also used to indicate non-nasality of vowels when followed by ⟨n⟩ [@faraclas2010orthography, p. 101].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         ""         "u  ũ"
Mid       "e"         ""         "o  õ"
""      "ɛ  ɛ̃"        ""          "ɔ"
Low       ""        "a  ã"         ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```  

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme  
"a"           "/a/"
"b"           "/b/"
"d"           "/d/"
"e"           "/e/"
"è"           "/ɛ/"
"f"           "/f/"
"g"           "/ɡ/"
"h"           "/h/"
"i"           "/i/"
"j"           "/ʒ/"
"k"           "/k/"
"l"           "/l/"
"m"           "/m/"
"n"           "/n/"
"o"           "/o/"
"ò"           "/ɔ/"
"p"           "/p/"
"r"           "/ɣ/"
"s"           "/s/"
"t"           "/t/"
"v"           "/v/"
"w"           "/w/"
"y"           "/j/"
"z"           "/z/"
**Digraph**    ""
"ch"          "/ʃ/"
"ng"          "/ŋ/"
"ou"          "/u/"
"ui"          "/wi/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* /j/ deletes word-finally following /i/ [@Gurevich2004, p. 112].
* /ʒ/ becomes /j/ word-finally (ibid.).

# References
