---
title: "Kabardian"
author: "Emily Strand"
bibliography: kbd.bib
output: html_document
---

Last Updated: 2019-10-18

# Background

 **Language Family**: North Caucasian / Northwest / Circassian

* Kabardian is spoken in Western Russia, specifically in the Kabardino-Balkar republic, North Ossetia-Alania republic, and Stavropol Krai.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
 "Manner of Articulation"   Labial       Alveolar      Alveopalatal     Postalveolar   Palatal      Velar            Uvular         Pharyngeal    Glottal
 Stops                    "p  pʼ  b"     "tʼ  d"            ""               ""          ""      "kʷ  kʷʼ  ɡʷ"  "q  qʷ  qʼ  qʷʼ"        ""        "ʔ  ʔʷ"
 Affricates                   ""       "ts  tsʼ  dz"        ""         "tʃ  tʃʼ  dʒ"     ""          ""                ""               ""          ""
 Fricatives               "f  fʼ  v"     "s  z"         "ɕ  ɕʼ  ʑ"        "ʃ  ʒ"         ""        "xʷ  ɣ"       "χ  χʷ  ʁ  ʁʷ"         "ħ"         "h"   
 Nasals                      "m"           "n"              ""               ""          ""          ""                ""               ""          ""
 Trills                      ""            "r"              ""               ""          ""          ""                ""               ""          ""
 Approximants                ""        "ɬ  ɬʼ  ɮ"           ""               ""          "j"         ""                ""               ""          ""
 '), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Alveopalatal", "Postalveolar", "Palatal", "Velar", "Uvular", "Pharyngeal", "Glottal"), align = 'c') %>%
 kable_styling("bordered") %>%
 add_header_above(c("", "Place of Articulation" = 9)) %>%
 column_spec(1, bold = TRUE) %>%
 footnote(general = "Note: Phonemes that have the diacritic /ʼ/ are ejectives; phonemes that have the diacritic /ʷ/ are labialized; phonemes that have /ʷʼ/ are labialized ejectives. The alveolar approximants are lateral fricatives.", general_title = "")
 ```

## Vowels

* Vowel height is the only contrastive feature among Kabardian vowels (backness is not phonemic) [@gordon_phonetic_2006, pg. 28].  
* There is some debate as to whether or not /aː/ is a phoneme in Kabardian, however, @gordon_phonetic_2006 explains that it's not entirely predictable (pg. 28). I will treat it as a phoneme.   
* A lot of free variation occurs with the production of vowels, so many of the vowels that are phonemic in other languages do arise in Kabardian [@pittman_phoneme_1963, pg. 22].  
* /ə/ can also be epenthetic given certain environments [@colarusso_grammar_1992, pg. 18].  

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
          Central  
 High      ""      
 Mid       "ə"     
 Low     "a  aː"   
 '), TRUE)

kable(vowels, align = 'c') %>%
 kable_styling("bordered") %>%
 column_spec(1, bold = TRUE) %>%
 footnote(general = "Note: The low central vowel on the right is long.", general_title = "")
 ```

# Alphabet

* [e] often appears to represent the grapheme ⟨e⟩; however, it's realized as an allophone of /a/. I will use /a/ for the transcription of ⟨e⟩.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
 Grapheme      Phoneme                                         Comment        
 "а"           "/aː/"                                            ""
 "э"           "/a/"                                             ""
 "б"           "/b/"                                             ""
 "в"           "/v/"                                             ""
 "г"           "/ɣ/"                                             ""
 "д"           "/d/"                                             ""
 "е"           "/a/"                                             ""
 "ё"          "/jo/"                               "only used in Russian loanwords"         
 "ж"           "/ʒ/"                                             ""
 "з"           "/z/"                                             ""
 "и"          "/jə/"                                             ""
 "й"          "/j/"                                              ""
 "к"          "/k/"                                              ""
 "л"           "/ɮ/"                                             ""
 "м"           "/m/"                                             ""
 "н"          "/n/"                                              ""
 "о"         "/aw/"                                         "rarely used"       
 "п"          "/p/"                                              ""
 "р"          "/r/"                                              ""
 "с"          "/s/"                                              ""
 "т"          "/t/"                                              ""
 "у"          "/w/"                                   "indicates labialization"      
 "ф"          "/f/"                                              ""
 "х"          "/x/"                                              ""
 "ц"         "/ts/"                                              ""
 "ч"         "/tʃ/"                                              ""
 "ш"          "/ʃ/"                                              ""
 "щ"          "/ɕ/"                                              ""                 
 "ъ"           ""           "used in combination to indicate uvularization (otherwise used in Russian loanwords)"        
 "ы"          "/ə/"                                              ""      
 "ь"           ""           "used in combination to indicate palatalization (otherwise used in Russian loanwords)"       
 "ю"         "/ju/"                                              ""
 "я"         "/jaː/"                                             ""
 "ӏ"         "/ʔ/"                             "indicates ejectivization in combination"      
 **Multigraph**   ""                                             ""  
 "гу"        "/ɡʷ/"                                              ""        
 "гъ"        "/ʁ/"                                               ""
 "гъу"      "/ʁʷ/"                                               ""
 "дж"       "/dʒ/"                                               ""
 "дз"       "/dz/"                                               ""
 "жь"        "/ʑ/"                                               ""
 "ку"       "/kʷ/"                                               ""
 "къ"       "/q/"                                                ""
 "къу"      "/qʷ/"                                               ""
 "кхъ"      "/qχ/"                                               ""
 "кхъу"     "/qχʷ/"                                              ""
 "кӏ"       "/tʃʼ/"                                              ""
 "кӏу"      "/kʷʼ/"                                              ""
 "лъ"        "/ɬ/"                                               ""
 "лӏ"       "/ɬʼ/"                                               ""
 "пӏ"       "/pʼ/"                                               ""
 "тӏ"       "/tʼ/"                                               ""
 "фӏ"       "/fʼ/"                                               ""
 "ху"       "/xʷ/"                                               ""
 "хъ"       "/χ/"                                                ""
 "хъу"      "/χʷ/"                                               ""
 "хь"       "/ħ/"                                                ""
 "цӏ"       "/tsʼ/"                                              ""
 "чӏ"      "/tʃʼ/"                                               ""
 "Щӏ"       "/ɕʼ/"                                               ""
 "ӏу"      "/ʔʷ/"                                                ""
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
 ```

# Lenition Rules

* /ɕ'/, /ts'/, and /tʃ'/ are sometimes voiced in the word-medially [@pittman_phoneme_1963, pg.19]

# References
