---
title: "Georgian"
author: "Maddie Critz"
bibliography: ka.bib
output: html_document
---

Last updated: 2019-10-31

# Background

**Language Family:** Kartvelian / Karto-Zan / Georgian

* Georgian is spoken throughout the country of Georgia.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"      Labial      Alveolar    Postalveolar       Velar       Uvular   Glottal   
Stops                      "pʰ  pʼ  b"  "tʰ  tʼ  d"       ""          "kʰ  kʼ  ɡ"     "q’"      ""
Affricates                     ""      "ts  tsʼ  dz" "tʃ  tʃʼ  dʒ"        ""          ""        ""
Fricatives                     "v"        "s  z"        "ʃ  ʒ"           "x  ɣ"       ""        "h"
Nasals                         "m"         "n"            ""              ""          ""        ""
Flaps                          ""          "ɾ"            ""              ""          ""        ""
Approximants                   ""          "l"            ""              ""          ""        ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Velar", "Uvular", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: Phonemes that have the diacritic (ʰ) are aspirated and phonemes that have the diacritic (’) ejective.", general_title = "")
```

## Vowels

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Back
High      "i"       "u"
Mid       "ɛ"       "ɔ"
Low        ""       "ɑ"
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```  

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme  
"ა"		         "/ɑ/"
"ბ"		         "/b/"
"გ"		         "/ɡ/"
"დ"		        "/d/"
"ე"		         "/ɛ/"
"ვ"		         "/v/"
"ზ"		        "/z/"
"თ"		        "/tʰ/"
"ი"	           "/i/"
"კ"	          "/kʼ/"
"ლ"	          "/l/"
"მ"	          "/m/"
"ნ"		        "/n/"
"ო"		        "/ɔ/"
"პ"		        "/pʼ/"
"ჟ"		        "/ʒ/"
"რ"		        "/ɾ/"
"ს"		        "/s/"
"ტ"		       "/tʼ/"
"უ"		        "/u/"
"ფ"		       "/pʰ/"
"ქ"		       "/kʰ/"
"ღ"		       "/ɣ/"
"ყ"		       "/qʼ/"
"შ"		        "/ʃ/"
"ჩ"		       "/tʃ/"
"ც"		       "/ts/"
"ძ"		       "/dz/"
"წ"		       "/tsʼ/"
"ჭ"		       "/tʃʼ/"
"ხ"		       "/x/"
"ჯ"	         "/dʒ/"
"ჰ"		        "/h/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* /v/ becomes a glide [w] following consonants and in coda positions [@hewitt1995, p. 21].
* /q/ is often pronounced with frication (ibid.).

# Misc. Rules

* Word-final voiced stops are realized as their aspirated counterparts [@aronson_kiziria_1999, p. 15].

# References
