---
title: "Naro"
author: "Bill Mizgerd"
bibliography: nhr.bib
output: html_document
---

Last Updated: 2019-07-02

# Background

**Language Family**: Khoe-Kwadi / Khoe / Kalahari Khoe / Southwest / Naro

* Also spelled Nharo and Naron, it is spoken in the Ghanzi district of Botswana.

# Phonology

## Consonants (Pulmonic)

* /p/ is very rare in native words, so much so that @NhrKagaya1978 argues that it only appears in loans [@NhrKagaya1978, p. 34; @NhrVisser1998, p. 119].
* Alveolar affricates and fricatives are produced in free variation with their postalveolar counterparts [@NhrKagaya1978, pp. 36-37].
* Loanwords also include /f/ and /l/ [@NhrVisser1998, p. 120].
* /x/ and /h/ only contrast word-initially [@NhrKagaya1978, p. 37].
* /m/ is the only consonant that can occur root-finally [@NhrVisser1998, p. 123].
* @NhrKagaya1978 diverges from Visser's analysis (which I have followed) in several places:
    - He finds /s/ to be an allophone of /ʃ/.
    - He identifies the aspirated stops (as well as /t'/) as allophones of the plain voiceless stops. (Largely based on frequency, per @NhrKagaya1978 (pp. 35-36), although he does find /tʰ/ and /t/ to exist in complementary distribution.)
    - He does not see the velar-released stops as occurring phonemically.
    - He identifies /h/ as an allophone of /x/.
    - He sees syllabic nasals for /m/, /n/, and /ŋ/ as existing contrastively [@NhrKagaya1978, p. 39].
    
    **(All aforementioned bullet points on @NhrKagaya1978, p. 34, except where mentioned)**
* The position of /w/ and /j/ is unclear; they only occur intervocalically, so their realization may be just an epenthetic process rather than the production of a distinct phoneme [@NhrKagaya1978, pp. 37-38].

```{r echo = FALSE, message = FALSE, warning = FALSE, results = 'asis'}
library(dplyr)
library(knitr)
library(kableExtra)
consonants <- read.table(textConnection('
"Manner of Articulation"               Labial     Alveolar         Palatal    Velar       Glottal
"Stops (normal release)"                "p pʰ b"   "t tʰ tʼ d"      ""         "k kʰ ɡ"    ""
"Stops (velar fricative release)"       ""         "tx"             ""         "kx kxʼ"    ""
"Affricates (normal release)"           ""         "ts tsʰ tsʼ dz"  ""         ""          ""
"Affricates (velar fricative release)"  ""         "tsx"            ""         ""          ""
Fricatives                              ""         "s"              ""         "x"         "h"
Nasals                                  "m"        "n"              ""         ""          ""
Flaps                                   ""         "ɾ"              ""         ""          ""
Approximants                            "w"        ""               "j"        ""          ""
'), header = TRUE)
kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 5)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote("Note: Phonemes with the diacritics (ʰ) and (ʼ) are aspirates and ejectives, respectively.", general_title = "")
```

## Clicks

* Clicks only occur root-initially. Each click has seven types of release: voiceless, voiced, prenasalized, ejective, aspirated, released into a fricative, and released into an ejective fricative [@NhrVisser1998, p. 121].
* @NhrKagaya1978 classes the types of releases as voiceless, voiced, prenasalized, ejective, aspirated, and glottalized (p. 41); he excludes the fricative-released clicks due to not finding minimal pairs between them and the aspirated clicks (p. 44). As for the ejective-fricative clicks, Kagaya argues that they could be allophones of either the ejective or the aspirated clicks. (p. 46)

```{r echo = FALSE, message = FALSE, warning = FALSE, results = 'asis'}
library(dplyr)
library(knitr)
library(kableExtra)
consonants <- read.table(textConnection('
"Manner of Release"     Dental     Alveolar     Palatal    Lateral
"Voiceless"             "ǀ"        "ǃ"          "ǂ"        "ǁ"
"Voiced"                "ᶢǀ"       "ᶢǃ"         "ᶢǂ"       "ᶢǁ"
"Prenasalized"          "ⁿǀ"       "ⁿǃ"         "ⁿǂ"       "ⁿǁ"
"Ejective"              "ǀʼ"       "ǃʼ"         "ǂʼ"       "ǁʼ"
"Aspirated"             "ǀʰ"       "ǃʰ"         "ǂʰ"       "ǁʰ"
"Fricative"             "ǀx"       "ǃx"         "ǂx"       "ǁx"
"Eject. Fric."          "ǀxʼ"      "ǃxʼ"        "ǂxʼ"      "ǁxʼ"
'), header = TRUE)
kable(consonants, col.names = c("Manner of Release", "Dental", "Alveolar", "Palatal", "Lateral"), align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE) %>%
  add_header_above(c("", "Place of Articulation" = 4))
```

## Vowels

* Vowel length is contrastive [@NhrKagaya1978, p. 31]; however, to better reflect the various suprasegmental features of Naro vowels, I will be following the patterns of the authors I have referenced by representing long vowels as sequences of consecutive vowels rather than by using the ː symbol.
* Vowel nasalization is also contrastive. Although my sources have argued that any vowel can be nasalized, nasalized high vowels are not attested in the Crúbadán corpus, so I have excluded them from my ruleset. Nasal vowels are marked with a tile (◌̃).
* Only the first vowel in a Naro word can be pharyngealized [@NhrVisser1998, p. 124].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
       Front    Central   Back
High   "i"      ""        "u"
Mid    "e"      ""        "o oˤ"
Low    ""       "a aˤ"    ""
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE) %>%
  footnote("Note: Phonemes on the left of each cell are plain, whereas phonemes on the right of each cell are pharyngealized", general_title = "")
```

## Tone

* Tone is carried on vowels, as well as on syllabic /m/. Naro has three tones: high, mid, and low. Orthographically, high tones are marked with an acute accent (´), while low tones are marked with a grave accent (\`). Mid tones are unmarked [@NhrVisser1998, p. 132].
* Voiced consonants tend to lower the tone of following vowels, while ejectives tend to raise it [@NhrVisser1998, p. 133].

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme
"a"            "/a/"  
"a̱"            "/aˤ/"  
"b"            "/b/"  
"c"            "/ǀ/"   
"c\'"          "/ǀʼ/"   
"d"            "/d/"  
"e"            "/e/"  
"f"            "/f/"  
"g"            "/x/"   
"h"            "/h/"  
"i"            "/i/"  
"k"            "/k/"   
"m"            "/m/"  
"n"            "/n/"  
"o"            "/o/"  
"o̱"            "/oˤ/"  
"p"            "/p/"
"q"            "/ǃ/"   
"q\'"          "/ǃʼ/"    
"r"            "/ɾ/"  
"s"            "/s/"  
"t"            "/t/"   
"u"            "/u/"  
"w"            "/w/"   
"x"            "/ǁ/"   
"x\'"          "/ǁʼ/"   
"y"            "/j/"  
"z"            "/dz/"
**Multigraph**  ""  
"cg"           "/ǀx/"  
"cg\'"         "/ǀxʼ/"  
"ch"           "/ǀʰ/"
"dc"           "/ᶢǀ/"  
"nc"           "/ⁿǀ/"
"gh"           "/ɡ/"
"kg"           "/kx/"  
"kg\'"         "/kxʼ/"  
"kh"           "/kʰ/"
"ph"           "/pʰ/"  
"qg"           "/ǃx/"  
"qg\'"         "/ǃxʼ/"  
"qh"           "/ǃʰ/"
"dq"           "/ᶢǃ/"  
"nq"           "/ⁿǃ/"
"tg"           "/tx/"  
"th"           "/tʰ/"  
"tc"           "/ǂ/"  
"tcg"          "/ǂx/"  
"tcg\'"        "/ǂxʼ/"  
"tch"          "/ǂʰ/"  
"tc\'"         "/ǂʼ/"  
"dtc"          "/ᶢǂ/"  
"ntc"          "/ⁿǂ/"  
"ts"           "/ts/"  
"tsh"          "/tsʰ/"
"xg"           "/ǁx/"  
"xg\'"         "/ǁxʼ/"  
"xh"           "/ǁʰ/"
"dx"           "/ᶢǁ/"  
"nx"           "/ⁿǁ/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* Some speakers spirantize /dz/ to [z] [@NhrVisser1998, p. 120].
* /d/ becomes [ɾ] root-medially [@NhrVisser1998, p. 122].
  * @NhrKagaya1978 observes the same process, but argues that it's a phonemic /ɾ/ that only occurs intervocalically (p. 37).
* The contrasts between /i e/ and /u o/ collapse when the vowels are nasalized [@NhrVisser1998, p. 124].
* In words that begin with a labial or alveolar consonant and have a front vowel in the second syllable, /a/ in the first syllable tends to raise [@NhrVisser1998, p. 128].
  * This consequently collapses the vowel sequences /ai/ and /ae/ to [ii] and [ee] respectively (ibid.).
* Intervocalic /b/ spirantizes to [β] [@NhrKagaya1978, p. 34].

# Misc. Rules

* Some interrogative words feature word-initial syllabic /n/ [@NhrVisser1998, p. 123].

# References
