---
title: "Yine"
author: "Emily Strand"
bibliography: pib.bib
output: html_document
---

Last Updated: 2020-03-05

**Slightly Compromised: possible conflation between /n/ and /h̃/ (see Misc. Rules)**

# Background

**Language family**: Arawakan / Maipuran / Southern Maipuran / Purus

* Also referred to as Pira, Yine is spoken in Peru.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"     Labial   Alveolar Postalveolar  Palatal  Velar  Glottal
Stops                        "p"      "t"      ""            ""       "k"    ""
Affricates                   ""       "ts"     "tʃ"          "tç"     ""     ""
Fricatives                   ""       "s"      "ʃ"           "ç"      ""     "h̃"
Nasals                       "m"      "n"      ""            ""       ""     ""
Flaps                        ""       "ɾ"      ""            ""       ""     ""
Approximants                 ""       "l"      ""            "j"      "w"    ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* Vowel length is said to be a phonemic feature, but rather marginal [@Sebastian2008, p. 367]. Long vowels are indicated by duplicate graphemes.
    - Other than duplicate like-vowels, consecutive vowels do not occur in Yine.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         ""          "ɯ"
Mid       "e"         ""          "o"         
Low       ""          "a"         ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```  

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme
"a"              "/a/"         
"e"              "/e/"
"g"              "/h̃/"      
"i"              "/i/"
"j"              "/ç/"
"k"              "/k/"
"l"              "/l/"
"m"              "/m/"
"n"              "/n/"
"o"              "/o/"
"p"              "/p/"
"r"              "/ɾ/"
"s"              "/s/"
"t"              "/t/"
"u"              "/ɯ/"
"w"              "/w/"
"x"              "/tʃ/"
"y"              "/j/"
**Digraph**      ""
"ch"             "/tç/"
"sh"             "/ʃ/"
"ts"             "/ts/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```  

# Misc. Rules

* /ɡ/ is realized as [n] preceding /ç/ and /k/ [@Bisso2009, p. 66]. In this environment, both ⟨n⟩ and ⟨g⟩ are sometimes used to indicate the phonetic realization, thus, there is some conflation between /n/ and /h̃/.
* Vowels nasalize following nasals or /h̃/ [@Sebastian2008, p. 367].
* /ɾ/ may be trilled word-initially (ibid.).
* /w/ is realized as [β] preceding front vowels [@Nies2008, p. 1].

# References
