---
title: "Chayahuita"
author: "Bill Mizgerd"
bibliography: cbt.bib
output: html_document
---

Last Updated: 2020-06-26

# Background

**Language Family**: Cahuapanan

* The language is known by several names, the most common of which are Chayahuita and Shawi.
* Chayahuita is spoken within the Loreto and San Martin regions in northern Peru, and along the Paranapura, Cahuapanas, Sillay, and Shanusi rivers.

# Phonology

## Consonants

* /β/ is attested as phonemic in @CbtPhonemesUCB, but is described as allophonic in @CbtBarraza2005 (see the Misc. Rules section).
* @CbtHart1976 and @CbtBarraza2005 both list /r/ as a "vibrant," a general term that could mean either a trill or a flap; because @CbtPhonemesUCB comes down decisively on the side of the flap, I have opted for that analysis.
* /ʔ/ and /h/ only contrast before voiceless consonants; otherwise, only /ʔ/ is attested [@CbtBarraza2005, p. 53]. Because of this, and because of the lack of clear attestation of /h/ in the orthography (i.e. sometimes, but not always, ⟨h⟩ is marked), I have opted to only use /ʔ/.
    - Neither glottal consonant is attested as phonemic in @CbtHart1976.
* /ɾ/ and the glottal consonants do not occur word-initially [@CbtBarraza2005, p. 47].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"      Labial   Alveolar  Postalveolar  Palatal   Velar  Glottal
Stops                          "p"      "t"       ""            ""        "k"    "ʔ"
Affricates                     ""       ""        "tʃ"          ""        ""     ""
Fricatives                     ""       "s"       "ʃ"           ""        ""     ""
Nasals                         "m"      "n"       ""            ""        ""     ""
Flaps                          ""       "ɾ"       ""            ""        ""     ""
Approximants                   "w"      ""        ""            "j"       ""     ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* @CbtHart1976 list four suprasegmental features that can be added to vowels: (length) /ː/, (aspiration) /ʰ/, (glottalization) /ˀ/, and (nasalization) /◌̃/ (p. 2). These do not seem, however, to be phonemic.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
       Front   Central   Back
High   "i"     "ɨ"       "u"
Low    ""      "a"       ""
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

* Spanish loanwords into Chayahuita incorporate several letters not used in native words: ⟨d e f j l ñ u z⟩ [@CbtOrthography]. However, these rarely occur, so I have omitted them from the ruleset.
* The acute accent indicates "emphasis or intensity" but does not carry phonemic value [@CbtOrthography].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme     Phoneme           Comment
"a"           "/a/"              ""
"c"           "/k/"              ""
"ë"           "/ɨ/"              ""
"i"           "/i/"              ""
"m"           "/m/"              ""
"n"           "/n/"              ""
"o"           "/u/"              ""
"p"           "/p/"              ""
"q"           "/k/"     "typically followed by ⟨u⟩"  
"r"           "/ɾ/"              ""
"s"           "/s/"              ""
"t"           "/t/"              ""
"y"           "/j/"              ""
"\'"          "/ʔ/"              ""
**Digraph**    ""                ""
"ch"          "/tʃ/"             ""
"hu"          "/w/"              ""
"sh"          "/ʃ/"              ""
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Chayahuita has a (C)V(C) syllable structure [@CbtBarraza2005, p. 43].

# Lenition Rules

* Root-final nasals delete in suffixed words [@CbtBarraza2005, p. 45-46].
* Consonants voice when (1) preceded by a nasal or (2) occurring in the third or later syllable of a word [@CbtBarraza2005, pp. 48-49; @CbtHart1976, pp. 5-6].
    * However, consonants do not voice in /ʔ/-final syllables (ibid.).
    * For some speakers, especially near the Cahuapanas River, consonants in such positions weaken even farther, either becoming a glide or deleting completely [@CbtBarraza2005, p. 49].
* Suffix-initial consonants are deleted when between two identical vowels. In such cases, the adjacent vowels are produced as a single long vowel [@CbtBarraza2005, p. 50].
* Word-final nasals delete, and nasalize their preceding vowel [@CbtBarraza2005, p. 56].

# Misc. Rules

* /w/ deletes in the following sequences: /uwa, uwi/ [@CbtBarraza2005, p. 44].
* /w/ spirantizes to [β] before /i/ [@CbtBarraza2005, pp. 51-52; @CbtHart1976, p. 7].
* /j/ deletes when preceded by /i/ and followed by /a/ (ibid.).
* Various minimal pairs can be found on pages 39-42 of @CbtBarraza2005.
* When a vowel-initial suffix is added onto the end of a vowel-final root, the consecutive vowels are reanalyzed into a diphthong [@CbtBarraza2005, p. 45].
* /p/ and /m/ are produced as [pʷ] and [mʷ] when before /ɨ/ [@CbtBarraza2005, p. 51; @CbtHart1976, p. 5].
* Epenthesis of /ʔ/ occurs at the end of monosyllabic roots, as well as between root-medial identical vowels [@CbtBarraza2005, p. 53].
* Nasals assimilate to the place of articulation of following consonants [@CbtBarraza2005, p. 51; @CbtHart1976, p. 7].
* /s/ palatalizes to [ʃ] before /i/; in this environment, the contrast between /s/ and /ʃ/ is neutralized [@CbtBarraza2005, p. 52].
    * Similarly, word-medial /t/ goes to [tʃ] before /i/ [@CbtBarraza2005, pp. 52-53].
* When occurring after a consonant-final syllable, /ɾ/ nasalizes to [n] [@CbtBarraza2005, p. 55].
* In the sequence /tuɾu/, the first /u/ deletes [@CbtBarraza2005, p. 57].
* The sequence /wɨ/ collapses to [u] word-finally (ibid.).

# References
