---
title: "Central Bikol"
author: "Bill Mizgerd"
bibliography: bcl.bib
output: html_document
---

Last Updated: 2019-06-17

**SLIGHTLY COMPROMISED: glottal stops not transcribed consistently**

# Background

**Language Family:** Austronesian / Malayo-Polynesian / Western Malayo-Polynesian / Meso Philippine / Central Philippine / Bikol / Coastal / Naga

* Central Bikol, or Bikol, is spoken throughout the Bikol provinces within the Philippines.

# Phonology

## Consonants

* Loans from Spanish and English have introduced /f/, /v/, /z/, /ʃ/, /ʒ/, /ɲ/, /ʎ/, /tʃ/, and /dʒ/ to Bikol, although not all speakers use those sounds [@BclMattes2014, p. 8].

```{r echo = FALSE, message = FALSE, warning = FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial   Alveolar   Palatal    Velar    Glottal
Stops                      "p  b"    "t  d"      ""       "k  ɡ"     "ʔ"
Nasals                      "m"       "n"        ""        "ŋ"       ""
Fricatives                  ""        "s"        ""        ""        "h"
Flaps                       ""        "ɾ"        ""        ""        ""
Approximants                "ʋ"       "l"        "j"       ""        ""
'), header = TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "") %>%
  add_header_above(c("", "Place of Articulation" = 5))

```

## Vowels

* Although the orthography includes ⟨e⟩ and ⟨o⟩, [e] and [o] only exist as allophones of /i/ and /u/ respectively [@BclMattes2014, p. 8].

```{r echo = FALSE}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         ""          "u"
Mid       ""          ""          ""
Low       ""         "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)


diphthongs <- read.table(textConnection('
        Diphthongs   
"/iu/,  /ui/,  /ai/,  /au/"
'), TRUE)

kable(diphthongs, align = 'c') %>%
  kable_styling("bordered")
```

# Alphabet

* Glottal stops are not always reflected in the spelling of a word [@BclMattes2014, p. 12]. Occurrences in at least the intervocalic positions are predictable (i.e. phonetic), which isn't problematic for the language (as we don't account for them); however, occurrences of glottal stops elsewhere aren't transcribed consistently, which compromises the language to some degree.

```{r echo = FALSE}

alphabet <- read.table(textConnection('
Grapheme  Phoneme    
  "a"      "/a/"       
  "b"      "/b/"       
  "d"      "/d/"       
  "e"      "/i/"       
  "g"      "/ɡ/"       
  "h"      "/h/"       
  "i"      "/i/"        
  "k"      "/k/"       
  "l"      "/l/"        
  "m"      "/m/"        
  "n"      "/n/"       
  "o"      "/u/"        
  "p"      "/p/"       
  "r"      "/ɾ/"        
  "s"      "/s/"       
  "t"      "/t/"        
  "u"      "/u/"       
  "w"      "/ʋ/"        
  "y"      "/j/"
  "\' ; -" "/ʔ/"
  **Digraph** ""        
  "ng"     "/ŋ/"
  "aw"     "/au/"
  "ay"     "/ai/"
  "iw"     "/iu/"
  "oy"     "/ui/"
  "uy"     "/ui/"'), header = TRUE)

knitr::kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Bikol syllable structure is CV(C), where V can be either a single vowel or a diphthong [@BclMattes2014, p. 10].

# Misc. Rules

* Epenthesis of [h] occurs between stem-final and suffix-initial vowels [@BclMattes2014, p. 9].
* /u/ is realized as [o] in the final syllable of a word [@BclMintzD1971, p. 17].
* Glottal stops are always inserted between orthographically adjacent vowels [@BclMattes2014, p. 12].
* Prefix-final /ŋ/ tends to assimilate, to varying degrees, to the first consonant of the stem [@BclMattes2014, p. 9].

# References
