---
title: "Ignaciano"
author: "Abi Creighton"
bibliography: ign.bib
output: html_document
---

Last Updated: 2020-08-05

# Background

**Language Family:** Arawakan / Maipuran / Southern Maipuran / Bolivia-Parana / Moxo

* Ignaciano is one of two existing Moxo languages along with Trinitario. It spoken in the lowlands of Bolivia.

# Phonology

## Consonants

* @ott_fonemas_1959 include /w/ in the phonemic inventory (p. 2). However, more recent sources [e.g. @ott_phonemes_1967, p. 1] do not report it as a phoneme, instead treating it as an allophone of /v/.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"    Labial   Alveolar  Postalveolar  Palatal  Velar   Glottal
Stops                        "p"       "t"         ""          ""      "k"      "ʔ"   
Affricates                   ""        "ts"       "tʃ"         ""      ""       ""
Fricatives                   "v"       "s"        "ʃ"          ""      ""       "h"
Nasals                       "m"       "n"         ""          "ɲ"     ""       ""       
Taps                         ""        "ɾ"         ""          ""      ""       ""       
Approximants                 ""        ""          ""          "j"     ""       ""   
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE)
```  

## Vowels

* Vowels with an acute accent (´) are stressed [@ott_fonemas_1959, p. 9]. Stress, however, is not accounted for in this project, and such vowels will be transcribed to their plain representations.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
             Front      Central       Back
High          "i"         ""           "u"
Mid           "e"          ""          ""
Low           ""          "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)

```     

# Alphabet

* Ignaciano commonly uses graphemes from the Spanish alphabet, especially ⟨l⟩, in borrowings [@ott_diccionario_1983, p. 5].
* @ott_diccionario_1983 and @danielsen_2011 use slightly different orthographic conventions:
    - @ott_diccionario_1983 represent /k/ with ⟨c⟩ and /ʔ/ with ⟨h⟩ (p. 5).
    - With reference to Baure, another Southern Arawaken language, @danielsen_2011 represents /k/ with ⟨k⟩ and /ʔ/ with an apostrophe (p. 497).
    - Because the different orthographic conventions do not conflict and both are represented in the Crúbadán corpus, I have included both.
* Although no documentation indicates ⟨w⟩ as being a part of the Ignaciano alphabet, there is reason to believe that it represents /v/, given that nearly all occurrences of it in the Crúbadán corpus precede /a/ or /u/, which is where the phonological process from /v/ to [w] occurs [@ott_phonemes_1967, p. 1]. In the rules, I have restricted the transcription to only occur preceding /a/ and /u/.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme                     Comment         
"a"              "/a/"                        ""
"c"              "/k/"                        ""
"e"              "/e/"                        ""
"h"              "/ʔ/"                        ""
"i"              "/i/"                        ""
"j"              "/h/"                        ""
"k"              "/k/"                        ""
"m"              "/m/"                        ""
"n"              "/n/"                        ""
"ñ"              "/ɲ/"                        ""
"p"              "/p/"                        ""
"r"              "/ɾ/"                        ""
"s"              "/s/"                        ""
"t"              "/t/"                        ""
"u"              "/u/"                        ""
"v"              "/v/"                        ""
"w"              "/v/"       "preceding /a/ and /u/; see comment above"   
"y"              "/j/"                        ""
"\'"             "/ʔ/"                        ""
**Digraph**       ""                          ""                   
"ch"             "/tʃ/"                       ""
"qu"             "/k/"                        ""
"sh"             "/ʃ/"                        ""
"ts"             "/ts/"                       ""
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Ignaciano syllables include the following structures [@ott_fonemas_1959, p. 9]:

    - V
    - CV
    - CVV
    - CVC

# Lenition Rules

* /v/ approximantizes to [w] preceding non-front vowels [@ott_diccionario_1983, p. 6].
* Word-initial stops in unstressed syllables tend to be lenis [@ott_phonemes_1967, p. 56].

# Misc. Rules

* For some speakers, /v/ is realized as [b] [@ott_phonemes_1967, p. 56].

# References
