---
title: "Russia Buryat"
author: "Abi Creighton"
bibliography: bxr.bib
output: html_document
---

Last Updated: 2020-07-27

# Background

**Language Family:** Mongolian / Eastern / Oirat-Khalkha / Khalkha-Buriat / Buriat

* Russia Buryat is the most widely spoken dialect of Buryat (also spelled Buriat). It is spoken along Russia's border with Mongolia in the Buryat Republic, Ust-Orda Buryatia, and Aga Buryatia. Speakers of other Buryat dialects live in Mongolia and China.

# Phonology

## Consonants

* /p/, /f/, /v/, /fʲ/, /vʲ/, /ts/, /tʃ/, and /k/ only appear in loanwords [@staroverov_2020, p. 45].
    - @graber_2012 argues, however, that /k/ appears in so many loanwords that it has become part of Buryat's phonemic inventory (p. xviii). I have therefore opted to include it in the inventory and rules.
    - I have opted also to include /p/, just as @Radnaeva2003 and @Sanzheeva1962 do (p. 2738; p. 12).
    - Although I closely follow the analysis done by @staroverov_2020, I have chosen to diverge from it and exclude palatalized /s/ and /z/, just as @Radnaeva2003, @Radnaeva2006, and @Sanzheeva1962 do.
* The palatalized consonants are recognized as phonemes, but based upon how infrequently they appear in Crúbadán, they seem rather marginal.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"    Bilabial    Alveolar      Postalveolar  Palatal  Velar      Glottal
Stops                      "p pʲ b bʲ" "t tʲ d dʲ"    ""            ""       "k ɡ ɡʲ"   ""
Fricatives                  ""         "s z"          "ʃ ʒ"         ""       "x xʲ"     "ɦ"
Nasals                      "m mʲ"     "n nʲ"         ""            ""       ""         ""  
Trills                      ""         "r rʲ"         ""            ""       ""         ""
Approximants                ""         "l lʲ"         ""            "j"      ""         ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Bilabial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes without diacritics that share a cell, those on the left are voiceless and those on the right are voiced. Phonemes with the diacritic (ʲ) are palatalized.", general_title = "")
```  

## Vowels

* Vowel length is phonemic in Buryat [@khabtagateva_2013, p. 161].
* There is much disagreement around how to represent the mid-front vowel. I have chosen to follow @staroverov_2020 and use /ɘ/ (p. 45).

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
             Front      Central       Back
High          "i"         "ʉ"          "u"
Mid           "ɘ"         "ɵ"          "o"
Low           ""          "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)


diphthongs <- read.table(textConnection('
Diphthongs
"/ai/, /oi/, /ui/, /ʉi/, /ei/"
'), TRUE)

kable(diphthongs, align = 'c') %>%
  kable_styling("bordered")

```     

# Alphabet

* **Note:** /ɡʲ/ is not represented in the Crúbadán corpus.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme                    Comment
"а"            "/a/"                       ""
"б"            "/b/"                       ""
"в"            "/v/"                "occurs in loanwords"
"г"            "/ɡ/"                       ""
"д"            "/d/"                       ""
"е"            "/jɘ/"                      ""
"ё"            "/jo/"                      ""
"ж"            "/ʒ/"                       ""
"з"            "/z/"                       ""
"и"            "/i/"                       ""
"й"            "/j/"            "often appears as the second element of a diphthong [@graber_2012, p. xviii]"
"к"            "/k/"                 "occurs in loanwords"
"л"            "/l/"                       ""
"м"            "/m/"                       ""
"н"            "/n/"                       ""
"о"            "/o/"                       ""
"ө"            "/ɵ/"                       ""
"п"            "/p/"                       ""
"р"            "/r/"                       ""
"с"            "/s/"                       ""
"т"            "/t/"                       ""
"у"            "/u/"                       ""
"ү"            "/ʉ/"                       ""
"ф"            "/f/"                  "occurs in loanwords"
"х"            "/x/"                       ""
"ц"            "/ts/"                 "occurs in loanwords"
"ч"            "/tʃ/"                 "occurs in loanwords"
"ш"            "/ʃ/"                        ""
"щ"            "/ʃtʃ/"                "occurs in loanwords"
"ъ"             ""           "indicates lack of palatalization of proceeding consonant"
"ы"            "/i/"                        ""
"ь"            "/ʲ/"         "marks palatalization of preceding consonant"  
"э"            "/ɘ/"                       ""
"ю"            "/ju/"                      ""  
"я"            "/ja/"                      ""
"һ"            "/ɦ/"                       ""

'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* The structure of syllables in Russia Buryat consist of CV(C)(C) word-medially and (C)V(C)(C) word-initially [@staroverov_2020, p. 46].

# Lenition Rules

* /ɡ/ spirantizes to [ɣ] intervocalically in words with front vowels and to [ʁ] intervocalically in words with back vowels [@staroverov_2020, p. 46].

# Misc. Rules

* Vowel harmony is present in Buryat [@khabtagateva_2013, p. 161].
* When a vowel-initial suffix is added to a short vowel stem, one of the vowels is deleted [@staroverov_2020, p. 47].
* When a vowel-initial suffix is added to a long vowel stem, a dorsal consonant is epenthesized (ibid.).

# References
