---
title: "Palauan"
author: "Bill Mizgerd"
edited by: "Emily Strand"
bibliography: pau.bib
output: html_document
---

Last Updated: 2020-01-17

# Background

**Language Family:** Austronesian / Malayo-Polynesian / Western Malayo-Polynesian / Palauan

* Palauan is spoken in Palau, an archipelagic country off the coast of the Philippines.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Bilabial Alveolar  Velar   Glottal
Stops                      "b"      "t d"     "k"     "ʔ"
Fricatives                 ""       "s"       ""      ""
Nasals                     "m"      ""        "ŋ"     ""
Taps                       ""       "ɾ"       ""      ""
Trills                     ""       "r"       ""      ""
Approximants               ""       "l"       ""      ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Bilabial", "Alveolar", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 4)) %>%
  footnote(general = "Note: Phonemes to the left of each cell are voiceless and phonemes to the right of each cell are voiced.", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* /ə/ only appears in unstressed syllables, and consequently, it is not universally considered to be phonemic in Palauan [@PauJosephs1975, p. 22]. It is interpreted as an allophone of /ɛ/, or a phonetic realization based on context [@Carlson1968, p. 36]. It is represented by ⟨ȩ⟩ in the alphabet; however, this grapheme is not present in the Crúbadán corpus. Nonetheless, I have chosen to represent it in the rule set as /ɛ/.
* Vowel length is contrastive for non-central vowels. Long front vowels glide into [j], while long back vowels glide into [w] [@PauJosephs1975, pp. 25-26]. Long vowels are represented by doubling the vowel character.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
          Front   Central   Back
High      "i"     ""        "u"
Mid       "ɛ"     ""        "o"
Low       ""      "a"       ""
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme           Comment  
"a"              "/a/"               ""
"b"              "/b/"               ""
"d"              "/d/"               ""
"e"              "/ɛ/"               ""
"ȩ"              "/ɛ/"        "realized as [ə]"
"i"              "/i/"               ""
"k"              "/k/"               ""
"l"              "/l/"               ""
"m"              "/m/"               ""
"n"              "/n/"        "only in loanwords"
"o"              "/o/"               ""
"r"              "/ɾ/"               ""
"s"              "/s/"               ""
"t"              "/t/"               ""
"u"              "/u/"               ""
**Digraph**       ""                 ""
"ch"             "/ʔ/"               ""
"ll"             "/lː/"              ""
"ng"             "/ŋ/"               ""
"rr"             "/r/"               ""
'), TRUE)

kable(alphabet, align = "c") %>%
  kable_styling("bordered")
```

# Lenition Rules

* Word-initial /d/ spirantizes to [ð], particularly in careful speech [@PauJosephs1975, p. 8].
* Intervocalic and word-final /d/ spirantize to [ð] [@PauJosephs1975, p. 9].
* /k/ voices to [ɡ] intervocalically [@PauJosephs1975, p. 10].

# Misc. Rules

* When adjacent to another consonant - except /l/ - /b/ devoices to [p] [@PauJosephs1975, p. 6].
* Word-final voiceless stops aspirate [@PauJosephs1975, pp. 7, 10].
* Word-initial preconsonantal /d/ devoices to [t]; this rule can also co-occur with the spirantization rule mentioned above, resulting in [θ] [@PauJosephs1975, p. 9].
* Epenthesis of [ʔ] occurs at the beginning of vowel-initial utterances [@PauJosephs1975, pp. 11-12].
* /ŋ/ assimilates to the place of subsequent alveolar consonants [@PauJosephs1975, p. 13].
* Word-initial preconsonantal sonorants are syllabic, except in the word ⟨mlai⟩ [@PauJosephs1975, pp. 17-18].
* Epenthesis of [ə] occurs after word-final consonant clusters [@PauJosephs1975, p. 24].
* In vowel clusters, unstressed vowels turn to glides in various contexts [@PauJosephs1975, pp. 27-28].

# References
