---
title: "Wayana"
author: "Emily Strand"
bibliography: way.bib
output: html_document
---

Last updated: 2019-11-05

# Background

**Language family**: Carib / Northern / East-West Guiana / Wayana-Trio

* Wayana is spoken in Suriname, a country off the coast of South America.

# Phonology

## Consonants

* /s/ is said to have neutralized with /h/ [@Jackson1972, p. 48]; however, I have chosen to include both in the phonemic inventory, given that the distinction is still made by the majority of speakers.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial  Alveolar  Retroflex  Palatal  Velar  Glottal
Stops                       "p"      "t"        ""        ""      "k"      ""
Fricatives                  ""       "s"        ""        ""       ""      "h"
Nasals                      "m"      "n"        ""        ""       ""      ""
Flaps                       ""       ""         "ɽ"       ""       ""      ""
Approximants                "w"      ""         ""        "j"      ""      ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Retroflex", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE)
```

# Vowels

* Although, long vowels (duplicate graphemes) are apparent in the language, @SilvaTavares2005 explains that there are no minimal pairs between long and short vowels (p. 11). Long vowels are likely to result as a by-product from syllable reduction. In the rules, instances of long vowels will be transcribed as short.
* Vowels that occur consecutively are interpreted as sequences rather than complex phonemes [@SilvaTavares2005, p. 29].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         "ɨ"          "u"
Mid       "e"         "ə"          "o"
Low        ""         "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```   

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme
"a"            "a"
"e"            "e"
"ë"            "ə"
"h"            "h"
"i"            "i"
"ï"            "ɨ"
"j"            "j"
"k"            "k"
"l"            "ɽ"
"m"            "m"
"n"            "n"
"o"            "o"
"p"            "p"
"s"            "s"
"t"            "t"
"u"            "u"
"w"            "w"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Syllables in Wayana include the following structures [@Jackson1972, p. 25]:
    - V
    - VC
    - CV
    - CVC

# Lenition Rules

* /p/, /t/, and /k/ are voiced preceding voiced consonants [@Jackson1972, p. 48].
* /p/, /t/, and /k/ primarily debuccalize preceding /p/, /t/, and /k/, respectively (ibid.).
    - /p/ can also debuccalize preceding (sometimes) /m/.
    - /t/ can also debuccalize preceding /l/ and (sometimes) /n/.
* /t/ occasionally palatalizes preceding /i/ [@SilvaTavares2005, p. 18].
* /w/ and /j/ are deleted preceding /u/ and /i/, respectively [@SilvaTavares2005, p. 30].
* Vowels and whole syllables are often deleted, which is consistent with other Carib languages [@SilvaTavares2005, p. 31].

# Misc. Rules

* /h/ varies freely with several different places of articulation (e.g. voiceless bilabial fricative preceding /p/, voiceless dental fricative preceding /m/, and voiceless velar fricative preceding velar consonants) [@Jackson1972, p. 48].
* /p/, /t/, and /k/ can change to nasals, maintaining place of articulation (e.g. /p/ goes to /m/), preceding other nasal consonants [@Jackson1972, pp. 48-49].
* Vowels often nasalize preceding nasal consonants [@SilvaTavares2005, p. 14].
* Consonants do not geminate in Wayana [@SilvaTavares2005, p. 17]. Phonological processes, such as the debuccalization, mentioned above, seem to remedy these potential consonant clusters. @SilvaTavares2005 explains that there is one instance of consonant gemination (⟨kk⟩); however, he does posit that it's merely due to a mis-transcription (p. 17). Therefore, geminates will not be accounted for in the rules.
* [r] and [l] vary freely with /ɽ/ [@SilvaTavares2005, p. 24].
* /w/ varies freely with /β/ preceding front vowels (ibid.).

# References
