---
title: "kk.rmd"
author: "Abi Creighton"
date: "September 24, 2019"
output: html_document
---

# Background

**Language family**: Turkic / Common Turkic / Kipchak / Kipchak-Nogai / Kazakh

* Kazakh is spoken within Central Asia, specifically in China, Kazakhstan, Mongolia, Turkmenistan, and Uzbekistan.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial    Alveolar    Alveolo-palatal     Palatal     Velar     Uvular
Stops                      "p  b"    "t  d"             ""              ""      "k  ɡ"      "q"
Affricates                   ""        ""              "tɕ"             ""         ""        ""
Fricatives                 "f  v"    "s  z"          "ɕ  ʑ"             ""         ""      "χ  ʁ"
Nasals                       "m"       "n"              ""              ""        "ŋ"        ""
Flaps                        ""        "ɾ"              ""              ""         ""        ""
Approximants                 ""        "l"              ""              "j"        "w"       ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Alveolo-palatal", "Palatal", "Velar", "Uvular"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```

## Vowels


```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front     Central      Back
High      "i"        "ʉ"         ""
"High-Mid""ʏ"        ""          "ʊ"
Mid       "e"        ""          "o"
"Low-Mid" "æ"        "ə"         ""
Low       ""         ""          "ɑ"
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

* A Cyrllic script has tradtionally been used for Kazakh in Kazakhstan and Mongolial, but in 2017, the President of Kazakhstan announced that they would transition to using a Latin alphabet by 2025. The 2017 Latin script was revised again in 2018. Parts of China, Iran, and Afghanistan also use an Arabic script. The Cyrllic script and the 2018 Latin script are addressed here.

## Latin

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

l_alphabet <- read.table(textConnection('
Grapheme      Phoneme   
"a"            "/ɑ/"
"á"            "/æ/"
"b"            "/b/"
"d"            "/d/"
"e"            "/jɪ/"
"f"            "/f/"
"g"            "/ɡ/"
"ǵ"            "/ʁ/"
"h"            "/χ/"
"i"            "/ɪ/"
"ı"            "/ɪj/ (explain this)"
"j"            "/ʑ/"
"k"            "/k/"
"l"            "/l/"
"m"            "/m/"
"n"            "/n/"
"ń"            "/ŋ/"
"o"            "/wʊ/"
"ó"            "/wʏ/"
"p"            "/p/"
"q"            "/q/"
"r"            "/ɾ/"
"s"            "/s/"
"t"            "/t/"
"u"            "/ʊ/"
"ú"            "/ʏ/"
"v"            "/v/"
"y"            "/ə/"
"ý"            "/w/"
"z"            "/z/"
**Multigraph**  ""
"ch"           "/tɕ/"
"sh"           "/ɕ/"
"shsh"         "/ɕtɕ/"
"ts"           "/ts/"
"ıý"           "/jʉw/"
"ıa"           "/jɑ/"
"ıo"           "/jo/"
'), TRUE)

kable(l_alphabet, align = 'c') %>%
  kable_styling("bordered")
```

## Cyrillic

* ⟨в⟩, ⟨ё⟩, ⟨ф⟩, ⟨х⟩, ⟨ц⟩, ⟨ч⟩, ⟨щ⟩, ⟨ъ⟩, ⟨ь⟩, ⟨э⟩ appear in loanwords [@vajda, pg. 608, 609].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

c_alphabet <- read.table(textConnection('
Grapheme      Phoneme                         Comment
"а"            "/ɑ/"                            ""
"ә"            "/æ/"                            ""
"б"            "/b/"                            ""
"в"            "/v/"                            ""
"г"            "/ɡ/"                            ""
"ғ"            "/ʁ/"                            ""
"д"            "/d/"                            ""
"е"            "/jɪ/"                           ""
"ё"            "/jo/"                           ""
"ж"            "/ʑ/"                            ""
"з"            "/z/"                            ""
"и"            "/ɪj/"                           ""
"й"            "/j/"                            ""
"к"            "/k/"                            ""
"қ"            "/q/"                            ""
"л"            "/l/"                            ""
"м"            "/m/"                            ""
"н"            "/n/"                            ""
"ң"            "/ŋ/"                            ""
"о"            "/wʊ/"                           ""
"ө"            "/wʏ/"                           ""
"п"            "/p/"                            ""
"р"            "/ɾ/"                            ""
"с"            "/s/"                            ""
"т"            "/t/"                            ""
"у"            "/w/"                            ""
"ұ"            "/ʊ/"                            ""
"ү"            "/ʏ/"                            ""
"ф"            "/f/"                            ""
"х"            "/χ/"                            ""
"һ"            "/h/"                            ""
"ц"            "/ts/"                           ""
"ч"            "/tɕ/"                           ""
"ш"            "/ɕ/"                            ""
"щ"            "/ɕtɕ/"                          ""
"ъ"            "/ə/"    "Originally denoted ultra short or reduced middle rounded vowel"
"ы"            "/ə/"                            ""
"і"            "/ɪ/"                            ""
"ь"            "/ə/"        "Originally denoted short or reduced front vowel"
"э"            "/ɛ/"                            ""
"ю"            "/jʉw/"                          ""
"я"            "/jɑ/"                           ""
'), TRUE)

kable(c_alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* /b/, /ɡ/, and /q/ may spirantize intervocalically [@kirchner_kazakh_1998].

# Misc. Rules

* There is some alternation with /ʒ/ and /̌dʒ/ [@manczak-wohlfeld_studies_2010].
* ⟨и⟩ transcribes to /əj/ before or after ⟨ғ⟩ & ⟨қ⟩ and back vowels.
* Before or after ⟨г⟩ & ⟨к⟩ and back vowels, ⟨у⟩ = /ʊw~əw/, ⟨и⟩ = /ɘj/ and ⟨у⟩ = /ʉw~ɘw/

# References
