---
title: "Dedua"
author: "Jordan Schuster"
bibliography: ded.bib
output: html_document
---

Last Updated: 2020-06-02

# Background

**Language Family:** Trans-New Guinea, Main Section, Central and Western, Huon-Finisterre, Huon, Eastern

* Dedua is spoken in the Morobe province of Papua New Guinea.

# Phonology

## Consonants

* Based on the comparison with other languages in the same language family, @McEvoy2005 concludes that voiceless stops in syllable codas are neutralized and realized as glottal stops (p. 269). Thus, the phonemic status of the glottal stop is called into question. However, I have opted to treat it as a phoneme here because this seems to be the common practice found in the literature.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial     Alveolar  Alveopalatal    Palatal     Velar    Glottal  
Stops                      "p  b"     "t  d"        ""            ""     "k kp ɡ ɡb"   "ʔ"
Affricates                  ""         ""           "dʑ"          ""          ""       ""
Fricatives                  "f"        "s"          ""            ""          ""       ""
Nasals                      "m"        "n"          ""            ""          "ŋ"      ""
Trills                      ""         "r"          ""            ""          ""       ""
Approximants                "w"        ""           ""            "j"         ""       ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Alveopalatal", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced. /kp/ and /ɡb/ are labiovelar stops.", general_title = "")
```

## Vowels

* According to @McEvoy2005, the diphthongs listed below are always interpreted as complex nuclei (p. 270).

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         ""          "u"
Mid       "e"         ""          "o"
Low        ""         "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)


diphthongs <- read.table(textConnection('
Diphthongs
"/ei/, /eu/, /ai/, /au/, /oi/, /ou/"
'), TRUE)

kable(diphthongs, align = 'c') %>%
  kable_styling("bordered")
```

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme       Phoneme        Comment
"a"             "/a/"           ""
"b"             "/b/"           ""
"c"             "/ʔ/"        "syllable-finally"
"d"             "/d/"           ""
"e"             "/e/"           ""
"f"             "/f/"           ""
"g"             "/ɡ/"           ""
"h"             "/ʔ/"        "syllable-initially"
"i"             "/i/"           ""
"k"             "/k/"           ""
"m"             "/m/"           ""
"n"             "/n/"           ""
"o"             "/o/"           ""
"p"             "/p/"           ""
"r"             "/r/"           ""
"s"             "/s/"           ""
"t"             "/t/"           ""
"u"             "/u/"           ""
"w"             "/w/"           ""
"y"             "/j/"           ""
**Digraph**      ""             ""
"dz"            "/dʑ/"          ""
"gb"            "/ɡb/"          ""
"kp"            "/kp/"          ""
"ng"            "/ŋ/"           ""
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Syllables in Dedua consist of the following structures [@ceder_dedua_2003, p. 3]:
    - V
    - VV
    - VC
    - CV
    - CVV
    - CVC
    - CVVC
* Glottal stops and /ŋ/ are the only phonemes that appear in syllable codas [@McEvoy2005, p. 268]. This makes for the inference that complex consonants are always realized as such rather than two separate phonemes.

# Lenition Rules

* /ʔ/ becomes [h] syllable-initially [@ceder_dedua_2003, p. 3].

# Misc. Rules

* /r/ varies freely with [l] intervocalically [@ceder_dedua_2003, p. 3].
* /w/ varies freely with [v] (ibid.).
* In syllable coda position, the velar nasal undergoes assimilation to the place of articulation of the following consonant if present. This assimilation is reflected in written Dedua [@McEvoy2005, p. 268]; however, it has been accounted for in the rules to reflect /ŋ/ as underlying.

# References
