---
title: "Mapudungun"
author: "Emily Strand"
bibliography: arn.bib
output: html_document
---

Last updated: 2020-06-26

**COMPROMISED: ambiguity due to non-standard alphabet; conflation of dental and alveolar consonants /t̪/ and /t/ (most likely with the others as well)**

# Background

**Language Family:** Araucanian

* Also referred to as Mapuche or Araucana, it is spoken throughout Chile.

# Phonology

## Consonants

* In some dialects of Mapudungun, the distinction between the dental and alveolar phonemes have merged [@sadowsky_mapudungun_2013, p. 89], leaving the dentals to appear as a result of allophonic variation.
* /ʃ/ is rather infrequent, often alternating with /s/, yet it is still considered in most sources to be a phoneme [@smeets_grammar_2008, p. 23].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"     Labial     Dental    Alveolar   Postalveolar   Retroflex  Palatal   Velar
Stops                         "p"        "t̪"         "t"         ""            ""        ""       "k"  
Affricates                    ""         ""          ""         "tʃ"          "ʈʂ"       ""       ""    
Fricatives                    "f"        "θ"         "s"         "ʃ"          "ʐ"        ""       "ɣ"    
Nasals                        "m"        "n̪"         "n"         ""            ""        "ɲ"      "ŋ"     
Approximants                  ""         "l̪"         "l"         ""            ""       "j ʎ"     "w"   
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Dental", "Alveolar", "Postalveolar", "Retroflex", "Palatal", "Velar"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 7)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: The palatal approximant on the right is lateral.", general_title = "")
```

## Vowels

* Diphthongs aren't prevalent in Mapudungun; however, /ae/ is generally realized as one [@smeets_grammar_2008, p. 52]. Because it can also occur as a sequence of vowels, it will not be transcribed in the rules.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
      Front    Central    Back
High   "i"       ""       "u"
Mid    "e"       "ɘ"      "o"
Low     ""       "a"      ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)

```

# Alphabet

* Although there isn't a standardization of the alphabet, the three alphabets commonly referred to provide a rather comprehensive picture in terms of depicting each phoneme. These alphabets include the Alfabeto Mapuche Unificado, the Frafemario Raguileo, and the Azumchefi [@bronzino_loanword_nodate, p. 22]. Listed below is the Alfabeto Mapuche Unificado, but following it, I have provided the orthographic variations from the other two.
* Regarding the corresponding phoneme of ⟨g⟩, the Unified Alphabet uses /ɣ/ whereas the Raguileo and Azumchefi alphabets use /ŋ/. I went with the ⟨g⟩ to /ŋ/ correspondence due to the inclusion of ⟨q⟩ in the Crúbadán corpus, which based on the Raguileo and Azumchefi alphabets corresponds to /ɣ/.
* Based on the Alfabeto Mapuche Unificado, the dentals are not represented in the Crúbadán corpus. With the grapheme correspondences of the other two alphabets, they appear (except /t̪/, as the distinction between it at /t/ is not maintained in either). Based on this, conflation most likely occurs between all alveolar consonants and their dental counterparts, compromising the language.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme      Comment
"a"            "/a/"         ""
"d"            "/θ/"         ""
"e"            "/e/"         ""
"f"            "/f/"         ""
"g"            "/ŋ/"         ""
"i"            "/i/"         ""
"ï"            "/ɘ/"         ""
"k"            "/k/"         ""
"l"            "/l/"         ""
"ḻ"            "/l̪/"         ""
"m"            "/m/"         ""
"n"            "/n/"         ""
"ṉ"            "/n̪/"         ""
"ñ"            "/ɲ/"         ""
"o"            "/o/"         ""
"p"            "/p/"         ""
"r"            "/ʐ/"         ""
"s"            "/s/"         ""
"t"            "/t/"         ""
"ṯ"            "/t̪/"         "not reflected in the Crúbadán corpus"
"u"            "/u/"         ""
"ü"            "/ɘ/"         ""
"w"            "/w/"         ""
"y"            "/j/"         ""
**Digraph**     ""           ""
"ch"           "/tʃ/"        ""
"tr"           "/ʈʂ/"        ""
"sh"           "/ʃ/"         ""
"ng"           "/ŋ/"         ""
"ll"           "/ʎ/"         ""
"**Orthographic Variation**" "" ""
"c"            "/tʃ/"    "Raguileo"  
"x"            "/ʈʂ/"    "Raguileo"  
"tx"           "/ʈʂ/"    "Azumchefi"  
"z"            "/θ/"     "Raguileo and Azumchefi"   
"h"            "/n̪/"     "Raguileo"  
"nh"           "/n̪/"     "Azumchefi"  
"q"            "/ɣ/"     "Raguileo and Azumchefi"  
"b"            "/l̪/"     "Raguileo"  
"lh"           "/l̪/"     "Azumchefi"  
"j"            "/ʎ/"     "Raguileo"   
"v"            "/ɘ/"     "Raguileo"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* Geminates occur in the language, but they are often realized as single consonants [@smeets_grammar_2008, p. 51].
* /ʐ/ can approximate to /ɻ/ [@sadowsky_mapudungun_2013, p. 90].

# Misc. Rules

* Aspiration of some of the stops can occur [@sadowsky_mapudungun_2013, p. 89].
* Unstressed vowels are often deleted in word-final positions following voiceless consonants [@sadowsky_mapudungun_2013, p. 94].
* /ɘ/ may be deleted following a stressed syllable (ibid.).
* Consonant clusters are only allowed intervocalically [@smeets_grammar_2008, pp. 37, 38].
* The velar phonemes /k/, /ŋ/, /ɣ/ tend to be fronted preceding front vowels [@sadowsky_mapudungun_2013, p. 89].


# References
