---
title: "Ilocano"
author: "Emily Strand"
bibliography: ilo.bib
output: html_document
---

Last Updated: 2019-10-17

# Background

**Language Family**: Austronesian / Malayo-Polynesian / Northern Luzon

* Also known as Iloko, Ilocano is spoken throughout the Northern Philippines.
* The northern dialect is represented below.

# Phonology

## Consonants

* /h/ appears in only one native word "haán," but it is not recognized as a phoneme in the language [@yamamoto_phonological_2017, pg. 23].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial     Dental   Alveolar    Palatal     Velar    Glottal
Stops                      "p  b"     "t̪  d̪"      ""         ""        "k  ɡ"     "ʔ"
Fricatives                   ""         ""        "s"        ""         ""        ""
Nasals                       "m"        "n̪"       ""         ""         "ŋ"       ""
Taps                         ""         ""        "ɾ"        ""         ""        ""
Approximants                 "w"        ""        "l"        "j"        ""        ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Dental", "Alveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```            

## Vowels

* Although /o/ was historically recognized as an allophone of /u/, it is now considered a phoneme due to the adoption of several Spanish loanwords [@rubino_ilocano_2000, pg. xxviii].
* There is some debate as to how many dipthongs are present in Ilocano. @constantino_ilokano_1971 specifies 14 (pg. 6), while @rubino_ilocano:_1998 specifies 5 (pg. xxix). @yamamoto_phonological_2017 argues that such combinations actually represent sequences of consonants and vowels rather than dipthongs (pg. 32). Due to a lack of consensus, I will not account for them.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         ""           "u"
Mid       "ɛ"         ""           "o"
Low        ""         "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```  

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme
"a"            "/a/"     
"b"            "/b/"           
"d"            "/d̪/"     
"e"            "/ɛ/"             
"g"            "/ɡ/"          
"i"            "/i/"       
"k"            "/k/"        
"l"            "/l/"      
"m"            "/m/"           
"n"            "/n̪/"              
"o"            "/o/"      
"p"            "/p/"        
"r"            "/ɾ/"         
"s"            "/s/"           
"t"            "/t̪/"          
"u"            "/u/"        
"w"            "/w/"         
"y"            "/j/"      
"\\-"          "/ʔ/"
**Digraph**     ""
"ng"           "/ŋ/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```       

# Lenition Rules

* /k/ and /t/ sometimes debuccalizes to /ʔ/ preceding consonants [@rubino_ilocano_2000, pg.xxiv]    
* /k/ often spirantizes intervocalically [@rubino_ilocano_2000, pg. xxiv].
* /t/,/d/, and /s/ palatalize to /tʃ/, /dʒ/, and /ʃ/ respectively preceding /j/ or /i/ [@rubino_ilocano_2000, pg.xxvii]. /t/ also palatalizes to /tʃ/ in borrowed words [@rubino_ilocano_2000, pg. xxv].   

# Misc Rules

* Glottal stops occur preceding word initial vowels [@rubino_ilocano:_1998, pg.12].       
* Glottal stops occur intervocalically (except with ia, ie, iu, and ua) [@rubino_ilocano:_1998, pg. 12].      
* Gemination can occur with every consonant except /ʔ/ and /h/ [@rubino_ilocano_2000, pg. xxv, xxxiv].        
* /u/ is sometimes realized as /o/ word finally [@constantino_ilokano_1971, pg. 4]

# References
