---
title: "Ilocano"
author: "Emily Strand"
bibliography: ilo.bib
output: html_document
---

Last Updated: 2020-06-30

# Background

**Language Family**: Austronesian / Malayo-Polynesian / Western Malayo-Polynesian / Northern Philippine / Northern Luzon / Ilocano

* Also known as Iloko, Ilocano is spoken throughout the Northern Philippines.
* The northern dialect is represented below.

# Phonology

## Consonants

* /h/ appears in loanwords but only one native word "haán," so it is not recognized as a phoneme in the language [@yamamoto_phonological_2017, p. 23].
* Not all analyses consider /ʔ/ to be a phoneme; instead, it is argued that it predictably epenthesized. However, @yamamoto_phonological_2017 argues that epenthesis only explains the presence of glottal stops word-initially and intervocalically. Because it is also found in other environments, it must have phonemic status (pp. 23-26).
* /r/ is typically pronounced as a dental flap, but is realized as a trill in careful speech [@rubino_ilocano_2000, p. xxvi].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial     Dental   Alveolar    Palatal     Velar    Glottal
Stops                      "p  b"     "t̪  d̪"      ""         ""        "k  ɡ"     "ʔ"
Fricatives                   ""         ""        "s"        ""         ""        ""
Nasals                       "m"        "n̪"       ""         ""         "ŋ"       ""
Flaps                        ""         ""        "ɾ"        ""         ""        ""
Approximants                 "w"        ""        "l"        "j"        ""        ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Dental", "Alveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```            

## Vowels

* The southern dialect has 6 phonemic vowels, while the northern dialect has 5 [@rubino_ilocano:_1998, pp. 11-12].
* Although /o/ was historically recognized as an allophone of /u/, it is now considered a phoneme due to the adoption of many Spanish loanwords [@rubino_ilocano_2000, p. xxviii].
* There is some debate as to how many diphthongs are present in Ilocano. @constantino_ilokano_1971 specifies 14 (p. 6), while @rubino_ilocano:_1998 specifies 5 (p. xxix). @yamamoto_phonological_2017 argues that such combinations actually represent sequences of glides and vowels rather than diphthongs (p. 32). Due to a lack of consensus, I have chosen not to account for them.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         ""           "u"
Mid       "ɛ"         ""           "o"
Low        ""         "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```  

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme      
"a"            "/a/"       
"b"            "/b/"       
"d"            "/d̪/"      
"e"            "/ɛ/"       
"g"            "/ɡ/"       
"i"            "/i/"       
"k"            "/k/"       
"l"            "/l/"       
"m"            "/m/"       
"n"            "/n̪/"      
"o"            "/o/"       
"p"            "/p/"       
"r"            "/ɾ/"       
"s"            "/s/"       
"t"            "/t̪/"      
"u"            "/u/"       
"w"            "/w/"       
"y"            "/j/"       
"\\-"          "/ʔ/"       
**Digraph**     ""         
"ng"           "/ŋ/"       
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```       

# Lenition Rules

* /k/ and /t/ sometimes debuccalize to /ʔ/ preceding consonants [@rubino_ilocano_2000, p.xxiv]    
* /k/ often spirantizes intervocalically [@rubino_ilocano_2000, p. xxiv].

# Misc. Rules

* Glottal stops predictably occur preceding word initial vowels [@rubino_ilocano:_1998, p.12].       
* Glottal stops occur intervocalically (except with ia, ie, iu, and ua) [@rubino_ilocano:_1998, p. 12].
* Gemination can occur with every consonant except /ʔ/ and /h/ [@rubino_ilocano_2000, pp. xxv, xxxiv].
* /t/,/d/, and /s/ palatalize to /tʃ/, /dʒ/, and /ʃ/ respectively preceding /j/ or /i/. /j/ and /i/ also must be followed by a vowel [@rubino_ilocano_2000, p. xxvii]. /t/ also palatalizes to /tʃ/ in borrowed words [@rubino_ilocano_2000, p. xxv].   
* /u/ is sometimes realized as [o] word finally [@constantino_ilokano_1971, p. 4]
* Prefixes ending with /ŋ/ undergo nasal assimilation according to the place of articulation of the follow obstruent [@yamamoto_phonological_2017, p. 37].


# References
