---
title: "Slovak"  
author: "Emily Strand"  
bibliography: sk.bib  
output: html_document  
---

Last updated: 2019-07-26

**Compromised: orthography unreliably marks palatal consonants; ambiguity in terms of when a digraph represents one phoneme or two**  

# Background  

**Language Family**: Indo-European / Slavic / West/ Czech-Slovak  

* Also known as Slovakian, it is predominantly spoken in Slovakia, but also in Canada, Hungary, Poland, Romania, Ukraine, and the U.S.

# Phonology  

## Consonants  

* @rubach_lexical_1993 states that the palatal consonants of ⟨ť⟩, ⟨ď⟩, ⟨ľ⟩, and ⟨ň⟩ are in fact prepalatal (p. 30); however, given that @hanulikova_slovak_2010 and @short_slovak_1993 didn't make this distinction, I went with the standard palatal IPA symbols (p. 374; p. 535).  
* [@hanulikova_slovak_2010] makes the point that even though [lː] and [rː] are often considered allophones of /l/ and /r/ respectively, they don't appear in complementary distribution (p. 374). Based on this, I will keep the phonemic distinction.  

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial  Alveolar  Postalveolar  Palatal   Velar   Laryngeal
Stops                      "p b"   "t d"         ""         "c ɟ"    "k ɡ"       ""
Affricates                 ""      "ts dz"     "tʃ dʒ"        ""       ""        ""
Fricatives                 "f v"   "s z"        "ʃ ʒ"         ""       "x"       "ɦ"
Nasals                     "m"     "n"           ""           "ɲ"      ""        ""
Trills                     ""      "r rː"        ""           ""       ""        ""
Approximants               ""      "l lː"        ""           "j ʎ"    ""        ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Laryngeal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  footnote("Note: For phonemes that share a cell, those on the left are voiceless, whereas those on the right are voiced. Both palatal approximants are voiced, but the one on the right is lateral. Both alveolar trills and approximants are voiced, but the ones on the right are phonemically long.", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

## Vowels  

* @hanulikova_slovak_2010 and @short_slovak_1993 express that /æ/ (⟨ä⟩) is virtually obsolete in the language and is realized more often as /ɛ/ (p. 375; p. 534). Therefore, I will use /ɛ/ to represent the grapheme ⟨ä⟩.  
* Slovak makes a distinction between long and short vowels [@hanulikova_slovak_2010, p. 534]. Long vowels are indicated with acute accents (´).  
* /ʊɔ/ is the only dipthong that is realized by an individual grapheme [@hanulikova_slovak_2010, p. 376].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
          Front   Central   Back
High      "i"     ""        "u"
Mid       "ɛ"     ""        "ɔ"
Low       ""      "a"       ""
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)


diphthongs <- read.table(textConnection('
Diphthongs
"/ia/, /iɛ/, /iu/, /ʊɔ/"
'), TRUE)

kable(diphthongs, align = 'c') %>%
  kable_styling("bordered")
```  

# Alphabet  

* Regarding the graphemes of the palatal consonants ⟨ť⟩, ⟨ď⟩, ⟨ľ⟩, and ⟨ň⟩, the diacritics aren't present when preceding front vowels [@rubach_lexical_1993, p. 30]. This creates ambiguity with the transcription, and is therefore somewhat compromised.
* Graphemes containing two characters (ie. dz) represent one phoneme when not situated at a morpheme boundary. If situated at a morpheme boundary, they are independent graphemes [@altmann_phoneme-grapheme_2008, p. 80].  
* Notice that the uppercase correspondents of ⟨d'⟩ and ⟨tˈ⟩ are ⟨Ď⟩ and ⟨Ť⟩.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Uppercase     Lowercase        Phoneme         Comment
"A"              "a"            "/a/"            ""
"Á"              "á"            "/aː/"           ""
"Ä"              "ä"            "/ɛ/"            ""
"B"              "b"            "/b/"            ""
"C"              "c"            "/ts/"           ""
"Č"              "č"            "/tʃ/"           ""
"D"              "d"            "/d/"            ""
"Ď"              "ď"            "/ɟ/"            ""
"E"              "e"            "/ɛ/"            ""
"É"              "é"            "/ɛː/"           ""
"F"              "f"            "/f/"            ""
"G"              "g"            "/ɡ/"            ""
"H"              "h"            "/ɦ/"            ""
"I"              "i"            "/i/"            ""
"Í"              "í"            "/iː/"           ""
"J"              "j"            "/j/"            ""
"K"              "k"            "/k/"            ""
"L"              "l"            "/l/"            ""
"Ĺ"              "ĺ"            "/lː/"           ""
"L\'"            "ľ"            "/ʎ/"            ""
"M"              "m"            "/m/"            ""
"N"              "n"            "/n/"            ""
"Ň"              "ň"            "/ɲ/"            ""
"O"              "o"            "/ɔ/"            ""
"Ó"              "ó"            "/ɔː/"           ""
"Ô"              "ô"            "/ʊɔ/"           ""
"P"              "p"            "/p/"            ""
"Q"              "q"          "/k/; /kv/"  "/kv/: only in loanwords"  
"R"              "r"            "/r/"            ""
"Ŕ"              "ŕ"            "/rː/"           ""
"S"              "s"            "/s/"            ""
"Š"              "š"            "/ʃ/"            ""
"T"              "t"            "/t/"            ""
"Ť"              "tˈ"           "/c/"            ""
"U"              "u"            "/u/"            ""
"Ú"              "ú"            "/uː/"           ""
"V"              "v"          "/v/; /w/"    "/w/: only in loanwords"  
"W"              "w"            "/v/"            ""
"X"              "x"         "/ks/; /gz/"   "/gz/: only in loanwords"
"Y"              "y"            "/i/"            ""
"Ý"              "ý"            "/iː/"           ""
"Z"              "z"            "/z/"            ""
"Ž"              "ž"            "/ʒ/"            ""
**Digraph**      ""              ""              ""
"Dz"             "dz"           "/dz/"           ""
"Dž"             "dž"           "/dʒ/"           ""
"CH"             "ch"           "/x/"            ""
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("Grapheme" = 2, "", ""))
```

# Lenition Rules

* /ɡ/ spirantizes to [ɣ] [@short_slovak_1993, p. 536]
* /v/ approximates to [ʋ] in syllable onsets preceding a vowel or liquid [@hanulikova_slovak_2010, p. 374]. It also approximates to [ʋ] or [w] in coda positions.
* Long vowels shorten if preceded by a long vowel and glides [@rubach_lexical_1993, pp. 43-4].
* In verbs, the preceding vowel in a vowel cluster deletes [@rubach_lexical_1993, p. 45].

# Misc. Rules

* Velars and coronals palatalize before front vowels and glides [@rubach_lexical_1993, p. 44].
* Alveolar fricatives palatalize and alveolar stops become affricates preceding /j/. However, if such stops are preceded by an alveolar fricative, they palatalize [@rubach_lexical_1993, p. 44].
* There is final devoicing of voiced obstruents [@hanulikova_slovak_2010, p 376]. /ɦ/ is realized as [x] in this case.
* /n/ assimilates to [ɲ] prior to velar sounds and to [m] prior to labials [@hanulikova_slovak_2010, p. 374].
* Obstruents assimilate in voicing (the preceding obstruent must assimilate to the following obstruent) [@rubach_lexical_1993, p. 30]. /ch/ becomes [ɣ] in this case if voicing is required.

# References
