---
title: "Shilha"
author: "Bill Mizgerd"
bibliography: shi.bib
output: html_document
---

Last Updated: 2019-07-09

# Background

**Language Family**: Afro-Asiatic / Berber / Northern / Atlas

* The name of the language is spelled in a variety of ways, including Shilha, Tashelheet, and Tachelhit.
* Shilha is spoken in Morocco.

# Phonology

## Consonants

* All Shilha consonants can be geminated [@ShiRidouane2008, p. 323].
* @ShiApplegate1958 argues for the labialized consonants as allophones that occur before /u/, although he also describes them as having relatively free variation (pp. 2-3).
* @ShiKossmann1996 argues for the absence of /nˤ/ and /ʒˤ/ (p. 467).
* @ShiKossmann1996 argue for velar rather than uvular fricatives (ibid.).
* @ShiKossmann1996 state that no minimal pairs exist between labialized consonants and their non-labialized counterparts (p. 468). I have chosen to include labialized consonants in the rule set for those that have been attested, thus represented below.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial  Alveolar  "Alveolar (pharyngealized)"  Postalveolar  "Postalveolar (pharyngealized)"  Palatal  Velar  Labiovelar  Uvular  Labio-uvular  Pharyngeal   Glottal
Stops                      "b"      "t d"             "tˤ dˤ"                ""                      ""                    ""     "k ɡ"   "kʷ ɡʷ"     "q"      "qʷ"         ""           ""
Fricatives                 "f"      "s z"             "sˤ zˤ"                "ʃ ʒ"                   "ʒˤ"                  ""     ""      ""          "χ ʁ"    "χʷ ʁʷ"      "ħ ʕ"       "h"
Nasals                     "m"      "n"               "nˤ"                   ""                      ""                    ""     ""      ""          ""       ""           ""           ""
Trills                     ""       "r"               "rˤ"                   ""                      ""                    ""     ""      ""          ""       ""           ""           ""
Approximants               ""       "l"               "lˤ"                   ""                      ""                    "j"    ""      "w"         ""       ""           ""           ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Alveolar (pharyngealized)", "Postalveolar", "Postalveolar (pharyngealized)", "Palatal", "Velar", "Labiovelar", "Uvular", "Labio-uvular", "Pharyngeal", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 12)) %>%
  footnote(general = "Note: Phonemes to the left of each cell are voiceless and phonemes to the right of each cell are voiced.", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* [ə] is added through epenthesis in some contexts, but is not a phoneme in its own right [@ShiKossmann1996, p. 468].
* There are no diphthongs or vowel clusters in Shilha [@ShiKossmann1996, p. 463].
* @ShiApplegate1958 attests two tonal contours, but does not appear to find them contrastive (p. 6).

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
          Front   Central   Back
High      "i"     ""        "u"
Low       ""      "a"       ""
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered")
```

# Alphabet

* **Note:** /ʒˤ/, /nˤ/, /rˤ/, and /lˤ/ are not represented in Crúbadán.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme               Comment
"a"              "/a/"                  ""
"b"              "/b/"                  ""
"d"              "/d/"                  ""
"ḍ"              "/dˤ/"                 ""
"e"              "[ə]"     "⟨e⟩ will go to null in the ruleset due to [ə]\'s non-phonemic status in Shilha"  
"ɛ"              "/ʕ/"                  ""
"f"              "/f/"                  ""
"g"              "/ɡ/"                  ""
"ġ"              "/ʁ/"                  ""
"h"              "/h/"                  ""
"ḥ"              "/ħ/"                  ""
"h̬"              "/χ/"          "in some orthographies"  
"i"              "/i/"                  ""
"j"              "/ʒ/"                  ""
"j̣"              "/ʒˤ/"                 ""
"k"              "/k/"                  ""
"l"              "/l/"                  ""
"ḷ"              "/lˤ/"                 ""
"m"              "/m/"                  ""
"n"              "/n/"                  ""
"ṇ"              "/nˤ/"                 ""
"q"              "/q/"                  ""
"r"              "/r/"                  ""
"ṛ"              "/rˤ/"                 ""
"s"              "/s/"                  ""
"ṣ"              "/sˤ/"                 ""
"š"              "/ʃ/"                  ""
"t"              "/t/"                  ""
"ṭ"              "/tˤ/"                 ""
"u"              "/u/"                  ""
"w"              "/w/"                  ""
"ʷ"               ""          "marks labialization of preceding consonant"  
"x"              "/χ/"          "in some orthographies"
"y"              "/j/"                  ""
"z"              "/z/"                  ""
"ẓ"              "/zˤ/"                 ""
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* Some dialects spirantize bilabial and velar stops; other dialects spirantize the alveolar stops [@ShiRidouane2008, p. 323].

# Misc. Rules

* Epenthetic [ə] occurs interconsonantally, except in VCCV or #CCV sequences, where adding [ə] would yield an open syllable [@ShiKossmann1996, p. 463].
    - Where there exist more than two consecutive consonants, epenthesis first occurs between the final two consonants in the cluster, and proceeds backwards, adding [ə] in any permissible positions. For instance, the word /iʃrz/ becomes [iʃrəz] rather than \*[iʃərəz] or \*[iʃərz] (ibid.).
* A few consonants undergo fortition when they lengthen. /dː/ devoices to [tː], /ʁː/ despirantizes and devoices to [qː], and /wː/ deapproximantizes to [ɡʷː] [@ShiKossmann1996, p. 465].

# References
