---
title: "Maldivian"
author: "Emily Strand"
bibliography: dv.bib
output: html_document
---

Last Updated: 2020-06-29

# Background

**Language Family**: Indo-European / Indo-Iranian / Indo-Aryan / Sinhalese-Maldivian

* Also referred to as Dhivehi, Maldivian is spoken in a multitude of places, but primarily in the Republic of Maldives, a country in South Asia southwest of Sri Lanka.

# Phonology

## Consonants

* Although the glottal stop appears in Maldivian, it is not considered phonemic [@DeSilva1969, p. 204]. Rather, only appearing word-finally, it is treated as an allophone of /k/.
* The palatal nasal appears as well; however, @DeSilva1969 explains that it is not phonemic. @Cain2000 explain that it is a fusion of /n̪/ and /i/, and I have chosen to treat the corresponding grapheme as such.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial    Dental   Retroflex   Palatal    Velar   Glottal
"Stops (plain)"            "p b"      "t̪ d̪"     "ʈ ɖ"      "c ɟ"     "k ɡ"   ""
"Stops (prenasalized)"     "ᵐb"       "ⁿd̪"      "ᶯɖ"        ""       "ᵑɡ"    ""
Fricatives                 "f"        "s̪ z̪"      "ʂ"        ""        ""     "h"
Nasals                     "m"        "n̪"        ""         ""        ""     ""
Flaps                      ""         ""         "ɽ"        ""        ""     ""
Approximants               "ʋ"        "l̪"        "ɭ"        "j"       ""     ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Dental", "Retroflex", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  footnote("Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* Vowel length is contrastive in Maldivian [@DeSilva1969, p. 204]. Long vowels are indicated by duplicate vowel diacritics called fili.
* Although many diphthongs may occur [@DeSilva1969, p. 205], only /ai/ and /au/ are considered to be complex nuclei [@Cain2000, p. 11].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
          Front   Central   Back
High      "i"     ""        "u"
Mid       "e"     ""        "o"
Low       ""      "a"       ""
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered")


diphthongs <- read.table(textConnection('
Diphthongs
"/ai/, /au/"
'), TRUE)

kable(diphthongs, align = 'c') %>%
  kable_styling("bordered")
```

# Alphabet

* The Thaana script is used to represent Maldivian [@Cain2000, p. 9]. It is read from right to left.
* Consonant graphemes do not carry an inherent vowel [@Cain2000, p. 9]. The vowels are represented as diacritics above or below the consonant.
* Diphthongs are represented with an inflected alifu ⟨އ⟩ following a consonant with a fili, or with two fili on an alifu (one above and one below) [@Gnanadesikan2012, pp. 94, 96].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
"Consonant Grapheme"        Phoneme                    Comment
"ހ"                          "/h/"                       ""
"ށ"                         "/ʂ/"                        ""
"ނ"                         "/n̪/"                        ""
"ރ"                         "/ɽ/"                        ""
"ބ"                         "/b/"                        ""
"ޅ"                         "/ɭ/"                        ""
"ކ"                         "/k/"                        ""
"އ"                          ""       "called an alifu, this grapheme needs inflection from fili to have phonemic value [@Cain2000, p. 9]"
"ވ"                         "/ʋ/"                        ""
"މ"                         "/m/"                        ""
"ފ"                         "/f/"                        ""
"ދ"                         "/d̪/"                        ""
"ތ"                         "/t̪/"                        ""
"ލ"                         "/l̪/"                        ""
"ގ"                         "/ɡ/"                        ""
"ޏ"                         "/n̪i/"                       ""
"ސ"                        "/s̪/"                        ""
"ޑ"                         "/ɖ/"                        ""
"ޒ"                         "/z̪/"                        ""
"ޓ"                         "/ʈ/"                        ""
"ޔ"                         "/j/"                        ""
"ޕ"                         "/p/"                        ""
"ޖ"                         "/ɟ/"                        ""
"ޗ"                         "/c/"                        ""
**Digraph**                  ""                          ""
"ނބ"                      "/ᵐb/"                        ""
"ނދ"                      "/ⁿd̪/"                        ""
"ނޑ"                      "/ᶯɖ/"                        ""
"ނގ"                      "/ᵑɡ/"                        ""
"**Vowel Grapheme**"        ""                           ""
"އަ"                        "/a/"                         ""
"އާ"                       "/aː/"                         ""
"އި"                        "/i/"                         ""
"އީ"                       "/iː/"                         ""
"އު"                        "/u/"                         ""
"އޫ"                       "/uː/"                         ""
"އެ"                       "/e/"                          ""
"އޭ"                       "/eː/"                         ""
"އޮ"                       "/o/"                          ""
"އޯ"                       "/oː/"                         ""
"އް"                        ""               "this grapheme is a geminate marker or /k/ word-finally (see Misc. Rules below)"
"**Vowel Diacritic**"        ""                            ""
"ަ"                        "/a/"                          ""
"ާ"                       "/aː/"                          ""
"ި"                        "/i/"                          ""
"ީ"                       "/iː/"                          ""
"ު"                       "/u/"                           ""
"ޫ"                       "/uː/"                          ""
"ެ"                       "/e/"                           ""
"ޭ"                       "/eː/"                          ""
"ޮ"                       "/o/"                           ""
"ޯ"                       "/oː/"                          ""
"ް"                        ""           "called a sukun, this diacritic indicates that the base character is vowel-less [@Cain2000, p. 9]"
'), TRUE)

kable(alphabet, col.names = c("Consonant Grapheme", "Phoneme", "Comment"), align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Syllables in Maldivian generally follow the order of CV [@Cain2000, p. 11]; however, the following structures are present:
    - V
    - VC
    - CVC
    - CVV
    - VV
    - CVVC
    - VVC
* Consonant clusters only appear word-medially (ibid.).

# Lenition Rules

* /k/, /ʂ/, and /t̪/ debuccalize to glottal stops prepausally [@Cain2000, pp. 11-12].
* /s̪/ becomes /h/ morpheme-finally preceding a vowel-initial suffix [@Gnanadesikan2016, p. 44].

# Misc. Rules

* After /k/, /ʂ/, and /t̪/ debuccalize, the allophonic glottal stop becomes a velar nasal word-finally preceding word-initial vowels or /h/ [@Cain2000, pp. 11-12].
* Consonants, except /f/, /ʂ/, and /h/, geminate [@Cain2000, p. 13]. This is represented orthographically by duplicate graphemes, with a sukun on the first consonant (no inflection for vowel). They may also be represented by an alifu sukun followed by a consonant. Otherwise, the empty vowel marker with a sukun indicates a word-final /k/.
* Word-initial /ʋ/ becomes [w] preceding /a/ [@Cain2000, p. 7]. It also becomes [w] following /u/.
* Nasals assimilate in place to following consonants (excluding /h/), even across word boundaries [@Cain2000, p. 12].
* Morpheme-final /u/ deletes preceding a vowel-initial suffix  [@Gnanadesikan2016, p. 44].
* Consonants /t̪/, /d̪/, /ⁿd̪/, /n̪/, and /l̪/ palatalize preceding /i/ [@Gnanadesikan2016, p. 46]. In doing so they become geminates.
* /h/ and /ʂ/ become geminated [s] and [t] preceding /i/ [@Gnanadesikan2016, pp. 45-46].

# References
