---
title: "Arosi"
author: "Emily Strand"
bibliography: aia.bib
output: html_document
---

Last Updated: 2020-06-24

# Background

**Language Family**: Austronesian / Malayo-Polynesian / Central-Eastern / Eastern Malayo-Polynesian / Oceanic / Central-Eastern Oceanic / Southeast Solomonic / Malaita-San Cristobal / San Cristobal

* Arosi is spoken in the Makira province of the Solomon Islands.

# Phonology

## Consonants

* It's interesting that the Arosi language includes the glide /w/ but not /j/. Although @Capell1971 [as cited in @Lynch2001, p. 562] includes /j/ as a separate phoneme (along with /kʷ/ and /ɡʷ/), @Lynch2001 explain that, due to a lack of evidence, they are not explicitly contrastive, but rather allophones of intervocalic /i/, /k/, and /ɡ/, respectively.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"    Bilabial    Alveolar  Velar   Glottal
Stops                     "p pʷ b bʷ"    "t d"    "k ɡ"     "ʔ"
Fricatives                     ""         "s"      ""       "h"
Nasals                       "m mʷ"       "n"      "ŋ"      ""
Flaps                          ""         "ɾ"      ""       ""
Approximants                   "w"        ""       ""       ""  
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Bilabial", "Alveolar", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 4)) %>%
  footnote(general = "Note: The alveolar and velar stops are ordered voiceless and voiced. Phonemes that have the diacritic (ʷ) are labialized.", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* Vowel length is contrastive in Arosi [@Lynch2001, p. 562]. Long vowels are indicated by duplicate graphemes.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
          Front   Central   Back
High     "i"      ""        "u"
Mid      "e"      ""        "o"
Low       ""      "a"       ""
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```  

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme         Phoneme
"a"               "/a/"
"b"               "/b/"
"d"               "/d/"
"e"               "/e/"
"g"               "/ɡ/"
"h"               "/h/"
"i"               "/i/"
"k"               "/k/"
"m"               "/m/"
"n"               "/n/"
"o"               "/o/"
"p"               "/p/"
"r"               "/ɾ/"
"s"               "/s/"
"t"               "/t/"
"u"               "/u/"
"w"               "/w/"
"\'"              "/ʔ/"
**Digraph**        ""
"pw"              "/pʷ/"
"bw"              "/bʷ/"
"mw"              "/mʷ/"
"ng"              "/ŋ/"
'), TRUE)

kable(alphabet, align = "c") %>%
  kable_styling("bordered")
```

# Syllable Structure

* Syllables in Arosi consist of either /V/ or /CV/ [@Capell1971, p. 6].
* Consonant clusters are not permitted in Arosi (ibid.).


# Misc. Rules

* /k/ and /ɡ/ are labialized preceding back vowels [@Lynch2001, p. 562].
* /i/ is realized as [j] intervocalically (ibid.).
* /t/ is realized as [s] preceding /i/ (ibid).

# References
