---
title: "Mussau-Emira"
author: "Bill Mizgerd"
bibliography: emi.bib
output: html_document
---

Last Updated: 2019-07-23

# Background

**Language Family**: Austronesian / Malayo-Polynesian / Eastern Malayo-Polynesian / Oceanic / St. Matthias / Mussau-Emira

* Mussau-Emira is spoken in Papua New Guinea, specifically in the province of New Ireland.

# Phonology

## Consonants

* Consonant gemination occurs contrastively [@EmiBrownie2007, p. 13].
* @EmiBrownie2007 argues for the flap /ɾ/ rather than /r/ (p. 13).
* @EmiRoss2002 argues that the phonemes I have listed as voiced fricatives are actually stops in their underlying forms (p. 149).

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Bilabial   Alveolar   Velar
Stops                       "p"        "t"        "k"
Fricatives                  "β"        "s"        "ɣ"
Nasals                      "m"        "n"        "ŋ"
Trills                      ""         "r"        ""
Approximants                ""         "l"        ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Bilabial", "Alveolar", "Velar"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 3)) %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* Vowel length is contrastive [@EmiBrownie2007, p. 13].
* @EmiRoss2002 argues for the more traditional five-vowel system of /i u e o a/ (p. 149).
* Because Crubadan has instances of three vowel sequences, leaving diphthongs to be determined by syllable structure, they aren't included in the rule set.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
vowels <- read.table(textConnection('
       Front   Back
High   "i"     "u"
Mid    "ɛ"     "ɔ"
Low    ""      "ɑ"
'), TRUE)

kable(vowels, align = ('c')) %>%
  kable_styling("bordered")


diphthongs <- read.table(textConnection('
Diphthongs
"/ɑi/, /ɑu/, /ɑɛ/, /ɑɔ/, /ɛi/, /ɔi/, /ɔu/"
'), TRUE)

kable(diphthongs, align = 'c') %>%
  kable_styling("bordered")
```

# Alphabet

* Gemination and vowel length are written by doubling the relevant character.
* The Seventh-Day Adventist orthography varies slightly, using ⟨q⟩ for /ɣ/ and ⟨g⟩ for /ŋ/ [@EmiRoss2002, p. 151]; I have opted not to support that orthography by default, given that it is unused by the Crubadan corpus.


```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme   
"a"           "/ɑ/"  
"b"           "/β/"  
"e"           "/ɛ/"  
"g"           "/ɣ/"   
"i"           "/i/"  
"k"           "/k/"  
"l"           "/l/"  
"m"           "/m/"  
"n"           "/n/"   
"o"           "/ɔ/"  
"p"           "/p/"  
"r"           "/r/"  
"s"           "/s/"  
"t"           "/t/"  
"u"           "/u/"  
"v"           "/β/"
**Digraph**    ""  
"gh"          "/ɣ/"
"ng"          "/ŋ/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* Final vowels are frequently dropped [@EmiBrownie2007, p. 15; @EmiBrownie2009, p. 11].
* In casual speech, syncope of vowels occurs between identical consonants [@EmiBrownie2007, p. 16].

# Misc. Rules

* Voiced fricatives become stops when geminated, and sometimes when word-initial [@EmiBrownie2007, p. 13; @EmiRoss2002, p. 149].

# References
