---
title: "Maltese"
author: "Bill Mizgerd"
bibliography: mt.bib
output: html_document
---

Last Updated: 2020-07-01

**COMPROMISED: conflation between /ʃ/ and /ʒ/ and between /ts/ and /dz/**

# Background

**Language Family:** Afro-Asiatic / Semitic / Central / South / Arabic

* Maltese is spoken in Malta and throughout the European Union.

# Phonology

## Consonants

* /ʒ/ and /dz/ are commonly included in the phonemic inventory of Maltese through loanword integration; however, their native voiceless counterparts are represented by the same graphemes, which compromised the language [@MtBorg1997, p. 248]. The ruleset only accounts for the voiceless variants.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial     Dental   Alveolar    Postalveolar   Palatal     Velar     Pharyngeal     Glottal
Stops                      "p  b"     "t̪  d̪"      ""           ""           ""       "k  ɡ"        ""           "ʔ"
Affricates                   ""        ""      "ts  dz"     "tʃ  dʒ"        ""         ""           ""           ""
Fricatives                 "f  v"      ""       "s  z"       "ʃ  ʒ"         ""         ""           "ħ"          ""
Nasals                       "m"       ""         "n"          ""           ""         ""           ""           ""
Trills                       ""        ""         "r"          ""           ""         ""           ""           ""
Approximants                 ""        ""         "l"          ""           "j"        "w"          ""           ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Dental", "Alveolar", "Postalveolar", "Palatal", "Velar", "Pharyngeal", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 8)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```  

## Vowels

* Vowel length is contrastive [@MtBorg1997, p. 251].
* Authors seem to vary in terms of the vowel inventory:
    - @MtHume1996:             /a/ /e/ /i/ /o/ /u/  
    - @MtOrtho1996:            /a/ /ɛ/ /i/ /o/ /u/  
    - @MtBorg1997:             /a/ /e/ /i/ /o/ /u/  
    - @MtHumeEtAl2009:         /ɐ/ /ɛ/ /i/ /ɔ/ /u/  
    - @MtPuech2016:            /a/ /ɛ/ /ɪ/ /ɔ/ /ʊ/
    - NB: @MtHumeEtAl2009 only use phonetic transcriptions, rather than phonemic, so it is ambiguous whether her choices of vowels reflect phonemes rather than realizations.
* Vowels with grave accents (´) tend to appear word-finally, denoting stress; however, we do not account for stress, so accented vowels in this position will be transcribed to their plain representations.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
            Front      Central       Back
High       "i  iː"       ""         "u  uː"
Near-High   "ɪː"         ""           ""
Mid         "e"          ""           "o"
Low          ""          "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)


diphthongs <- read.table(textConnection('
Diphthongs
"/aj/, /ej/, /oj/, /aw/, /ew/, /iw/, /ow/"
'), TRUE)

kable(diphthongs, align = 'c') %>%
  kable_styling("bordered")
```

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme                   Comment
"a"            "/a/"             "some authors use /ɐ/"
"b"            "/b/"                      ""
"ċ"            "/tʃ/"                     ""
"d"            "/d̪/"                      ""
"e"            "/e/"             "some authors use /ɛ/"
"f"            "/f/"                      ""
"ġ"            "/dʒ/"                     ""
"g"            "/ɡ/"                      ""  
"h"            "/ː/"         "lengthens associated vowels"  
"ħ"            "/ħ/"   "realized by some speakers as [x] or [h]"    
"i"         "/i/; /iː/"                   ""
"j"            "/j/"                      ""
"k"            "/k/"                      ""
"l"            "/l/"                      ""
"m"            "/m/"                      ""
"n"            "/n/"                      ""
"o"            "/o/"             "some authors use /ɔ/"  
"p"            "/p/"                      ""
"q"            "/ʔ/"                      ""
"r"            "/r/"                      ""
"s"            "/s/"                      ""
"t"            "/t̪/"                      ""
"u"         "/u/, /uː/"          "some authors use /ʊ/"
"v"            "/v/"                      ""
"w"            "/w/"                      ""
"x"            "/ʃ/"         "/ʒ/ appears in certain loanwords"  
"ż"            "/z/"                      ""
"z"            "/ts/"       "/dz/ appears in certain loanwords"
**Digraph**     ""                        ""
"għ"           "/ː/"         "lengthens associated vowels"
"ie"           "/ɪː/"           "some authors use /ie/"
"aj"          "/aj/"                      ""
"aw"          "/aw/"                      ""
"ew"          "/ew/"         "occasionally realized as [ow]"  
"għi"         "/ej/"                      ""
"għu"         "/ow/"                      ""
"iw"          "/iw/"                      ""
"oj"          "/oj/"                      ""
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Misc. Rules

* Clusters of obstruents assimilate to be either fully voiced or fully voiceless [@MtBorg1997, p. 250].
* Consonants geminate [@MtBorg1997, pp. 251-252].
    - Word-final long consonants are sometimes shortened (ibid.).
* In the definite article, ⟨il-⟩, the /l/ typically assimilates to subsequent coronal segments (except /dʒ/) [@MtBorg1997, p. 255].
* Long vowels are shortened word-finally [@MtBorg1997, p. 267].
* Metathesis of the definite article occurs before Italian-derived words with initial sibilants [@MtBorg1997, p. 273].
* The ⟨h⟩ in the suffixed pronouns (⟨-ha⟩, ⟨-hom⟩) is replaced by a glide after stressed stem-final vowels [@MtBorg1997, p. 261].
* Elision of short vowels occurs in unstressed open syllables [@MtBorg1997, p. 273].
* CCC clusters are pronounced as CVCC when the middle C is a sonorant [@MtBorg1997, p. 274].
* ⟨għ⟩ represents the Old Arabic phones [ʕ] or [ɣ], which were deleted at some point during the evolution of modern Maltese [@MtBorg1997, p. 261].
* ⟨h⟩ represents the Old Arabic phone [h], which was deleted at some point during the evolution of modern Maltese [@MtBorg1997, p. 259].

# References
