---
title: "Rikbaktsa"
author: "Emily Strand"
bibliography: rkb.bib
output: html_document
---

Last Updated: 2020-02-04

# Background

**Language family**: Macro-Ge / Rikbaktsa

* Rikbaktsa is spoken in Brazil.

# Phonology

## Consonants

* Both @Sheila2007 and @Silva2005 [as cited in @Lobato2018, p. 7] make the distinction between a dental and an alveolar flap (p. viii). However, it is unclear how to mark this phoneme given that /ɾ/ can represent both. @Silva2005 [as cited in @Lobato2018, p. 7] represents it with the retroflex flap, but this doesn't accurately reflect the place of articulation. I have chosen to represent this phoneme with the alveolar flap, with the dental diacritic below.
* @Boswood1978 [as cited in @Lobato2018, p. 7] doesn't include /tʃ/ in the phonemic inventory, based on the interpretation of the corresponding digraph ⟨ts⟩ as two separate consonants. I have chosen to follow the analysis of the more current source, including /tʃ/ below.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial   Dental  Alveolar  Postalveolar     Palatal    Velar   Glottal   
Stops                      "p b"      ""     "t d"        ""             ""        "k"      ""
Affricates                  ""        ""      ""         "tʃ"            ""        ""       ""
Fricatives                  ""        ""      ""         "ʃ"             ""        ""       "h"
Nasals                      "m"       ""      "n"         ""             ""        ""       ""
Trills                      ""        "ɾ̪"     "ɾ"         ""             ""        ""       ""
Approximants                "w"       ""      ""          ""             "j"       ""       ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Dental", "Alveolar", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 7)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```

# Vowels

* Nasal vowels are contrastive in Rikbaktsa [@Sheila2007, p. viii]. They are indicated with a tilde or with a following nasal consonant (the tilde is omitted in these cases).

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
           Front      Central       Back
High        "i"         "ɨ"          "u"
Mid         "e"         ""           "o"
Low         ""          "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)

```

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme       Phoneme            Comment
"a"             "/a/"               ""
"y"             "/ɨ/"               ""
"b"             "/b/"               ""
"d"             "/d/"               ""
"e"             "/e/"               ""
"h"             "/h/"               ""
"i"             "/i/"               ""
"j"             "/j/"               ""
"k"             "/k/"               ""
"m"             "/m/"               ""
"n"             "/n/"               ""
"o"             "/o/"               ""
"p"             "/p/"               ""
"r"             "/ɾ/"               ""
"s"             "/ʃ/"               ""
"t"             "/t/"               ""
"u"             "/u/"               ""
"w"             "/w/"               ""
"z"             "/ɾ̪/"               ""
**Digraph**      ""                 ""
"ts"            "/tʃ/"              ""
"**Orthographic Variation**" ""     ""
"g"             "/k/"         "preceding voiced consonants; intervocalically"
"c"             "/k/"               ""
"qu"            "/k/"               ""
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* Given that ⟨g⟩ represents /k/ preceding voiced consonants and intervocalically, it would support the lenition process where /k/ becomes voiced in these positions [@Sheila2007, p. viii].

# Misc. Rules

* Vowel epenthesis occurs preceding both the dental and alveolar flap when they occur word-initially or following /k/ [@Sheila2007, p. viii].

# References
