---
title: "Mam"
author: "Emily Strand"
bibliography: mam.bib
output: html_document
---

Last updated: 2020-03-02

# Background

**Language Family:** Mayan / Quichean-Mamean / Greater Mamean / Mamean

* Mam is spoken in Guatemala.

# Phonology

## Consonants

* /b/, /d/, /ɡ/, and /f/ appear in Spanish loanwords [@Baayil2001, pp. 17-18].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)    


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial    Alveolar  Postalveolar  Retroflex  Palatal  Velar    Uvular   Glottal
Stops                      "p ɓ"     "t tʼ"    ""            ""         "c cʼ"   "k kʼ"   "q qʼ"   "ʔ"
Affricates                 ""        "ts tsʼ"  "tʃ tʃʼ"      "ʈʂ ʈʂʼ"   ""       ""       ""       ""
Fricatives                 ""        "s"       "ʃ"           "ʂ"        ""       ""       "χ"      ""
Nasals                     "m"       "n"       ""            ""         ""       ""       ""       ""
Trills                     ""        "r"       ""            ""         ""       ""       ""       ""
Approximants               "w"       "l"       ""            ""         "j"      ""       ""       ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Retroflex", "Palatal", "Velar", "Uvular", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 8)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are plain and those on the right are ejectives. The labial stop on the right is an implosive.", general_title = "")
```

## Vowels

* Vowel length is contrastive in Mam [@Baayil2001, p. 6]. Long vowels are represented by duplicate graphemes.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High     "i"          ""           "u"         
Mid      "e"          ""           "o"
Low      ""           "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```  

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme
"a"            "/a/"
"b\'"          "/ɓ/"
"e"            "/e/"
"i"            "/i/"
"j"            "/χ/"                   
"k"            "/k/"
"k\'"          "/kʼ/"
"l"            "/l/"
"m"            "/m/"
"n"            "/n/"
"o"            "/o/"
"p"            "/p/"
"q"            "/q/"
"q\'"          "/qʼ/"
"r"            "/r/"
"s"            "/s/"
"t"            "/t/"
"t\'"          "/tʼ/"
"u"            "/u/"
"w"            "/w/"
"x"            "/ʂ/"
"y"            "/j/"
"\'"           "/ʔ/"
**Digraph**    ""
"ch"           "/tʃ/"
"ch\'"         "/tʃʼ/"
"ky"           "/c/"
"ky\'"         "/cʼ/"
"tx"           "/ʈʂ/"
"tx\'"         "/ʈʂʼ/"
"tz"           "/ts/"
"tz\'"         "/tsʼ/"
"xh; ẍ"        "/ʃ/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* /tʼ/ tends to vary freely with /ɗ/ [@Baayil2001, p. 11].

# Misc. Rules

* In some dialects of Mam, the palatalized stop and ejective are the result of predictable palatalization [@Baayil2001, p. 10].
* A phonetic glottal stop occurs preceding word-initial vowels (ibid.).
* /n/ assimilates (not always) in place to following consonants [@Baayil2001, p. 15].

# References
