---
title: "Kagulu"
author: "Emily Strand"
bibliography: kki.bib
output: html_document
---

Last updated: 2019-10-31

# Background

**Language family**: Niger-Congo / Atlantic-Congo / Volta-Congo / Benue-Congo / Bantoid / Southern / Narrow Bantu / East / Northeast Savanna Bantu / Northeast Coastal Bantu / Ruvu

* Kagulu is spoken in Eastern Tanzania.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial      Alveolar    Postalveolar   Palatal     Velar     Glottal
Stops                    "p  b  ᵐb"   "t  d  ⁿd"        ""          ""     "k  ɡ  ᵑɡ"     ""
Affricates                   ""          ""       "tʃ  dʒ  ⁿdʒ"     ""         ""         ""
Fricatives                  "f"         "s"            "ʃ"          ""         ""         "h"     
Nasals                     "m̥  m"      "n̥  n"           ""          "ɲ"      "ŋ̥  ŋ"       ""
Approximants                "w"         "l"            "j"          ""         ""         ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: Where three phonemes share a cell, they are ordered voiceless, voiced, prenasalized. Where two phonemes share a cell, they are ordered voiceless and voiced.", general_title = "")
```

## Vowels

* Unlike other Bantu languages, Kagulu does not possess phonemic tone [@Petzell2008, p. 40].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         ""           "u"
Mid       "e"         ""           "o"
Low        ""         "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```   

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme  
"a"             "a"
"b"             "b"
"d"             "d"
"e"             "e"
"f"             "f"
"g"             "ɡ"
"h"             "h"
"i"             "i"
"j"             "dʒ"
"k"             "k"
"l"             "l"
"m"             "m"
"n"             "n"
"o"             "o"
"p"             "p"
"s"             "s"
"t"             "t"
"u"             "u"
"w"             "w"
"y"             "j"
**Multigraph**  ""
"ch"            "tʃ"
"mb"            "ᵐb"
"mh"            "m̥"
"nd"            "ⁿd"
"nj"            "ⁿdʒ"
"nh"            "n̥"
"ng\'"           "ŋ"
"ng"            "ᵑɡ"
"ng\'h"          "ŋ̥"
"ny"            "ɲ"
"sh"            "ʃ"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Misc. Rules

* Excluding the digraphs listed above, consonant clusters are not permitted in Kagulu [@Petzell2008, p. 41]. Adjacent consonants (as in loanwords) are separated by an epenthetic vowel.
* Adjacent vowels are also not allowed in Kagulu [@Petzell2008, pp. 42-45]. If this occurs, such that /i/ or /u/ precedes a vowel, they become the glides [j] and [w]. Also, if this occurs through affixation, the first vowel may be deleted or the two vowels coalesce.
* Vowel harmony is present in the language [@Petzell2008, p. 45].

# References
