---
title: "Korean"
author: "Justin Bai"
bibliography: ko.bib
output: html_document
---
# Background

**Language family**: Language isolate

* Korean a language isolate, but it is heavily influenced by Mandarin.
* Korean is spoken in Korea, but also in China, Russia, and other regions of Central Asia.

# Phonology

## Consonants

* There is some disagreement within the literature about the nature of the affricates, with some sources considering it postalveolar [@lee1993korean, p. 28] and others considering it alveopalatal [@ahn2009korean, p. 26; @brown2015handbook, p. 7]. We have decided to categorized the affricate as postalveolar.
* @lee1993korean explains that /j/ and /w/ are not separate phonemes, only appearing as a by-product of diphthongs (p. 30); however, I have chosen to include them in the inventory just as @ahn2009korean and @Kim1999 do (p. 41; p. 315).

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Bilabial      Alveolar   Postalveolar   Palatal     Velar         Glottal
Stops                     "p  p͈  pʰ"   "t  t͈  tʰ"       ""            ""     "k  k͈  kʰ"        ""
Affricates                   ""            ""      "tʃ  t͈ʃ  tʃʰ"      ""        ""             ""
Fricatives                   ""          "s  s͈"         ""            ""        ""             "h"
Nasals                       "m"           "n"          ""            ""        "ŋ"            ""
Approximants                 ""            "l"          ""            "j"       "w"            ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Bilabial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: Where three phonemes share a cell, they are ordered plain, tense, and aspirated.", general_title = "")
```

## Vowels

* In some dialects of Korean, /e/ and /ɛ/ have merged [@choo2003sounds, pp. 11-12]. Our rules preserve the contrast.
* Vowel length is traditionally contrastive, but it is not obvious from the orthography [@KoreanGrammarLee]. So, our rules will only output normal length vowels (and no long vowels). This reflects the Seoul dialect as spoken by young speakers who do not consistently distinguish length.
* There are two types of diphthongs marked in the Korean orthography.
    - One type of diphthong is sometimes referred to as "y" diphthongs because they have a /j/ sound [@choo2003sounds].
    - The other type of diphthong is the "w" diphthong.
    - However, these can be interpreted as vowel-glide sequences [@Berrjod2014, p. 18], so they will be separated in the transcription.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"        "ɨ"          "u"
High-Mid "e  ø"      ""           "o"
Low-Mid   "ɛ"        ""           "ʌ"
Low       ""         "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```    

# Alphabet

* The Korean alphabet is known as Hangul. The individual letters are known as jamo,
which are composed into syllable blocks.
* There are some notes to be made about how certain sounds have been transcribed
as there is some debate in the literature about how best to transcribe particular
segments in Korean.
    - As for representing a transcription for the tense stops, there are various methods employed within the literature. Some authors use an apostrophe (e.g. /p’/). However, we have decided to use /◌͈/, which is a diacritic used to indicate "strong articulation."
    - ⟨ㅚ⟩ is sometimes transcribed as /wɛ/. However, following the traditional vowel
  system of Korean [@brown2015handbook], our rules transcribe it as /ø/.
    - ⟨ㅡ⟩ has been transcribed in one of two ways within the literature: /ɨ/ or /ɯ/. Our rules use /ɨ/ as "ㅡ is produced with the tongue more forward [than ㅜ]" [@choo2003sounds].
    - ⟨ㅢ⟩ is often transcribed as /ɰi/; however, /ɰi/ is only found at the beginning
  of words. Elsewhere, ⟨ㅢ⟩ represents /i/, which is reflected in our rules [@choo2003sounds].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
"**Consonant Grapheme**" ""
"ㄱ"                   "/k/"
"ㄲ"                   "/k͈/"
"ㅋ"                   "/kʰ/"
"ㄷ"                   "/t/"
"ㄸ"                   "/t͈/"
"ㅌ"                   "/tʰ/"
"ㅂ"                   "/p/"
"ㅃ"                   "/p͈/"
"ㅍ"                   "/pʰ/"
"ㅅ"                   "/s/"
"ㅆ"                   "/s͈/"
"ㅈ"                   "/tʃ/"
"ㅉ"                   "/t͈ʃ/"
"ㅊ"                   "/tʃʰ/"
"ㅁ"                   "/m/"
"ㄴ"                   "/n/"
"ㅇ"                   "/ŋ/"
"ㄹ"                   "/l/"
"ㅎ"                   "/h/"
"Vowel Grapheme"      Phoneme   
"ㅏ"                   "/a/"
"ㅔ"                   "/e/"
"ㅐ"                   "/ɛ/"
"ㅣ"                   "/i/"
"ㅗ"                   "/o/"
"ㅜ"                   "/u/"
"ㅡ"                   "/ɨ/"
"ㅓ"                   "/ʌ/"
"ㅚ"                   "/ø/"
"**Diphthong Grapheme**" ""
"ㅑ"                   "/ja/"
"ㅖ"                   "/je/"
"ㅒ"                   "/jɛ/"
"ㅛ"                   "/jo/"
"ㅠ"                   "/ju/"
"ㅕ"                   "/jʌ/"
"ㅢ"                   "/ɨi/"
"ㅟ"                   "/wi/"
"ㅞ"                   "/we/"
"ㅙ"                   "/wɛ/"
"ㅘ"                   "/wa/"
"ㅝ"                   "/wʌ/"
'), TRUE)

kable(alphabet, col.names = c("Consonant Grapheme", "Phoneme"), align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* Voiceless stops become voiced following sonorants and preceding vowels [@ahn2009korean, pp. 26-31]. They also become voiced intervocalically.
* /l/ is realized as a flap intervocalically [@ahn2009korean, p. 28].

# References
