---
title: "Rawa"
author: "Emily Strand"
bibliography: rwo.bib
output: html_document
---

Last Updated: 2020-04-17

**SLIGHTLY COMPROMISED: conflation between /l/ and /r/**

# Background

**Language Family:** Trans-New Guinea / Main Section / Central and Western / Huon-Finisterre / Finisterre / Gusap-Mot

* Rawa is spoken in the Madang province of Papua New Guinea.
* Rawa has two dialects: Rawa and Karo. The main differences are [@Toland1991, p. 2]:
    - The Karo dialect has voiced stops, whereas the Rawa dialect has aspirated voiceless stops.
    - /l/ is much more frequent than /r/ in the Karo dialect, and /r/ is much more frequent than /l/ in the Rawa dialect.
* I have opted to follow the analysis that includes voiced stops, meaning the Karo dialect is reflected more closely below.

# Phonology

## Consonants

* /h/ is considered a marginal phoneme [@Toland1992, p. 3].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"       Bilabial   Alveolar    Palatal    Velar    Glottal
"Stops (plain)"                 "p b"      "t d"        ""       "k ɡ"      ""
"Stops (prenasalized)"           "ᵐb"       "ⁿd"        ""       "ᵑɡ"       ""
Fricatives                       ""         "s"         ""       ""         "h"
Nasals                           "m"        "n"         ""       "ŋ"        ""
Trills                           ""         "r"         ""       ""         ""
Approximants                     "w"        "l"         "j"      ""         ""

'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Bilabial", "Alveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 5))  %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```

## Vowels

* Vowels do not occur in succession [@Toland1991, p. 2].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
            Front   Central   Back
High         "i"      ""      "u"
High-Mid     "e"      ""      "o"
Low-Mid      ""       ""      "ɔ"
Low          ""       "a"      ""
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

* Depending on the dialect, /r/ can be written with either ⟨l⟩ or ⟨r⟩ [@Toland1992, p. 3]. ⟨r⟩ is said to be used to transcribe /r/ in the Rawa dialect, whereas ⟨l⟩ is said to be used to transcribe /r/ in the Karo dialect. This is somewhat suspect, as the corresponding Crúbadán corpus of the Karo dialect contains instances of both ⟨r⟩ and ⟨l⟩ (only ⟨l⟩ is expected). Given this, I think there is high likelihood of both ⟨r⟩ and ⟨l⟩ being used to represent /r/ in the other Crúbadán corpora for the language as well (the general corpus and the Rawa dialect corpus). This means that conflation between /r/ and /l/ cannot be avoided, resulting in a somewhat compromised language.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme          Comment
"a"              "/a/"             ""
"b"              "/b/"             ""
"d"              "/d/"             ""
"e"              "/e/"             ""
"g"              "/ɡ/"             ""
"h"              "/h/"             ""
"i"              "/i/"             ""
"k"              "/k/"             ""
"l"           "/l/; /r/"    "/l/: default in rule set"
"m"              "/m/"             ""
"n"              "/n/"             ""
"o"              "/o/"             ""
"p"              "/p/"             ""
"r; l"           "/r/"       "⟨r⟩: default in rule set"
"s"              "/s/"             ""
"t"              "/t/"             ""
"u"              "/u/"             ""
"w"              "/w/"             ""
"y"              "/j/"             ""
**Multigraph**    ""               ""
"mb"             "/ᵐb/"            ""
"nd"             "/ⁿd/"            ""
"ng"             "/ŋ/"             ""
"ngg"            "/ᵑɡ/"            ""
"oo"             "/ɔ/"             ""
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Syllables in Rawa generally have a CV structure [@Toland1991, p. 2].
    - Syllables consisting of only a vowel can occur word-initially.

# Lenition Rules

* /h/ may be elided word-initially [@Toland1992, p. 3].

# Misc. Rules

* /r/ may be realized as [l] [@Toland1992, p. 3].

# References
