---
title: "Swahili"
author: "Delphine Morse Mahos"
edited by: "Emily Strand"
bibliography: sw.bib
output: html_document
---

Last Updated: 2020-01-10

**Slightly Compromised: possible conflation among syllabic nasals and non-syllabic counterparts**

# Background

**Language family**: Niger-Congo / Atlantic-Congo / Volta-Congo / Benue-Congo / Bantoid / Southern / Narrow Bantu / Central / G / Swahili (G. 40)

* Also referred to as Kiswahili, it is spoken throughout Africa.

# Phonology

## Consonants

* There is some controversy around the phonemic status of aspirated stops. @Polome1967 states that there are inconsistencies among dialects, where some maintain the distinction and others do not (pp. 39-41). He also states that the contrast is neutralizing, especially among educated speakers, largely dictated by stress patterns and word placement. I have chosen to not include them in the phonemic inventory below.
* @Polome1967 states that the phonemes /θ/, /ð/, and /ɣ/ appear only in loanwords; however, they have become part of the phonemic system, especially in that of educated speakers (more uneducated speakers or speakers that are less influenced by Arabic substitute native phonemes) (p. 42). I have chosen to include them below.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial   Dental   Alveolar  Postalveolar  Palatal     Velar   Glottal
Stops                     "p b ᵐb"   ""      "t d ⁿd"       ""         ""      "k ɡ ᵑɡ"    ""
Affricates                 ""        ""         ""      "tʃ dʒ ⁿdʒ"    ""         ""       ""
Fricatives                "f v ᶬv"   "θ ð"    "s z ⁿz"      "ʃ"        ""        "x ɣ"     "h"
Nasals                     "m"       ""         "n"         ""         "ɲ"        "ŋ"      ""
Trills                     ""        ""         "r"         ""         ""         ""       ""
Approximants               ""        ""         "l"         ""         "j"        "w"      ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Dental", "Alveolar", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 7)) %>%
  footnote("Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced. Where three phonemes share a cell, they are ordered voiceless, voiced, and prenasalized.", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* Double vowels occur; however, they are independent of one another and do not constitute a single (long) phoneme [@Maw1999, p. 44].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
          Front    Back
High      "i"      "u"
Mid       "ɛ"      "ɔ"
Low       ""       "ɑ"
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme
"a"              "/ɑ/"
"b"              "/b/"
"d"              "/d/"
"e"              "/ɛ/"
"f"              "/f/"
"g"              "/ɡ/"
"h"              "/h/"
"i"              "/i/"
"j"              "/dʒ/"
"k"              "/k/"
"l"              "/l/"
"m"              "/m/"
"n"              "/n/"
"o"              "/ɔ/"
"p"              "/p/"
"r"              "/r/"
"s"              "/s/"
"t"              "/t/"
"u"              "/u/"
"v"              "/v/"
"w"              "/w/"
"y"              "/j/"
"z"              "/z/"
**Digraph**       ""
"mb"             "/ᵐb/"
"nd"             "/ⁿd/"
"nj"             "/ⁿdʒ/"
"ny"             "/ɲ/"
"mv"             "/ᶬv/"
"nz"             "/ⁿz/"
"ng"             "/ᵑɡ/"
"ng\'"           "/ŋ/"
"ch"             "/tʃ/"
"dh"             "/ð/"
"gh"             "/ɣ/"
"kh"             "/x/"
"sh"             "/ʃ/"
"th"             "/θ/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Syllables in Swahili have the following structures [@mohamed2001, pp. 11-12]:
    - V
    - CV
    - CCV
    - CCCV
* Although @mohamed2001 also indicates that a syllable can consist of just a nasal (p. 11), @Polome1967 states that such occurrences are merely variants of the non-syllabic nasals, predictable by stress patterns (p. 43). However, there are occurrences where syllabic /m/ and non-syllabic /m/ contrast, making the distinction between (ex. /mb/ and /m b/) unpredictable [@Polome1967, p. 43].

# Lenition Processes

* /n/ becomes the glide /j/ preceding a noun or adjective with a vowel stem [@mohamed2001, p. 24].
* /l/ deletes preceding /u/ [@Meinhof1932, p. 118].
* /v/ becomes the glide /w/ preceding /i/ (ibid.).
* /r/ is often realized as a flap [@Polome1967, p. 44].
* The contrast between voiced and voiceless consonants is sometimes (unpredictably) neutralized [@Polome1967, p. 46].

# Misc. Rules

* The voiced stops vary freely with their implosive variants [@Polome1967, p. 41].
* /v/ and /z/ vary freely in some unpredictable environments [@Polome1967, p. 46].

# References
