---
title: "Aymara"
author: "Bill Mizgerd"
bibliography: ay.bib
output: html_document
---

Last Updated: 2019-02-26

# Background

**Language Family**: Aymaran / Aymara

* Aymara is a Native American language spoken along the Andean Mountains in Bolivia, Peru, and Chile.

# Phonology

## Consonants

* Voiced stops are not native to Aymara, but do appear in some loanwords [@AyWexler1967, p. xiv].
* Wexler argues for /r/ rather than /ɾ/ [@AyWexler1967, p. xiv]; Briggs, meanwhile, argues that [r] exists as an allophone of /ɾ/ [@AyBriggs1976, p. 82].
* /ŋ/ is phonemic in a small number of dialects [@AyBriggs1976, pp. 82-83].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"     Labial     Alveolar  Postalveolar  Palatal   Velar     Uvular
Stops                        "p pʰ pʼ"  "t tʰ tʼ" ""            ""        "k kʰ kʼ" "q qʰ qʼ"
Affricates                   ""         ""        "tʃ tʃʰ tʃʼ"  ""        ""        ""
Fricatives                   ""         "s"       ""            ""        "x"       "χ"
Nasals                       "m"        "n"       ""            "ɲ"       ""        ""
Flaps                        ""         "ɾ"       ""            ""        ""        ""
Approximants                 "w"        "l"       ""            "j ʎ"     ""        ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Uvular"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  footnote(general = "Note: From left to right, phonemes in a given cell are plain, aspirated, and ejective; the exception is in the approximants row, where the phoneme on the right of a cell is lateral.", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* Older sources also include [e] and [o], but these are just allophones of /i/ and /u/ that emerge next to uvular obstruents [@AyWexler1967, p. xiv].
* Vowel length is phonemic [@AyBriggs1976, p. 80]. Long vowels are marked with umlauts [@AyColer2014, p. 31].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
       Front   Central   Back
High   "i"     ""        "u"
Low    ""      "a"       ""
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

## Stress

* Stress falls on the penultimate syllable of words [@AyBriggs1976, p. 89].
* Stress is not phonemic in Aymara [@AyHardman1981, p. 8].

# Alphabet

* The orthography I will be using is the Unified Alphabet, which was made official by the governments of Bolivia and Peru in 1984 and 1985 respectively [@AyColer2014, pp. 31-32]. Older sources, such as Wexler and Hardman, will use different orthographies.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme     Comment
"a"            "/a/"        ""
"i"            "/i/"        ""
"j"            "/x/"        ""
"k"            "/k/"        ""
"k\'"          "/kʼ/"        ""
"l"           "/l/"         ""
"m"           "/m/"         ""
"n"           "/n/"         ""
"ñ"           "/ɲ/"         ""
"p"           "/p/"         ""
"p\'"         "/pʼ/"         ""
"q"          "/q/"          ""
"q\'"         "/qʼ/"         ""
"r"          "/ɾ/"          ""
"s"          "/s/"          ""
"t"         "/t/"           ""
"t\'"        "/tʼ/"          ""
"u"         "/u/"           ""
"w"         "/w/"           ""
"x"         "/χ/"           ""
"y"         "/j/"           ""
**Multigraph** ""           ""
"ch"       "/tʃ/"           ""
"chh"      "/tʃʰ/"          ""
"ch\'"      "/tʃʼ/"         ""
"kh"       "/kʰ/"           ""
"ll"       "/ʎ/"            ""
"nh"       "/ŋ/" "only phonemic in some dialects"
"ph"       "/pʰ/"           ""
"qh"       "/qʰ/"           ""
"th"       "/tʰ/"           ""
'), TRUE)
kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

## Syllable structure

* Roots are typically either CV(C)CV or (less commonly) V(C)CV. Most suffixes are CV but there are several other options [@AyColer2014, p. 29].
* Vowel clusters are avoided [@AyHardman1981, p. 190].

# Lenition Rules

* Word-final vowels sometimes are sometimes devoiced or dropped [@AyHardman1981, p. 190].
* In certain morphological contexts, word-medial vowels are dropped [@AyHardman1981, p. 8].
* /x/ debuccalizes to [h] in some dialects [@AyBriggs1976, p. 84].
* High vowels are lowered in uvular environments [@AyBriggs1976, p. 85].
* Nasals assimilate to following consonants [@AyColer2014, pp. 41-42].

# Misc. Rules

* /χ/ and /ɾ/ rarely if ever occur word-initially [@AyBriggs1976, p. 87].
* Glides and vowels do not occur adjacent to each other [@AyBriggs1976, p. 89].

# References
