---
title: "Erzya"
author: "Abi Creighton"
bibliography: myv.bib
output: html_document
---

Last Updated: 2020-07-15

**COMPROMISED: conflation between /n/ and /ŋ/**

# Background

**Language Family:** Uralic / Finno-Ugric / Finno-Permic / Finno-Cheremisic / Finno-Mordvinic / Mordvinic

* Erzya, also known as Erzian or Mordvin, is spoken in Russia in the Republic of Mordovia and nearby regions. There are also communities of speakers in Armenia, Estonia, and Central Asia.

# Phonology

## Consonants

* @aasmae_2014 implies that the alveolar consonants may reflect a more dental articulation (p. 10).
* There is some debate as to whether /ŋ/ is a distinct phoneme in Erzya or an allophone of /n/. Because @rueter_2010 provides examples of minimal pairs with /ŋ/ and /n/ (p. 58), I have decided to include it as a phoneme. However, because the same grapheme ⟨н⟩ is used to represent both /ŋ/ and /n/, and it is not apparent by context which sound the grapheme represents in a given instance, the language is compromised.  

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial   Alveolar   Postalveolar  Palatal  Velar
Stops                      "p b"  "t tʲ d dʲ"      ""          ""     "k ɡ"     
Affricates                  ""     "ts tsʲ"       "tʃ"         ""      ""     
Fricatives                 "f v"  "s sʲ z zʲ"     "ʃ ʒ"        ""      "x"
Nasals                      "m"     "n nʲ"         ""          ""      "ŋ"
Trills                      ""      "r rʲ"         ""          ""      ""   
Approximants                ""      "l lʲ"         ""          "j"     ""   
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 5)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes without diacritics that share a cell, those on the left are voiceless and those on the right are voiced. Phonemes with the diacritic (ʲ) are palatalized.", general_title = "")
```  

## Vowels

* Vowel length is not phonemic in Erzya [@aasmae_2014, p. 10].
* Erzya does not have diphthongs (ibid.).

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
             Front      Central       Back
High          "i"         ""          "u"
Mid           "e"         ""          "o"
Low           ""          "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```     

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme                    Comment
"а"            "/a/"                       ""
"б"            "/b/"                       ""
"в"            "/v/"                       ""
"г"            "/ɡ/"                       ""
"д"            "/d/"                       ""
"е"            "/je/"        "marks palatalization of preceding consonant"
"ё"            "/jo/"        "marks palatalization of preceding consonant"
"ж"            "/ʒ/"                       ""
"з"            "/z/"                       ""
"и"            "/ji/"        "marks palatalization of preceding consonant"
"й"            "/j/"                       ""
"к"            "/k/"                       ""
"л"            "/l/"                       ""
"м"            "/m/"                       ""
"н"            "/n/; /ŋ/"        "/n/: default in the rules"
"о"            "/o/"                       ""
"п"            "/p/"                       ""
"р"            "/r/"                       ""
"с"            "/s/"                       ""
"т"            "/t/"                       ""
"у"            "/u/"                       ""
"ф"            "/f/"                       ""
"х"            "/x/"                       ""
"ц"            "/ts/"                      ""
"ч"            "/tʃ/"                      ""
"ш"            "/ʃ/"                       ""
"щ"            "/ʃtʃ/"                     ""
"ъ"            "∅"                         ""
"ы"            "/i/"                       ""
"ь"            "/ʲ/"         "marks palatalization of preceding consonant"  
"э"            "/e/"                       ""
"ю"            "/ju/"        "marks palatalization of preceding consonant"  
"я"            "/ja/"        "marks palatalization of preceding consonant"

'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Misc. Rules

* Vowel harmony is present in Erzya [@aasmae_2014, p. 11].
* Consonants often palatalize when adjacent to other palatalized consonants [@aasmae_2014, p. 12].
* The voiced stops and /z/ become voiceless at a morpheme boundaries if they precede a voiceless consonant (ibid.).
* Adding a suffix to a stem may cause a stem-final vowel to elide [@aasmae_2014, p. 13].
* An epenthetic vowel may be inserted between two consonants at morpheme boundaries (ibid.).

# References
