---
title: "Yucuna"
author: "Bill Mizgerd"
bibliography: ycn.bib
output: html_document
---

Last Updated: 2019-06-28

# Background

**Language Family**: Arawakan / Maipuran / Northern Maipuran / Upper Amazon / Western Nawiki / Yucuna

* Yucuna is spoken in southern Columbia.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial  Alveolar  Postalveolar  Palatal  Velar  Glottal
Stops                      "p pʰ"  "t tʰ"    ""            ""       "k"    "ʔ"
Affricates                 ""      ""        "tʃ"          ""       ""     ""
Fricatives                 ""      "s"       ""            ""       ""     "h"
Nasals                     "m"     "n"       ""            "ɲ"      ""     ""
Flaps                      ""      "ɾ"       ""            ""       ""     ""
Approximants               "w"     "l"       ""            "j"      ""     ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  footnote("Note: For phonemes that share a cell, those on the left are plain, whereas those on the right are aspirated.", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* Although vowel nasalization is contrastive in a few rare cases, overall it is not considered a phonemic feature [@YcnSchauer1967, p. 66].
* Adjacent vowels do not form diphthongs [@YcnSchauer1967, p. 67].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
          Front   Central   Back
High      "i"     ""        "u"
Mid       "e"     ""        "o"
Low       ""      "a"       ""
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered")
```

# Alphabet

* Orthographies disagree slightly at a few points; older, more Spanish-influenced texts use ⟨hu, c, qu⟩ where newer orthographies tend to use ⟨w, k, k⟩ respectively. Based on the Crubadan corpus, both patterns are still in wide usage, so I have included rules for both.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme  
"a"             "/a/"  
"c"             "/k/"    
"e"             "/e/"  
"\'"             "/ʔ/"   
"i"             "/i/"  
"j"             "/h/"  
"k"             "/k/"  
"l"             "/l/"  
"m"             "/m/"  
"n"             "/n/"  
"ñ"             "/ɲ/"  
"o"             "/o/"  
"p"             "/p/"   
"r"             "/ɾ/"  
"s"             "/s/"  
"t"             "/t/"    
"u"             "/u/"  
"w"             "/w/"  
"y"             "/j/"
**Digraph**      ""
"ch"            "/tʃ/"
"hu"            "/w/"
"ph"            "/pʰ/"  
"qu"            "/k/"
"th"            "/tʰ/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Yucuna has a syllable structure of (C)V(C) [@YcnSchauer1967, pp. 66-67].
    - The only consonants that appear in syllable codas are /n/ and /h/.

# Lenition Rules

* In some words, /pʰ/ spirantizes to [ɸ] [@YcnSchauer1967, p. 64; @YcnSchauerEtAl2005, p. 15].
* Many speakers debuccalize /s/ to [h] [@YcnSchauerEtAl2005, p. 15].
* Word-initial /j/ is sometimes deleted (ibid.).

# Misc. Rules

* /k/ labializes to [kʷ] after /u/, and palatalizes to [kʲ] after /i/ [@YcnSchauer1967, p. 64].
* /k/ and [kʰ] exist in free variation in some words (ibid.).
* /n/ assimilates to the place of following velars [@YcnSchauer1967, p. 65].
* Before consonants, /h/ is realized as [x] (ibid.).
* Word-final vowels nasalize in syllables of the pattern V or hV [@YcnSchauer1967, p. 66; @YcnSchauerEtAl2005, p. 15].
* Epenthesis of [ʔ] occurs intervocalically [@YcnSchauerEtAl2005, p. 15].

# References
