---
title: "Hmong"
author: "Bill Mizgerd"
bibliography: hmn.bib
output: html_document
---

Last Updated: 2019-07-24

# Background

**Language Family**: Hmong-Mien / Hmongic / Chuanqiandian

* The Hmong language comprises a number of dialects with some phonological differences. Where they diverge, I will be prioritizing Hmong Daw (also known as White Hmong).
* Hmong Daw is predominantly spoken in Laos and the United States.

# Phonology

## Consonants

* The only consonant clusters to occur in Hmong are obstruent-lateral sequences [@HmnMortensen2004, p. 3]. Some authors [e.g. @HmnGolston2001] argue that, rather than being a cluster, these are just single consonants with lateral release; however, I follow Mortensen in analyzing these as clusters.
* In prenasalized stops, the prenasalization fully assimilates to the place of the consonant [@HmnGolston2001, p. 4].
* Although the series that I have described as retroflex fricatives and affricates are widely described as postalveolar, I have found that the retroflex place of articulation is more in keeping with the actual sounds described by the authors. Similarly, although the sound I have described as /lʰ/ is frequently transcribed as /ɬ/, I have found /lʰ/ to be more in keeping with the characteristics of the sound.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"      Labial   Alveolar    Retroflex  Palatal   Velar     Uvular    Glottal
"Stops (unaspirated)"         "p ᵐb"   "t d ⁿd"    "ʈ ᶯɖ"     "c ᶮɟ"    "k ᵑɡ"    "q ᶰɢ"    ""
"Stops (aspirated)"           "pʰ ᵐbʱ" "tʰ dʱ ⁿdʱ" "ʈʰ ᶯɖʱ"   "cʰ ᶮɟʱ"  "kʰ ᵑɡʱ"  "qʰ ᶰɢʱ"  ""
"Affricates (unaspirated)"    ""       "ts ⁿdz"    "tʂ ⁿdʐ"   ""        ""        ""        ""
"Affricates (aspirated)"      ""       "tsʰ ⁿdzʱ"  "tʂʰ ⁿdʐʱ" ""        ""        ""        ""
"Fricatives"                  "f v"    "s"         "ʂ ʐ"      "ç ʝ"     ""        ""        "h"
"Nasals"                      "m̥ m"    "n̥ n"       ""         "ɲ̥ ɲ"     ""        ""        ""
"Approximants"                ""       "l lʰ"      ""         ""        ""        ""        ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Retroflex", "Palatal", "Velar", "Uvular", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 7)) %>%
  footnote(general = "Note: Phonemes to the left of each cell are voiceless and phonemes to the right of each cell are voiced. The alveolar stops (unaspirated and aspirated) are ordered voiceless, voiced, pre-nasalized.", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* @HmnBurt2010 argues for /ɔ/ rather than /o/ (p. 245).

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
vowels <- read.table(textConnection('
      Front  Central  Back
High  "i"    "ɨ"      "u"
Mid   "e ẽ"  ""       "o õ"
Low   ""     "a ã"    ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are oral and those on the right are nasal.", general_title = "")

diphthongs <- read.table(textConnection('
Diphthongs
"/ai/, /aɨ/, /au/, /ia/, /ua/"
'), TRUE)

kable(diphthongs, align = 'c') %>%
  kable_styling("bordered")
```

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme   
"a"           "/a/"    
"c"           "/c/"
"d"           "/d/"   
"e"           "/e/"    
"f"           "/f/"  
"h"           "/h/"  
"i"           "/i/"  
"k"           "/k/"    
"l"           "/l/"   
"m"           "/m/"    
"n"           "/n/"     
"o"           "/o/"   
"p"           "/p/"   
"q"           "/q/"    
"r"           "/ʈ/"   
"s"           "/ʂ/"  
"t"           "/t/"      
"u"           "/u/"  
"v"           "/v/"  
"w"           "/ɨ/"  
"x"           "/s/"    
"y"           "/ʝ/"  
"z"           "/ʐ/"
**Multigraph** ""
"aa"          "/ã/"
"ch"          "/cʰ/"
"dh"          "/dʱ/"  
"nc"          "/ᶮɟ/"  
"nch"         "/ᶮɟʱ/"
"ee"          "/ẽ/"
"kh"          "/kʰ/"  
"nk"          "/ᵑɡ/"  
"nkh"         "/ᵑɡʱ/"
"hl"          "/lʰ/"
"hm"          "/m̥/"
"hn"          "/n̥/"
"ny"          "/ɲ/"  
"hny"         "/ɲ̥/"
"oo"          "/õ/"
"ph"          "/pʰ/"  
"np"          "/ᵐb/"  
"nph"         "/ᵐbʱ/"
"qh"          "/qʰ/"  
"nq"          "/ᶰɢ/"  
"nqh"         "/ᶰɢʱ/"
"rh"          "/ʈʰ/"  
"nr"          "/ᶯɖ/"  
"nrh"         "/ᶯɖʱ/"
"th"          "/tʰ/"  
"nt"          "/ⁿd/"  
"nth"         "/ⁿdʱ/"
"ts"          "/tʂ/"  
"tsh"         "/tʂʰ/"  
"nts"         "/ⁿdʐ/"  
"ntsh"        "/ⁿdʐʱ/"  
"tx"          "/ts/"  
"txh"         "/tsʰ/"  
"ntx"         "/ⁿdz/"  
"ntxh"        "/ⁿdzʱ/"
"xy"          "/ç/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Tone

* Hmong has seven contrastive tones: high, high falling, high rising, mid, low, falling breathy, and falling creaky. There also exists an eighth tone, low rising, that is morphologically formed in some scenarios [@HmnMortensen2004, p. 4].
* All tone-marking characters appear after the vowel they modify. Because consonants cannot occur in Hmong syllable codas, consonants that precede other consonants can thus be identified as tone-markers; this allows us to distinguish whether a given ⟨v⟩, ⟨s⟩, or ⟨m⟩ represents a consonant or a tone.
* Some analyses describe the creaky falling tone as terminating in a glottal stop [@HmnHuffman1987, p. 496; @HmnMortensen2004, p. 4].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

tone <- read.table(textConnection('
Grapheme     Tone                       Comment
"b"          "/˦/"                    "high tone"  
"j"          "/˥˨/"                 "high falling tone"  
"v"          "/˨˦/"                 "high rising tone"  
"s"          "/˨/"                    "low tone"
"g"          "/˦˨/"                "breathy falling tone"  
"m"          "/˨˩/"                "creaky falling tone"
"d"          "/˨˩/"      "low rising tone (allophone of creaky falling)"  
""           "/˧/"      "mid tone (unmarked) [@HmnBurt2010, p. 246]"
'), TRUE)

kable(tone, align = 'c') %>%
  kable_styling("bordered")
```

# Misc. Rules

* Epenthesis of [ŋ] occurs after word-final nasal vowels [@HmnWhitelock1982, p. 24].
* Epenthetic glottal stops occur word-initially preceding vowels in borrowed words [@HmnGolston2001, p. 5].

# References
