---
title: "Bulgarian"
author: "Bill Mizgerd"
bibliography: bg.bib
output: html_document
---

Last Updated: 2020-06-26

# Background

**Language Family**: Indo-European / Slavic / South / Eastern

* Bulgarian is spoken in Bulgaria, Greece, Serbia, and Turkey.

# Phonology

## Consonants

* All consonants (except for postalveolars) also have palatalized forms in Bulgarian. The palatalization distinctions are frequently held to be phonemic, but due to their relative scarcity (fairly few words with > 100 instances in Crúbadán) and their limited distribution (the contrast only occurs before non-front vowels, per @BgScatton1983, p. 64), such an analysis would be problematic for this project. Instead, I have opted to follow an alternate analysis that treats the palatalized consonants as sequences of the non-palatalized form and /j/.
* @BgHauge1999 lists the non-palatalized lateral approximant in Bulgarian as /ɫ/, while most other authors prefer /l/; however, @BgScatton1993 describes the same approximant as "markedly velarized except before front vowels" (p. 190). The combination of these opinions has made the /ɫ/ analysis more compelling to me.
* Outside of the aforementioned palatalization sequences, /j/ mostly appears immediately after vowels [@BgScatton1983, p. 65], but can also appear word-initially [@BgScatton1993, p. 191].
* /ʒ/ is relatively rare, often being realized as /z/ [@BgScatton1993, p. 191].


```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"      Labial     Alveolar  Postalveolar  Palatal   Velar
Stops                         "p b"      "t d"         ""          ""      "k ɡ"
Affricates                     ""         "ts"       "tʃ dʒ"       ""       ""
Fricatives                    "f v"      "s z"        "ʃ ʒ"        ""       "x"
Nasals                         "m"        "n"          ""          ""       ""
Trills                         ""         "r"          ""          ""       ""
Approximants                   ""         "ɫ"          ""          "j"      ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 5)) %>%
  footnote(general = "Note: Phonemes to the left of each cell are voiceless and phonemes to the right of each cell are voiced.", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* Klagstad describes many of the listed vowel phonemes as not in fact being the underlying allophones of their phonemes: his analysis sees [ɛ], [ɔ], and [ɑ] as being underlying rather than /e/, /o/, and /a/. He also describes /ə/ as being realized as [ɯ] in "all positions" [@BgKlagstad1958, pp. 44-45].
* The mid central vowel in Bulgarian is described in a wide variety of ways by different sources, including /ŭ/ (or /ɨ/) [@BgScatton1983], /ā/ [@BgScatton1993], and /ə/ [@BgKlagstad1958; @BgGroen1988]. I have opted for /ɤ/ rather than /ə/, given that @BgKlagstad1958 describes it as a mid-back unrounded vowel, present in both stressed and unstressed syllables (p. 44).
* Geminate vowels do not phonemically reduce in Bulgarian [@BgGroen1988, p. 186].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
       Front   Central   Back
High   "i"     ""        "u"
Mid    "ɛ"     ""       "ɤ ɔ"
Low    ""      "a"       ""
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: The mid-back vowel on the left is unrounded.", general_title = "")
```

# Alphabet

* Non-palatalized variants of each character are listed as the "default" in this section. For rules on how to mark palatalization, see the following subsection.
    - Palatalized consonants are not represented by specific standalone graphemes; instead, specific characters are used to mark that the preceding consonant should be palatalized [@BgHauge1999, p. 8]. Such examples are detailed below in the chart.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme                    Comment
"а"            "/a/"                       ""
"б"            "/b/"                       ""
"в"            "/v/"                       ""
"г"            "/ɡ/"                       ""
"д"            "/d/"                       ""
"е"            "/e/"                       ""
"ж"            "/ʒ/"                       ""
"з"            "/z/"                       ""
"и"            "/i/"                       ""
"й"            "/j/"                       ""
"к"            "/k/"                       ""
"л"            "/ɫ/"                       ""
"м"            "/m/"                       ""
"н"            "/n/"                       ""
"о"            "/o/"                       ""
"п"            "/p/"                       ""
"р"            "/r/"                       ""
"с"            "/s/"                       ""
"т"            "/t/"                       ""
"у"            "/u/"                       ""
"ф"            "/f/"                       ""
"х"            "/x/"                       ""
"ц"           "/ts/"                       ""
"ч"           "/tʃ/"                       ""
"ш"           "/ʃ/"                        ""
"щ"          "/ʃtʃ/"                       ""
"ъ"            "/ɤ/"                       ""
"ь"             ""         "marks palatalization of preceding consonant"  
"ю"           "/ju/"       "marks palatalization of preceding consonant"  
"я"           "/ja/"       "marks palatalization of preceding consonant"
**Digraph**     ""                         ""
"дж"          "/dʒ/"                       ""
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```  

# Lenition Rules

* /j/ deletes before front vowels [@BgHauge1999, p. 11], as well as after palatalized consonants [@BgScatton1983, pp. 76-77].
* Geminate consonants tend to simplify across morpheme boundaries [@BgScatton1983, p. 71].
* /s/ deletes when preceded by a sibilant and followed by /k/ [@BgScatton1983, p. 77].

# Misc. Rules

* Vowels nasalize when followed by first a nasal consonant and then a fricative [@BgHauge1999, p. 1].
* In consonant clusters, obstruents assimilate to the voicing of the final obstruent in the cluster, and may yield non-phonemic sounds (e.g. [dz]); however, this pattern does not consistently occur before /v/ or /vʲ/ [@BgKlagstad1958, pp. 47-48; @BgScatton1983, pp. 58, 73].
* /u/ fronts to [y] after palatalized consonants [@BgKlagstad1958, p. 44].
* /l/ fronts before front vowels [@BgKlagstad1958, p. 48].
* Palatalized consonants followed by /a/ tend to alternate with their non-palatalized counterparts before /e/ [@BgHauge1999, p. 9].
* Consonants devoice word-finally [@BgScatton1983, p. 73].
* Voiceless stops are aspirated in word-initial prevocalic contexts [@BgKlagstad1958, pp. 47, 49].
* /n/ assimilates to [ŋ] before velars [@BgKlagstad1958, p. 48]; /m/ assimilates to [ɱ] before labiodentals [@BgHauge1999, p. 4].

# References
