---
title: "South Tairora"
author: "Bill Mizgerd"
bibliography: omw.bib
output: html_document
---

Last Updated: 2019-07-19

# Background

**Language Family**: Trans-New Guinea / Kainantu-Goroka / Kainantu / Tairora / South Tairora

* South Tairora is spoken in the Eastern Highlands province of Papua New Guinea.

# Phonology

## Consonants

* The contrast between /s/ and /h/ is minor, and is most clearly realized in loanwords [@OmwLVincent2010, p. v].
* The palatalized consonants only occur in some dialects [@OmwAVincent2003, p. 1].
* **/tʲ/ and /ⁿtʲ/ do not appear in the Crúbadán corpus.**

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Bilabial   Alveolar  Velar     Glottal
"Stops (plain)"            "p ᵐp"     "t ⁿt"    "k ᵑk"    "ʔ"
"Stops (palatalized)"      ""         "tʲ ⁿtʲ"  "kʲ ᵑkʲ"  ""
Fricatives                 "β β̃"      "s"       ""        "h"
"Nasals (plain)"           "m"        "n"       ""        ""
"Nasals (palatalized)"     ""         "nʲ"      ""        ""
Flaps                      ""         "ɾ ɾ̃"     ""        ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Bilabial", "Alveolar", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 4)) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left of each cell are plain, and phonemes to the right of each cell are nasalized or prenasalized.", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

## Vowels

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
          Front   Central   Back
High     "i"      ""        "u"
Mid      "ɛː"     "ɘ"       "ɔː"
Low       ""      "aː"      ""
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)

diphthongs <- read.table(textConnection('
Diphthongs
"/ɘj/, /ɘw/"
'), TRUE)

kable(diphthongs, align = 'c') %>%
  kable_styling("bordered")
```  

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme
"a"              "/ɘ/"   
"e"              "/ɛː/"  
"h"              "/h/"  
"i"              "/i/"  
"k"              "/k/"   
"m"              "/m/"  
"n"              "/n/"   
"o"              "/ɔː/"  
"p"              "/p/"  
"q"              "/ʔ/"  
"r"              "/ɾ/"  
"s"              "/s/"  
"t"              "/t/"   
"u"              "/u/"  
"v"              "/β/"  
**Multigraph**    ""
"ai"             "/ɘj/"  
"au"             "/ɘw/"  
"aa"             "/aː/"
"ky"             "/kʲ/"
"mp"             "/ᵐp/"  
"mw"             "/β̃/"  
"nk"             "/ᵑk/"  
"nky"            "/ᵑkʲ/"  
"nr"             "/ɾ̃/"  
"nt"             "/ⁿt/"  
"nty"            "/ⁿtʲ/"  
"ny"             "/nʲ/"
"ty"             "/tʲ/"
'), TRUE)

kable(alphabet, align = "c") %>%
  kable_styling("bordered")
```

# Lenition Rules

* Prenasalization is often omitted in the Suwaira dialect [@OmwLVincent2010, p. 578].
* /ᵐp/ and /ⁿt/ can be realized with voicing in the Suwaira dialect [@OmwLVincent2010, p. 578].

# Misc. Rules

* Word-initial /β/ is realized as [b] [@OmwLVincent2010, p. 578].

# References
