---
title: "Tuvan"
author: "Abi Creighton"
bibliography: tyv.bib
output: html_document
---

Last Updated: 2020-02-12

# Background

**Language family**: Turkic / Common Turkic / Siberian Turkic / South Siberian / Sayan Turkic / Tuvan

* Tuvan, also know as Tuvinian, Tyvan, or Tuvin, is spoken in the Tuva Republic in Russia.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"          Labial  Alveolar Palatal Velar     
Stops                             "p b"    "t d"    ""     "k ɡ"
Affricates                         ""       ""     "cç"     ""      
Fricatives                         ""      "s z"   "ç ʝ"    "x"     
Nasals                             "m"      "n"     ""      "ŋ"    
Trills                             ""       "r"     ""      ""      
Approximants                       "ʋ"      "l"     "j"     ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Palatal", "Velar"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 4)) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* Vowel length is contrastive in Tuvan [@harrison_topics_2000, p. 11]. Long vowels are indicated by duplicate graphemes.
* Low tone (vs. no tone) is contrastive syllable-initially [@OBrien2010, pp. 1-2]; however, it's rather marginal, such that very few words display this distinction (although the prevalence depends on the dialect). It is not marked in Crúbadán, so I have chosen to proceed with the language, not accounting for it in the rules.
* Vowel harmony is present in Tuvan [@harrison_topics_2000, p. 111].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
          Front      Back
High      "i y"      "ɯ u"
Mid       "e ø"      "o"
Low       ""         "ɑ"
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are unrounded and those on the right are rounded.", general_title = "")
```

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme                           Comment
"а"            "/ɑ/"                              ""
"б"            "/b/"                              ""
"в"            "/ʋ/"                              ""
"г"            "/ɡ/"                              ""
"д"            "/d/"                              ""
"e"            "/e/; /je/"                              "/e/: word-medially and word-finally; /je/: word-initially in Russian loanwords"
"ё"            "/jo/"                             ""
"ж"            "/ʝ/"                              ""
"з"            "/z/"                              ""
"и"            "/i/"                              ""
"й"            "/j/"                              ""
"к"            "/k/"                              ""
"л"            "/l/"                              ""
"м"            "/m/"                              ""
"н"            "/n/"                              ""
"ң"            "/ŋ/"                              ""
"о"            "/o/"                              ""
"ө"            "/ø/"                              ""
"п"            "/p/"                              ""
"р"            "/r/"                              ""
"с"            "/s/"                              ""
"т"            "/t/"                              ""
"у"            "/u/"                              ""
"ү"            "/y/"                              ""
"ф"            "/f/"                    "appears in Russian loanwords"
"х"            "/x/"                              ""
"ц"            "/ts/"                   "appears in Russian loanwords"
"ч"            "/cç/"                             ""
"ш"            "/ç/"                              ""
"щ"            "/çcç/"                  "appears in Russian loanwords"
"ъ"            ""                        "indicates pitch accent"
"ы"            "/ɯ/"                              ""
"ь"            ""                 "represents where there was a reduced vowel"
"э"            "/e/"                              "word-initially"
"ю"            "/ju/"                    "appears in Russian loanwords"
"я"            "/jɑ/"                    "appears in Russian loanwords"
'), TRUE)

kable(alphabet) %>%
  kable_styling("bordered")
```

# Syllable Structure

* Syllables in Tuvan follow the order of (C)V(V)(C)(C) [@harrison_topics_2000, p. 18].

# Lenition Rules

* /ɡ/ may spirantize to [ɣ] [@harrison_topics_2000, p. 13].
* Velar consonants are often elided intervocalically [@harrison_topics_2000, p. 87].
* Voiceless stops become voiced intervocalically [@harrison_topics_2000, p. 83].
    - In this position /b/ and /ɡ/ become [ʋ] and [ɣ], respectively.

# Misc. Rules

* Voiceless stops (and voiced velar stops) can geminate [@harrison_topics_2000, p. 82].

# References
