---
title: "Huarijío"
author: "Emily Strand"
bibliography: var.bib
output: html_document
---

Last Updated: 2020-07-02

# Background

**Language Family:** Uto-Aztecan / Southern / Sonoran / Tarahumaran / Guarijio

* Also referred to as Guarijío, Huarijío is spoken in northwestern Mexico.

# Phonology

## Consonants

* Consonants do not geminate [@Caballero2007, p. 5].
* Glottal stops only appear word-medially [@FelixArmendariz2005, p. 11].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial       Alveolar     Postalveolar     Palatal     Velar      Glottal    
Stops                       "p"          "t"              ""             ""        "k"         "ʔ"
Affricates                  ""            ""             "tʃ"            ""        ""          ""
Fricatives                  ""            "s"             ""             ""        "h"         ""
Nasals                      "m"           "n"             ""             ""        ""          ""
Flaps                       ""            "ɾ"             ""             ""        ""          ""
Approximants                "w"            ""             ""             "j"       ""          ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* Vowel length is not contrastive [@Caballero2007, p. 5]. However, free variation among long and short vowels does occur [@FelixArmendariz2005, p. 15].
    - When the repetition of vowels occurs, if the second one is stressed, they are realized as individual syllabic vowels rather than a long vowel [@FelixArmendariz2005, p. 15]. Duplicate vowels will be transcribed as short if the second one does not have a stress marking.
* Accents are put over the vowels to indicate stress [@Miller1993, p. 25].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
             Front      Central       Back
High          "i"         ""           "u"
Mid           "e"         ""           "o"
Low           ""          "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```      

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme
"a"              "/a/"  
"e"              "/e/"  
"h"              "/h/"  
"i"              "/i/"  
"j"              "/h/"  
"k"              "/k/"  
"m"              "/m/"  
"n"              "/n/"  
"o"              "/o/"  
"p"              "/p/"  
"r"              "/ɾ/"  
"s"              "/s/"  
"t"              "/t/"  
"u"              "/u/"  
"w"              "/w/"  
"y"              "/j/"  
"\'"              "/ʔ/"  
**Digraph**       ""
"ch"             "/tʃ/"  
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Syllables in Huarijío generally consist of CV or CVV [@FelixArmendariz2005, pp. 19-20]. Holistically, however, they are said to follow the form of (C)V(C) [@Miller1993, p. 23], the codas of which mainly consist of /h/, /ʔ/, /m/, and /n/. Word-final consonants are also prohibited.
* Consonant clusters are rare, appearing as a result of contractions [@Miller1993, p. 24] or syllable boundaries [@FelixArmendariz2005, p. 10].

# Lenition Rules

* Voiceless stops /p/ and /k/ become voiced intervocalically if the following vowel is not accented [@Miller1993, p. 25].
* /t/ becomes /ɾ/ intervocalically in environments following a stressed syllable [@Miller1993, p. 25].

# Misc. Rules

* Vowel disharmony plays a role in Huarijío when concerning the addition of suffixes [@FelixArmendariz2005, p. 26]: If the vowel of the root is low, the vowel of the suffix will be high and vice versa.
* Reduplication of prefixes is abundant in Huarijío [@Haugen2014, p. 5].
* Word-initial vowels may be preceded by /h/ or a glottal stop (rare) [@Miller1993, p. 28].
* /w/ word-initially can be realized as [b] [@Miller1993, p. 27].
* /s/ palatalizes preceding /i/ (ibid.).
* Preceding /i/, /e/, and /u/, /ɾ/ may be realized as retroflex (ibid.).

# References
