---
title: "Ese"
author: "Bill Mizgerd"
bibliography: mcq.bib
output: html_document
---

Last Updated: 2019-08-06

# Background

**Language Family**: Trans-New Guinea / Koiarian / Baraic / Ese

* The language is also known as Managalasi and is spoken in the Oro province of Papua New Guinea.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Bilabial      Alveolar      Alveopalatal     Velar      Glottal
Stops                        "p"           "t"             ""            "k"          "ʔ"
Affricates                    ""           ""            "tɕ  dʑ"         ""          ""
Fricatives                   "β"           "s"             ""             ""          "h"
Nasals                       "m"           "n"             ""             ""          ""
Flaps                        ""            "ɾ"             ""             ""          ""   
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Alveopalatal", "Palatal", "Velar"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 5)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced. Both palatal approximants are voiced.", general_title = "")
```

## Vowels

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"        "ʉ"           "u"
Mid       "e"         ""           "o"
Low        ""         ""           "ɑ"
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```  

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme  
"a"            "/ɑ/"    
"e"            "/e/"  
"h"            "/h/"  
"i"            "/i/"  
"j"            "/dʑ/"  
"k"            "/k/"  
"m"            "/m/"  
"n"            "/n/"  
"o"            "/o/"  
"p"            "/p/"  
"r"            "/ɾ/"  
"s"            "/s/"  
"t"            "/t/"  
"u"            "/u/"  
"ú"            "/ʉ/"  
"v"            "/β/"  
"\'"           "/ʔ/"
**Digraph**     ""
"ch"           "/tɕ/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Ese has a (C)V syllable structure [@McqParlier1992, p. 2].

# Lenition Rules

* Postvocalic /a/ reduces to [ə], except when followed by /ne/ or /nɑ/ [@McqParlier1963, p. 19].

# Misc. Rules

* Word-initial /ɾ/ lateralizes to [ɺ] [@McqParlier1992, p. 3].
* @McqParlier1963 also attests this process as occurring after /i/ (p. 10).
* /u/ becomes [w] before stressed vowels [@McqParlier1992, p. 3; @McqParlier1981, p. viii].
* /β/ goes to [b] before unstressed /u/ [@McqParlier1963, p. 7].
* /tɕ/ alternates with [ts] before /ɑ/ [@McqParlier1963, p. 8].
* /dʑ/ retroflexes (to [ɖʑ], supposedly) before non-high back vowels [@McqParlier1963, p. 9].
* Word-final mid vowels lower to be mid-open [@McqParlier1963, pp. 18-20].

# References
