---
title: "Kiribati"
author: "Justin Bai"
bibliography: gil.bib
output: html_document
---

Last-updated: 2020-06-29

# Background

**Language Family**: Austronesian / Malayo-Polynesian / Central-Eastern / Eastern Malayo-Polynesian / Oceanic / Central-Eastern Oceanic / Remote Oceanic / Micronesian / Micronesian Proper / Ikiribati

* Also referred to as Gilbertese, Kiribati is spoken on the Solomon Islands, located just east of Papua New Guinea.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Bilabial     Alveolar    Velar
Stops                      "p pˠ"        "t"        "k"       
Fricatives                  "βˠ"         ""          ""
Nasals                   "m  mː  mˠ"    "n  nː"    "ŋ  ŋː"
Flaps                        ""          "ɾ"         ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Bilabial", "Alveolar", "Velar"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 3)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: Phonemes that share a cell match in voicing. Those transcribed with /ˠ/ are velarized and those transcribed with /ː/ are phonemically long.", general_title = "")
```

## Vowels

* Vowel length is contrastive, where long vowels are represented orthographically by doubling the vowel [@blevins1999trimoraic, p. 206].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         ""           "u"
Mid       "e"         ""           "o"
Low        ""         "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```   

# Alphabet


```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme
"a"           "/a/"   
"b"           "/p/"
"e"           "/e/"
"i"           "/i/"
"k"           "/k/"
"m"           "/m/"
"n"           "/n/"
"o"           "/o/"
"r"           "/ɾ/"
"t"           "/t/"
"u"           "/u/"
"w"           "/βˠ/"
**Multigraph**  ""
"bw"          "/pˠ/"
"mm"          "/mː/"
"mw"          "/mˠ/"
"nn"          "/nː/"
"ng"          "/ŋ/"
"ngng"        "/ŋː/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Misc. Rules

* /t/ spirantizes to [s] when followed by /i/ in all dialects [@blevins1999trimoraic, p. 206].

# References
