---
title: "Tongan"
author: "Justin Bai"
bibliography: to.bib
output: html_document
---

Last Updated: 2019-11-21

# Background

**Language family**: Austronesian / Malayo-Polynesian / Central-Eastern / Eastern Malayo-Polynesian / Oceanic / Central-Eastern Oceanic / Remote Oceanic / Central Pacific / East Fijian-Polynesian / Polynesian / Tongic

* Tongan is spoken throughout the Kindgdom of Tonga, a sovereign state northeast of New Zealand comprised of over one hundred islands.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial  Alveolar  Velar  Glottal
Stops                       "p"      "t"      "k"     "ʔ"
Affricates                  ""       ""       ""      ""
Fricatives                "f  v"     "s"      ""      "h"
Nasals                      "m"      "n"      "ŋ"     ""
Approximants                ""       "l"      ""      ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 4)) %>%
  footnote("Note: For phonemes that share a cell, those on the left are voiceless, whereas those on the right are voiced.", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* Vowel length is phonemic in Tongan [@Anderson2006, p. 27]. Long vowels are indicated by a macron (¯).

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
          Front   Central   Back
High      "i"     ""        "u"
Mid       "e"     ""        "o"
Low       ""      "a"       ""
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered")
```

# Alphabet

* An acute accent is also used in the orthography to indicate stress, but it will be transcribed out in the rule set [@Anderson2006, p. 23].
* The grapheme indicating the glottal stop is called a fakauʻa - it is not a standard apostrophe [@Schutz1994, p. 267].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme
"a"              "/a/"
"e"              "/e/"
"f"              "/f/"
"h"              "/h/"
"i"              "/i/"
"k"              "/k/"
"l"              "/l/"
"m"              "/m/"
"n"              "/n/"
"o"              "/o/"
"p"              "/p/"
"s"              "/s/"
"t"              "/t/"
"v"              "/v/"
"ʻ"              "/ʔ/"
**Digraph**       ""
"ng"             "/ŋ/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* /t/ regularly changes to /s/ preceding /i/ [@Morton1962 as cited in @Garellek2019, p. 2].

# Misc. Rules  

* Tongan has a rare three-way laryngeal contrast between /ʔ/, /h/, and vowels in initial positions [@Garellek2019, p. 3].

# References
