---
title: "Sepik Iwam"
author: "Emily Strand"
bibliography: iws.bib
output: html_document
---

Last Updated: 2020-01-07

# Background

**Language Family**: Sepik-Ramu / Sepik / Upper Sepik / Iwam

* Sepik Iwam is spoken in the East Sepik province of Papua New Guinea.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial    Alveolar    Palatal   Velar   Glottal
Stops                        "p"      "t d"         ""     "k ɡ"     ""
Affricates                   ""        ""           ""      ""       ""
Fricatives                   ""        "s"          ""      ""       "h"
Nasals                       "m"       "n"          ""      ""       ""
Trills                       ""        "r"          ""      ""       ""
Approximants                 "w"       ""           "j"     ""       ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 5)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```  

## Vowels

* Adjacent vowel are interpreted as sequences [@2011, p. 3].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         "ɨ"           "u"
Mid       "e"         ""           "o"
Low        ""         "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme  
"a"             "/a/"
"b"             "/b/"
"d"             "/d/"
"e"             "/e/"
"g"             "/ɡ/"
"h"             "/h/"
"i"             "/i/"
"ɨ"             "/ɨ/"
"k"             "/k/"
"m"             "/m/"
"n"             "/n/"
"o"             "/o/"
"p"             "/p/"
"r"             "/r/"
"s"             "/s/"
"t"             "/t/"
"u"             "/u/"
"w"             "/w/"
"y"             "/j/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* Word-medially, /p/ may become voiced, fricated, or voiced and fricated [@2011, p. 4].
* Word medially, /k/ may become [x] or [ɣ]  intervocalically (ibid.).

# Misc. Rules

* /r/ varies freely with [l] [@2011, p. 4].
* Orthographically, the combinations /hw/, /hm/, /hn/, and /hy/ are written ⟨wh⟩, ⟨mh⟩, ⟨nh⟩, and ⟨yh⟩, respectively [@2011, p. 4; @Laszlo1970, p. 31].

# References
