---
title: "Sepik Iwam"
author: "Emily Strand"
bibliography: iws.bib
output: html_document
---

Last Updated: 2020-01-07

# Background

**Language Family**: Sepik-Ramu / Sepik / Upper Sepik / Iwam

* Sepik Iwam is spoken in the East Sepik province of Papua New Guinea.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial    Alveolar    Palatal   Velar   Glottal
Stops                       "p b"      "t d"        ""     "k ɡ"     ""
Fricatives                   ""        "s"          ""      ""       "h"
Nasals                       "m"       "n"          ""      ""       ""
Trills                       ""        "r"          ""      ""       ""
Approximants                 "w"       ""           "j"     ""       ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 5)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```  

## Vowels

* Adjacent vowels (including like-vowels) are interpreted as sequences [@2011, p. 3].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         "ɨ"          "u"
Mid       "e"         ""           "o"
Low        ""         "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

* The documentation is somewhat unclear regarding the phonemic status of voiceless /w/, /m/, /n/, and /y/. @Laszlo1970 indicate that the language has these voiceless variants (p. 8), but just like @2011 and @Laycock1965, they do not actually specify them as phonemes (p. 1; p. 115). It appears that they are consonant sequences (/hw/, /hm/, /hn/, and /hy/) that result in phonetic devoicing. Orthographically, these combinations (/hw/, /hm/, /hn/, and /hy/) are written ⟨wh⟩, ⟨mh⟩, ⟨nh⟩, and ⟨yh⟩, respectively [@2011, p. 4; @Laszlo1970, p. 31]. The reason for a following rather than a preceding ⟨h⟩ is to distinguish between the sequence /hw/ and the labialized realization [hʷ]; to maintain consistency, all "voiceless" consonants are written with a following ⟨h⟩.    

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme  
"a"             "/a/"
"b"             "/b/"
"d"             "/d/"
"e"             "/e/"
"g"             "/ɡ/"
"h"             "/h/"
"i"             "/i/"
"ɨ"             "/ɨ/"
"k"             "/k/"
"m"             "/m/"
"n"             "/n/"
"o"             "/o/"
"p"             "/p/"
"r"             "/r/"
"s"             "/s/"
"t"             "/t/"
"u"             "/u/"
"w"             "/w/"
"y"             "/j/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* Word-medially, /p/ may become voiced, fricated, or voiced and fricated [@2011, p. 4].
* Word medially, /k/ may become [x] or [ɣ]  intervocalically (ibid.).

# Misc. Rules

* /r/ varies freely with [l] [@2011, p. 4].

# References
