---
title: "Asturian"
author: "Emily Strand"
bibliography: ast.bib
output: html_document
---

Last updated: 2019-08-28

# Background

**Language family**: Indo-European / Italic / Romance / Italo-Western / Western / Gallo-Iberian / Ibero-Romance / West Iberian / Asturo-Leonese

* Asturian is spoken in northwestern Spain.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial     Dental   Alveolar    Palatal     Velar   
Stops                      "p  b"     "t̪  d̪"     ""          ""       "k  ɡ"
Affricates                   ""         ""       ""         "tʃ"        ""
Fricatives                  "f"        "θ"       "s"        "ʃ"         ""
Nasals                      "m"        ""        "n"        "ɲ"         ""
Trills                      ""         ""        "r"         ""         ""
Flaps                       ""         ""        "ɾ"         ""         ""
Approximants                ""         ""        "l"       "j  ʎ"       ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Dental", "Alveolar", "Palatal", "Velar"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 5)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced. Both palatal approximants are voiced, but the one on the right is lateral.", general_title = "")
```


## Vowels

* /i/ and /u/ become glides [j] and [w] respectively when unstressed, forming diphthongs and triphthongs with adjoining vowels [@muniz-cachon_asturian_2018, p. 6].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         ""           "u"
Mid       "e"         ""           "o"
Low        ""         "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```    

# Alphabet

* In some dialects the digraphs ⟨ḷḷ⟩, ⟨ḥ⟩, ⟨ts⟩, ⟨yy⟩ appear, corresponding to a variety of different phonemes [@academia_de_la_llingua_asturiana_gramatica_2001, p. 15].  
* ⟨j⟩, ⟨k⟩, and ⟨w⟩ appear in loanwords [@academia_de_la_llingua_asturiana_gramatica_2001, p. 15].   


```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme                              Comment
"a"            "/a/"                                 ""
"b"            "/b/"                                 ""
"c"          "/k/; /θ/"    "/k/: preceding back vowels; /θ/: preceding front vowels"    
"d"            "/d̪/"                                 ""
"e"            "/e/"                                 ""
"f"            "/f/"                                 ""
"g"            "/ɡ/"                                 ""
"h"            "∅"                                   ""
"i"            "/i/"                                 ""
"l"            "/l/"                                 ""
"m"            "/m/"                                 ""
"n"            "/n/"                                 ""
"ñ"            "/ɲ/"                                 ""
"o"            "/o/"                                 ""
"p"            "/p/"                                 ""
"r"          "/r/; /ɾ/"                      "/r/: word-initially"    
"s"            "/s/"                                 ""
"t"            "/t̪/"                                 ""
"u"            "/u/"                                 ""
"ü"            "/u/"                                 ""
"v"            "/b/"                                 ""
"x"            "/ʃ/"                                 ""
"y"            "/j/"                                 ""
"z"            "/θ/"                                 ""
**Digraph**    ""                                   ""
"ch"           "/tʃ/"                                ""                          
"gu"           "/ɡ/"                     "preceding front vowels"
"ll"           "/ʎ/"                                 ""
"qu"           "/k/"                     "preceding front vowels"
"rr"           "/r/"                                 ""
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* Fricatives /θ/, /s/, and /ʃ/ tend to become voiced in coda positions [@academia_de_la_llingua_asturiana_gramatica_2001, pp. 39-40].  
* /d/ spirantizes intervocalically and followed by /r/ [@academia_de_la_llingua_asturiana_gramatica_2001, p. 38].  
* /b/, /d/, /ɡ/ become approximates word-medially, following obstruents [@muniz-cachon_asturian_2018, p. 3].    
* In addition to approximating, /ɡ/ often spirantizes to /ɣ/ (ibid.).  

# Misc. Rules

* An apostrophe indicates a sound is missing - synonymous with a contraction [@academia_de_la_llingua_asturiana_gramatica_2001, p. 18].  
* Dashes are used to separate sounds that would otherwise be altered due to specified digraphs (e.g. retention of pronouns) [@academia_de_la_llingua_asturiana_gramatica_2001, p. 18].  
* /r/ and /ɾ/ are contrastive word-medially. Otherwise, /r/ occurs word-initially and /ɾ/ occurs word-finally [@muniz-cachon_asturian_2018, p. 3].  

# References
