---
title: "Greek"
author: "Shiying Yang"
date: "4/29/2019"
output:
  pdf_document:
    dev: cairo_pdf
    keep_tex: yes
    latex_engine: xelatex
    number_sections: yes
  html_document:
    df_print: paged
geometry: margin=1.2in
header-includes:
- \usepackage{setspace}\doublespacing
- \usepackage{tipa}
mainfont: DejaVu Serif
fontsize: 12pt
tables: yes
bibliography: el.bib
---
# Background

**Language family:** 

* Indo-European
  + Hellenic/Greek

**Adopted variety:** 
Standard Modern Greek

# Phonology
```{r, echo=F, include=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
```

## Consonants [@IPA, p 1] {#con}
general mapping of phonemes and orthography

```{r el_con, echo=F, tab.cap = NULL, results='asis'}
el_c <- read.csv('el_c.csv')
#colnames(jp_c)[c(1:2,5)] <- c(' ', '  ', 'alveolo-palatal')
c_position <- c('labial', 'dental', 'alveolar', 'velar')
c_colname <- paste(c_position, 'O', sep = ',')
c_colname_in_table <- unlist(strsplit(c_colname, ','))

kable(el_c, caption = 'Consonants', 
      align = "c", longtable = T, 
      digits = 3, format = 'latex', 
      booktabs = T, col.names = c(' ', 'Voicing', c_colname_in_table)) %>% 
  collapse_rows(columns = c(1), valign = 'middle', latex_hline = 'custom', custom_latex_hline = 1) %>%
  kable_styling("striped", full_width = T)
```

## Vowels [@IPA] {#v}
/i ε  ɐ o u/

# Orthography
Α,α,ά  
Β,β  
Γ,γ   
Δ,δ  
Ε,ε,Έ,έ    
Ζ,ζ  
Η,η,Ή,ή  
Θ,θ  
Ι,ι,Ί,ί,Ϊ,ϊ,Ḯ,ΐ  
Κ,κ  
Λ,λ   
Μ,μ  
Ν,ν  
Ξ,ξ  
Ο,ο,Ό,ό  
Π,π  
Σ,ς/σ  
Τ,τ  
Υ,υ,Ύ,ύ,Ϋ,ϋ,Ϋ́,ΰ      
Φ,φ  
Χ,χ  
Ψ,ψ  
Ω,ω,Ώ,ώ   

## Correspondence between orthography and phonemes
* Most consonant mappings are shown in table \ref{tab:el_con}, here are mappings for vowels, as well as consonant combinations that do not correspond to single phonemes.
* Stressed vowels (shown by stress markings as seen in ά,έ,ή,ί,ό,ύ,ώ) are represented as the ipa symbol followed by a "1". e.g. /i1/  

|orthography|pronunciation|
|-----------|-------------|
|ξ|/ks/|
|ψ|/ps/|
|α|/ɐ/|
|ά|/ɐ1/|
|ω/ο|/o/|
|ώ/ό|/o1/|
|ι/ϊ/η/υ/ϋ/οι/ει|/i/|
|ί/ΐ/ή/ύ/ΰ/οί/εί|/i1/
|ε/αι|/ε/|
|έ/αί|/ε1/|
|ου|/u/|
|ού|/u1/|
|αυ|/ɐf/|
|αύ|/ɐ1f/|
|ευ|/εv/|
|εύ|/ε1v/|

# Rules

## Issues and representative phonological processes
1. All but two diacritics are kept in the Standard Modern Greek
- stress marking
- umlaut
2. Palatalization of velar consonants
- Velar consonants /k/, /g/, /x/, /ɣ/ surface as palatal allophones [c], [ɟ], [ç], [ʝ] in front of high vowels /i/ and /ε/
- [+velar, +consonantal] -> [+palatal]/__[+high, -consonantal]{\$ or [-back, -consonantal]}
- [+velar, +consonantal][+high, -consonantal] -> [+palatal]/__[+back, -consonantal]

## Lenition Processes

# References {-}
