---
title: "Greek"
author: "Shiying Yang"
edited by: "Emily Strand"
bibliography: el.bib
output: html_document
---

Last Updated: 2020-06-29

# Background

**Language family**: Indo-European / Hellenic / Greek

* The adopted variety is Standard Modern Greek.
* Speakers reside in a multitude of places, including Albania, Macedonia, Montenegro, Cyprus, Greece, and Ukraine.

# Phonology

## Consonants

* There is some controversy around the phonemic status of voiced stops. Although some authors categorize them surfacing allophonically from homorganic nasal-voiceless stop sequences [@Newton1961 as cited in @Arvaniti2007, p. 7], @Antoniou2008 explain that diachronically speaking, native speakers have come to perceive voiced stops as phonemic (p. 1). Given that @Arvaniti2007, @Mirambel1959, and @Householder1964 [both cited in @Arvaniti2007, p. 7] identify the voiced stops as independent phonemes, I have chosen to do so as well.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"    Labial   Dental    Alveolar   Velar
Stops                       "p  b"   "t̪  d̪"       ""      "k  ɡ"
Fricatives                  "f  v"   "θ  ð"     "s  z"    "x  ɣ"
Nasals                        "m"      "n̪"        ""        ""
Flaps                         ""       ""         "ɾ"       ""
Approximants                  ""       ""         "l"       ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Dental", "Alveolar", "Velar"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 4)) %>%
  column_spec(1, bold = TRUE)
```

## Vowels

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
       Front   Central   Back
High   "i"     ""        "u"
Mid    "ε"     ""        "o"
Low    ""      "ɐ"       ""
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

* Modern Greek includes two diacritics: the acute accent (΄) and the diaresis (¨) [@Ktori2008, p. 774]. The acute accent marks stress, and the diaresis marks the hiatus or the enunciation of two independent adjacent vowels.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
"Uppercase"   "Lowercase"  "Phoneme"                   "Comment"
"Α"              "α"          "/ɐ/"                       ""
"Β"              "β"          "/v/"                       ""
"Γ"              "γ"          "/ɣ/"                       ""
"Δ"              "δ"          "/ð/"                       ""
"Ε"              "ε"          "/ε/"                       ""
"Ζ"              "ζ"          "/z/"                       ""
"Η"              "η"          "/i/"                       ""
"Θ"              "θ"          "/θ/"                       ""
"Ι"              "ι"          "/i/"                       ""
"Κ"              "κ"          "/k/"                       ""
"Λ"              "λ"          "/l/"                       ""
"Μ"              "μ"          "/m/"                       ""
"Ν"              "ν"          "/n̪/"                       ""
"Ξ"              "ξ"          "/ks/"                      ""
"Ο"              "ο"          "/o/"                       ""
"Π"              "π"          "/p/"                       ""
"Ρ"              "ρ"          "/ɾ/"                       ""
"Σ"            "ς; σ"         "/s/"                       ""
"Τ"              "τ"          "/t̪/"                       ""
"Υ"              "υ"          "/i/"                       ""
"Φ"              "φ"          "/f/"                       ""
"Χ"              "χ"          "/x/"                       ""
"Ψ"              "ψ"          "/ps/"                      ""
"Ω"              "ω"          "/o/"                       ""
"**Digraph**"     ""           ""                         ""
"ΕΙ"             "ει"         "/i/"                       ""
"ΟΙ"             "οι"         "/i/"                       ""
"ΥΙ"             "υι"         "/i/"                       ""
"ΑΙ"             "αι"         "/ɛ/"                       ""
"ΟΥ"             "ου"         "/u/"                       ""
"ΑΥ"             "αυ"         "/ɐv/; /ɐf/"   "/ɐv/: preceding vowels or voiced consonants [@Dawkins1908, p. 218]"
"ΕΥ"             "ευ"         "/εv/; /εf/"   "/εv/: preceding vowels or voiced consonants [@Dawkins1908, p. 218]"
"ΜΠ"             "μπ"         "/b/"                       ""
"ΝΤ"             "ντ"         "/d̪/"                       ""
"ΓΚ"             "γκ"         "/ɡ/"                       ""
"ΓΓ"             "γγ"         "/ɡ/"                       ""
'), TRUE)

kable(alphabet, col.names = c("Uppercase", "Lowercase", "Phoneme", "Comment"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("Grapheme" = 2, "", ""))

```

# Lenition Rules

* Word-final /s/ becomes voiced when preceding a word-initial voiced consonant [@Arvaniti1999, p. 4].
* Although rare, in fast speech, voiced stops may be realized as fricatives [@Arvaniti2000, as cited in @Arvaniti1999, p. 2].
* /ɾ/ may be realized as an alveolar approximant [@Arvaniti1999, pp. 2-3].
* /p/, /t̪/, and /k/ are often voiced intervocalically [@Arvaniti2007, p. 102].

# Misc. Rules

* /k/, /ɡ/, /x/, /ɣ/ surface as palatal allophones [c], [ɟ], [ç], [ʝ] in front of high vowels /i/ and /ε/ [@Arvaniti1999, p. 3].
* In formal speech, voiced stops are pronounced with some prenasalization [@Arvaniti1999, p. 2], although this is becoming more obselete with younger generations.
* In consonant clusters, nasals often assimilate in place to the following consonant [@Arvaniti1999, p. 2].
* In fast speech, adjacent identical vowels across word boundaries degeminate [@Arvaniti1999, p. 4].

# References
