---
title: "Jamaican Creole"
author: "Justin Bai"
bibliography: jam.bib
output: html_document
---
# Background

**Language family**: English creole

* Also referred to as Jamaican Patois, Patwa, or Patwah, Jamaican Creole is spoken in a multitude of places. These include Jamaica, the United Kingdom, the United States, Canada, other Caribbean islands, and throughout the Caribbean coast of Central America.

# Phonology

## Consonants

* There is some debate about the status of /h/ in Jamaican Creole. It appears that "[h] may be phonetic in Eastern varieties" but "is phonemic in the Western varieties" [@harry2006jamaican, p. 126]. I have chosen to include it below.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial     Alveolar   Postalveolar   Palatal     Velar    Glottal
Stops                      "p  b"     "t  d"         ""           ""       "k  ɡ"      ""
Affricates                   ""         ""        "tʃ  dʒ"        ""         ""        ""
Fricatives                 "f  v"     "s  z"         "ʃ"          ""         ""        "h"
Nasals                       "m"        "n"          ""           ""         "ŋ"       ""
Approximants                 "w"      "ɹ  l"         ""           "j"        ""        ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced. Both alveolar approximants are voiced, but the one on the right is lateral.", general_title = "")
```

## Vowels

* Jamaican Creole makes a vowel length distinction. Long vowels are indicated by repeated graphemes (/e/ and /o/ do not have long counterparts) [@harry2006jamaican, p. 128].
* Jamaican Creole has diphthongs. The only diphthong that is not transparent orthographically is /au/, which is spelled ⟨ou⟩ (ibid.).
* There is some uncertainty regarding the concept of contrastive tone in Jamaican Creole. @Lawton1968 argues that tone is contrastive, whereas @Parsard explains that speakers of creole languages do not make use of it (p. 23; p. 7). Due to this uncertainty, I will proceed with the language not accounting for tone.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         ""           "u"
Mid       "e"         ""           "o"
Low        ""         "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)


diphthongs <- read.table(textConnection('
Diphthongs
"/ai/, /ua/, /ia/, /au/"
'), TRUE)

kable(diphthongs, align = 'c') %>%
  kable_styling("bordered")
```   

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme
"a"            "/a/"
"b"            "/b/"
"d"            "/d/"
"e"            "/e/"
"f"            "/f/"
"g"            "/ɡ/"
"h"            "/h/"
"i"            "/i/"
"j"            "/dʒ/"
"k"            "/k/"
"l"            "/l/"
"m"            "/m/"
"n"            "/n/"
"o"            "/o/"
"p"            "/p/"
"r"            "/ɹ/"
"s"            "/s/"
"t"            "/t/"
"u"            "/u/"
"v"            "/v/"
"w"            "/w/"
"y"            "/j/"
"z"            "/z/"
**Digraph**     ""
"ch"           "/tʃ/"
"ng"           "/ŋ/"
"sh"           "/ʃ/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* Obstruents become palatalized preceding high vowels /i/ and /u/ [@harry2006jamaican, p. 127].
* /b/, /d/, and /ɡ/ are pronounced lenis (specifically as implosives) word-initially (ibid.).

# References
