---
title: "Paumarí"
author: "Emily Strand"
bibliography: pad.bib
output: html_document
---

Last updated: 2020-03-06

# Background

**Language Family:** Arauan

* Paumarí is spoken in the Amazonas state of Brazil.

# Phonology

## Consonants

* The Paumarí language is very interesting as it contrasts stops on three different dimensions, two of which (implosives and aspirates) are not homogeneous [@Everett2003, p. 23].
* @Chapman1998 and @Chapman1991 (p. 346) include /ʃ/ in the phonemic inventory, whereas @Everett2003 (p. 23) doesn't. I have chosen to include it, given that it is represented specifically in the orthography, but it seems extremely marginal (only present in 0.1% of the Crúbadán corpus).

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial    Alveolar  Postalveolar  Velar   Glottal
"Stops (plain)"            "p b"     "t d"     ""            "k ɡ"   "ʔ"
"Stops (aspirated)"        ""        "tʰ"      ""            "kʰ"    ""
"Stops (implosive)"        "ɓ"       "ɗ"       ""            ""      ""
Affricates                 ""        ""        "tʃ dʒ"       ""      ""
Fricatives                 "f"       "s"       "ʃ"           ""      "h"
Nasals                     "m"       "n"       ""            ""      ""
Flaps                      ""        "ɾ"       ""            ""      ""
Approximants               ""        "w"       ""            ""      ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 5)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```

## Vowels

* Adjacent vowels are interpreted as separate nuclei [@Chapman1991, p. 346].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High     "i"          ""           ""         
Mid      ""           ""           "o"
Low      ""           "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```  

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme
"a"            "/a/"
"b"            "/b/"
"\'b"          "/ɓ/"
"d"            "/d/"
"\'d"          "/ɗ/"
"f"            "/f/"
"g"            "/ɡ/"
"h"            "/h/"
"i"            "/i/"
"j"            "/dʒ/"
"k"            "/k/"
"m"            "/m/"
"n"            "/n/"
"o"            "/o/"
"p"            "/p/"
"r"            "/ɾ/"
"s"            "/s/"
"t"            "/t/"
"v"            "/w/"
"x"            "/ʃ/"
"\'"           "/ʔ/"
**Digraph**    ""
"kh"           "/kʰ/"
"th"           "/tʰ/"
"tx"           "/tʃ/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Misc. Rules

* Predictable stress patterns elicit phonetic lengthening of both vowels and consonants [@Chapman1991, p. 346].

# References
