---
title: "Ankave"
author: "Jordan Schuster"
bibliography: aak.bib
output: html_document
---

Last Updated: 2019-12-16

# Background

**Language Family:** Trans-New Guinea / Main Section / Central and Western / Angan / Angan Proper

* Also referred to as Angave, this language is spoken in the Gulf Province of Papua New Guinea.

# Phonology

## Consonants


```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"       Bilabial   Alveolar    Palatal    Velar    Glottal
"Stops (plain)"                  "p"        "t"          ""       "k"       "ʔ"
"Stops (prenasalized)"           "ᵐb"       "ⁿd"         ""       "ᵑɡ"      ""
"Affricates (prenasalized)"      ""         "ⁿdz"        ""       ""        ""
Fricatives                       ""         "s"          ""       "x"       ""
Nasals                           "m"        "n"          ""       "ŋ"       ""
Flaps                            ""         "ɾ"          ""       ""        ""
Approximants                     ""         ""           "j"      "w"       ""

'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Bilabial", "Alveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 5))  %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* @speece1992 transcribes ⟨ɨ⟩ as /ə/; however, it is not a true schwa, so we have chosen to represent it as /ɘ/.


```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
          Front   Central   Back
High      "i"      ""        "u"
Mid       "e"      "ɘ"       "o"
Low       ""       "a"       ""
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

**Note:** ⟨v⟩ appears in Crúbadán only within "Ankave". Our guess is that it acts similarly to other languages in the family, such that it varies freely with [w]. However, we have not yet found evidence to support this.


```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme  
"a"              "/a/"
"b"              "/ᵐb/"
"d"              "/ⁿd/"
"e"              "/e/"
"g"              "/ᵑɡ/"
"i"              "/i/"
"ɨ"              "/ɘ/"
"j"              "/ⁿdz/"
"k"              "/k/"
"x"              "/x/"
"m"              "/m/"
"n"              "/n/"
"ŋ"              "/ŋ/"
"o"              "/o/"
"p"              "/p/"
"r"              "/ɾ/"
"s"              "/s/"
"t"              "/t/"
"u"              "/u/"
"w"              "/w/"
"y"              "/j/"
"\'"             "/ʔ/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Ankave syllables include the following structures [@speece1992, p. 3]:

    - VV
    - VC
    - CV
    - CVV
    - CVVV

# Misc Rules

* Vowels may be followed optionally by a glottal stop [@speece1992, p. 3].

# References
