---
title: "Iduna"
author: "Bill Mizgerd"
bibliography: viv.bib
output: html_document
---

Last Updated: 2019-07-15

# Background

**Language Family**: Austronesian / Malayo-Polynesian / Eastern Malayo-Polynesian / Oceanic / Western Oceanic / Papuan Tip / Nuclear Papuan Tip / North Papuan Mainland-D'Entrecasteaux / Bwaidoga / Iduna

* Also known as Vivigani, Iduna is spoken in the Milne Bay province of Papua New Guinea.

# Phonology

## Consonants

* [p] occurs in free variation with /b/ [@VivSampson1965, p. 10].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial  Alveolar  Palatal  Velar  Glottal
Stops                      "b"     "t d"     ""       "k ɡ"  "ʔ"
Fricatives                 "f v"   ""        ""       ""     "h"
Nasals                     "m"     "n"       ""       ""     ""
Approximants               "w"     "l"       "j"      ""     ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 5)) %>%
  footnote("Note: For phonemes that share a cell, those on the left are voiceless, whereas those on the right are voiced.", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* @VivSampson1965 argues for /a/ rather than /ɑ/ (p. 21).

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
          Front   Back
High      "i"     "u"
Mid       "e"     "o"
Low       ""      "ɑ"
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered")
```

# Alphabet

* /u/ (rather than /w/) is held to be the underlying phoneme for consonant labialization, because Iduna disallows consonant clusters [@VivSampson1965, p. 30; @VivLucht1992, p. 3].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme                 Comment
"a"            "/ɑ/"                    ""
"b"            "/b/"                    ""
"d"            "/d/"                    ""
"e"            "/e/"                    ""
"f"            "/f/"                    ""
"g"            "/ɡ/"                    ""
"h"            "/h/"                    ""
"i"            "/i/"                    ""
"k"            "/k/"                    ""
"l"            "/l/"                    ""
"m"            "/m/"                    ""
"n"            "/n/"                    ""
"o"            "/o/"                    ""
"s"            "/t/"      "[s] is just a spirantized allophone of /t/"  
"t"            "/t/"                    ""
"u"            "/u/"                    ""
"v"            "/v/"                    ""
"w"            "/u/; /w/"     "/u/: in C_V position; /w/: elsewhere"
"y"            "/j/"                    ""
"\'"           "/ʔ/"                    ""
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* /t/ spirantizes to [s] before front vowels [@VivLucht1992, p. 3; @VivSampson1965, p. 9].
* Word-medial /ɡ/ spirantizes to [ɣ] (ibid.).
* In C_V contexts, /u/ is realized as simply a labialization of the preceding consonant [@VivSampson1965, p. 6].

# Misc. Rules

* Phrase-final /d/ devoices and aspirates, yielding [tʰ] [@VivLucht1992, p. 3].
* /u/ nasalizes after /m/ [@VivSampson1965, p. 20].
* Unstressed /ɑ/ raises to [ʌ] [@VivSampson1965, p. 21].

# References
