---
title: "Nepali"
author: "Becky Mathew"
edited by: "Emily Strand"
bibliography: ne.bib
output: html_document
---

Last Updated: 2020-02-27

**Compromised: certain diacritics used interchangeably and inconsistently marked**

# Background

**Language family**: Indo-European / Indo-Iranian / Indo-Aryan / Northern Zone / Eastern Pahari

* Nepali is the official language of Nepal and is also spoken in India and Bhutan and by members of the Nepalese diaspora [@JIPA2009, p. 373]. My rules are written based on the eastern Nepali dialect. Like many other Indic languages it has a left-to-right abugida alphabet.

# Phonology

## Consonants

* @Acharya1991 identifies palatal stops rather than alveolar affricates (p. 19).
    - I opted for the analysis done by @JIPA2009, given that it was the most recent.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"             Labial     Dental   Alveolar     Retroflex   Palatal     Velar     Glottal
"Stops (plain)"                      "p  b"     "t̪  d̪"      ""         "ʈ  ɖ"       ""       "k  ɡ"       ""
"Stops (aspirated)"                 "pʰ  bʱ"    "t̪ʰ d̪ʱ"     ""         "ʈʰ  ɖʱ"     ""       "kʰ  ɡʱ"     ""
"Affricates (plain)"                  ""          ""      "ts  dz"       ""         ""         ""         ""
"Affricates (aspirated)"              ""          ""      "tsʰ dzʱ"      ""         ""         ""         ""
Fricatives                            ""          ""        "s"          ""         ""         ""         "ɦ"
Nasals                                "m"        "n̪"        ""           ""         ""         "ŋ"        ""
Trills                                ""         ""         "r"          ""         ""         ""         ""
Approximants                          "w"        ""         "l"          ""         "j"        ""         ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Dental", "Alveolar", "Retroflex", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 7)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```  

## Vowels

* Nasal vowels are considered phonemic in Nepali [@JIPA2009, p. 377]. /o/ does not have a contrastive nasal counterpart, but it may vary freely with it.
* I have chosen to follow the vowel outline by @JIPA2009, such that /ʌ/ is used rather than /ə/ (p. 377). This is based on the description that the vowel in question is lower and more back than schwa.
* Vowel length is not considered a phonemic feature [@JIPA2009, p. 378].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         ""           "u"
Mid       "e"         ""           "o"
Low-Mid    ""         ""           "ʌ"
Low        ""         "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)


diphthongs <- read.table(textConnection('
Diphthongs
"/ai/, /au/"
'), TRUE)

  kable(diphthongs, align = 'c') %>%
    kable_styling("bordered")
  ```

# Alphabet

* Nepali is written using the Devanagari script [@Bal2004, pp. 333-334]. Each consonant comes with an inherent /ʌ/ following.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme                   Comment
"क"           "/kʌ/"                      ""
"ख"           "/kʰʌ/"                      ""
"ग"           "/ɡʌ/"                       ""
"घ"           "/ɡʱʌ/"                       ""
"ङ"           "/ŋʌ/"                       ""
"च"           "/tsʌ/"                      ""
"छ"           "/tsʰʌ/"                      ""
"ज"           "/dzʌ/"                      ""
"झ"           "/dzʱʌ/"                      ""
"ट"           "/ʈʌ/"                       ""
"ठ"           "/ʈʰʌ/"                      ""
"ड"           "/ɖʌ/"                       ""
"ढ"           "/ɖʱʌ/"                      ""
"त"           "/t̪ʌ/"                       ""
"थ"           "/t̪ʰʌ/"                      ""
"द"           "/d̪ʌ/"                       ""
"ध"           "/d̪ʱʌ/"                      ""
"न; ञ; ण"    "/n̪ʌ/"     "the palatal and retroflex nasals are considered allophones of /n̪/ [@Acharya1991, p. 13]"
"प"           "/pʌ/"                       ""
"फ"           "/pʰʌ/"                     ""
"ब"           "/bʌ/"                      ""
"भ"           "/bʱʌ/"                     ""     
"म"           "/mʌ/"                      ""
"य"           "/jʌ/"                      ""       
"र"           "/rʌ/"                      ""
"ल"           "/lʌ/"                      ""
"व"           "/wʌ/"                      ""  
"श; ष; स"    "sʌ/"     "the voiceless retroflex and postalveolar fricatives are considered allophones of /s/ [@Acharya1991, p. 13]"
"ह"           "/ɦʌ/"                      ""
"**Independent Character**"    ""         ""       
"अ"           "/ʌ/"                      ""
"आ"          "/a/"                       ""
"इ; ई"        "/i/"          "includes both the short and long variants given that vowel length is not contrastive"
"उ; ऊ"       "/u/"          "includes both the short and long variants given that vowel length is not contrastive"
"ऋ"          "/ri/"                      ""
"ए"           "/e/"                       ""
"ओ"          "/o/"                       ""
"ऐ"          "/ai/"                       ""
"औ"         "/au/"                       ""
"**Diacritic**" ""                        ""
"ा"           "/a/"                       ""
"ि; ी"       "/i/"         "includes both the short and long variants given that vowel length is not contrastive"
"ु; ू"        "/u/"         "includes both the short and long variants given that vowel length is not contrastive"
"ृ"           "/ri/"                       ""
"े"           "/e/"                        ""
"ै"           "/ai/"                       ""
"ो"          "/o/"                         ""
"ौ"          "/au/"                        ""
"ँ"            ""                 "called a candrabindu, this diacritic indicates nasalization of the preceding vowel [@Bal2004, p. 333]"
"ं"           "/n/"               "called an anusvar or sirabindu, this diacritic indicates a homoorganic nasal with following stops (/n/ appears to be underlying); however, it is inconsistently marked and used interchangeably with the candrabindu [@Bal2004, p. 333] - this compromises the language to some extent"
"्"           "∅"                "called a viram or halanta, this diacritic removes the inherent vowel /ʌ/ after a consonant [@Bal2004, p. 334]"
"ः"          "/ɦ/"              "called the visarga, this diacritic only appears in Sanskrit loanwords, so it won\'t be transcribed in the rule set [@Bal2004, p. 334]"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* /ɖ/ is realized as a retroflex flap when preceded by a vowel [@JIPA2009, p. 374].
* /pʰ/, /bʱ/, /kʰ/, and /ɡʱ/ may spirantize in spontaneous speech [@JIPA2009, p. 376].
* /r/ may be realized as a flap [@JIPA2009, p. 377]:
    - Intervocalically [@Pokharel1989 as cited in @JIPA2009, p. 377]
    - Word-initially
    - Word-finally
* Voiced aspirates may be realized without aspiration intervocalically and word-finally [@JIPA2009, p. 376].
* /b/, /d/, and /ɡ/ may be realized as nasals postvocalically, preceding nasal vowels [@Acharya1991, p. 26].
* One consonant within a cluster may be deleted (unclear whether this is only word-initially) [@Acharya1991, p. 33].

# Misc. Rules

* Consonants, apart from /ɦ/, /j/, and /w/, undergo gemination [@JIPA2009, p. 376].
* /d/ and /ɡ/ may be devoiced word-finally [@Acharya1991, p. 26].
* Velar stops become fronted preceding front vowels [@Acharya1991, p. 26].
* Non-syllabic /i/ and /u/ are realized as [j] and [w], respectively [@JIPA2009, p. 373].
* Vowel epenthesis occurs within consonant clusters [@Acharya1991, p. 32].

# References
