---
title: "Inuktitut"
author: "Bill Mizgerd"
bibliography: iu.bib
output: html_document
---

Last Updated: 2019-07-10

# Background

**Language Family**: Eskimo-Aleut / Eskimo / Inuit / Inuktitut

* Where dialects diverge, this analysis will be preferring Eastern Canadian Inuktitut.
* Inuktitut is spoken in Canada, specifically in the provinces of Newfoundland and Labrador, Nunavut, and Quebec.

# Phonology

## Consonants

* /r/ survives in a small number of dialects [@IuDorais1990, p. 215].
* /h/ is not normally included in the phonemic inventory given that it is quite infrequent and only appears in some dialects [@IuDorais1986, p. 29]. I have chosen not to include it in the inventory here, and likewise it will not be represented in the rules.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial     Alveolar    Palatal     Velar     Uvular
Stops                       "p"         "t"          ""        "k"       "q"
Fricatives                  "v"        "s  ɬ"        ""        "ɣ"       "ʁ"
Nasals                      "m"         "n"          ""        "ŋ"       ""
Approximants                 ""         "l"          "j"       ""        ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Palatal", "Velar", "Uvular"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 5)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: The alveolar fricative on the right is lateral.", general_title = "")
```

## Vowels

* Vowel length is contrastive [@IuDorais1990, p. 214]. Long vowels are indicated by doubling the vowel character [@IuDorais2010, p. 56].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"        ""           "u"
Mid       ""         ""           ""
Low       ""         "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```   

# Alphabet

* Regarding the transcription, there are two rulesets (and, accordingly, two verify files): one for the Latin-alphabet orthography of Inuktitut, and one for the Inuktitut syllabary.

## Inuit Syllabary

* A dot can be appended above a character to indicate a long syllable.
* See @IuDorais1990 (p. 241) for details.

| ᐃ - /i/  | ᐅ - /u/  | ᐊ - /a/  | ᐁ - /ai/  |  Final   |
|:--------:|:--------:|:--------:|:---------:|:--------:|
| ᐱ	- /pi/ | ᐳ - /pu/ | ᐸ - /pa/ | ᐯ - /pai/ | ᑉ - /p/ |
| ᑎ - /ti/ | ᑐ - /tu/ | ᑕ - /ta/ | ᑌ - /tai/ | ᑦ - /t/ |
| ᑭ - /ki/ | ᑯ - /ku/ | ᑲ - /ka/ | ᑫ - /kai/ | ᒃ - /k/ |
| ᒋ - /ɣi/ | ᒍ - /ɣu/ | ᒐ - /ɣa/ | ᒉ - /ɣai/ | ᒡ - /ɣ/ |
| ᒥ - /mi/ | ᒧ - /mu/ | ᒪ - /ma/ | ᒣ - /mai/ | ᒻ - /m/ |
| ᓂ - /ni/ | ᓄ - /nu/ | ᓇ - /na/ | ᓀ - /nai/ | ᓐ - /n/ |
| ᓯ - /si/ | ᓱ - /su/ | ᓴ - /sa/ | ᓭ - /sai/ | ᔅ - /s/ |
| ᓕ - /li/ | ᓗ - /lu/ | ᓚ - /la/ | ᓓ - /lai/ | ᓪ - /l/ |
| ᔨ - /ji/ | ᔪ - /ju/ | ᔭ - /ja/ | ᔦ - /jai/ | ᔾ - /j/ |
| ᕕ - /vi/ | ᕗ - /vu/ | ᕙ - /va/ | ᕓ - /vai/ | ᕝ - /v/ |
| ᕆ - /ʁi/ | ᕈ - /ʁu/ | ᕋ - /ʁa/ | ᕃ - /ʁai/ | ᕐ - /ʁ/ |
| ᕿ - /qi/ | ᖁ - /qu/ | ᖃ - /qa/ | ᙯ - /qai/ | ᖅ - /q/ |
| ᖏ - /ŋi/ | ᖑ - /ŋu/ | ᖓ - /ŋa/ | ᙰ - /ŋai/| ᖕ - /ŋ/ |
| ᖠ - /ɬi/ | ᖢ - /ɬu/ | ᖤ - /ɬa/ |            | ᖦ - /ɬ/ |

## Roman Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme       
"a"            "/a/"         
"g"            "/ɣ/"          
"i"            "/i/"         
"j"            "/j/"         
"k"            "/k/"         
"l"            "/l/"         
"&"            "/ɬ/"         
"m"            "/m/"         
"n"            "/n/"       
"p"            "/p/"       
"q"            "/q/"      
"r"            "/ʁ/"      
"s"            "/s/"      
"t"            "/t/"     
"u"            "/u/"      
"v"            "/v/"       
**Digraph**     ""          
"ng"           "/ŋ/"       
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* In some dialects, /s/ debuccalizes to [h] [@IuDorais1990, pp. 215-216].
* In some dialects, /t/ spirantizes to [s] - or reduces even farther to [h] - when preceded by /i/ [@IuDorais1990, pp. 216-217].
* In some Alaskan dialects, /p/ spirantizes to [v] and /v/ approximantizes to [w] when preceded by unstressed syllables [@IuDorais1986, p. 25].
* In some dialects, /q/ spirantizes to [χ] or [ʁ] (ibid.).
* In some Greenlandic dialects, intervocalic /ŋ/ deletes [@IuDorais1986, p. 27].
* In dialects in Greenland and western Canada, word-final /i/ deletes [@IuDorais1986, p. 31].

# Misc. Rules

* Regressive assimilation occurs in consonant clusters, with the extent of the assimilation varying based on dialect [@IuDorais1986, pp. 34-35].

# References
