---
title: "Huallaga Huánuco Quechua"
author: "Emily Strand"
bibliography: qub.bib
output: html_document
---

Last Updated: 2019-08-21

**Compromised: unclear documentation pertaining to the orthography; possible conflation with vowels and glides**

# Background

**Language Family**: Quechuan / Quechua I

* Huallaga Huánuco Quechua is spoken in the Huanuco region of Peru.

# Phonology

## Consonants

* /b/, /d/, /ɡ/, /f/, and /ʐ/ appear in loanwords [@weber_grammar_1989, p. 450].
* Some sources indicate that /r/ is in fact an alveolar flap. Although @weber_rimaycuna_2008 and @weber_grammar_1989 don't make this distinction, I have opted to treat it as a flap, given that other dialects of Quechua (e.g. Cuzco Quechua) classify it as such.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial    Alveolar  Postalveolar   Palatal  Velar   Uvular  Glottal
Stops                      "p"         "t"        ""            ""      "k"     "q"      ""
Affricates                 ""          ""         "tʃ"          ""      ""      ""       ""
Fricatives                 ""          "s"        "ʃ"           ""      ""      ""       "h"
Nasals                     "m"         "n"        ""            "ɲ"     ""      ""       ""
Flaps                      ""          "ɾ"        ""            ""      ""      ""       ""
Approximants               ""          "l"        ""           "j ʎ"    "w"     ""       ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Uvular", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 7)) %>%
  footnote(general = "Note: The palatal approximant on the right is lateral.", general_title = "") %>%
  column_spec(1, bold = TRUE)
```   

## Vowels

* Vowel length is contrastive. Long vowels are either marked with a diareses (¨) or duplicate vowels, depending on the orthography [@weber_rimaycuna_2008, p. 14; @weber_grammar_1989, p. 456].
* [e] and [o] are allophones of /i/ and /u/ respectively, and used in loanwords [@weber_grammar_1989, p. 450].
* Apart from long vowels (duplicate graphemes), vowels do not occur in succession [@weber_grammar_1989, p. 455].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
           Front      Central       Back
High        "i"         ""           "u"
Low         ""          "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

* It's somewhat unclear as to what constitutes the Huallaga Huánuco Quechua alphabet. The Peruvian Ministry of Education publicly recognized the language in 2015, but the resources don't seem to be accessible. @weber_grammar_1989 lists variations between the Peruvian Ministry of Education (PME), the Departmental Office of Education (DOE), and Hispanic orthography (p. 456), of which I have noted below. Crúbadán uses graphemes from all three systems, so all those listed below will be represented in the rule set.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme            Comment
"a"              "/a/"               ""
"ä"              "/aː/"             "PME"        
"c"              "/k/"           "Hispanic"         
"e"              "/i/"               ""
"ë"              "/iː/"              ""
"g"              "/q/"          "DOE; Hispanic"
"h"              "/h/"              "PME"
"i"              "/i/"               ""
"i"              "/j/"            "Hispanic"
"ï"              "/iː/"              ""
"j"              "/h/"          "DOE; Hispanic"
"k"              "/k/"            "PME; DOE"
"l"              "/l/"               ""
"m"              "/m/"               ""
"n"              "/n/"               ""
"ñ"              "/ɲ/"               ""
"o"              "/u/"               ""
"ö"              "/uː/"              ""
"p"              "/p/"               ""
"q"              "/q/"              "PME"
"r"              "/ɾ/"               ""
"s"              "/s/"               ""
"t"              "/t/"               ""
"u"              "/u/"               ""
"u"              "/w/"            "Hispanic"
"ü"              "/uː/"              ""
"w"              "/w/"               ""
"y"              "/j/"               ""
**Digraph**       ""                 ""
"ch"             "/tʃ/"              ""
"ll"             "/ʎ/"               ""
"qu"             "/k/"     "Hispanic: preceding front vowels"
"sh"             "/ʃ/"               ""
"hu"             "/w/"           "Hispanic"
"gu"             "/q/"     "Hispanic: preceding front vowels"
"aa"             "/aː/"             "DOE"
"ee"             "/iː/"             "DOE"
"ii"             "/iː/"             "DOE"
"oo"             "/uː/"             "DOE"
"uu"             "/uː/"             "DOE"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* The syllable structure consists of (C)V(C). The onset is required except in an initial syllable [@weber_grammar_1989, p. 454].

# Lenition Rules

* /q/ fricates word-finally or preceding a voiceless obstruent [@weber_grammar_1989, p. 452].
* /q/ becomes voiced intervocalically, spirantizes preceding a voiceless consonant, and deletes word-finally [@weber_grammar_1989, p. 452].

# Misc. Rules

* /n/ assimilates in place to the following consonant [@weber_grammar_1989, p. 452].
* /r/ becomes a retroflex fricative word-initially and in sequences ⟨rr⟩ [@weber_grammar_1989, p. 453].

# References
