---
title: "Huichol"
author: "Emily Strand"
bibliography: hch.bib
output: html_document
---

Last updated: 2019-10-17

# Background

**Language family**: Uto-Aztecan / Southern / Sonoran / Corachol

* Huichol is spoken in the Nayarit and Jalisco states, located near Mexico City.

# Phonology

* Although varying tones are found in Huichol [@grimes_huichol_1959], whether they are contrastive or not is rather unclear [@banerji_huichol_2014, p. 15]. @banerji_huichol_2014 explains that they are most likely not contrastive, but rather different phonological realizations (p. 18).

## Consonants

* There is not a voicing contrast in Huichol [@bierge_clause_2017, p. 39].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Bilabial   Alveolar    Palatal     Velar    Glottal
Stops                       "p"         "t"         ""       "k  kʷ"    "ʔ"
Affricates                   ""         "ts"        ""         ""       ""
Fricatives                   ""         ""          ""         ""       "h"
Nasals                       "m"        "n"         ""         ""       ""
Flaps                        ""         "ɾ"         ""         ""       ""
Trills                       ""         "r"         ""         ""       ""
Approximants                 "w"        ""          "j"        ""       ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 5)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: The velar stop on the right is labialized.", general_title = "")
```

## Vowels

* Although very few minimal pairs exist, vowel length is contrastive [@bierge_clause_2017, p. 36]. Long vowels are indicated by repeated graphemes.
* 13 diphthongs can be found in the Huichol language (@leza_gramatica_1999, as cited in @bierge_clause_2017, pp. 37-38).

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         "ɨ"          "u"
Mid       "e"         ""           ""
Low        ""         "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)

diphthongs <- read.table(textConnection('
Diphthongs
"/ai/, /au/, /aɨ/, /ua/, /ui/, /ei/, /eɨ/, /eu/, /ie/, /iɨ/, /iu/, /ɨi/, /ɨa/"
'), TRUE)

kable(diphthongs, align = 'c') %>%
  kable_styling("bordered")

```  

# Alphabet

* There have been several accounts concerning the Huichol orthography. The orthography immediately below represents that of the Departamento de Lenguas Indígenas (DELI) (@leza_gramatica_1999 as cited in @bierge_clause_2017, pp. 44-45). Immediately following, I have also included the different orthographic variations.
  * Most of the variations were indicated by @leza_gramatica_1999, following either the Secretaría de Educación Pública (SEP) or the Summer Institute of Linguistics (SIL) (as cited in @bierge_clause_2017, pp. 44-45).

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme              Comment
"a"            "/a/"                 ""
"e"            "/e/"                 ""
"h"            "/h/"                 ""
"i"            "/i/"                 ""
"ɨ"            "/ɨ/"                 ""
"k"            "/k/"                 ""
"m"            "/m/"                 ""
"n"            "/n/"                 ""
"p"            "/p/"                 ""
"r"            "/ɾ/"                 ""
"t"            "/t/"                 ""
"u"            "/u/"                 ""
"w"            "/w/"                 ""
"x"            "/r/"                 ""
"y"            "/j/"                 ""
"\'"            "/ʔ/"                ""
**Digraph**    ""                    ""
"kw"          "/kʷ/"                 ""
"ts"          "/ts/"                 ""
"**Orthographic Variations**"  ""    ""
"qu; c"       "/k/"                "(SIL)"
"ku"          "/kʷ/"               "(SEP)"
"cu"          "/kʷ/"               "(SIL)"  
"s"           "/ts/"               "(SIL)"  
"tz"          "/ts/"               "(SEP)"  
"xr; rr"      "/r/"                "(SEP)"  
"j"           "/h/"                "(SEP)"  
"v"           "/w/"                "(SIL)"  
"ü"           "/ɨ/"          "[@haugen_kill_2015]"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```  

# Syllable Structure

* The syllable structures permitted in Huichol follow the order of CV(V) [@banerji_huichol_2014, p. 6].
* Complex consonant clusters are not permitted (ibid.).

# Lenition Rules

* /p/ and /k/ can become voiced intervocalically [@leza_gramatica_1999, as cited in @bierge_clause_2017, p. 42].
* /h/ is frequently deleted word-initially [@mcintosh_huichol_1945, p. 31].
* /r/ often alternates with [l] [@mcintosh_huichol_1945, p. 32].
* /w/ fricates to [β] preceding high-central and mid-front vowels [@mcintosh_huichol_1945, p. 32].
* /k/ is palatalized preceding /e/ [@mcintosh_huichol_1945, p. 31].
* /ts/ changes to /s/ preceding a consonant [@mcintosh_huichol_1945, p. 31].


# Misc. Rules

* /e/ is realized as [ɛ] preceding /k/ [@bierge_clause_2017, p. 38].
* /u/ and /o/ often alternate with [o] (ibid.).
* Although stress is indicated orthographically by a grave accent over the vowel [@bierge_clause_2017, p. 37], it doesn't appear to be marked in the Crubadan corpus, thus no rules will be written to account for such a diacritic.

# References
