---
title: "Burarra"
author: "Emily Strand"
bibliography: bvr.bib
output: html_document
---

Last updated: 2020-02-04

# Background

**Language family**: Australian / Gunwingguan / Burarran

* Burarra is spoken in the Northern Territory of Australia.

# Phonology

## Consonants

* Although @Glasgow1967 attest that the voiced stops are allophones of the corresponding voiced stops (pp. 3-4), @Green1987 states that they are contrastive word-medially (initial stops realized as voiced, final stops realized as voiceless) (p. 3). I have chosen to follow the more recent analysis of the language. Therefore, I do include the voicing distinction below.
* The alveolar and retroflex consonants are not contrastive word-initially, with the retroflex consonants being realized (though not orthographically represented) by default [@Green1987, p. 4]. Although the distinction is neutralized, the rules represent the retroflex consonants appropriately given that they are technically underlying in this position.


```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial   Alveolar  Retroflex   Palatal    Velar     
Stops                      "p b"     "t d"      "ʈ ɖ"      "c ɟ"     "k ɡ"   
Nasals                      "m"       "n"        "ɳ"        "ɲ"       "ŋ"   
Trills                      ""        "r"        "ɽ"        ""        ""    
Approximants                "w"       "l"        "ɭ"        "j"       ""    
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Retroflex", "Palatal", "Velar"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 5)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```

## Vowels

* Vowels do not occur in succession [@Glasgow1967, p. 8].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
           Front      Central       Back
High        "i"         ""           "u"
Mid         "e"         ""           "o"
Low         ""          "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)

```

# Alphabet

* Where the sequence ⟨ng⟩ refers to separate consonants as opposed to /ŋ/, a fullstop (.) appears between them [@Glasgow1981, p. 94]. These sequences, however, are extremely rare, and the convention does not appear in the Crúbadán corpus.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme       Phoneme
"a"             "/a/"
"b"             "/b/"
"d"             "/d/"
"e"             "/e/"
"g"             "/ɡ/"
"i"             "/i/"
"j"             "/ɟ/"
"k"             "/k/"
"l"             "/l/"
"m"             "/m/"
"n"             "/n/"
"o"             "/o/"
"p"             "/p/"
"r"             "/ɽ/"
"t"             "/t/"
"u"             "/u/"
"w"             "/w/"
"y"             "/j/"
**Digraph**      ""
"ch"            "/c/"
"ng"            "/ŋ/"
"ny"            "/ɲ/"
"rr"            "/r/"
"rt"            "/ʈ/"
"rd"            "/ɖ/"
"rn"            "/ɳ/"
"rl"            "/ɭ/"
"**Orthographic Exception**"   ""
"rnd"           "/ɳɖ/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* Unstressed vowels are deleted when occurring in homoorganic nasal-stop and stop-nasal sequences [@Glasgow1981, p. 84].
* /b/ may be realized as [w] within a prefix when preceding a stem-initial /b/ or /m/ [@Glasgow1981, p. 86].

# References
