---
title: "Nunggubuyu"
author: "Bill Mizgerd"
bibliography: nuy.bib
output: html_document
---

Last Updated: 2019-08-07

# Background

**Language Family:** Australian / Gunwinyguan / Nunggubuyu

* Nunggubuyu is spoken in the Northern Territory of Australia.

# Phonology

## Consonants

* /n̪/ appears only rarely, in a handful of animal-related nouns [@NuyHeath1984, p. 12].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)

consonants <- read.table(textConnection('
"Manner of Articulation"  Labial  Dental   Alveolar  Retroflex  Palatal   Velar
Stops                     "b"     "d̪"      "d"       "ɖ"        "ɟ"       "ɡ"
Nasals                    "m"     "n̪"      "n"       "ɳ"        "ɲ"       "ŋ"
Taps                      ""      ""       "ɾ"       ""         ""        ""
Approximants              "w"     ""       ""        "ɻ"        "j"       ""
"Lateral approximants"    ""      "l̪"      "l"       "ɭ"        ""        ""
'), TRUE)
kable(consonants, col.names = c("Manner of Articulation", "Labial", "Dental", "Alveolar", "Retroflex", "Palatal", "Velar"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* Vowel lengthening is present in Nunggubuyu, but only /aː/ has minimal pairs with its short counterpart [@NuyHeath1984, p. 9].
* Tone does not appear to be contrastive [@NuyHeath1984, p. 32].
* @NuyHughes1971 list /æ/ as part of the vowel inventory, albeit only marginally (p. 72). Other authors do not share this view, so I have opted not to include it below.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"       ""            "u"
Low       ""        "a aː"        ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: Where phonemes share a cell, the phoneme on the left is short and the one on the right is long.", general_title = "")
```    

# Alphabet

* **Note:** /n̪/ is not represented in the Crúbadán corpus.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme          Phoneme
"a"                "/a/"    
"b"                "/b/"  
"d"                "/d/"    
"ḏ"                "/ɖ/"  
"g"                "/ɡ/"  
"i"                "/i/"  
"j"                "/ɟ/"   
"l"                "/l/"    
"ḻ"                "/ɭ/"  
"m"                "/m/"  
"n"                "/n/"    
"ṉ"                "/ɳ/"   
"r"                "/ɻ/"  
"u"                "/u/"  
"w"                "/w/"
"y"                "/j/"  
**Digraph**         ""
"aa"               "/aː/"
"dh"               "/d̪/"
"lh"               "/l̪/"
"nh"               "/n̪/"
"ny"               "/ɲ/"  
"ng"               "/ŋ/"
"rr"               "/ɾ/"  
'), TRUE)

kable(alphabet, align = "c") %>%
  kable_styling("bordered")
```

# Syllable Structure

* Nunggubuyu has a (C)V(C)(C) syllable structure [@NuyHughes1971, pp. 75-76].

# Lenition Rules

* /ɟ/ undergoes an affricate release when non-syllable-final [@NuyHeath1984, p. 13].
* Word-final /ɡ/ is weakly realized when preceded by a liquid or vowel (ibid.).

# Misc. Rules

* /au/ sequences are sometimes realized as [oː] [@NuyHeath1984, p. 33].
* /ɾ/ trills to [r] in preconsonantal and word-final positions [@NuyHughes1971, p. 75].

# References
