---
title: "Guhu-Samane"
author: "Emily Strand"
bibliography: ghs.bib
output: html_document
---

Last updated: 2020-01-31

# Background

**Language family**: Trans-New Guinea / Main Section / Eastern / Binanderean / Guhu-Samane

* Guhu-Samane is spoken in the Oro and Morobe provinces of Papua New Guinea.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial   Dental   Alveolar   Retroflex  Velar   Glottal   
Stops                      "p  b"     "t̪"      "t"          ""     "k ɡ"     "ʔ"
Affricates                  ""        ""       "dz"         ""      "kx"      ""
Fricatives                  ""        ""       "s"          ""       ""       "h"
Nasals                      "m"       ""       "n"          ""       "ŋ"      ""
Flaps                       ""        ""       ""           "ɽ"      ""       ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Dental", "Alveolar", "Retroflex", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```

## Vowels

* Vowel length is phonemic [@Richert2002, p. 11]. Long vowels are represented by duplicate graphemes.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
           Front      Central       Back
High        "i"         ""           "u"
Mid         "e"         ""           "o"
Low         ""          "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)


diphthongs <- read.table(textConnection('
Diphthongs
"/ae/, /ai/, /ao/, /au/, /ei/, /eo/, /eu/, /iu/, /oe/, /oi/, /ou/, /ui/"
'), TRUE)

kable(diphthongs, align = 'c') %>%
  kable_styling("bordered")
```

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme       Phoneme
"a"             "/a/"
"b"             "/b/"
"e"             "/e/"
"g"             "/ɡ/"
"h"             "/h/"
"i"             "/i/"
"k"             "/k/"
"m"             "/m/"
"n"             "/n/"
"o"             "/o/"
"p"             "/p/"
"q"             "/ʔ/"
"r"             "/ɽ/"
"s"             "/s/"
"t"             "/t/"
"u"             "/u/"
**Digraph**      ""
"dz"            "/dz/"
"kh"            "/kx/"
"ng"            "/ŋ/"
"tt"            "/t̪/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* /b/ is realized as /β/ intervocalically [@Richert2002, p. 12].
* /t/ varies with /s/ frequently in the Garaina dialect (ibid.).

# Misc. Rules

* /ɽ/ is realized as [ɗ] following /i/ [@Richert2002, p. 13].
* /ɽ/ is realized as [ɭ] following /u/ (ibid.).

# References
