---
title: "Hawaiian"
author: "Justin Bai"
bibliography: haw.bib
output: html_document
---

Last updated: 2019-10-17

# Background

**Language family**: Austronesian / Malayo-Polynesian / Central-Eastern / Eastern Malayo-Polynesian / Oceanic / Central Pacific linkage / Tokelau-Fijian / Polynesian / Nuclear / East / Central / Marquesic

* Hawaiian is spoken throughout the Hawaiian Islands, located in the Pacific Ocean.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial   Alveolar   Velar   Glottal
Stops                       "p"        ""       "k"       "ʔ"
Fricatives                  "v"        ""       ""        "h"
Nasals                      "m"        "n"      ""        ""
Liquids                     ""         "l"      ""        ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 4)) %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* Vowel length is contrastive in Hawaiian [@elbert1986hawaiian, p. 14]. Long vowels are indicated with a macron ⟨¯⟩.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         ""           "u"
Mid       "e"         ""           "o"
Low        ""         "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```  

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme
"a"            "/a/"
"e"            "/e/"
"h"            "/h/"
"i"            "/i/"
"k"            "/k/"
"l"            "/l/"
"m"            "/m/"
"n"            "/n/"
"o"            "/o/"
"p"            "/p/"
"u"            "/u/"
"w"            "/v/"
"ʻ"            "/ʔ/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* /h/ deletes intervocalically [@Lavoie1996, p. 287].
* In a few instances /k/ may debuccalize to /ʔ/ [@elbert1986hawaiian, p. 12].

# Misc. Rules

* /k/ varies freely with /t/ [@jones2018hawaiian, pp. 105-106]. Because most sources don't indicate /t/ as being part of the phonemic inventory, I have chosen to transcribe ⟨k⟩ as /k/. Similarly, /p/ varies freely with [b], /w/ with [v], and /l/ with [d] and [r].

# References
