---
title: "Hawaiian"
author: "Justin Bai"
bibliography: haw.bib
output: html_document
---

Last updated: 2020-06-03

# Background

**Language family**: Austronesian / Malayo-Polynesian / Central-Eastern / Eastern Malayo-Polynesian / Oceanic / Central-Eastern Oceanic / Remote Oceanic / Central Pacific / East Fijian-Polynesian / Nuclear / East / Central / Marquesic

* Hawaiian is spoken throughout the Hawaiian Islands, located in the Pacific Ocean.
* Hawaiian contains many borrowings from English. Some loanwords are not fully adapted to fit Hawaiian's phonological system, retaining sounds that do not exist in native Hawaiian words [@jones2018hawaiian, p. 107].

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial   Alveolar   Velar   Glottal
Stops                       "p"        ""       "k"       "ʔ"
Fricatives                  "v"        ""       ""        "h"
Nasals                      "m"        "n"      ""        ""
Liquids                     ""         "l"      ""        ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 4)) %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* Vowel length is contrastive in Hawaiian [@elbert1986hawaiian, p. 14]. In the orthography, long vowels are indicated with a macron ⟨¯⟩.
* According to @jones2018hawaiian, Hawaiian may have diphthongs. This is uncertain because some speakers separate the vowels (p. 110). For this reason, diphthongs are not accounted for here.
    - The possible short-vowel diphthongs are: /ai/, /ae/, /au/, /ao/, /ei/, /eu/, /iu/, /oi/, and /ou/.
    - The possible long-vowel diphthongs are: /aːe/, /aːi/, /aːo/, /aːu/, /eːi/, and /oːu/.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         ""           "u"
Mid       "e"         ""           "o"
Low        ""         "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```  

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme
"a"            "/a/"
"e"            "/e/"
"h"            "/h/"
"i"            "/i/"
"k"            "/k/"
"l"            "/l/"
"m"            "/m/"
"n"            "/n/"
"o"            "/o/"
"p"            "/p/"
"u"            "/u/"
"w"            "/v/"
"ʻ"            "/ʔ/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* /h/ deletes intervocalically [@Lavoie1996, p. 287].
* In a few instances /k/ may debuccalize to /ʔ/ [@elbert1986hawaiian, p. 12].

# Misc. Rules

* /k/ varies freely with /t/ [@jones2018hawaiian, pp. 105-106]. Because most sources don't indicate /t/ as being part of the phonemic inventory, I have chosen to transcribe ⟨k⟩ as /k/. Similarly, /p/ varies freely with [b], /w/ with [v], and /l/ with [d] and [r].
* The glottal stop is predictably found before utterance-initial /a/, /e/, and /i/ [@elbert1986hawaiian, p. 10].

# References
