---
title: "Colorado"
author: "Emily Strand"
bibliography: cof.bib
output: html_document
---

Last Updated: 2020-05-04

**Compromised: orthographic ambiguity around glottal stops**

# Background

**Language Family**: Barbacoan / Cayapa-Colorado

* Also known as Tsafiki, Colorado is spoken in Ecuador.

# Phonology

## Consonants

* @Adelaar2007 as well as @Curnow1998 express the ambiguous phonemic status of /ʔ/ (p. 145; p. 387); however, they do not expand on the matter or state how it's ambiguous. @Moore1962 [as cited in @Curnow1998, p. 387] classifies it as a phoneme, but @Moore1966 doesn't even mention it. It is clear from @Dickinson2011 that glottal stops occur intervocalically (p. 311), which may be the point of confusion, but there isn't a clear consensus on whether its a phoneme in Colorado or not. Nevertheless, I have chosen to include it below
* There is also mention of aspiration being a phonemic suprasegmental feature, preceding consonants [@Moore1966, p. 96]. @Adelaar2007 note that because of this, the inventory might include other fricatives or preaspirated consonants; however, although this feature used to show contrast, it has become entirely predictable [@Moore1962 as cited in @Curnow1998, p. 387; @Moore1966, p. 96].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"     Labial   Alveolar  Palatal   Velar  Glottal
Stops                         "p b"    "t d"      ""       "k"     "ʔ"
Affricates                    ""       "ts"       ""       ""      ""
Fricatives                    "ɸ"      "s"        ""       ""      "h"
Nasals                        "m"      "n"        ""       ""      ""
Trills                        ""       "r"        ""       ""      ""
Approximants                  "w"      "l"        "j"      ""      ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 5)) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "") %>%
  column_spec(1, bold = TRUE)

```

# Vowels

* Although @Adelaar2007 see vowel nasalization as a phonetic feature, conditioned by syllable-final nasals (p. 143), most sources indicate that it is contrastive  [@Aikhenvald2007, p. 194; @Moore1966 as cited in @Curnow1998, p. 387; @Aikhenvald2011, p. 127].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
       Front   Central   Back
High   "i"     ""        "u"
Mid    "e"     ""        "o"
Low    ""      "a"       ""
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```


# Alphabet

* Contributing to the ambiguity centered around the phonemic status of /ʔ/, there is also uncertainty in terms of how its represented orthographically. @Dickinson2011 states that (non-intervocalic/supposedly phonemic) glottal stops are indicated by apostrophes (p. 311); however, @Curnow1998 state that apostrophes mark the once contrastive aspiration and glottal stops aren't marked (p. 387). In the rule set I have opted to transcribe apostrophes as glottal stops.   
* @Dickinson2011 states that nasal vowels are indicated by a following, syllable-final ⟨n⟩ (p. 311). This is made clear through the syllable structure, given that syllables consist of V or CV [@Walls, p. 15]. ⟨VnC⟩ sequences are transcribed as /ṼC/, and ⟨VnhV⟩ sequences are transcribed as /ṼV/; the ⟨h⟩ following the ⟨n⟩ in intervocalic positions merely serves as an indicator for the nasalization of the preceding vowel [@Moore1966, p. 96]. That is, ⟨VnV⟩ sequences (i.e. those without the ⟨h⟩) are transcribed as /VnV/.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme     Phoneme          Comment
"a"           "/a/"             ""
"b"           "/b/"             ""
"c"           "/k/"             ""
"d"           "/d/"             ""
"e"           "/e/"             ""
"f"           "/ɸ/"             ""
"i"           "/i/"             ""
"k"           "/k/"             ""
"j"           "/h/"             ""
"l"           "/l/"             ""
"m"           "/m/"             ""
"n"           "/n/"             ""
"o"           "/o/"             ""
"p"           "/p/"             ""
"r"           "/r/"             ""
"s"           "/s/"             ""
"t"           "/t/"             ""
"u"           "/u/"             ""
"w"           "/w/"             ""
"y"           "/j/"             ""
"\'"          "/ʔ/"             ""
**Digraph**    ""               ""
"ch"          "/ts/"       "phonetically [tʃ]: precedes /i/ or /u/"
"qu"          "/k/"       "precedes front vowels"
"sh"          "/s/"         "phonetically [ʃ]: precedes /i/ or /u/"
"ts"          "/ts/"            ""
"hu"          "/w/"             ""
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* /r/ is realized as a flap word-initially [@Moore1966, p. 96].
* /k/ is voiced intervocalically [@Walls, p. 11].

# Misc. Rules

* /w/ is realized as [ʋ] preceding /i/ [@Moore1966, p. 95].
* /b/ and /d/ are realized as implosives word-medially (ibid.).
* /ts/ and /s/ are realized as [tʃ] and [ʃ], respectively, preceding high vowels (ibid.).
    - /s/ is also realized as [ʃ] following /i/.
* Glottal stop epenthesis occurs intervocalically [@Dickinson2011, p. 311].

# References
