---
title: "Mapos Buang"
author: "Emily Strand"
bibliography: bzh.bib
output: html_document
---

Last Updated: 2020-05-06

**Slightly Compromised: conflation between /ɛ/ and other vowels**

# Background

**Language Family:** Austronesian / Malayo-Polynesian / Central-Eastern / Eastern Malayo-Polynesian / Oceanic / Western Oceanic / North New Guinea / Huon Gulf / South / Hote-Buang / Buang

* Mapos Buang, also referred to as Central Buang, is spoken in the Morobe province of Papua New Guinea.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"       Bilabial  Dental  Postalveolar Retroflex  Palatal  Velar  Labiovelar  Uvular
Stops                           "p ᵐb"   "t̪ ⁿd̪"      ""          ""        ""    "k ᵑɡ"   "kʷ ᵑɡʷ"   "q ᶰɢ"
Affricates                       ""        ""     "tʃ ⁿdʒ"       ""        ""      ""       ""         ""
Fricatives                       "β"       ""        ""          ""        ""      "ɣ"      ""         "ʁ"
Nasals                           "m"       "n̪"       ""          ""        "ɲ"     "ŋ"      "ŋʷ"       "ɴ"
Flaps                            ""        ""        ""          "ɽ"       ""      ""       ""         ""
Approximants                     ""        "l̪"       ""          ""        "j"     ""       "w"        ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Bilabial", "Dental", "Palatal", "Retroflex", "Palatal", "Velar", "Labiovelar", "Uvular"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 8))  %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```  

## Vowels

* Vowel length is contrastive in Mapos Buang [@Hooley2006, pp. 5-6]. Long vowels are indicated by duplicate graphemes.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
             Front   Central   Back
High         "i"      ""        "u"
High-Mid     "e"      ""        "o"
Low-Mid      "ɛ"      ""        "ɔ"
Low          ""       "a"       ""
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

* Schwa is represented orthographically by ⟨e⟩, however it is not contrastive with other vowels [@Hooley2006, p. 6]; ⟨e⟩ also represents /ɛ/, and has been used as the default transcription in the rule set. Because of this, the rule set is somewhat skewed in its representation of /ɛ/, conflating it with other vowels. Thus, the language is compromised to some degree.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme
"a"              "/a/"
"b"              "/ᵐb/"     
"d"              "/ⁿd̪/"
"e"              "/ɛ/"
"ë"              "/e/"
"g"              "/ᵑɡ/"
"ġ"              "/ᶰɢ/"
"h"              "/ʁ/"
"i"              "/i/"
"j"              "/ⁿdʒ/"
"k"              "/k/"
"ḳ"              "/q/"
"l"              "/l̪/"
"m"              "/m/"
"n"              "/n̪/"
"o"              "/ɔ/"
"ö"              "/o/"
"p"              "/p/"
"r"              "/ɽ/"
"s"              "/tʃ/"
"t"              "/t̪/"
"u"              "/u/"
"v"              "/β/"
"w"              "/w/"
"y"              "/j/"
**Multigraph**    ""
"kw"             "/kʷ/"
"gg"             "/ɣ/"
"gw"             "/ᵑɡʷ/"
"ng"             "/ŋ/"
"nġ"             "/ɴ/"
"ny"             "/ɲ/"
"ngw"            "/ŋʷ/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* Vowels are realized as [ə] in unstressed syllables [@Hooley2006, p. 6].

# Misc. Rules

* /ɽ/ may be realized by some speakers as [ɭ] [@Hooley2006, p. 4].
* /l̪/ and /j/ may be realized with some frication [@Hooley2006, pp. 4-5].

# References
