---
title: "Mapos Buang"
author: "Emily Strand"
bibliography: bzh.bib
output: html_document
---

Last Updated: 2020-05-06

**SLIGHTLY COMPROMISED: possible misrepresentation of /ɛ/**

# Background

**Language Family:** Austronesian / Malayo-Polynesian / Central-Eastern / Eastern Malayo-Polynesian / Oceanic / Western Oceanic / North New Guinea / Huon Gulf / South / Hote-Buang / Buang

* Mapos Buang, also referred to as Central Buang, is spoken in the Morobe province of Papua New Guinea.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"       Bilabial  Dental  Postalveolar Retroflex  Palatal  Velar  Labiovelar  Uvular
Stops                           "p ᵐb"   "t̪ ⁿd̪"      ""          ""        ""    "k ᵑɡ"   "kʷ ᵑɡʷ"   "q ᶰɢ"
Affricates                       ""        ""     "tʃ ⁿdʒ"       ""        ""      ""       ""         ""
Fricatives                       "β"       ""        ""          ""        ""      "ɣ"      ""         "ʁ"
Nasals                           "m"       "n̪"       ""          ""        "ɲ"     "ŋ"      "ŋʷ"       "ɴ"
Flaps                            ""        ""        ""          "ɽ"       ""      ""       ""         ""
Approximants                     ""        "l̪"       ""          ""        "j"     ""       "w"        ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Bilabial", "Dental", "Palatal", "Retroflex", "Palatal", "Velar", "Labiovelar", "Uvular"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 8))  %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```  

## Vowels

* Vowel length is contrastive in Mapos Buang [@Hooley2006, pp. 5-6]. Long vowels are indicated by duplicate graphemes.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
             Front   Central   Back
High         "i"      ""        "u"
High-Mid     "e"      ""        "o"
Low-Mid      "ɛ"      ""        "ɔ"
Low          ""       "a"       ""
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

* In unstressed syllables, ⟨e⟩ represents [ə]; it is not contrastive with other vowels, generally manifesting in consonant sequences [@Hooley2006, p. 6]. ⟨e⟩ also represents /ɛ/ and, for the purposes of this project, shouldn't necessarily compromise the language. However, it seems that, based on the description provided in @Hooley2006, ⟨e⟩ represents a surface realization that may in some instances stem from nothing underlyingly. In the rules, I have transcribed ⟨e⟩ to /ɛ/ by default in all instances, but this may skew the representation of /ɛ/, such that it appears more frequently than it actually should.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme
"a"              "/a/"
"b"              "/ᵐb/"     
"d"              "/ⁿd̪/"
"e"              "/ɛ/"
"ë"              "/e/"
"g"              "/ᵑɡ/"
"ġ"              "/ᶰɢ/"
"h"              "/ʁ/"
"i"              "/i/"
"j"              "/ⁿdʒ/"
"k"              "/k/"
"ḳ"              "/q/"
"l"              "/l̪/"
"m"              "/m/"
"n"              "/n̪/"
"ö"              "/ɔ/"
"o"              "/o/"
"p"              "/p/"
"r"              "/ɽ/"
"s"              "/tʃ/"
"t"              "/t̪/"
"u"              "/u/"
"v"              "/β/"
"w"              "/w/"
"y"              "/j/"
**Multigraph**    ""
"kw"             "/kʷ/"
"gg"             "/ɣ/"
"gw"             "/ᵑɡʷ/"
"ng"             "/ŋ/"
"nġ"             "/ɴ/"
"ny"             "/ɲ/"
"ngw"            "/ŋʷ/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Misc. Rules

* /ɽ/ may be realized by some speakers as [ɭ] [@Hooley2006, p. 4].
* /l̪/ and /j/ may be realized with some frication [@Hooley2006, pp. 4-5].

# References
