---
title: "Zaza"
author: "Bill Mizgerd"
bibliography: zza.bib
output: html_document
---

Last Updated: 2020-07-06

**COMPROMISED: conflicting orthographies make for possible conflation of vowels**

# Background

**Language Family:** Indo-European / Indo-Iranian / Iranian / Western / Northwestern / Zaza-Gorani

* The language is also sometimes known as Zazaki or Dimli and it is spoken throughout Turkey.

# Phonology

## Consonants

* /sˤ/, /t̪ˤ/, and /ʕ/ are marginal [@ZzaTodd2002, p. 1].
    - /t̪ˤ/ and /ʕ/ only occur in Arabic loanwords, but they are considered to be part of the phonemic inventory.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial     Dental     Alveolar    Postalveolar   Palatal     Velar   Uvular  Pharyngeal    Glottal
Stops                      "p  b"   "t̪  t̪ˤ  d̪"     ""            ""           ""       "k  ɡ"    "q"       ""            ""
Affricates                   ""         ""         ""         "tʃ  dʒ"        ""         ""      ""        ""            ""
Fricatives                 "f  v"       ""      "s  sˤ  z"     "ʃ  ʒ"         ""       "x  ɣ"    ""       "ħ ʕ"          "h"
Nasals                      "m"         ""         "n"           ""           ""         ""      ""        ""            ""
Trills                      ""          ""         "r"           ""           ""         ""      ""        ""            ""
Flaps                       ""          ""         "ɾ"           ""           ""         ""      ""        ""            ""
Approximants                "w"         ""       "l  ɫ"          ""           "j"        ""      ""        ""            ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Dental", "Alveolar", "Postalveolar", "Palatal", "Velar", "Uvular", "Pharyngeal", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 9)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For cells that have two phonemes, those on the left are voiceless and those on the right are voiced. Phonemes that have the diacritic (ˤ) are pharyngealized. Both alveolar approximants are voiced, but the one on the right is velarized.", general_title = "")
```

## Vowels

* @ZzaTodd2002 includes /ə/ rather than /ɜ/ (p. 21); however, I opted to use /ɜ/ given that they describe it as a "mid open central vowel," which is lower than schwa.
* Zaza has a number of diphthongs, all of which are spelled as a sequence of the two sounds that comprise them (e.g. /əj/ is written as ⟨ey⟩ and so forth) [@ZzaTodd2002, p. 16].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
              Front      Central       Back
High          "i"         "ɨ"          "u"
"Near-High"   ""          ""           "ʊ"
Mid           "e"         ""           "o"
Low-Mid       ""          "ɜ"          ""
Low           ""          ""           "ɑ"
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)


diphthongs <- read.table(textConnection('
Diphthongs
"/oj/, /ʊj/, /ɑj/, /ɑw/, /ɜj/, /ɜw/"
'), TRUE)

kable(diphthongs, align = 'c') %>%
  kable_styling("bordered")
```  

# Alphabet

* This orthography is somewhat compromised because the transcription of ⟨i⟩ differs depending on the orthography. By default, my ruleset only supports the orthography that maps ⟨i⟩ to /i/; however, rules for the other orthography are included but commented out in the ruleset, so by uncommenting those lines, the other orthography can be represented.
* **NOTE:** /ħ/ is not represented in the Crúbadán corpus.


```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme          Comment
"a"            "/ɑ/"             ""
"b"            "/b/"             ""
"c"            "/dʒ/"            ""
"ç"            "/tʃ/"            ""
"d"            "/d̪/"             ""
"e; é"         "/ə/"             ""
"ê"            "/e/"             ""
"f"            "/f/"             ""
"g"            "/ɡ/"             ""
"ğ"            "/ɣ/"             ""
"h"            "/h/"             ""
"\'h"           "/ħ/"             ""
"ı"            "/ɨ/"             ""
"i"            "/i/"      "in orthographies with ⟨ı⟩"  
"i"            "/ɨ/"      "in orthographies with ⟨î⟩"  
"î"            "/i/"       "in some orthographies"  
"j"            "/ʒ/"             ""
"k"            "/k/"             ""
"l"            "/l/"             ""
"\'l"          "/ɫ/"       "in some orthographies"   
"m"            "/m/"             ""
"n"            "/n/"             ""
"o"            "/o/"             ""
"p"            "/p/"             ""
"q"            "/q/"             ""
"r"            "/ɾ/"             ""
"s"            "/s/"             ""
"\'s"          "/sˤ/"        "in some orthographies"    
"ş"	           "/ʃ/"             ""
"t"            "/t̪/"             ""
"\'t"          "/t̪ˤ/"        "in some orthographies"    
"u"            "/ʊ/"             ""
"û"            "/u/"         "in some orthographies"  
"ü"            "/u/"         "in some orthographies"  
"v"            "/v/"             ""
"w"            "/w/"             ""
"x"            "/x/"             ""
"y"            "/j/"             ""
"z"            "/z/"             ""
"\'"           "/ʕ/"         "in loanwords"
**Digraph**     ""               ""
"lh"           "/ɫ/"         "in some orthographies"
"rr"           "/r/"             ""
"sh"           "/sˤ/"        "in some orthographies"
"th"           "/t̪ˤ/"        "in some orthographies"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Zaza has a (C)(C)V(C)(C) syllable structure.
    - Note, however, that - with the exception of the word <psing> - Zaza does not allow syllables to contain both an onset cluster and a coda cluster [@ZzaTodd2002, pp. 24-25].

# Lenition Rules

* /b/ spirantizes to [v] in postvocalic position, as well as in a handful of loanwords [@ZzaTodd2002, pp. 10-11].
* /dʒ/ spirantizes to [ʒ] postvocalically [@ZzaTodd2002, p. 11].
* Word-final /n/, if preceded by /ɑ/ or /ə/, is realized only as a nasalization of the vowel. This pattern also holds if there's an intervening /j/ (ibid.).
* Word-final /n/, if preceded by a vowel *not* mentioned in the previous rule, deletes entirely [@ZzaTodd2002, p. 13].

# Misc. Rules

* Word-final /ng/ collapses to [ŋ] [@ZzaTodd2002, p. 12].
* Vowels become marginally lower and farther back when following a pharyngeal or pharyngealized consonant, or when following /ɫ/ [@ZzaTodd2002, p. 17].
* Postvocalic /i/ is realized as [j] (ibid.).
* Epenthesis of [w] occurs between /u/ and front vowels (ibid.).
* The sequence /ii/ shortens to [i] [@ZzaTodd2002, p. 18].

# References
