---
title: "Bargam"
author: "Emily Strand"
bibliography: mlp.bib
output: html_document
---

Last updated: 2020-02-20

**Slightly Compromised: conflation between /n/ and /ŋ/**

# Background

**Language family**: Trans-New Guinea / Madang-Adelbert Range / Adelbert Range / Pihom-Isumrud-Mugil / Mugil

# Phonology

## Consonants

* /ŋ/ is marginal [@Hepner1989, p. 18].
* /p/ and /f/ are also rare (not to the extent of /ŋ/) [@Hepner1989, p. 25].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial  Alveolar  Palatal  Velar   Glottal   
Stops                       "p b"   "t d"      ""     "k ɡ"     "ʔ"
Fricatives                  "f"     "s z"      ""      ""       "h"
Nasals                      "m"      "n"       ""      "ŋ"      ""
Flaps                       ""       "ɾ"       ""      ""       ""
Approximants                "w"      "l"       "j"     ""       ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 5)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```

## Vowels

* Glides /w/ and /j/ preceding or following vowels are always interpreted as consonants and never part of complex nuclei with adjacent vowels [@Hepner1992, p. 3].
    - Consecutive vowels constitute separate syllables [@Hepner1989, p. 24].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
           Front      Central       Back
High        "i"         ""           "u"
Mid         "e"         ""           "o"
Low         ""          "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)

```

# Alphabet

* The orthographic contrast for /n/ versus /ŋ/ is somewhat convoluted. @Hepner1992 represents /ŋ/ with ⟨ng⟩ (p. 1). @Hepner1989, however, state that /n/ preceding velar consonants is phonetically realized as [ŋ]. While @Hepner1989 identify /ŋ/ as a distinctive phoneme, it appears that the sequence ⟨ng⟩ may represent both the underlying form and surface realization of /ŋ/, compromising the language. Given that /ŋ/ is rare underlyingly in this language, I have chosen to transcribe ⟨ng⟩ sequences as /nɡ/.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme       Phoneme
"a"             "/a/"      
"b"             "/b/"
"d"             "/d/"
"e"             "/e/"
"f"             "/f/"
"g"             "/ɡ/"  
"h"             "/h/"
"i"             "/i/"
"k"             "/k/"
"l"             "/l/"
"m"             "/m/"
"n"             "/n/"
"o"             "/o/"
"p"             "/p/"
"q"             "/ʔ/"
"r"             "/ɾ/"
"s"             "/s/"
"t"             "/t/"
"u"             "/u/"
"w"             "/w/"
"y"             "/j/"
"z"             "/z/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure  

* Syllables in Bargam follow the order of (C)(C)V(C)(C) [@Hepner1989, p. 31].

# Misc. Rules

* /ɾ/ may vary freely with /r/ word-initially and word-medially [@Hepner1989, p. 19].
* Phonetic glottal stops occur preceding word-initial vowels [@Hepner1989, p. 8].
* Consonant clusters generally are realized with an intervening epenthetic vowel [@Hepner1989, p. 15].

# References
