---
title: "Djambarrpuyngu"
author: "Bill Mizgerd"
bibliography: djr.bib
output: html_document
---

Last Updated: 2019-07-03

# Background

**Language Family**: Australian / Pama-Nyungan / Yuulngu / Dhuwal

* Djambarrpuyngu is spoken in the Northern Territory of Australia.

# Phonology

## Consonants

* Voicing distinctions are only contrastive intervocalically or between sonorants [@DjrWilkinson1991, p. 42].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"    Labial  Dental   Alveolar  Retroflex  Palatal   Velar  Glottal
Stops                        "p b"   "t̪ d̪"    "t d"     "ʈ ɖ"      "c ɟ"     "k ɡ"  "ʔ"
Nasals                        "m"     "n̪"      "n"       "ɳ"        "ɲ"       "ŋ"    ""
Trills                        ""      ""       "r"       ""         ""        ""     ""
Approximants                  "w"     ""       "l"       "ɻ  ɭ"     "j"       ""     ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Dental", "Alveolar", "Retroflex", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 7)) %>%
  footnote(general = "Note: Phonemes to the right of a cell are either voiced (in the stop row) or lateral (in the approximant row).", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* Vowel length is contrastive in Djambarrpuyngu [@DjrWilkinson1991, p. 44].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
       Front   Central   Back
High   "i"     ""        "u"
Low    ""      "a"       ""
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme
"a"            "/a/"  
"ä"            "/aː/"  
"b"            "/b/"  
"d"            "/d/"   
"ḏ"            "/ɖ/"  
"e"            "/iː/"  
"g"            "/ɡ/"  
"i"            "/i/"  
"k"            "/k/"  
"l"            "/l/"  
"ḻ"            "/ɭ/"  
"m"            "/m/"  
"n"            "/n/"  
"ṉ"            "/ɳ/"
"ŋ"            "/ŋ/"  
"o"            "/uː/"  
"p"            "/p/"  
"r"            "/ɻ/"    
"t"            "/t/"   
"ṯ"            "/ʈ/"  
"u"            "/u/"  
"w"            "/w/"  
"y"            "/j/"  
"\'"           "/ʔ/"
**Digraph**    ""
"dh"           "/d̪/"  
"dj"           "/ɟ/"
"nh"           "/n̪/"  
"ny"           "/ɲ/"
"rr"           "/r/"  
"th"           "/t̪/"  
"tj"           "/c/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Djambarrpuyngu has a CV(C)(C)(ʔ) syllable structure [@DjrWilkinson1991, p. 45].

# Lenition Rules

* Intramorphemic labial and velar stops often approximantize to [w], while intramorphemic dental and palatal stops often approximantize to [j] [@DjrWilkinson1991, p. 42].
* /ɖ/ usually becomes the flap [ɾ] in the sequences /jɖ/ and /rɖ/ (ibid.).

# References
