---
title: "Apurinã"
author: "Emily Strand"
bibliography: apu.bib
output: html_document
---

Last Updated: 2019-09-11

# Background

**Language Family:** Arawakan / Maipuran / Southern Maipuran / Purus

* Apurinã is spoken throughout the Amazon Basin in Central Brazil.

# Phonology

## Consonants

* /ɲ/ and /h/ are sometimes omitted from the consonantal inventory depending on the dialect; I have chosen to include them below @facundes_language_2000 does (pp. 55, 71).
* Voicing is not a contrastive feature in Apurinã [@facundes_language_2000, p. 70].


```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"    Bilabial   Alveolar    Postalveolar   Palatal   Velar    Glottal
Stops                         "p"        "t"           ""            ""      "k"        ""
Affricates                    ""         "ts"         "tʃ"           ""      ""         ""
Fricatives                    ""         "s"          "ʃ"            ""      ""         "h"
Nasals                        "m"        "n"           ""            "ɲ"     ""         ""
Flaps                         ""         "ɾ"           ""            ""      ""         ""
Approximants                  ""         ""            ""            "j"     "ɰ"        ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Bilabial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* Vowels are contrastive in terms of length and nasality [@facundes_language_2000, p. 55].
    - Long vowels (both oral and nasal) are indicated by repeated graphemes.
* In addition to oral diphthongs, Apurinã includes nasal diphthongs, although these are extremely infrequent (I have not accounted for the nasal diphthongs in the rules) [@facundes_language_2000, pp. 85-87].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
      Front    Central    Back
High   "i ĩ"   "ɨ ɨ̃"       ""
Mid    "e ẽ"    ""       "o õ"
Low     ""     "a ã"       ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: Phonemes on the left of each cell are oral and phonemes on the right are nasal.", general_title = "")


diphthongs <- read.table(textConnection('
        Diphthongs   
"/ai/,  /ei/,  /oi/,  /ao/,  /io/"
'), TRUE)

kable(diphthongs, align = 'c') %>%
  kable_styling("bordered")
```

# Alphabet

* **Note:** /ɨ̃/, /ɲ/, /j/, and /ɰ/ are not represented in the Crúbadán corpus.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme
"a"            "/a/"
"ã"            "/ã/"
"e"            "/e/"
"ẽ"            "/ẽ/"
"h"            "/h/"
"i"            "/i/"
"ĩ"            "/ĩ/"
"k"            "/k/"
"m"            "/m/"
"n"            "/n/"
"o"            "/o/"
"õ"            "/õ/"
"p"            "/p/"
"r"            "/ɾ/"
"s"            "/s/"
"t"            "/t/"
"u"            "/ɨ/"
"ũ"            "/ɨ̃/"
"w"            "/ɰ/"
"x"            "/ʃ/"
"y"            "/j/"
**Digraph**    ""
"nh"           "/ɲ/"
"ts"           "/ts/"
"tx"           "/tʃ/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* The possible syllable structures include [@facundes_language_2000, pp. 87-88]:
    - CV  
    - CVV  
    - VV  
    - V  
* Based on this, it would seem that Apurinã does not allow for complex onsets and consonant clusters. They also seem to require empty codas.

# Lenition Rules

* Voiceless stops can become voiced following nasal vowels [@facundes_language_2000, p. 73].
* /h/ deletes when followed by a vowel [@facundes_language_2000, p. 83].

# Misc. Rules

* Vowels are often devoiced word-finally in unstressed positions [@facundes_language_2000, p. 60].
* There is free variation of oral and nasal vowels preceding nasal consonants (the contrast is neutralized) [@facundes_language_2000, p. 64].
* Vowels preceding nasal vowels become nasal [@facundes_language_2000, p. 66].
* /k/ palatalizes when preceding [ɛ], which is an allophone of /e/ [@facundes_language_2000, p. 78].
* Aspiration of stops occurs preceding devoiced vowels in unstressed positions [@facundes_language_2000, pp. 78-79].
* /n/ palatalizes preceding /i/ [@facundes_language_2000, p. 82].
* Apostrophes indicate vowel omission [@104611, p. 3].

# References
