---
title: "Huehuetla Tepehua"
author: "Abi Creighton"
bibliography: tee.bib
output: html_document
---

Last Updated: 2020-08-13

# Background

**Language Family:** Totonacan / Tepehua / Huehuetla Tepehua

* Huehuetla Tepehua is a language spoken in Huehuetla, a municipality in Hildago, Mexico.

# Phonology

## Consonants

* For younger speakers, the uvular and glottal stops have merged [@kung_2007, p. 31]. Because older speakers still contrast /q/ with /ʔ/, I have chosen to include both below.
* Although /r/ and /ɾ/ are said to occur only in Spanish loanwords and ideophones, they are understood to be part of Huehuetla Tepehua's phonemic inventory [@kung_2002, p. 167]. I have chosen to include both below.


```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Bilabial   Alveolar  Postalveolar  Palatal  Velar    Uvular  Glottal
Stops                      "p pʼ"      "t tʼ"        ""         ""     "k kʼ"    "q"      "ʔ"
Affricates                  ""        "ts tsʼ"    "tʃ tʃʼ"      ""      ""       ""       ""
Fricatives                  ""         "s ɬ"        "ʃ"         ""      ""       ""       "h"
Nasals                      "m"         "n"          ""         ""      ""       ""       ""
Liquids                     ""          "l"          ""         ""      ""       ""       ""
Trills                      ""          "r"          ""         ""      ""       ""       ""
Flaps                       ""          "ɾ"          ""         ""      ""       ""       ""
Approximants                "w"          ""          ""         "j"     ""       ""       ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Bilabial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Uvular", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 7)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are plain and those on the right are glottalized. Both alveolar fricatives are plain, but the one on the right is lateral.", general_title = "")
```  

## Vowels

* Vowel length is phonemic [@kung_2007, pp. 31, 64]. Long vowels are indicated with duplicate graphemes.
* Like other modern Totonacan languages, Huehuetla Tepehua evolved from a three vowel system to contrast five vowels, the mid vowels of which gained phonemic status through Spanish influence and ideophones [@kung_2007, p. 50].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
             Front      Central       Back
High          "i"         ""           "u"
Mid           "e"         ""           "o"
Low           ""          "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)

```     

# Alphabet

* @garcia_agustin_2012 appears to include a glottalized /q/ in the orthography (p. 13). As @kung_2002 notes, however, the glottalized uvular stop has merged with its plain counterpart (p. 167).
* @garcia_agustin_2012 represents /w/ with ⟨v⟩ and /ts/ with ⟨ts⟩ (p. 13), while @kung_2007 uses ⟨w⟩ and ⟨tz⟩, respectively (pp. 33-34). Because there is no overlap with the different transcriptions that would conflate the different phonemes, I have included the orthographic variations from both sources below.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme  
"a"              "/a/"              
"c"              "/k/"  
"c\'"            "/kʼ/"
"e"              "/e/"              
"i"              "/i/"
"j"              "/h/"
"k"              "/k/"  
"l"              "/l/"
"m"              "/m/"              
"n"              "/n/"
"o"              "/o/"
"p"              "/p/"
"p\'"            "/pʼ/"
"r"              "/ɾ/"              
"s"              "/s/"              
"t"              "/t/"
"t\'"            "/tʼ/"
"u"              "/u/"              
"v"              "/w/"  
"w"              "/w/"
"x"              "/ʃ/"
"y"              "/j/"
"\'"             "/ʔ/"
**Digraph**      ""                   
"ch"             "/tʃ/"
"ch\'"           "/tʃʼ/"
"qu"             "/q/"
"q\'u"           "/q/"
"rr"             "/r/"
"lh"             "/ɬ/"
"ts"             "/ts/"
"ts\'"           "/tsʼ/"
"tz"             "/ts/"
"tz\'"           "/tsʼ/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Syllables in HT follow the order (C)CV(V)(C)(C) [@kung_2007, p. 81].

# Lenition Rules

* /p/, /t/, and /k/ are optionally voiced between voiced phonemes [@kung_2007, p. 38].
* For younger speakers, /q/ debuccalizes to /ʔ/ [@kung_2007, p. 67].
* /k/ spirantizes to [h] when preceded by a vowel and followed by a velar or uvular stop [@kung_2007, p. 137].
* When two identical consonants become adjacent because of morpheme concatenation, the first consonant is deleted [@kung_2007, p. 139].
* Stem final-glottal stops are deleted when followed by a suffix or enclitic [@kung_2007, p. 140].
* Intervocalic /h/ is deleted when followed by an unstressed vowel [@kung_2007, p. 143].
* When /n/ is part of the prefix kin- or min-, it deletes before an approximant [@kung_2007, p. 163].

# Misc. Rules

* When short vowels occur utterance-finally, they are devoiced or deleted [@kung_2007, pp. 124-126].
* /l/ neutralizes to [ɬ] in the coda position [@kung_2007, p. 126].
* A glottal stop is inserted preceding syllable-initial vowels [@kung_2007, p. 129].
* Nasals assimilate to the place of a following consonant, and velars assimilate to the place of a following uvular [@kung_2007, p. 132].
* When the sequence /kh/ occurs after /a/, metathesis occurs and the sequence becomes /hk/ [@kung_2007, p. 135].

# References
