---
title: "Yucatec Maya"
author: "Justin Bai"
bibliography: yua.bib
output: html_document
---

Last Updated: 05-12-2020

# Background

**Language family**: Mayan / Yucatecan / Yucatec-Lacandon

* Yucatec Maya is spoken in the Quintana Roo, Yucatán, Campeche, and Tabasco states of Mexico.

# Phonology

## Consonants

* /b/, /d/, /ɡ/, /f/, and /r/ occur in Spanish loanwords [@bohnemeyer2002grammar, p. 65].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial       Alveolar     Postalveolar   Palatal   Velar    Glottal   
Stops                     "p pʼ ɓ"      "t  tʼ"         ""            ""     "k  kʼ"    "ʔ"
Affricates                   ""         "ts  tsʼ"     "tʃ  tʃʼ"       ""       ""       ""
Fricatives                   ""           "s"            "ʃ"          ""       ""       "h"
Nasals                       "m"          "n"            ""           ""       ""       ""
Flaps                        ""           "ɾ"            ""           ""       ""       ""
Approximants                 "w"          "l"            ""           "j"      ""       ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are plain and those on the right are ejctive. The labial stops are ordered plain, ejective, and implosive.", general_title = "")
```

## Vowels

* Vowels can occur with four contrastive sets of suprasegmental features [@frazier2009production, pp. 18-19].
    - As summarized by @frazier2009production, these are the four vowel shapes in terms of length, pitch, and glottalization (p. 19):
        - SHORT: short, unmarked for tone, modal voice
        - LOW TONE: long, low tone, modal voice
        - HIGH TONE: long, high tone, modal voice
        - GLOTTALIZED: long, high tone followed by creaky voice
            - @frazier2009production also states that the glottalized vowels are produced with an intervening glottal stop (i.e. [v́ʔv]); however, in fast speech, this glottal stop is not always realized (p. 20)
* Per discussions with Scott AnderBois, an expert on the language, the rules will represent "glottalized" vowels as a long, high tone vowel followed by creaky voice rather than a vowel interrupted by a glottal stop.


```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         ""           "u"
Mid       "e"         ""           "o"
Low        ""         "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)


  vowel_shapes <- read.table(textConnection('
  "Vowel Shapes"  "Orthographic Representation"     "Our Transcription"
  Short                      "V"                            "V"
  "Low Tone"                "VV"                          "V ː ˨"
  "High Tone"               "V́V"                          "V ː ˦"
  "Glottalized"            "V\'V"                         "V ː ˦ ˷"

  '), TRUE)

  kable(vowel_shapes, col.names = c("Vowel Shapes", "Orthographic Representation", "Our Transcription"), align = 'c') %>%
    kable_styling("bordered")%>%
    column_spec(1, bold = TRUE)

```

# Alphabet

* **Note:** /tsʼ/ does not appear in the Crúbadán corpus.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme
"a"            "/a/"
"b"            "/ɓ/"
"e"            "/e/"
"i"            "/i/"
"j"            "/h/"
"k"            "/k/"
"k\'"          "/kʼ/"
"l"            "/l/"
"m"            "/m/"
"n"            "/n/"
"o"            "/o/"
"p"            "/p/"
"p\'"          "/pʼ/"
"r"            "/ɾ/"  
"s"            "/s/"  
"t"            "/t/"  
"t\'"          "/tʼ/"
"u"            "/u/"  
"x"            "/ʃ/"  
"w"            "/w/"  
"y"            "/j/"  
"\'"           "/ʔ/"
**Digraph**     ""  
"ch"           "/tʃ/"
"ch\'"         "/tʃʼ/"
"tz"           "/ts/"  
"tz\'"         "/tsʼ/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* /k/ debuccalizes to /h/ preceding another consonant [@rogers2016, p. 81].

# References
