---
title: "Yucatec Maya"
author: "Justin Bai"
bibliography: yua.bib
output: html_document
---
# Background

**Language family**: Mayan / Yucatecan / Yucatec-Lacandon

* Yucatec Maya is spoken in the Quintana Roo, Yucatán, Campeche, and Tabasco states of Mexico.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial       Alveolar     Postalveolar   Palatal   Velar    Glottal   
Stops                     "pʰ pʼ ɓ"     "tʰ  tʼ"         ""            ""    "kʰ  kʼ"    "ʔ"
Affricates                   ""        "tsʰ  tsʼ"     "tʃʰ  tʃʼ"       ""       ""       ""
Fricatives                  ""            "s"             "ʃ"          ""       ""       "h"
Nasals                      "m"           "n"             ""           ""       ""       ""
Flaps                       ""            "ɾ"             ""           ""       ""       ""
Approximants                "w"           "l"             ""          "j"       ""       ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: Consonants with the diacritics (ʰ) and (ʼ) are aspirates and ejectives, respectively. The labial stop on the right is an implosive.", general_title = "")
```

## Vowels

* Although there are only five vowel qualities, for every vowel, "each vowel quality can occur with four contrastive sets of suprasegmental features, resulting in 20 unique possible syllable heads" [@frazier2009production, pp. 18-19].
* As summarized by @frazier2009production, these are the four vowel shapes in terms of length, pitch, and glottalization (p. 19):
    - SHORT: short, unmarked for tone, modal voice
    - LOW TONE: long, low tone, modal voice
    - HIGH TONE: long, high tone, modal voice
    - GLOTTALIZED: long, high tone followed by creaky voice
* @frazier2009production also states, "The GLOTTALIZED vowels are unanimously described as being produced with initial high pitch and modal voice that is interrupted by a glottal stop (i.e. [v́ʔv]), though B&VS and B et al. recognized that this glottal stop does not always have a full realization, especially in rapid speech" (p. 20).
* Per discussions with Scott AnderBois, an expert on the language, the rules will represent "glottalized" vowels as a long, high tone vowel followed by creaky voice rather than a vowel interrupted by a glottal stop.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         ""           "u"
Mid       "e"         ""           "o"
Low        ""         "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)


  vowel_shapes <- read.table(textConnection('
  "Vowel Shapes"  "Orthographic Representation"     "Our Transcription"
  Short                      "V"                            "V"
  "Low Tone"                "VV"                          "V ː ˨"
  "High Tone"               "V́V"                          "V ː ˦"
  "Glottalized"            "V\'V"                         "V ː ˦ ˷"

  '), TRUE)

  kable(vowel_shapes, col.names = c("Vowel Shapes", "Orthographic Representation", "Our Transcription"), align = 'c') %>%
    kable_styling("bordered")%>%
    column_spec(1, bold = TRUE)

```

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme
"a"            "/a/"
"b"            "/ɓ/"
"e"            "/e/"
"i"            "/i/"
"j"            "/h/"
"k"            "/kʰ/"
"k\'"          "/kʼ/"
"l"            "/l/"
"m"            "/m/"
"n"            "/n/"
"o"            "/o/"
"p"            "/pʰ/"
"p\'"          "/pʼ/"
"r"            "/ɾ/"  
"s"            "/s/"  
"t"            "/tʰ/"  
"t\'"          "/tʼ/"
"u"            "/u/"  
"x"            "/ʃ/"  
"w"            "/w/"  
"y"            "/j/"  
"\'"           "/ʔ/"
**Digraph**     ""  
"ch"           "/tʃʰ/"
"ch\'"         "/tʃʼ/"
"tz"           "/tsʰ/"  
"tz\'"         "/tsʼ/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* /k/ debuccalizes to /h/ preceding another consonant [@rogers2016, p. 81].

# References
