---
title: "Benabena"
author: "Bill Mizgerd"
bibliography: bef.bib
output: html_document
---

Last Updated: 2019-09-30

# Background

**Language Family**: Trans-New Guinea / Main Section / Central and Western / East New Guinea Highlands / East-Central / Gahuku-Benabena

* Some sources hyphenate the language as Bena-Bena.
* Benabena is spoken throughout the Goroka district of the Eastern Highlands province of Papua New Guinea.  

# Phonology

## Consonants

* @BefYoung1961 and @BefYoung1975 argue for [β] rather than /ɸ/ (p. 4; p. 1). However, I have preferred the analysis of @BefYoung1992, which seems to show the most recent form of Young's understanding of the phoneme.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"     Bilabial  Labiodental Alveolar  Retroflex  Palatal  Velar  Glottal
Stops                          "p"      ""           "t"       ""         ""       "k"    "ʔ"
Fricatives                     "ɸ"      "f"          "s"       ""         ""       "ɣ"    "h"
Nasals                         "m"      ""           "n"       ""         ""       ""     ""
Approximants                   ""       ""           ""        "ɭ"        "j"      ""     ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Bilabial", "Labiodental", "Alveolar", "Retroflex", "Palatal", "Velar", "Glottal"), align = ('c')) %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 7)) %>%
  column_spec(1, bold = TRUE)
```

## Vowels

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
       Front   Back
High   "i"     "u"
Mid    "e"     "o"
Low    ""      "ɑ"
'), TRUE)

kable(consonants, align = ('c')) %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme
"a"           "/ɑ/"  
"b"           "/ɸ/"  
"e"           "/e/"  
"f"           "/f/"  
"g"           "/ɣ/"  
"h"           "/h/"  
"i"           "/i/"  
"k"           "/k/"  
"l"           "/ɭ/"  
"m"           "/m/"  
"n"           "/n/"  
"o"           "/o/"  
"p"           "/p/"  
"s"           "/s/"  
"t"           "/t/"  
"u"           "/u/"  
"y"           "/j/"  
"\'"           "/ʔ/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Benabena has a (C)V(ʔ) syllable structure [@BefYoung1992, p. 2].

# Lenition Rules

* In a small number of words, word-initial /ɸ/ approximantizes to [w] [@BefYoung1961, p. 9].
* /ʔ/ may be deleted in certain words (ibid.).

# Misc. Rules

* /ɸ/ alternates with [b] word-initially [@BefYoung1992, p. 2].
* /j/ undergoes some word-initial alternation: in stressed syllables, it alternates with [ʝ], while in unstressed syllables it alternates with [z] [@BefYoung1992 pp. 10-11].
* Some speakers realize /ɣ/ as [ɡ] [@BefYoung1992, p. 3; @BefYoung1961, p. 12].
* Some speakers realize /ɭ/ as [ɽ] [@BefYoung1992, p. 3; @BefYoung1961, p. 14].
* Non-glottal stops are sometimes realized with aspiration [@BefYoung1961, p. 8].
* Preconsonantal glottal stops may adopt the place of articulation of the following consonant; when this occurs, the stop is produced as unreleased [@BefYoung1961, p. 9].
* Some speakers back /s/ to [ʃ] before high vowels [@BefYoung1961, p. 11].

# References
