---
title: "Matsés"
author: "Emily Strand"
bibliography: mcf.bib
output: html_document
---

Last updated: 2019-10-04

**COMPROMISED: Crúbadán doesn't distinguish orthographically between retroflex and alveolar consonants; conflation between /u/ and /w/ and between /u/ and /o/**

# Background

**Language Family:** Panoan / Northern

* Also referred to as Mayoruna, Matsés is spoken along the border of Brazil and Peru.

# Phonology

## Consonants

* Alveolar and retroflex phonemes don't contrast when in the environment of /e/ and /i/ (they are realized as alveolar in such cases) [@Fleck2006, p. 62].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial    Alveolar  Postalveolar   Retroflex   Palatal   Velar    
Stops                       "p b"     "t d"        ""            ""         ""       "k"     
Affricates                   ""       "ts"         "tʃ"         "tʂ"        ""       ""
Fricatives                   ""        "s"         "ʃ"           "ʂ"        ""       ""
Nasals                       "m"       "n"         ""            ""         ""       ""
Approximants                 "w"       ""          ""            ""        "j"       ""       
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Retroflex", "Palatal", "Velar"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```  

## Vowels

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
           Front      Central       Back
High        "i"         "ɨ"         "u"
Mid         "e"         ""          "o"
Low         ""          "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)

```  

# Alphabet

* There may be some conflation among the retroflex and alveolar consonants, given that some orthographies don't make a distinction between them [@Fleck2003, p. 84]. **The Crúbadán corpus does not represent the retroflex consonants.**
* There may also be some conflation between /u/ and /w/ and between /o/ and /u/ given that in some documentation, both pairs are represented by ⟨u⟩ and ⟨o⟩, respectively [@Fleck2003, pp. 64 91].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme           Comment
"a"              "/a/"              ""
"b"              "/b/"              ""
"c"              "/k/"              ""
"d"              "/d/"              ""
"e"              "/e/"              ""
"ë"              "/ɨ/"              ""
"i"              "/i/"              ""
"k"              "/k/"              ""
"m"              "/m/"              ""
"n"              "/n/"              ""
"o"              "/o/"              ""
"p"              "/p/"              ""
"s"              "/s/"              ""
"t"              "/t/"              ""
"u"              "/u/"              ""
"w"              "/w/"  "**not represented in the Crúbadán corpus**"
"y"              "/j/"              ""
**Digraph**       ""                ""
"sh"             "/ʃ/"              ""
"şh"             "/ʂ/"              ""
"ts"             "/ts/"             ""
"ch"             "/tʃ/"             ""
"çh"             "/tʂ/"             ""
"qu"             "/k/"   "preceding /e/, /i/, and /ɨ/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* /k/ is realized as a glottal stop word finally [@Fleck2003, p. 62].
* /d/ is realized as a flap intervocalically [ibid.].
* /b/ is said to spirantize [@Fleck2010, p. 34]; however, @Fleck2003 states that the frication of /b/ is very minimal and is more often realized as /w/ (intervocalically) (p. 74).

# Misc. Rules

* /k/ is the only stop that can geminate, realized phonetically as a glottal stop followed by /k/ [@Fleck2003, p. 76].
* /n/ assimilates to the place of articulation of following velar and bilabial stops [@Fleck2003, p. 78].
* /n/ can geminate, while /m/ does not [@Fleck2003, p. 79].
* Fricatives can geminate, whereas affricates and approximants don't [@Fleck2003, pp. 84, 87, 88].
* /i/ and /u/ are realized as [j] and [w] respectively preceding or following vowels (not intervocalically) [@Fleck2003, p. 95].
* /o/ and /e/ are often realized as /u/ and /i/, respectively, when adjacent to each other or /a/ [@Fleck2003, p. 96].

# References
