---
title: "Nabak"
author: "Emily Strand"
bibliography: naf.bib
output: html_document
---

Last Updated: 2019-11-19

# Background

**Language Family:** Trans-New Guinea / Main Section / Central and Western / Huon-Finisterre / Huon / Western

* Nabak is spoken in the Morobe province of Papua New Guinea.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial       Alveolar    Palatal        Velar     
Stops                      "p  b"       "t  d"        ""       "k  kʷ  ɡ  ɡʷ"
Fricatives                   ""         "s  z"        ""            ""  
Nasals                      "m"           "n"         ""           "ŋ"  
Approximants                "w"           "l"         "j"           ""   
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Palatal", "Velar"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 4)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced. The velar stops that have the diacritic (ʷ) are labialized.", general_title = "")
```

## Vowels

* Vowel sequences are interpreted as such, never as complex nuclei [@Fabian1971, p. 8].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
             Front      Central       Back
High          "i"         ""           "u"
"High-Mid"    "e"         ""           "o"
"Low-Mid"     ""          ""           "ɔ"
Low           ""          "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```  

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme         Phoneme
"a"               "/a/"
"â"               "/ɔ/"
"b"               "/b/"
"d"               "/d/"
"e"               "/e/"
"g"               "/ɡ/"
"i"               "/i/"
"k"               "/k/"
"l"               "/l/"
"m"               "/m/"
"n"               "/n/"
"ŋ"               "/ŋ/"
"o"               "/o/"
"p"               "/p/"
"s"               "/s/"
"t"               "/t/"
"u"               "/u/"
"w"               "/w/"
"y"               "/j/"
"z"               "/z/"
**Digraph**       ""
"gw"              "/ɡʷ/"
"kw"              "/kʷ/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Syllables in Nabak have the following structures [@Fabian1998, p. 3]:
    - V
    - VC
    - CV
    - CVC
    - C (see note below)

* Syllabic nasals are present in Nabak [@Fabian1971, p. 5]; however, in word-initial and word-medial positions, most speakers insert a vowel between it and the following consonant.

# Lenition Rules

* /ɡ/ spirantizes to [ɣ] between like vowels [@Fabian1998, p. 3].

# Misc. Rules

* /s/ and /z/ vary freely with [ts] and [dz], respectively [@Fabian1998, p. 3].
* /kʷ/ and /ɡʷ/ vary freely with [kp] and [gb], respectively (ibid.).
* /l/ varies freely with [r] word-medially (ibid.).
* /w/ varies freely with [v] (ibid.).
* ⟨kw⟩ and ⟨gw⟩ are always realized as /kʷ/ and /ɡʷ/ as such sequences do not occur across syllable boundaries [@Fabian1971, p. 7].

# References
