---
title: "Crimean Tatar"
author: "Allison Shao"
bibliography: crh.bib
output: html_document
---

Last updated: 2020-04-10

# Background

**Language family**: Turkic / Kipchak / West Kipchak

* Crimean Tatar is spoken primarily in Crimea and Uzbekistan, but small communities of speakers can also be found in Russia, Bulgaria, Romania, and Turkey [@Kavitskaya2010, p. 1].
* For an estimation of the number of speakers, around 250,000 Crimean Tatars reside in Crimea, and another 250,000 live outside Crimea due to massive deportation in 1944 [@Kavitskaya2010, p.1].
* Crimean Tatar is divided into Central, Northern, and Southern varieties. This account focuses on the Central variety, which is considered the standard [@Kavitskaya2010, p. 1].

# Phonology

## Consonants

* There is some controversy over the phonemic status of /q/ and /ɣ/, such that they are sometimes viewed as allophones of /k/ and /ɡ/, respectively, where /k/ and /ɡ/ precede front vowels and /q/ and /ɣ/ precede back vowels [@Berta1998, p. 302]. However, I have chosen to recognize all four phonemes, given that @Kavitskaya2010 finds instances where /k/ is found in words preceding back vowels (p. 10).
* Glottal stops and palatalized /nʲ/ can be found in some loanwords [@Kavitskaya2010, p. 10].
* @Kavitskaya2010 describes /lʲ/ and /kʲ/ as marginally phonemic, with contrasts occurring in loanwords; I have chosen to treat them as allophones of /l/ and /k/, respectively, given that they appear in complementary distribution with /l/ and /k/, respectively, in native words (pp. 10, 14, 17).

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"      Labial    Dental   Alveolar  Postalveolar   Palatal    Velar   Uvular
Stops                         "p b"     "t̪ d̪"       ""         ""           ""       "k ɡ"    "q"
Affricates                      ""        ""        ""       "tʃ dʒ"        ""        ""       ""
Fricatives                    "f v"       ""      "s z"        "ʃ"          ""       "x ɣ"     ""
Nasals                         "m"        ""       "n"         ""           ""       "ŋ"       ""
Trills                          ""        ""       "r"         ""           ""        ""       ""
Approximants                    ""        ""       "l"         ""           "j"       ""       ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Dental", "Alveolar", "Postalveolar", "Palatal", "Velar", "Uvular"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 7)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```

## Vowels

* Like other Turkic languages, Crimean Tatar exhibits vowel harmony based on backness and rounding. Vowel harmony varies moderately between the three dialects [@Kavitskaya2014, pp. 224-225].
* According to @Kavitskaya2010, long vowels are present in some loanwords. Some speakers treat them as two short vowels separated by a glottal stop (p. 8).

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front        Back
High     "i y"       "ɯ u"       
Mid      "e ø"        "o"
Low        ""         "ɑ"
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are unrounded and those on the right are rounded.", general_title = "")
```   

# Alphabet

* According to @Kavitskaya2010, the grapheme ⟨â⟩ represents palatalization of preceding consonants or the sequence /jɑ/ (p. 4). Since Crimean Tatar does not have phonemic palatalized consonants, I will use /jɑ/.
* The language can be written with both the Latin and Cyrillic scripts; however, due to ambiguity related to the correspondences between graphemes and phonemes using the Cyrillic script, I have chosen only to represent the Latin script below [@Kavitskaya2010, p. 4].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme                Comment
"a"	           "/ɑ/"                   ""
"â"	           "/jɑ/"                  ""
"b"	           "/b/"                   ""
"c"            "/dʒ/"                  ""
"ç"            "/tʃ/"                  ""
"d"	           "/d̪/"                   ""
"e"	           "/e/"                   ""
"f"            "/f/"                   ""
"g"	           "/ɡ/"                   ""
"ğ"            "/ɣ/"                   ""
"h"	           "/x/"                   ""
"j"            "/ʒ/"        "only appears in loanwords"
"i"	           "/i/"                   ""
"ı"            "/ɯ/"                   ""
"k"	           "/k/"                   ""
"l"	           "/l/"                   ""
"m"	           "/m/"                   ""
"n"	           "/n/"                   ""
"ñ"            "/ŋ/"                   ""
"o"	           "/o/"                   ""
"ö"	           "/ø/"                   ""
"p"	           "/p/"                   ""
"q"            "/q/"                   ""
"r"            "/r/"                   ""
"s"	           "/s/"                   ""
"ş"            "/ʃ/"                   ""
"t"	           "/t̪/"                   ""
"u"	           "/u/"                   ""
"ü"            "/y/"                   ""
"v"	           "/v/"                   ""
"y"	           "/j/"                   ""
"z"            "/z/"                   ""
**Digraph**     ""                     ""
"ts"           "/ts/"      "only appears in loanwords"
"şç"           "/ʃʲ/"      "only appears in loanwords"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* /p/, /k/, /x/, and /tʃ/ may become voiced intervocalically [@Berta1998, p. 303].
* The second consonant in a coda consonant-cluster may be deleted if the coda does not exhibit falling sonority [@Kavitskaya2010, p. 20]. This does not occur if a vowel immediately follows the consonant cluster.
    - If a coda consonant-cluster consists of more than two vowels, the medial one will be deleted.

# Misc. Rules

* Consonant gemination occurs in Crimean Tatar; however, it is extremely marginal (not present in the Crúbadán corpus), occurring primarily in loanwords [@Kavitskaya2010, pp. 13-14].
* Voiced stops and voiced fricatives are devoiced word-finally [@Kavitskaya2010, p. 10].
* There is evidence of monophthongization of diphthongs, as seen in sequences /uv/, which can be realized as [uw] or [u:], and /ij/, which is realized as [i:] [@Kavitskaya2010, pp. 8-9].
* Syncope of high vowels is common word-initially and word-medially [@Kavitskaya2014, p. 225].
* In suffixes, /l/ undergoes progressive nasal assimilation and surfaces as [n] [@Kavitskaya2010, p. 32-33].
* /v/ may be realized as [w] [@Kavitskaya2010, p.12].
* /x/ often surfaces as uvular [χ] (ibid.).
* /p/, /b/, /m/, /s/, /z/, /l/, /k/, and /ɡ/ have palatalized allophonic forms when preceding front vowels [@Kavitskaya2010, p. 14].

# References  
