---
title: "Tzotzil"
author: "Bill Mizgerd"
bibliography: tzo.bib
output: html_document
---

Last Updated: 2019-06-27

# Background

**Language Family**: Mayan / Core Mayan / Cholan-Tzeltalan / Tzeltalan / Tzotzil

* Tzotzil is spoken in the State of Chiapas located in southern Mexico.

# Phonology

## Consonants

* /d f ɡ β/ appear in some recent loanwords [@TzoLaughlin1975, p. 22].
* @TzoCowan1969 describes the affricates as sequences of two distinct phones, but she ultimately opts to "reserve judgment as to their phonemic status," so I have opted to follow Laughlin's more recent and decisive analysis (p. 4).
* @TzoCowan1969 argues for /x/ rather than /h/.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial      Alveolar    Postalveolar  Palatal  Velar    Glottal
Stops                      "p pʼ b"    "t tʼ"      ""            ""       "k kʼ"   "ʔ"
Affricates                 ""          "ts tsʼ"    "tʃ tʃʼ"      ""       ""       ""
Fricatives                 "v"         "s"         "ʃ"           ""       ""       "h"
Nasals                     "m"         "n"         ""            ""       ""       ""
Flaps                      ""          "ɾ"         ""            ""       ""       ""
Approximants               ""          "l"         ""            "j"      ""       ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  footnote("Note: For phonemes that share a cell, those on the left are plain and those on the right are ejective. The labial stops are ordered voiceless, ejective, and voiced.", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* Some speakers realize all of their vowels without rounding [@TzoLaughlin1975, p. 22].
* The claim has been made that tone is developing in Tzotzil, but as of 2010 there is no evidence for any tones in Tzotzil being contrastive [@TzoAvelino2011, pp. 1-2].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
          Front   Central   Back
High      "i"     ""        "u"
Mid       "e"     ""        "o"
Low       ""      "a"       ""
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered")
```

# Alphabet

* Some orthographies just use ⟨k⟩ and ⟨k'⟩ rather than ⟨c⟩, ⟨qu⟩, ⟨c'⟩, and ⟨q'u⟩ for representing velar stops.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme
"a"              "/a/"  
"b"              "/b/"  
"c"              "/k/"  
"c\'"            "/kʼ/"   
"e"              "/e/"  
"i"              "/i/"  
"j"              "/h/"  
"l"              "/l/"  
"m"              "/m/"  
"n"              "/n/"  
"o"              "/o/"  
"p"              "/p/"  
"p\'"            "/pʼ/"  
"r"              "/ɾ/"  
"s"              "/s/"  
"t"              "/t/"  
"t\'"            "/tʼ/"   
"u"              "/u/"  
"v"              "/v/"  
"x"              "/ʃ/"  
"y"              "/j/"  
"\'"             "/ʔ/"
**Digraph**      ""
"ch"             "/tʃ/"  
"ch\'"           "/tʃʼ/"
"qu"             "/k/"  
"q\'u"           "/kʼ/"  
"ts"             "/ts/"  
"ts\'"           "/tsʼ/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```   

# Lenition Rules

* /h/ is weakly realized when followed by [ʔ] [@TzoLaughlin1975, p. 22].
* /j/ deletes after /ʃ/ or before sibilant fricatives [@TzoCowan1969, p. 111].
* Before stressed syllables that begin with consonants (other than /ʔ/), /k/ spirantizes to [x] and /v/ deletes entirely (ibid.).
* /ʔ/ deletes after the following consonants: /k v j/ [@TzoCowan1969, p. 112].
* /ʔ/ and /ʃ/ delete when followed by two consecutive non-/ʔ/ consonants [@TzoCowan1969, pp. 112-113].
* Mid vowels delete before non-front vowels [@TzoCowan1969, p. 114].
* Unstressed /i/ deletes before /a/ or /u/ (ibid.).
* /t/ spirantizes to [ʃ] before the /-uk/ suffix [@TzoCowan1969, p. 115].
* /h/ deletes before consonant clusters, or when preceded by an unstressed vowel and followed by a stressed vowel (ibid.).

# Misc. Rules

* Epenthesis of [ʔ] occurs before intervocalic /b/ [@TzoLaughlin1975, p. 22].
* Preconsonantal or word-final /b/ is realized as [ʔm], and the vowel preceding that /b/ is nasalized and glottalized (ibid.).
* /b/ is generally realized as [ɓ] word-initially and intervocalically (ibid.).
* Word-final /ɾ/ is trilled to [r] (ibid.).
* Word-medial /v/ moves to [β] before /a/ (ibid.).
* Doubled consonants (e.g. identical root-final and suffix-initial consonants) are realized identically to their single counterparts, except for /b/ [@TzoCowan1969, p. 111].
* Before stressed syllables that begin with consonants (other than /ʔ/), /j/ deapproximantizes to /s/ (ibid.).
* /ʃ/ moves to [s] before affricates [@TzoCowan1969, p. 112].
* In the sequence /tʃkC/, where C is any consonant other than /ʔ/, /tʃ/ vocalizes to [ti] (ibid.).
* /n/ assimilates to the place of following /b/ [@TzoCowan1969, p. 115].
* /m/ moves to [n] after /VmV/ sequences [@TzoCowan1969, p. 116].

# References
