---
title: "Misima-Panaeati"
author: "Allison Shao"
bibliography: mpx.bib
output: html_document
---

Last updated: 2020-03-05

**Slightly Compromised: conflation between /e/ and /ɛ/ and between /o/ and /ɔ/**

# Background

**Language family**: Austronesian / Malayo-Polynesian / Oceanic / Kilivila-Misima

* Speakers of Misima-Panaeati reside in Papua New Guinea, mostly on Misima Island, which is part of the Calvados Chain. Speakers also live on the nearby Panaeti Island.

# Phonology

## Consonants

* For some speakers, /j/ is closer to [ʝ] [@Callister_1993, pp. 5-6].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation" Labial      Dental    Retroflex    Palatal    Velar     Glottal
"Stops (plain)"          "p  b"      "t̪  d̪"       ""          ""       "k  ɡ"      ""
"Stop (labialized)"     "pʷ  bʷ"      ""          ""          ""        ""         ""
Fricatives                "β"         "s̪"         ""          ""        ""         "h"
"Nasals (plain)"          "m"         "n̪"         ""          ""        ""         ""
"Nasals (labialized)"     "mʷ"        ""          ""          ""        ""         ""
Approximants              "w          ""          "ɭ"         "j"       ""         ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Dental", "Retroflex", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```

## Vowels

* @SIL_Misima and @Callister_1993 disagree on whether Misima-Panaeati has a low back vowel /ɑ/ or a low front vowel /a/ (p. 1; p. 2). I have opted to use /a/.
* Adjacent vowels are not interpreted as complex nuclei [@Callister_1993, pp. 3-4].  
* Phonemes /e/ and /ɛ/, and phonemes /o/ and /ɔ/ contrast in closed syllables [@SIL_Misima, p. 3]. However, this is rather marginal [@Callister_1993, pp. 6-7].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High     "i"          ""          "u"         
High-Mid "e"          ""          "o"
Low-Mid  "ɛ"          ""          "ɔ"
Low      ""           "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```    

# Alphabet

* Phonemes /e/ and /ɛ/ and phonemes /o/ and /ɔ/ are represented by ⟨e⟩ and ⟨o⟩, respectively [@SIL_Misima, p. 1]. Given this, the language is somewhat compromised. I have opted to use /e/ and /o/ as the default phonemes in the rule set.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme
"a"	           "/a/"
"b"	           "/b/"
"d"	           "/d̪/"
"e"	           "/e/; /ɛ/"
"g"	           "/ɡ/"
"h"	           "/h/"
"i"	           "/i/"
"k"	           "/k/"
"l"	           "/ɭ/"
"m"	           "/m/"
"n"	           "/n̪/"
"o"	           "/o/; /ɔ/"
"p"	           "/p/"
"s"	           "/s̪/"
"t"	           "/t̪/"
"u"	           "/u/"
"v"	           "/β/"
"w"	           "/w/"
"y"	           "/j/"
**Digraph**     ""
"bw"	         "/bʷ/"
"mw"	         "/mʷ/"
"pw"	         "/pʷ/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Syllables in Misima-Panaeati follow the order of (C)V(C) [@SIL_Misima, p. 3].

# Misc. Rules

* /n/ is realized as /ŋ/ before velar stops [@Callister_1993, p. 5].
* /i/ and /u/ will surface as [j] and [w], respectively, when a vowel-initial suffix is added to a stem ending in /i/ or /u/ [@Callister_1993, p. 15].
* Stem-final /m/ is labialized when followed by a vowel via reduplication or addition of a suffix. This is typically not that case for stem-final /p/ and /b/ [@Callister_1993,p. 15].
* Prefixes /pa/ and /ma/ surface as [pʷa] and [mʷa], respectively, when the word stem begins with a labialized consonant or /mu/, /mo/, /bu/ or /bo/ [@Callister_1993, p. 16].
* /a/ is typically realized as [ʌ]. It is realized as [a] when in a stressed syllable and preceding /β/, /w/, /k/, or /g/. When it is not in a stressed syllable and precedes /m/, /s̪/, /n̪/, or /ɭ/, it is realized as [ə] [@Callister_1993, p. 6].
* Stress is phonemic, but stress will not be accounted for here [@SIL_Misima, p. 3].
* Codas can only be occupied by nasals, /s̪/, /ɭ/, /h/, and nonlabialized stops (excluding /d/) [@Callister_1993, p. 18].

# References  
