---
title: "Piapoco"
author: "Emily Strand"
bibliography: pio.bib
output: html_document
---

Last updated: 2019-11-04

# Background

**Language Family:** Arawakan / Maipuran / Northern Maaipuran / Inland

* Piapoco is spoken in Guainía and Vichada departments of Eastern Colombia.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial  Alveolar  Palatal   Velar   Glottal
Stops                      "p  b"  "t  d"      ""       "k"      ""
Affricates                  ""      "ts"       ""       ""       ""
Fricatives                  ""      "s"        ""       ""       "h"
Nasals                      "m"     "n"        ""       ""       ""
Flaps                       ""      "ɾ"        ""       ""       ""
Approximants                "w"     "l"        "j"      ""       ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 5)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```

## Vowels

* /o/ appears in some loanwords [@Klumpp2019, p. 4].
* Vowel length is phonemic in Piapoco [@Klumpp1985, pp. 22-23]. Long vowels are indicated by duplicate graphemes.
* Stress and tone appear to be one in the same (i.e. the suprasegmental features overlap) [@Klumpp2019, p. 5]. That is, high stress/high tone is marked by an acute accent (´) over the vowel, while low stress/low tone is marked by a grave accent (\`) over the vowel. @Klumpp1995 explains that stress is the contrastive feature that is accompanied by changes in tone (p. vi), suggesting that tonal differences may be a by-product of stress. Although there is this ambiguity, I have opted to include tone as a contrastive feature in the rules.
    - Syllables (vowels) not marked with either accent fluctuate freely in terms of tone [@Klumpp2019, p. 2].
    - Accents are only atop the first vowel of a long vowel sequence [@Klumpp2019, p. 5].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         ""           "u"
Mid       "e"         ""           ""
Low        ""         "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```   

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme
"a"             "/a/"
"b"             "/b/"
"c"             "/k/"
"d"             "/d/"
"e"             "/e/"
"i"             "/i/"
"j"             "/h/"
"l"             "/l/"
"m"             "/m/"
"n"             "/n/"
"p"             "/p/"
"r"             "/ɾ/"
"s"             "/s/"
"t"             "/t/"
"u"             "/u/"
"w"             "/w/"
"y"             "/j/"
**Digraph**     ""
"ch"            "/ts/"
"qu"            "/k/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Syllables in Piapoco follow the order of (C)V [@Klumpp2019, p. 2].

# Lenition Rules

* /h/ is often deleted, especially by younger generations [@Klumpp2019, p. 3].

# Misc. Rules

* /k/ often palatalizes between /i/ and /a/ [@Klumpp2019, p. 2].
* /ts/ and [tʃ] are said to be in free variation [@Klumpp2019, p. 3].
* /w/ becomes a fricative [β] preceding front vowels [@Klumpp2019, p. 4].

# References
