---
title: "Yakut"
author: "Jordan Schuster"
edited by: "Emily Strand"
bibliography: sah.bib
output: html_document
---

Last Updated: 2019-08-22

**SLIGHTLY COMPROMISED: conflation between /j/ and /j̃/**

# Background

**Language Family:** Turkic / Common Turkic / Siberian / Northern / Yukutian

* Also referred to as Sakha, Saxa, Saqa, or Yakutian, it is spoken in Russia.

# Phonology

## Consonants

* There seems to be some debate as to whether the palatal affricates /cç/ and /ɟʝ/ are in fact stops /c/ and /ɟ/ respectively, due to a lack of frication. I opted for that classification of affricates as done by @tolkov_phonetics_2011 (p. 1).

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation" Bilabial Alveolar Palatal  Velar   Glottal
Stops                      "p b"   "t d"     ""     "k ɡ"     ""
Affricates                 ""        ""    "cç ɟʝ"   ""       ""
Fricatives                 ""       "s"      ""     "x ɣ"     "h"
Nasals                     "m"      "n"      "ɲ"     "ŋ"      ""
Flaps                      ""       "ɾ"      ""      ""       ""
Approximants               ""       "l"     "j j̃"    ""       ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Bilabial", "Alveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 5)) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced. The palatal approximant on the right is nasalized.", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* Vowel harmony exists in Yakut [@tolkov_phonetics_2011, p. 6-7].
* Vowel length is contrastive. Long vowels are indicated by repeated graphemes [@tarasova_translation_2013, p. 2].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
           Front      Central       Back
High       "i y"         ""        "ɯ u"
Mid        "e ø"         ""         "o"
Low         ""           "a"        ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are unrounded and those on the right are rounded.", general_title = "")


diphthongs <- read.table(textConnection('
Diphthongs
"/ie/, /yø/, /ɯa/, /uo/"
'), TRUE)

kable(diphthongs, align = "c") %>%
  kable_styling("bordered")
```

# Alphabet

* The distinction between /j/ and /j̃/ is not marked in the orthography [@tolkov_phonetics_2011, p. 2], compromising the language.
    - /j/ is transcribed by default in the rules, meaning /j̃/ is not represented in the Crúbadán corpus.
* Graphemes present in loanwords will be included in the transcription only if the corresponding phoneme is also a native phoneme in Yakut.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme             Comment
"а"	             "/a/"                ""
"б"	             "/b/"                ""
"в"              "/v/"       "only used in loanwords"
"г"	             "/ɡ/"                ""
"ҕ"	             "/ɣ/"                ""
"д"	             "/d/"                ""
"е"              "/e/"       "only used in loanwords"
"ё"              "/jo/"      "only used in loanwords"
"ж"              "/ʒ/"       "only used in loanwords"
"з"              "/z/"       "only used in loanwords"
"и"	             "/i/"                ""
"й"            "/j/; /j̃/"    "/j/: default in the rules"
"к"              "/k/"                ""
"л"	             "/l/"                ""
"м"	             "/m/"                ""
"н"	             "/n/"                ""
"ҥ"	             "/ŋ/"                ""
"о"	             "/o/"                ""
"ө"	             "/ø/"                ""
"п"	             "/p/"                ""
"р"	             "/ɾ/"                ""
"с"	             "/s/"                ""
"һ"	             "/h/"                ""
"т"              "/t/"                ""
"у"	             "/u/"                ""
"ү"	             "/y/"                ""
"ф"              "/f/"        "only used in loanwords"
"х"              "/x/"                ""
"ц"              "/ts/"       "only used in loanwords"
"ч"	             "/cç/"               ""
"ш"              "/ʃ/"        "only used in loanwords"
"щ"              "/ɕː/"       "only used in loanwords"
"ъ"                ""         "only used in loanwords"
"ы"	             "/ɯ/"                ""
"ь"                ""     "may indicate palatalization of preceding consonant; only used in loanwords or in the digraphs ⟨дь⟩ and ⟨нь⟩"
"э"	             "/e/"                ""
"ю"              "/ju/"       "only used in loanwords"
"я"              "/ja/"       "only used in loanwords"
**Digraph**       ""                  ""
"нь"             "/ɲ/"   "geminate ⟨нь⟩ is written as ⟨ннь⟩, rather than ⟨ньнь⟩"
"дь"             "/ɟʝ/"               ""
'), TRUE)

kable(alphabet, align = "c") %>%
  kable_styling("bordered")
```

# Lenition Rules

* /s/ debuccalizes to /h/ [@tolkov_phonetics_2011, p. 1].
* /p/ voices intervocalically [@Gurevich2004, p. 269-272].
* /k/ is said to spirantize to /χ/ [@Gurevich2004]; however, @Kingston2008 argues that it's merely due to place assimilation (p. 9).

# Misc. Rules

* Gemination of consonant occurs [@tolkov_phonetics_2011, p. 7].

# References
