---
title: "Apalaí"
author: "Emily Strand"
bibliography: apy.bib
output: html_document
---

Last Updated: 2020-06-25

# Background

**Language family**: Carib / Northern / East-West Guiana / Wayana-Trio

* Apalaí is spoken in Brazil.

# Phonology

## Consonants

* @Koehn1995 argues for /b/ as an extremely marginal phoneme (p. iii).

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"  Bilabial  Alveolar  Postalveolar  Palatal  Velar  Glottal
Stops                         "p"     "t"          ""         ""      "k"     "ʔ"
Affricates                    ""       ""          ""         ""      ""      ""  
Fricatives                    ""      "s z"        "ʃ"        ""      ""      ""  
Nasals                        "m"      "n"         ""         ""      ""      ""
Flaps                         ""       "ɾ"         ""         ""      ""      ""  
Approximants                  "w"      ""          ""         "j"     ""      ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Bilabial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: Where two phonemes share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```

## Vowels

* Nasal vowels are contrastive in Apalaí [@Koehn1971, p. 3]. Nasal vowels are indicated by a tilde (˜) [@Koehn1995, p. iii].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         "ɨ"          "u"
Mid       "e"         ""           "o"
Low        ""         "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme
"a"            "/a/"
"e"            "/e/"
"h"            "/ʔ/"
"i"            "/i/"
"j"            "/j/"
"k"            "/k/"
"m"            "/m/"
"n"            "/n/"
"o"            "/o/"
"p"            "/p/"
"r"            "/ɾ/"
"s"            "/s/"
"t"            "/t/"
"u"            "/u/"
"w"            "/w/"
"x"            "/ʃ/"
"y"            "/ɨ/"
"z"            "/z/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Syllables in Apalaí follow the order of (C)V(C) [@Koehn1971, p. 4].

# Misc. Rules

* /k/ may be realized as [ɣ] in unstressed final syllables preceding words beginning with stressed /a/ [@Koehn1971, p. 3].
* In some dialects, /t/ and /k/ may palatalize following /i/. The palatalization of /t/ may result in some frication (ibid).
* /n/ may be realized as [ɲ] when contiguous with /i/ (ibid.).

# References
