---
title: "Ixil"
author: "Emily Strand"
bibliography: ixl.bib
output: html_document
---

Last updated: 2020-02-18

**Slightly Compromised: word-initial glottal stop not always marked; slightly ambiguous orthography**

# Background

**Language family**: Mayan/ Quichean-Mamean / Greater Mamean / Ixilan

* Ixil is spoken in the Quiché department of Guatemala.
* The Chajul dialect (as opposed to the Nebaj dialect) is represented below (i.e. the Crúbadán corpus reflects it better).

# Phonology

## Consonants

* Chajul Ixil is rather unique in that it has alveo-postalveolar consonants, which are said not to be represented in any other language [@Adell2019, p. 62]. In terms of the terminology and IPA symbols, I have chosen to use those represented in both @Adell2019 (p. 38) and @Adell2016 (p. 3).
* /q/ is said to be undergoing sound change to /χ/, so the phonemic status of /χ/ is somewhat unclear [@Adell2019, p. 83]. @Ayres1991 treats it as a distinct phoneme (p. 2); however, @Adell2019 does not, stating that, depending on the age group, speakers often treat it as if it's in free variation with /q/ (p. 86).
      - With such sound change, /χ/, especially with younger generations, is said to be merging with /h/.
      - I have chosen to follow the more recent source, not treating /χ/ as a distinct phoneme [@Adell2019, p. 38].
* /d/ and /ɡ/ appear in loanwords [@Ayres1991, p. 1].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial   Alveolar  Alveo-Postalveolar  Postalveolar  Retroflex  Palatal  Velar    Uvular  Glottal   
Stops                      "p ɓ"     "t t’"          ""                ""          ""        ""     "k k’"   "q q’"    "ʔ"
Affricates                  ""      "ts ts’"      "tsʃ tsʃ’"         "tʃ tʃ’"    "ʈʂ ʈʂ’"    ""      ""        ""      ""
Fricatives                  ""        "s"           "sʃ"               ""          "ʂ"       ""      ""        ""     "h"
Nasals                      "m"       "n"            ""                ""          ""        ""      ""        ""      ""
Flaps                       ""        "ɾ"            ""                ""          ""        ""      ""        ""      ""
Approximants                "ʋ"       "l"            ""                ""          ""        "j"     ""        ""      ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Alveo-Postalveolar", "Postalveolar", "Retroflex", "Palatal", "Velar", "Uvular", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 9)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are plain and those on the right are ejectives. The labial stop on the right is an implosive.", general_title = "")
```

## Vowels

* Vowel length is contrastive in Ixil [@Ayres1991, p. 1]. Long vowels are indicated by duplicate vowels.
    - When two consecutive like-vowels appear that don't constitute a long vowel, a hyphen is inserted between them.
* In native words, adjacent vowels do not represent complex nuclei (ibid.).

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
           Front      Central       Back
High        "i"         ""           "u"
Mid         "e"         ""           "o"
Low         ""          "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)

```

# Alphabet

* Word-initial glottal stop (the phonemic status of which is in question in this position) is not always written, which compromises the language to some extent [@Ayres1991, p. 5; @Adell2019, pp. 89-96].
* @Adell2019 also indicates variation in the orthography for some consonants (p. 23), which is also problematic. However, I have chosen to use the orthography represented in @Ayres1991, which seems largely consistent with what is used in Crúbadán (p. 1).

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme       Phoneme
"a"            "/a/"
"b\'"          "/ɓ/"
"e"            "/e/"
"h"            "/h/"
"i"            "/i/"
"j"            "/h/"
"k"            "/k/"
"k\'"          "/k’/"
"l"            "/l/"
"m"            "/m/"
"n"            "/n/"
"o"            "/o/"
"p"            "/p/"
"q"            "/q/"
"q\'"          "/q’/"
"r"            "/ɾ/"
"s"            "/s/"
"t"            "/t/"
"t\'"          "/t’/"
"u"            "/u/"
"v"            "/ʋ/"
"x"            "/ʂ/"
"y"            "/j/"
"\'"           "/ʔ/"
**Multigraph**  ""
"ch"           "/tʃ/"
"ch\'"         "/tʃ’/"
"tch"          "/tsʃ/"
"tch\'"        "/tsʃ’/"
"tx"           "/ʈʂ/"
"tx\'"         "/ʈʂ’/"
"tz"           "/ts/"
"tz\'"         "/ts’/"
"xh"           "/sʃ/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* /h/ is often deleted [@Ayres1991, p. 20].
* /l/ sometimes deletes preceding consonants [@Ayres1991, p. 21].

# Misc. Rules

* /n/ assimilates in place to the following consonant [@Ayres1991, pp. 24-25].
* /ʋ/ has several phonetic realizations that frequently vary freely [@Adell2019, pp. 50-51].
* /ɾ/ is said to be realized as [r] in some environments, but it's not exactly clear when this occurs. @Adell2019 explains that [r] occurs word-initially (p. 53), but @Ayres1991 states that it occurs word-finally, following a vowel (p. 3). However, this specification is not very relevant to this project.
    - After a glottal stop, /ɾ/ varies freely with [r] [@Ayres1991, p. 3].

# References
