---
title: "Akawaio"
author: "Emily Strand"
bibliography: ake.bib
output: html_document
---

Last updated: 2020-02-04

**Compromised: graphemes normally representing voiced stops and /z/ are not present in the orthography - ambiguity when voiceless stops and /s/ transcribe to voiced counterparts (underlying vs. surface level)**

# Background

**Language family**: Carib / Northern / East-West Guiana / Macushi-Kapon / Kapon

* Akawaio is spoken in Guyana.

# Phonology

* Although voiced consonants are said to be originally conditioned (voiceless to voiced) intervocalically or between vowels and nasals, they are considered phonemes of the language [@Gildea2012, p. 450; @Edwards1978, p. 78].

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial   Alveolar   Palatal    Velar      
Stops                      "p  b"    "t  d"      ""      "k  ɡ"     
Fricatives                  ""       "s  z"      ""        ""       
Nasals                      "m"        "n"       ""        ""       
Flaps                       ""         "ɾ"       ""        ""       
Approximants                "w"        ""        "j"       ""       
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Palatal", "Velar"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 4)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```

## Vowels   

* Rather than using their symbol /ʌ/ to represent the mid-central unrounded vowel [@Edwards1978, p. 77], I have opted for /ɘ/. /ə/ is not a suitable option given that all underlying vowels in Akawaio occur in stressed syllables.
* Diphthongs are present in Akawaio; however, because they are often interpreted as being separated by off-glides, or part of separate syllables, I have chosen not to include them below (ibid).  

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
           Front      Central       Back
High        "i"         "ɨ"          "u"
Mid         "e"         "ɘ"          "o"
Low         ""          "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)

```

# Alphabet

* Apostrophes indicate glottal stops [@Stegeman2014]; however, the glottal stop is not phonemic (see Lenition Rules below). It will be transcribed as /k/.
* The orthography doesn't represent the voiced stops [@Stegeman2014, p. 2], and it is unclear when the voiceless stops should transcribe to the voiced stops (ie. there is too much ambiguity whether the resulting (voiced consonant) transcription is underlying or a surface representation - see Lenition Rules below), thus the languages is compromised. I have chosen to transcribe ⟨p⟩, ⟨t⟩, ⟨k⟩, and ⟨s⟩ to their voiceless variants by default given their originating status in the language, but this means that /b/, /d/, /ɡ/, and /z/ are not represented by the rules.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme       Phoneme         
"a"             "/a/"            
"e"             "/e/"            
"i"             "/i/"            
"ɨ"             "/ɨ/"            
"k"             "/k/"            
"m"             "/m/"            
"n"             "/n/"            
"o"             "/o/"            
"p"             "/p/"            
"r"             "/ɾ/"            
"s"             "/s/"            
"t"             "/t/"            
"u"             "/u/"            
"ʉ"             "/ɘ/"            
"w"             "/w/"            
"y"             "/j/"     
"\'"            "/k/"     
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* /k/ debuccalizes to [ʔ] syllable-finally [@Edwards1978, p. 79].
* Voiceless stops and /s/ become voiced intervocalically [@Edwards1978, p. 81].
* Unstressed vowels are often deleted, especially word-initially [@Edwards1978, p. 82].
* Voiced stops optionally spirantize [@Gildea2012, p. 450].

# Misc. Rules

* /n/ is realized as [ŋ] word-finally and syllable finally, preceding velar consonants and /w/ [@Edwards1978, p. 79].
* /k/ becomes palatalized or labialized when preceding /i/ or /u/, respectively (ibid.).
* /s/ and /z/ are realized as [tʃ] and [dʒ], respectively, preceding /i/ (ibid.).

# References
