---
title: "Candoshi-Shapra"
author: "Abi Creighton"
bibliography: cbu.bib
output: html_document
---

Last Updated: 2020-07-14

# Background

**Language Family:** Language isolate

* Candoshi-Shapra, also known as Candoshi, Kandoshi, Candoxi, or Murato, is spoken in Peru along the Chapuli, Huitoyacu, Pastaza, and Morona river valleys.

# Phonology

## Consonants

* The inventory in @surralles_2007 includes voiced stops; however, it represents a phonetic description rather than a phonemic one (p. 255). The voiced stops are simply allophones of their voiceless counterparts [@tuggy_1966, p. 233].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Bilabial   Alveolar  Postalveolar Retroflex Palatal Velar   Glottal
Stops                      "p ᵐp"     "t ⁿt"        ""          ""        ""    "k ᵑk"    ""   
Affricates                  ""       "ts ⁿts"    "tʃ ⁿtʃ"     "ʈʂ ᶯʈʂ"    ""     ""       ""
Fricatives                  ""         "s"          "ʃ"         "ʂ"       ""     ""       "h"
Nasals                      "m"        "n"          ""          ""        ""     ""       ""       
Taps                        ""         "ɾ"          ""          ""        ""     ""       ""       
Approximants                ""         ""           ""          ""        "j"    "w"      ""   
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Bilabial", "Alveolar", "Postalveolar", "Retroflex", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 7)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are plain and those on the right are prenasalized.", general_title = "")
```  

## Vowels

* There is some debate around the back vowel and whether it better reflects /o/ or /u/. I have chosen to follow @Michael2015 and use /o/.
* Vowel length is contrastive in Candoshi-Shapra. Long vowels are marked by duplicate graphemes [@tuggy_1966, p. 235].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
             Front      Central       Back
High          "i"         ""           ""
Mid           ""          ""           "o"
Low           ""          "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)


diphthongs <- read.table(textConnection('
Diphthongs
"/ia/, /oa/"
'), TRUE)

kable(diphthongs, align = 'c') %>%
  kable_styling("bordered")
```     

# Alphabet

* @tuggy_1966 includes an apostrophe (’) as part of the Candoshi-Shapra alphabet, representing glottal stops (p. 232). However, the same source, which is a dictionary of the language, provides no instances of it. There is no justification of the glottal stop being a phoneme in the language as both @Michael2015 and @surralles_2007 make no mention of it, so I have chosen to transcribe apostrophes, which make up only 0.1% of the language's Crúbadán corpus, to nothing in the rule set.


```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme           
"a"              "/a/"              
"b"              "/ᵐp/"              
"d"              "/ⁿt/"              
"g"              "/ᵑk/"              
"h"              "/h/"              
"i"              "/i/"
"k"              "/k/"
"m"              "/m/"              
"n"              "/n/"              
"o"              "/o/"          
"p"              "/p/"              
"r"              "/ɾ/"              
"s"              "/s/"              
"t"              "/t/"              
"u"              "/o/"              
"v"              "/w/"   
"x"              "/ʈʂ/"
"y"              "/j/"   
"z"              "/ʂ/"
**Multigraph**    ""                   
"ch"             "/tʃ/"    
"ll"             "/ⁿtʃ/"
"sh"             "/ʃ/"
"ts"             "/ts/"
"nts"            "/ⁿts/"
"nx"             "/ᶯʈʂ/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Misc. Rules

* Consonant clusters are permitted, but they may not be consist of a consonant followed by a sonorant consonant or /h/ [@tuggy_1966, p. 235].
    - Clusters consisting of /h/ followed by a consonant are also not permitted.

# References
