---
title: "Guayabero"
author: "Emily Strand"
bibliography: guo.bib
output: html_document
---

Last Updated: 2020-06-29

# Background

**Language Family**: Guahiboan

* Guayabero is spoken in the Meta and Guaviare departments of central and southern Colombia, respectively.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial    Alveolar   Postalveolar   Palatal    Velar    Glottal
Stops                      "p  b"    "t  d"         ""           ""        "k"       "ʔ"
Affricates                   ""        ""          "tʃ"          ""         ""       ""
Fricatives                   "ɸ"       "s"          ""           ""        "x"       "h"
Nasals                       "m"       "n"          ""           ""         ""       ""
Flaps                        ""        "ɾ"          ""           ""         ""       ""  
Approximants                 "w"       "l"          ""          "j"         ""       ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```

## Vowels

* /i/ and /a/ are the only vowels that have long counterparts in Guayabero [@Keels1985, p. 67]. Although /aː/ is said to be contrastive with /a/, nothing is mentioned about the contrastiveness of /iː/ with /i/. Based on this and /iː/ being relatively non-existent in Crúbadán (only four instances), I have opted to only account for /aː/, which is represented orthographically by duplicate graphemes.
* Several diphthongs occur in Guayabero [@Keels1985, pp. 66-67]; however, given that they can also appear as vowel sequences [@Waller1976, p. 53], they will not be included in the rules.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         "ɨ"          "u"
Mid       "e"         ""           "o"
Low        ""        "a aː"         ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: The low central vowels are ordered short and long.", general_title = "")
```    

# Alphabet

* Stress is denoted by accented vowels; however, we do not account for stress, so accented vowels will be transcribed to their plain counterparts.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme
"a"            "/a/"  
"b"            "/b/"  
"c"            "/k/"  
"d"            "/d/"  
"e"            "/e/"  
"f"            "/ɸ/"  
"i"            "/i/"  
"j"            "/h/"  
"k"            "/k/"  
"l"            "/l/"  
"m"            "/m/"  
"n"            "/n/"  
"o"            "/o/"  
"p"            "/p/"  
"r"            "/ɾ/"  
"s"            "/s/"  
"t"            "/t/"  
"u"            "/u/"  
"ʉ"            "/ɨ/"  
"w"            "/w/"  
"x"            "/x/"  
"y"            "/j/"  
"\'"           "/ʔ/"  
**Digraph**     ""
"ch"           "/tʃ/"  
"qu"           "/k/"  
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* The syllable structures permitted in Guayabero include [@Keels1985, p. 59]:
    - CV  
    - CVC  
    - CVCC  
    - CVV   
    - CVVC   
    - CVVCC  

# Lenition Rules

* Consonant gemination seems to occur through affixation [@Keels1985, p. 84]. However, such occurrences undergo degemination and are preceded by an epenthetic vowel phonetically.
* Unstressed vowels are deleted immediately preceding stressed syllables [@Keels1985, pp. 80-81].

# Misc. Rules

* Voiceless stops become aspirated preceding vowels [@Waller1976, p. 49].
* Consonants may be labialized preceding a stressed, rounded vowel [@Keels1985, pp. 71-72].
* Unstressed /a/ is deleted preceding /ɾ/ and following an obstruent [@Keels1985, p. 79].
* Vowel epenthesis occurs preceding voiced consonant suffixes [@Keels1985, p. 82].


# References
