---
title: "Guayabero"
author: "Emily Strand"
bibliography: guo.bib
output: html_document
---

Last updated: 2019-10-17

# Background

**Language family**: Guahiboan

* Guayabero is spoken in the Meta and Guaviare departments of central and southern Colombia, respectively.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial    Alveolar   Postalveolar   Palatal    Velar    Glottal
Stops                      "p  b"    "t  d"         ""           ""        "k"       "ʔ"
Affricates                   ""        ""          "tʃ"          ""         ""       ""
Fricatives                   "ɸ"       "s"          ""           ""        "x"       "h"
Nasals                       "m"       "n"          ""           ""         ""       ""
Flaps                        ""        "ɾ"          ""           ""         ""       ""  
Approximants                 "w"       "l"          ""          "j"         ""       ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```

## Vowels

* /i/ and /a/ are the only vowels that have long counterparts in Guayabero [@Keels1985, p. 67]. Although /aa/ is said to be contrastive with /a/, nothing is mentioned about the contrastiveness of /ii/ with /i/. Long vowels are represented by duplicate graphemes.
* Several diphthongs occur in Guayabero [@Keels1985, p. 66, 67]; however, given that they can also appear as vowel sequences [@Waller1976, p. 53], they will not be included in the rules.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         "ɨ"          "u"
Mid       "e"         ""           "o"
Low        ""        "a aː"         ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: The low central vowels are ordered short and long.", general_title = "")
```    

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme
"a"            "/a/"  
"b"            "/b/"  
"c"            "/k/"  
"d"            "/d/"  
"e"            "/e/"  
"f"            "/ɸ/"  
"i"            "/i/"  
"j"            "/h/"  
"k"            "/k/"  
"l"            "/l/"  
"m"            "/m/"  
"n"            "/n/"  
"o"            "/o/"  
"p"            "/p/"  
"r"            "/ɾ/"  
"s"            "/s/"  
"t"            "/t/"  
"u"            "/u/"  
"ʉ"            "/ɨ/"  
"w"            "/w/"  
"x"            "/x/"  
"y"            "/j/"  
"\'"            "/ʔ/"  
**Digraph**     ""
"ch"          "/tʃ/"  
"qu"          "/k/"  
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* The syllable structures permitted in Guayabero include [@Keels1985, p. 59]:
    - CV  
    - CVC  
    - CVCC  
    - CVV   
    - CVVC   
    - CVVCC  

# Lenition Rules

* Unstressed /a/ is deleted preceding /ɾ/ and following an obstruent [@Keels1985, p. 79].
* Unstressed vowels are deleted immediately preceding stress [@Keels1985, p. 80, 81].

# Misc Rules

* Voiceless stops become aspirated preceding vowels [@Waller1976, p. 49].
* Consonants may be labialized preceding a stressed, rounded vowel [@Keels1985, p. 71, 72].
* Vowel epenthesis occurs preceding voiced consonant suffixes [@Keels1985, p. 82].
* Geminate consonants seem to appear in the orthography through affixation; however long consonants are not present in the language [@Keels1985, p. 84]. Vowel epenthesis occurs to prevent the realization of such long consonants (ibid.).

# References
