---
title: "Armenian"
author: "Bill Mizgerd"
bibliography: hy.bib
output: html_document
---

Last Updated: 2020-06-29

# Background

**Language Family**: Indo-European / Armenian

* Where Armenian dialects diverge, I will be preferring Eastern Armenian, since it is the most widely spoken variant of the language.
* Speakers of Armenian reside in a multitude of places. They include Armenia, Azerbaijan, Georgia, Russia, the United States, France, Argentina, and many more countries.

# Phonology

## Consonants

* The plain voiceless stops glottalize in some dialects, but these are generally considered to be nonstandard realizations [@HyDumTragut2009, pp. 17-18].
* Minimal pairs of the phonemes below can be found in @HyDumTragut2009 (pp. 18-20).

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"    Labial       Alveolar    Postalveolar   Palatal    Velar     Uvular   Glottal
Stops                     "p  pʰ  b"    "t  tʰ  d"        ""          ""    "k  kʰ  ɡ"     ""        ""
Affricates                    ""      "ts  tsʰ  dz"  "tʃ  tʃʰ  dʒ"    ""        ""         ""        ""  
Fricatives                  "f  v"       "s  z"         "ʃ  ʒ"        ""        ""       "χ  ʁ"      "h"  
Nasals                        "m"          "n"            ""          ""        ""         ""        ""
Flaps                         ""           "ɾ"            ""          ""        ""         ""        ""  
Trills                        ""           "r"            ""          ""        ""         ""        ""  
Approximants                  ""           "l"            ""          "j"       ""         ""        ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Uvular", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 7)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: Where two phonemes share a cell, those on the left are voiceless and those on the right are voiced. Where three phonemes share a cell, they are ordered voiceless, aspirated, and voiced.", general_title = "")
```

## Vowels

* Armenian does not have diphthongs, but it does have pseudo-diphthongs through vowel-glide combinations [@HyDumTragut2009, p. 21].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
         Front      Central       Back
High      "i"         ""           "u"
Mid       "ɛ"         "ə"          "ɔ"
Low        ""         ""           "ɑ"
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```    

# Alphabet

* Older forms of the orthography used to include ⟨ւ⟩ as a standalone character representing /w/, but that has fallen out of usage; the sound has diverged, with different realizations of it merging with either ⟨վ⟩ /v/ or ⟨ու⟩ /u/ [@HyDumTragut2009, p. 12].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme      Phoneme  
"ա"           "/ɑ/"  
"բ"           "/b/"  
"գ"           "/ɡ/"  
"դ"           "/d/"  
"ե"           "/ɛ/"
"զ"           "/z/"  
"է"           "/ɛ/"  
"ը"           "/ə/"  
"թ"           "/tʰ/"  
"ժ"           "/ʒ/"  
"ի"           "/i/"  
"լ"           "/l/"  
"խ"           "/χ/"  
"ծ"           "/ts/"  
"կ"           "/k/"  
"հ"           "/h/"  
"ձ"           "/dz/"  
"ղ"           "/ʁ/"  
"ճ"           "/tʃ/"  
"մ"           "/m/"  
"յ"           "/j/"  
"ն"           "/n/"  
"շ"           "/ʃ/"  
"ո"           "/ɔ/"  
"չ"           "/tʃʰ/"  
"պ"           "/p/"  
"ջ"           "/dʒ/"  
"ռ"           "/r/"  
"ս"           "/s/"  
"վ"           "/v/"  
"տ"           "/t/"  
"ր"           "/ɾ/"  
"ց"           "/tsʰ/"  
"ու"          "/u/"  
"փ"           "/pʰ/"  
"ք"           "/kʰ/"  
"և"           "/ɛv/"  
"օ"           "/ɔ/"  
"ֆ"           "/f/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```  

# Lenition Rules

* The /ɛv/ cluster typically approximantizes to [ɛʋ] [@HyDumTragut2009, p. 20].
* In a few specific words, /h/ deletes after /ɾ/ [@HyDumTragut2009, p. 28].
* In colloquial speech, /ɾ/ deletes in the coda of closed syllables (ibid.).

# Misc. Rules

* /u/ centralizes to [ʉ] after /j/ [@HyDumTragut2009, p. 14].
* [j] is inserted before /ɛ/ when it follows /ɑ/ or is word-initial [@HyDumTragut2009, pp. 14-15].
* [ʋ] is inserted before /ɔ/ in the same circumstances [@HyDumTragut2009, p. 16].
* [j] is inserted before /ɛv/ word-initially [@HyDumTragut2009, p. 20].
* Stops devoice and aspirate in the following circumstances [@HyDumTragut2009, pp. 22-23; @HySakayan2007, pp. 37, 60, 83]:
    - After /ɾ/, usually
    - In postvocalic final or intervocalic positions, occasionally
    - Following nasals with the same place of articulation
* Voiceless stops sometimes devoice when followed by sibilants (ibid.).
* In consonant clusters, fricatives undergo voicing assimilation, and affricates undergo both voicing and aspiration assimilation [@HyDumTragut2009, p. 24].
* /ɾ/ trills to [r] before /n/ [@HyDumTragut2009, p. 27].
* /n/ assimilates to the place of following stops or affricates [@HyDumTragut2009, pp. 19, 27].
* Final flaps palatalize in some dialects, especially in the east [@HyDumTragut2009, p. 28].
* Alveolar stops palatalize before front vowels or /j/ [@HyDumTragut2009, p. 29].
* /l/ palatalizes preconsonantally (ibid.).
* During the inflection process, non-high back vowels delete in some circumstances [@HyDumTragut2009, p. 39].
* Initial /i/ reduces to [ə] in monosyllabic words [@HyDumTragut2009, p. 41].
* Unstressed high vowels reduce to [ə] or delete during word formation and inflection [@HyDumTragut2009, pp. 41-42; @HyKozintseva1995, p. 6].
* Schwas are frequently inserted to syllabify consonant clusters [@HyDumTragut2009, p. 30].
* Medial /e/ raises to [i] in inflected forms of monosyllabic roots [@HyDumTragut2009, p. 40].

# References
