---
title: "Xicotepec de Juárez Totonac"
author: "Emily Strand"
bibliography: too.bib
output: html_document
---

Last updated: 2020-02-07

**Slightly Compromised: correspondences between grapheme and phoneme are somewhat unclear; phonological analysis of the language within the acquired documentation was lacking**

# Background

**Language family**: Totonacan / Totonac

* This dialect of Totonac is spoken in the Puebla state of Mexico.

# Phonology

## Consonants

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial   Alveolar  Postalveolar  Palatal  Velar  Uvular  Glottal   
Stops                       "p"       "t"         ""          ""      "k"    "q"      "ʔ"
Affricates                  ""      "ts tɬ"       "tʃ"        ""      ""     ""       ""
Fricatives                  ""       "s ɬ"        "ʃ"         ""      ""     ""       "h"
Nasals                      "m"       "n"         ""          ""      ""     ""       ""
Approximants                "w"       "l"         ""          "j"     ""     ""       ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Uvular", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 7)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: The alveolar affricate and fricative on the right are lateral.", general_title = "")
```

## Vowels

* Vowel length is contrastive [@Reid1974, p. 6]. Long vowels are indicated by a macron (¯).

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
           Front      Central       Back
High        "i"         ""           "u"
Mid         "e"         ""           "o"
Low         ""          "a"          ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)

```

# Alphabet

* ⟨b⟩, ⟨d⟩, ⟨f⟩, ⟨g⟩, ⟨ñ⟩, ⟨r⟩, ⟨rr⟩, ⟨v⟩, and ⟨z⟩ appear in Spanish loanwords [@Reid1974, p. 6].
* There is some ambiguity in terms of the transcription, given that the correspondences weren't exactly clear. I have opted to classify this language as compromised, given that I am not 100% certain on the transcription.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme       Phoneme
"a"             "/a/"
"c"             "/k/"
"e"             "/e/"
"i"             "/i/"
"j"             "/h/"
"k"             "/q/"
"l"             "/l/"
"m"             "/m/"
"n"             "/n/"
"o"             "/o/"
"p"             "/p/"
"s"             "/s/"
"t"             "/t/"
"u"             "/u/"
"x"             "/ʃ/"
"y"             "/j/"
"\'"            "/ʔ/"
**Digraph**      ""
"ch"            "/tʃ/"
"hu"            "/w/"
"lh"            "/ɬ/"
"qu"            "/k/"
"tl"            "/tɬ/"
"tz"            "/ts/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Rules

* In other dialects of Totonac (eg. Misantla) it's found that [@MacKay1994, pp. 374-376]:
    - Voiceless stops and affricates may become voiced between voiced consonants.
    -  /q/ may spirantize following vowels.  
**It is unclear whether the same processes are observed in the dialect under analysis**

# Misc. Rules

* In other dialects of Totonac (eg. Misantla), /n/ assimilates in place to following consonants [@MacKay1994, p. 380].

# References
