---
title: "Hungarian"
author: "Shiying Yang"
date: "4/1/2019"
output:
  html_document:
    df_print: paged
  pdf_document:
    dev: cairo_pdf
    keep_tex: yes
    latex_engine: xelatex
    number_sections: yes
geometry: margin=1.2in
header-includes:
- \usepackage{setspace}\doublespacing
- \usepackage{tipa}
mainfont: DejaVu Serif
fontsize: 12pt
tables: yes
bibliography: hu.bib
---
# Background

**Language family:** 

* Uralic 
  + Finno-Ugric

**Adopted variety:** 
ECH (Educated Colloquial Hungarian) [@ST]

# Phonology and Orthography

## Consonants [@ST, p 18-19; @IPA] {#con}

```{r, echo=F, include=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
```

```{r hu_con, echo=F, tab.cap = NULL, results='asis'}
hu_c <- read.csv('hu_c.csv')
#colnames(jp_c)[c(1:2,5)] <- c(' ', '  ', 'alveolo-palatal')
c_position <- c('labial', 'alveolar', 'post-alveolar', 'palatal', 'velar', 'glottal')
c_colname <- paste(c_position, 'O', sep = ',')
c_colname_in_table <- unlist(strsplit(c_colname, ','))

kable(hu_c, caption = 'Consonants', 
      align = "c", longtable = T, 
      digits = 3, format = 'latex', 
      booktabs = T, col.names = c(' ', 'Voicing', c_colname_in_table)) %>% 
  collapse_rows(columns = c(1), valign = 'middle', latex_hline = 'custom', custom_latex_hline = 1) %>% 
  footnote(alphabet = c('placeless /N/ appears preconsonantally and receives place from the following consonant', 'The [O] columns stands for orthography.', '[vs] and [vd] in the second column stands for voiceless and voiced'),footnote_as_chunk = F) %>%
  kable_styling("striped", full_width = T)
```

## Vowels [@IPA] {#v}
||IPA | Writing | Example | Gloss|
|-|----|-------|---------|------|
1 |\textipa{A}|a |agy|brain|
2 |a\textipa{:}|á |ágy|bed|
3 |\textipa{E}|e |egy|one|
4 |e\textipa{:}|é |ért|understand|
5 |i|i |irt|eradicate|
6 |i\textipa{:}|í |ír|write|
7 |o|o |orr|nose|
8 |o\textipa{:}|ó |ól|sty|
9 |\o|ö |öl|kill|
10|\o\textipa{:}|ő |őr|guard|
11|u|u |ujj|finger|
12|u\textipa{:}|ú |úgy|like that|
13|y|ü |ügy|affair|
14|y\textipa{:}|ű |űrr|space|
Marginal[^1]|a, e, \textipa{O:}, \textipa{E:}|

[^1]: typically used for pronuncing letters in abbreviations and borrowed words [@ST, p 280]

## Note on the Alphabet

Letters q, x, y, w as well as the combination 'ch' are excluded in hu.rules since they only occur in borrowed words like homoszexuális, and traditional surnames e.g. Andrássy.

# Rules 
<!--(Correspondence between orthography and phonemes)-->

see hu.rules written based on Section \ref{con} and \ref{v}

## Controversies and choices
1. Whether "ty", "dy" are affricates or stops as seen in \ref{con}
- stops [see @ST, p 82]
- does not changed the result of translation
2. Whether there are dipthongs 
- no [see @ST, p 16-18]
3. Whether [dz] is a phoneme or just a surface sound
- surface [see @ST, p 87-89]
- underlying [see @IPA]
4. The placeless [N] as the underlying default nasal preconsonantally that undergoes progressive nasal assimilation
- [see @ST, p 207-212]

# Lenition Processes

- Degemination [@ST, p 83]: Across word boundaries, affricates remain unmerged in careful speech. In colloquial speech, the first affricate may lenite into a fricative due to OCP

|| Writing | careful speech | colloquial speech|
|-|--------|-------------|-----------|
  ||rác cég|[raːts-tseːg]|[raːs-tseːg]|
  ||bölcs csere|[bøltʃ-tʃɛrɛ]|[bølʃ-tʃɛrɛ]|

# References {-}
