---
title: "Hungarian"
author: "Shiying Yang"
edited by: "Emily Strand"
bibliography: hu.bib
output: html_document
---

Last Updated: 2020-03-18

# Background

**Language family:** Uralic / Finnic

* Hungarian is spoken in Hungary and in several surrounding countries as well as in the United States, Canada, and Israel.
* I have chosen to represent Educated Colloquial Hungarian (ECH), the dialect spoken in Budapest, Hungary.

# Phonology

## Consonants

* /m/ and /ɲ/ are consistently listed as phonemes in Hungarian [@Szende1994, p. 91; @Siptar2000, p. 207]; however, there is some debate around /n/. @Siptar2000 explain that rather than having a nasal phoneme specified for place, Hungarian has a placeless /N/ that preconsonantally assimilates (p. 207). I have chosen to follow @Szende1994 and use the alveolar nasal (p. 91).

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial   Alveolar   Postalveolar   Palatal  Velar  Glottal  
Stops                       "p b"    "t d"         ""          "c ɟ"   "k ɡ"    ""
Affricates                  ""      "ts dz"       "tʃ dʒ"       ""      ""      ""
Fricatives                  "f v"    "s z"         "ʃ ʒ"        ""      ""      "h"
Nasals                      "m"       "n"          ""           "ɲ"     ""      ""
Trills                      ""        "r"          ""           ""      ""      ""
Approximants                ""        "l"          ""           "j"     ""      ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```

## Vowels

* Vowel length is contrastive in Hungarian [@Szende1994, p. 92]. Long vowels are indicated by an acute (´) or double acute (˝) accent.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
              Front         Central       Back
High        "i iː y yː"       ""         "u uː"
High-Mid     "eː ø øː"        ""         "o oː"
Low-Mid        "ɛ"            ""          ""
Low            ""             "aː"        "ɑ"
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: Phonemes with the diacritic (ː) are phonemically long. The two high front vowels on the left are unrounded and the two on the right are rounded. The high-mid front vowel on the left is unrounded and the two on the right are rounded.", general_title = "")
```

# Alphabet

* ⟨q⟩, ⟨w⟩, ⟨x⟩, ⟨y⟩ appear only in loanwords and in traditional surnames [@Rounds2001, p. 3].  

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme       Phoneme
"a"             "/ɑ/"
"á"             "/aː/"
"b"             "/b/"
"c"             "/ts/"
"d"             "/d/"
"e"             "/ɛ/"
"é"             "/eː/"
"f"             "/f/"
"g"             "/ɡ/"
"h"             "/h/"
"i"             "/i/"
"í"             "/iː/"
"j"             "/j/"
"k"             "/k/"
"l"             "/l/"
"m"             "/m/"
"n"             "/n/"
"o"             "/o/"
"ó"             "/oː/"
"ö"             "/ø/"
"ő"             "/øː/"
"p"             "/p/"
"r"             "/r/"
"s"             "/ʃ/"
"t"             "/t/"
"u"             "/u/"
"ú"             "/uː/"
"ü"             "/y/"
"ű"             "/yː/"
"v"             "/v/"
"z"             "/z/"
**Multigraph**   ""
"cs"            "/tʃ/"
"dz"            "/dz/"
"dzs"           "/dʒ/"
"gy"            "/ɟ/"
"ly"            "/j/"
"ny"            "/ɲ/"
"sz"            "/s/"
"ty"            "/c/"
"zs"            "/ʒ/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Lenition Processes

* Degemination [@Siptar2000, p 83]: Across word boundaries, affricates remain unmerged in careful speech. In colloquial speech, the first affricate may lenite into a fricative.
* Geminates also weaken intervocalically [@Bauer1988, p. 383].

# Misc. Rules

* All consonant can geminate in Hungarian [@Magyar2015, p. 288]. Orthographically this involves repetition of graphemes. For multigraphs, geminates are indicated by duplication of the first consonant [@Korompay2012, p. 341].

# References
