---
title: "Mauwake"
author: "Emily Strand"
bibliography: mhl.bib
output: html_document
---

Last Updated: 2020-05-06

# Background

**Language Family:** Trans-New Guinea / Madang-Adelbert Range / Adelbert Range / Pihom-Isumrud-Mugil / Pihom / Kumilan

* Mauwake is spoken in the Madang Province of Papua New Guinea.

# Phonology

## Consonants


```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"        Labial   Alveolar    Palatal    Velar    
Stops                           "p b"     "t d"        ""       "k ɡ"    
Fricatives                       "ɸ"       "s"         ""        ""      
Nasals                           "m"       "n"         ""        ""      
Trills                           ""        "r"         ""        ""      
Approximants                     "w"       "l"         "j"       ""      

'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Palatal", "Velar"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 4))  %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are voiceless and those on the right are voiced.", general_title = "")
```

## Vowels

* Vowel length is contrastive in word-initial syllables [@Berghaell2015, pp. 33-34]. Long vowels are represented by duplicate graphemes.
    - Long vowels are retained in words when affixation or reduplication occurs [@Berghaell2015, pp. 42-43]. For example, in the word "mika-puukiman," the long /uː/ that is present word-initially in the root "puukiman" is maintained even when the prefix "mika" is added. This is worth noting because the Crúbadán corpus includes hyphens indicating morpheme boundaries.
* Adjacent vowels are interpreted as complex nuclei when the second vowel is of the same articulatory height or higher than that of the first vowel [@Berghaell2015, p. 40].
    - /u/ is interpreted as having an articulatory height lower than that of /i/, resulting in the diphthong /ui/.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
          Front   Central   Back
High      "i"      ""        "u"
Mid       "e"      ""        "o"
Low       ""       "a"       ""
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)


diphthongs <- read.table(textConnection('
Diphthongs
"/ai/, /ae/, /ao/, /au/, /ui/"
'), TRUE)

kable(diphthongs, align = 'c') %>%
  kable_styling("bordered")
```

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme        Phoneme  
"a"              "/a/"
"b"              "/b/"
"d"              "/d/"
"e"              "/e/"
"f"              "/ɸ/"
"g"              "/ɡ/"
"i"              "/i/"
"k"              "/k/"
"l"              "/l/"
"m"              "/m/"
"n"              "/n/"
"o"              "/o/"
"p"              "/p/"
"r"              "/r/"
"s"              "/s/"
"t"              "/t/"
"u"              "/u/"
"w"              "/w/"
"y"              "/j/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Syllables in Mauwake follow the order of (C)V(V)(C), with CV being the most frequent structure [@Berghaell2015, p. 39].

# Lenition Rules

* Vowels preceding stressed syllables may be elided [@Berghaell2015, p. 41].
* /r/ varies freely with [ɾ] [@Berghaell2015, p. 27].

# Misc. Rules

* In some dialects, /j/ is realized as [ʒ] [@Berghaell2015, p. 28].
* /w/ may be realized as either [v] or [β] (ibid.).

# References
