---
title: "Moose Cree"
author: "Emily Strand"
bibliography: crm.bib
output: html_document
---

Last updated: 2019-10-28

**SLIGHTLY COMPROMISED: /h/ represented orthographically only when contrast is required**

# Background

**Language Family:** Algic / Algonquian / Central / Cree-Montagnais-Naskapi

* Moose Cree is spoken in Ontario, Canada.

# Phonology

## Consonants

* /r/ appears in loanwords [@Ellis2000, p. 468].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"   Labial   Alveolar    Postalveolar    Palatal    Velar    Glottal
Stops                       "p"       "t"            ""           ""        "k"       ""
Affricates                  ""        ""            "tʃ"          ""        ""        ""
Fricatives                  ""        "s"           "ʃ"           ""        ""        "h"
Nasals                      "m"       "n"            ""           ""        ""        ""
Approximants                "w"       "l"            ""           "j"       ""        ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Labial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* Vowel length is contrastive in Moose Cree. A small dot atop each character indicates length [@gaudry2005].
    - Notice that /e/ is always long (a dot is never affixed to the character).

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

vowels <- read.table(textConnection('
            Front    Central     Back
High       "i  iː"     ""         ""
Mid         "eː"       ""       "o  oː"
Low          ""      "a aː"       ""
'), TRUE)

kable(vowels, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = "Note: The vowels marked with (ː) are long.", general_title = "")
```   

# Syllabary

* /h/ is said to be included orthographically only when contrast or emphasis is necessary [@Ellis2000, p. 465]. Thus, the language is compromised to some extent.

| ᐃ - /i/   | ᐅ - /o/  | ᐊ - /a/  | ᐁ - /eː/   |  Final  |
|:---------:|:---------:|:--------:|:----------:|:-------:|
| ᐱ	- /pi/  | ᐳ - /po/ | ᐸ - /pa/ | ᐯ - /peː/ | ᑉ - /p/  |
| ᑎ - /ti/  | ᑐ - /to/ | ᑕ - /ta/ | ᑌ - /teː/ | ᑦ - /t/  |
| ᑭ - /ki/  | ᑯ - /ko/  | ᑲ - /ka/ | ᑫ - /keː/  | ᒃ - /k/  |
| ᒋ - /tʃi/ | ᒍ - /tʃo/ | ᒐ - /tʃa/| ᒉ - /tʃeː/ | ᒡ - /tʃ/ |
| ᒥ - /mi/  | ᒧ - /mo/  | ᒪ - /ma/ | ᒣ - /meː/  | ᒻ - /m/  |
| ᓂ - /ni/ | ᓄ - /no/  | ᓇ - /na/ | ᓀ - /neː/ | ᓐ - /n/  |
| ᓯ - /si/  | ᓱ - /so/  | ᓴ - /sa/  | ᓭ - /seː/ | ᔅ - /s/  |
| ᔑ - /ʃi/ | ᔓ - /ʃo/  | ᔕ - /ʃa/ | ᔐ - /ʃeː/ | ᔥ - /ʃ/ |
| ᔨ - /ji/  | ᔪ - /jo/  | ᔭ - /ja/  | ᔦ - /jeː/ | ᔾ - /j/  |   
| ᓕ - /li/ | ᓗ - /lo/  | ᓚ - /la/ | ᓓ - /leː/ | ᓪ - /l/  |
| ᐎ - /wi/ | ᐒ - /wo/  | ᐗ - /wa/ | ᐌ - /weː/ | ᐤ - /w/  |
| ᐦᐃ - /hi/ | ᐦᐅ - /ho/ | ᐦᐊ - /ha/ | ᐦᐁ - /heː/ | ᐦ - /h/  |
|           |           |           |           | ᔉ - /sk/ |
|           |           |           |           |          |

* /w/ can also appear between a consonant and a vowel, the transcription of which involves a dot (like in the /w/ + vowel sequence) on the left of any character (ex. ᑄ = /pwa/) [@gaudry2005].
    - Based on [@syllabics], which also seems consistent with Crúbadán, /w/ does not appear following /n/ and /h/, except in the sequences /nwa/ and /nwaː/.
* A small ring above a word-final syllable indicates /j/; however, Crúbadán doesn't seem to follow this convention, opting for ⟨ᔾ⟩ instead, the convention of which is consistent with the other word-final consonants [@Ellis2000, p. 465].

# References
