---
title: "Bora"
author: "Bill Mizgerd"
bibliography: boa.bib
output: html_document
---

Last Updated: 2019-07-02

# Background

**Language Family**: Witotoan / Boran

* Bora is spoken in five villages within the Loreto Region of Peru.

# Phonology

## Consonants

* @BoaRoe2014 describes /x/ as a phoneme in Bora (p. 8); @BoaThiesen2012 lists /x ~ h/ in free variation (p. 33), and while he prefers /h/ as the underlying phoneme, the majority of the evidence seems to point to /x/.
* Different authors have interpreted the flap differently: opinions have ranged from /r/ to /ɾ/ to /ɽ/ [@BoaRoe2014, p. 9]. I have opted to prefer /ɾ/ in my ruleset.
* Consonants followed by /j/ palatalize, and the /j/ deletes. @BoaThiesen2012 does not make a firm decision on whether these palatalized consonants are contrastive (p. 33); @BoaRoe2014 argues against it (pp. 8-9), and I follow her lead in my ruleset.

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

library(dplyr)
library(knitr)
library(kableExtra)


consonants <- read.table(textConnection('
"Manner of Articulation"     Bilabial   Alveolar  Postalveolar  Palatal   Velar      Glottal
Stops                        "p pʰ"     "t tʰ"    ""            ""        "k kʰ kp"  "ʔ"
Affricates                   ""         "ts tsʰ"  "tʃ tʃʰ"      ""        ""         ""
Fricatives                   "β"        ""        ""            ""        "x"        ""
Nasals                       "m"        "n"       ""            "ɲ"        ""        ""
Flaps                        ""         "ɾ"       ""            ""        ""         ""
Approximants                 ""         ""        ""            "j"       ""         ""
'), TRUE)

kable(consonants, col.names = c("Manner of Articulation", "Bilabial", "Alveolar", "Postalveolar", "Palatal", "Velar", "Glottal"), align = 'c') %>%
  kable_styling("bordered") %>%
  add_header_above(c("", "Place of Articulation" = 6)) %>%
  footnote(general = "Note: For phonemes that share a cell, those on the left are unaspirated and those on the right are aspirated. The third phoneme in the velar stop cell is a voiceless labiovelar stop.", general_title = "") %>%
  column_spec(1, bold = TRUE)
```

## Vowels

* Bora has two contrastive tones, high and low [@BoaThiesen2012, p. 27].
    - High tone is marked with an acute accent (´) on the vowel, whereas low tone is unmarked [@BoaThiesen2012, p. 27].
* Vowel length is contrastive in Bora. Long vowels are written by repeating the vowel character [@BoaThiesen2012, p. 31].

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}
consonants <- read.table(textConnection('
       Front   Central   Back
High   "i"     "ɨ"       "ɯ"
Mid    "ɛ"     ""        "o"
Low    ""      "a"       ""
'), TRUE)

kable(consonants, align = 'c') %>%
  kable_styling("bordered") %>%
  column_spec(1, bold = TRUE)
```

# Alphabet

```{r echo=FALSE, message=FALSE, warning=FALSE, results = 'asis'}

alphabet <- read.table(textConnection('
Grapheme     Phoneme
"a"           "/a/"  
"b"           "/p/"  
"c"           "/kʰ/"  
"d"           "/t/"    
"e"           "/ɛ/"  
"g"           "/k/"  
"h"           "/ʔ/"  
"i"           "/i/"  
"ɨ"           "/ɨ/"  
"j"           "/x/"  
"k"           "/kʰ/"  
"m"           "/m/"  
"n"           "/n/"  
"ñ"           "/ɲ/"  
"o"           "/o/"  
"p"           "/pʰ/"  
"r"           "/ɾ/"  
"t"           "/tʰ/"    
"u"           "/ɯ/"  
"v"           "/β/"  
"w"           "/kp/"  
"y"           "/j/"
**Digraph**   ""
"ch"         "/tʃʰ/"
"ds"          "/ts/"
"ll"          "/tʃ/"
"ts"         "/tsʰ/"
'), TRUE)

kable(alphabet, align = 'c') %>%
  kable_styling("bordered")
```

# Syllable Structure

* Bora syllables mostly take the form of (C)V, although /x/ and /ʔ/ can appear as syllable codas [@BoaRoe2014, p. 9].

# Lenition Rules

* Postconsonantal /j/ tends to delete, and instead just palatalizes the previous consonant [@BoaThiesen2012, p. 27].
* In syllable codas, /x/ is simply produced as a preaspiration of the following consonant (written by Thiesen as ⟨ˣ⟩) [@BoaThiesen2012, p. 40].
* Syllable-final /ʔ/ is produced simply as a glottalization of the vowel (written as ⟨ˀ⟩) [@BoaThiesen2012, p. 28].
* /kp/ becomes [kʷ] in fast speech, particularly among younger speakers [@BoaRoe2014, p. 9].

# Misc. Rules

* /i/ lowers in /x/-final syllables, and centralizes to [ɨ] when /ɨ/ is present in the following syllable [@BoaThiesen2012, p. 30].
* /ɛ/ lowers to [æ] before syllables that contain /i/ [@BoaThiesen2012, p. 31].
* Epenthesis of [ʔ] occurs at the beginning of vowel-initial phrases, and the end of vowel-final phrases [@BoaThiesen2012, p. 36].
* In Bora, two low-tone vowels cannot occur consecutively unless they are word-final [@BoaThiesen2012, p. 69].

# References
